### 2016-06-10 Gradschool/quantum/QuantumFinal.tex
Added QM Final

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{cancel}
\usepackage{pdfpages}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Quantum Mechanics, Final Exam \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Problem 1}

Consider a wavefunction given by

\begin{align}
\Psi(\vec{x}) = (x + y + 3z)f(r)
\end{align}

I wish to check if $\Psi$ is an eigenfunction of the $L^2$ operator in three dimensions. I know that in the spherical coordinate basis, the $L^2$ operator can be represented as (Shankar Equation 12.5.36):

\begin{align}
L^2 = \hbar^2 \left[ \frac{1}{\sin(\theta)}\frac{\partial}{\partial \theta}\sin(\theta)\frac{\partial}{\partial \theta} + \frac{1}{\sin^2(\theta)}\frac{\partial ^2}{\partial \phi^2} \right]
\end{align}

Using the transformations from cartiesian to spherical coordinates

\begin{align}
x &= r \cos(\phi)\sin(\theta) \\
y &= r \sin(\phi)\sin(\theta) \\
z &= r \cos(\theta)
\end{align}

The radial wavefunction reads

\begin{align}
\Psi &= (r\cos(\phi)\sin(\theta) + r\sin(\phi)\sin(\theta) + 3 r \cos(\theta))f(r) \\
&= r(\cos(\phi)\sin(\theta) + \sin(\phi)\sin(\theta) + 3 \cos(\theta))f(r) 
\end{align}

I now act upon the whole wavefunction with the $L^2$ operator in the coordinate basis and attempt to simplify the results. Starting from the last term first, I need two derivatives of the wavefunction with respect to $\phi$.

\begin{align}
\frac{d}{d\phi}\Psi &= rf(r)\frac{\partial}{\partial \phi} (\cos(\phi)\sin(\theta) + \sin(\phi)\sin(\theta) + \cancel{3 \cos(\theta))} \\
&= rf(r)(-\sin(\phi)\sin(\theta) + \cos(\phi)\sin(\theta)) \\
\text{2nd. }&\rightarrow \ \text{need} \ \frac{\partial}{\partial \phi} (-\sin(\phi)\sin(\theta) + \cos(\phi)\sin(\theta))  \\
&= rf(r)(-\cos(\phi)\sin(\theta) - \sin(\phi)\sin(\theta)) \\ 
\text{Times $\frac{1}{\sin^2(\theta)}$} &\rightarrow -rf(r)\left[\frac{\cos(\phi)}{\sin(\theta)} + \frac{\sin(\phi)}{\sin(\theta)} \right]
\end{align}

We need a derivative with respect to $\theta$:

\begin{align}
\frac{\partial}{\partial \theta}\Psi &=  \frac{\partial}{\partial \theta} rf(r)(\cos(\phi)\sin(\theta) + \sin(\phi)\sin(\theta) + 3 \cos(\theta)) \\
&= rf(r)(\cos(\phi)\cos(\theta) + \sin(\phi)\cos(\theta) - 3 \sin(\theta)) \\
\text{Times $\sin(\theta)$} &\rightarrow rf(r)(\cos(\phi)\cos(\theta)\sin(\theta) + \sin(\phi)\cos(\theta)\sin(\theta) - 3 \sin^2(\theta)) \\
&\rightarrow \frac{\partial}{\partial \theta} rf(r)(\cos(\phi)\cos(\theta)\sin(\theta) + \sin(\phi)\cos(\theta)\sin(\theta) - 3 \sin^2(\theta))
\end{align}
\begin{align} 
&= rf(r)\left[\cos(\phi)(\cos(2 \theta)) + \sin(\phi)(\cos(2 \theta)) - 3\sin(2 \theta)\right] \\
\text{Times $\frac{1}{\sin(\theta)}$} &\rightarrow rf(r)\frac{1}{\sin(\theta)} \left[\cos(\phi)(\cos(2 \theta)) + \sin(\phi)(\cos(2 \theta)) - 3\sin(2 \theta)\right]
\end{align}

%Now, using the identity that $\cos^2(\theta) - \sin^2(\theta) = \cos(\2\theta)$

At this point I should be able to piece together the action of the whole operator on $\Psi$:

\begin{align}
rf(r)\left[ \frac{1}{\sin(\theta)} \left[\cos(\phi)(\cos(2 \theta)) + \sin(\phi)(\cos(2 \theta)) - 3\sin(2 \theta)\right] -  \left[\frac{\cos(\phi)}{\sin(\theta)} + \frac{\sin(\phi)}{\sin(\theta)} \right] \right]
\end{align}

I'm not sure how exactly the trig identities make this work out, but \textit{Mathematica} \textbf{(code included)} this simplifies to

\begin{align}
h&= -2 (\sin (\theta ) (\sin (\phi )+\cos (\phi ))+3 \cos (\theta )) \\
\end{align}

Multiplying by the $-\hbar^2$ in front of the $L^2$ operator, we arrive at the fact that

\begin{align}
L^2 \Psi = 2 \hbar^2 \Psi
\end{align}

Meaning that $\Psi$ is an eigenstate of $L^2$ with the eigenvalue $2 \hbar^2$. Since we must have eigenvalues of the form $l(l+1) \hbar^2$, we know $l(l+1) = 2$, which indicates that $l=1$ for this eigenstate. This in turn indicates that $m_l$ can run from $-l$ to $l$ or that $m_l$ can take on the values $-1, 0$, and $1$. We should be able to get the probabilities of each of these states by expanding the angular part of the original state as a linear combination of spherical harmonics:

\begin{align}
\Psi(\theta, \phi) &= (\cos(\phi)\sin(\theta) + \sin(\phi)\sin(\theta) + 3 \cos(\theta)) \\
&=\left[ \left(\frac{e^{i \phi} + e^{- i \phi}}{2} \right) \sin(\theta) + \left(\frac{e^{i \phi} - e^{- i \phi}}{2i}\right)\sin(\theta) + 3 \cos(\theta) \right] \\
&= \left[\frac{1}{2}e^{i \phi}\sin(\theta) + \frac{1}{2}e^{-i \phi}\sin(\theta) - \frac{i}{2}e^{i \phi}\sin(\theta) + \frac{i}{2}e^{- i \phi}\sin(\theta) + 3 \cos(\theta)\right] \\
&= \left[\frac{1}{2}(1 - i)e^{i \phi}\sin(\theta) + \frac{1}{2}(1 + i)e^{-i \phi}\sin(\theta) + 3 \cos(\theta)\right]
\end{align}

Where I plugged in the standard definitons of sine and cosine as complex exponentials. Now, the $l=0$ and $l=1$ spherical harmonics read:

\begin{align}
Y^0_0 &= \sqrt{\frac{3}{4 \pi}} \cos(\theta) \\
Y^{\pm 1}_1 &= \mp \sqrt{\frac{3}{8 \pi}}\sin(\theta)e^{\pm i \phi}
\end{align}

I want to substitute these into the wavefunction in its current state. To do this I note that, for instance:

\begin{align}
3\cos(\theta) &= A\sqrt{\frac{3}{4 \pi}}\cos(\theta) \\
\end{align}

This is like asking "how much of that particular spherical harmonic is this term", then we can simply solve for $A$ to get to correct coefficients.

\begin{align}
&= \left[\frac{\frac{1}{2}(1 - i)}{-\sqrt{\frac{3}{8 \pi}}}Y^1_1 + \frac{\frac{1}{2}(1 + i)}{\sqrt{\frac{3}{8 \pi}}}Y^{-1}_1 + \frac{3}{\sqrt{\frac{3}{4 \pi}}}Y^0_0 \right] \\
&= \left[(i - 1)\sqrt{\frac{2 \pi}{3}}Y^1_1 + (1 + i)\sqrt{\frac{2 \pi}{3}}Y^{-1}_1 + 2 \sqrt{3 \pi}Y^0_0 \right]
\end{align}

The probabilities of getting each of the states are the squares of the moduli of each of the coefficients on the spherical harmonics, where $\Gamma$ is the sum of the squares of the moduli of the coefficients. Each one is individual divided by total. Gamma is:

\begin{align}
\Gamma &= \left|(i - 1)\sqrt{\frac{2 \pi}{3}}\right|^2 + \left|2 \sqrt{3 \pi}\right|^2 + \left|(1 + i)\sqrt{\frac{2 \pi}{3}}\right|^2 \\ 
&= \frac{44 \pi}{3} 
\end{align}

\begin{align}
P(m_l = -1) &= \frac{\left|(i - 1)\sqrt{\frac{2 \pi}{3}}\right|^2}{\Gamma} = \frac{1}{11} \\
P(m_l = 0) &= \frac{\left|2 \sqrt{3 \pi}\right|^2}{\Gamma} = \frac{9}{11}\\
P(m_l = 1) &= \frac{\left|(1 + i)\sqrt{\frac{2 \pi}{3}}\right|^2}{\Gamma} = \frac{1}{11} \\
\end{align}

Note the probability of finding the system in \textit{some state} adds up to 1.

\hspace{2mm}

If we know that $\Psi$ is an energy eigenfunction with eigenvalue $E$, we can find the potential using the schrodinger equation. To do this we would rewrite the radial schrodinger equation as

\begin{align}
-\frac{\hbar^2}{2m}\frac{d^2 u}{dr^2} + \left[V(r) + \frac{L^2}{2m r^2}\right]u = Eu
\end{align}

Since we have the eigenvalues of $L^2$, all we would do is take a bunch of derivatives of the $r$-dependent part, and then solve algebraically for $V(r)$. We have a wavefunction $\Psi = rf(r)g(\theta,\phi) = u(r)g(\theta,\phi)$ using the standard convention for the definiton of $u(r)$. Substituting in $-2\hbar^2$ for $L^2$ and expanding, we get

\begin{align}
-\frac{\hbar^2}{2m}\frac{d^2 u }{dr^2} + V(r)u + \frac{\hbar^2 u }{m r^2} = E u \\  
E u + \frac{\hbar^2}{m}\left[\frac{1}{2}\frac{d^2 u}{dr^2}  -\frac{u^2}{r^2}\right] = V(r)
\end{align}

Completing the problem.

\section{Problem 2}

We are asked to investigate a two spin-$\frac{1}{2}$, mass $m$ particle system interacting through the potential

\begin{align}
V(r) = \frac{g}{r}\sigma_1 \cdot \sigma_2
\end{align}

Where $g$ is a constant greater then zero and $\sigma_i$ are the pauli spin matrices for the $i'th$ particle.

I am interpreting $\sigma_1$ and $\sigma_2$ as some kind of vector of pauli matrices including the $x,y$, and $z$ components. The total spin operator is given by

\begin{align}
S = \frac{\hbar}{2}(\sigma_1 + \sigma_2)
\end{align}

So that $S \cdot S$ is given by

\begin{align}
\frac{\hbar^2}{4}(\sigma_1^2 + \sigma_2^2 + 2 \sigma_1 \cdot \sigma_2)
\end{align}

But we know that $\sigma_i^2 = \sigma_i \cdot \sigma_i = \sigma_i^x \sigma_i^x + \sigma_i^y \sigma_i^y + \sigma_i^z \sigma_i^z$. Now, any component pauli matrix squared equals the identity. So we have that $(\sigma_i^j)^2 = I$ for all of the components. This gets us to the fact that $\sigma_i^2 = 3I$.

\hspace{2mm}

Now we can write $S \cdot S$ as:

\begin{align}
S \cdot S &= \frac{\hbar^2}{4}(3I +3I + 2 \sigma_1 \cdot \sigma_2) \\
&= \frac{\hbar^2}{4}(6I + 2\sigma_1\cdot\sigma_2)
\end{align}

Rewriting the left hand side using the known eigenvalues for the $S \cdot S$ operator we have

\begin{align}
\cancel{\hbar^2} S(S+1) = \frac{\cancel{\hbar^2}}{4}(6I + 2 \sigma_1 \cdot \sigma_2)
\end{align}

From here we can solve the equation for the term $\sigma_1 \cdot \sigma_2$:

\begin{align}
&4  S(S + 1) = 6I + 2 \sigma_1 \cdot \sigma_2 \\
&4  S(S + 1) - 6I = 2 \sigma_1 \cdot \sigma_2 \\
&2  S(S + 1) - 3I = \sigma_1 \cdot \sigma_2 
\end{align}

Substituting this expression into the potential we have that

\begin{align}
V(r) = \frac{g}{r}\left[2 S(S+1) - 3I\right]
\end{align}

We can plug in a few values of $S$ to examine bound states. Evidently, if the total spin $S=1$, then $V(r) = \frac{g}{r}\left[4 - 3I\right]$ which is greater then zero and gives rise to repulsive potentials. The $S=0$ state has potential $V(r) = -\frac{3g}{r}$ which is attractive and gives rise to bound states.

\hspace{2mm}

Since this is a quantum mechanical two body problem with an attractive potential, the energy eigenvalues for this system must be something like the same ones for a hydrogen atom. Those are

\begin{align}
E_n = \frac{m}{2 \hbar^2 n^2} \left[\frac{e^2}{4 \pi \epsilon_0}\right]^2
\end{align}

Where the potential was $V(r) = -\frac{1}{4 \pi \epsilon_0 r}$. Therefore it is not unreasonable to expect energy eigenvalues something like:

\begin{align}
E_n = \frac{m}{2 \hbar^2 n^2}\left[3g\right]^2 = \frac{9 m g}{2 \hbar^2 n^2}
\end{align}

For the spin dependent potential.

\section{Problem 3}

We are asked to find the eigenstates for $S^2$ and $S_z$ for a three spin-$\frac{1}{2}$ particle system including no orbital angular momentum. We look to express the $2x2x2 = 8$ eigenstates in terms of the individual basis states. Since $[S^2, S_z] = 0$, we know that the two operators share a common eigenbasis. So it will suffice just to find the eigenstates of $S^2$. Since there are 3 such particles, the maximum possible value of $S$ is equal to $\frac{3}{2}$. This means that $m$ can take on values of $-\frac{3}{2}, -\frac{1}{2}, \frac{1}{2}$, and $\frac{3}{2}$.

\hspace{2mm}

We search for linear combinations of the individual basis states:

\begin{align}
1.\ket{\uparrow \uparrow \uparrow} \\
2.\ket{\downarrow \uparrow \uparrow} \\
3.\ket{\uparrow \downarrow \uparrow} \\
4.\ket{\uparrow \uparrow \downarrow} \\ 
5.\ket{\downarrow \downarrow \uparrow} \\
6.\ket{\downarrow \uparrow \downarrow} \\
7.\ket{\uparrow \downarrow \downarrow} \\
8.\ket{\downarrow \downarrow \downarrow}
\end{align}

We begin by showing that $\ket{\uparrow \uparrow \uparrow}$ is an eigenstate of $S^2$. We can do so by showing that it is in fact just an eigenstate of $S_z$. If $S^T_z = S^1_z + S^2_z + S^3_z$, then

\begin{align}
S^T_z\ket{\uparrow \uparrow \uparrow} &= (S^1_z + S^2_z + S^3_z)\ket{\uparrow \uparrow \uparrow} \\
&= S^1_z\ket{\uparrow}\ket{\uparrow}\ket{\uparrow} + \ket{\uparrow}S^2_z\ket{\uparrow} + \ket{\uparrow}\ket{\uparrow}S^3_z\ket{\uparrow} \\
&= \frac{\hbar}{2}\ket{\uparrow}\ket{\uparrow}\ket{\uparrow} + \ket{\uparrow}\frac{\hbar}{2}\ket{\uparrow} + \ket{\uparrow}\ket{\uparrow}\frac{\hbar}{2}\ket{\uparrow} \\
&= \frac{3\hbar}{2}\ket{\uparrow\uparrow\uparrow}
\end{align}

Which confirms that $\ket{\uparrow \uparrow \uparrow}$ is already an eigenstate both of $S_z$ and $S^2$. We can now get at least three other eigenstates of $S^2$ by acting the lowering operator on the top state:

\begin{align}
S^T_-\ket{\uparrow \uparrow \uparrow} &= (S^1_- + S^2_- + S^3_-) \ket{\uparrow \uparrow \uparrow} \\
&= S^1_-\ket{\uparrow}\ket{\uparrow}\ket{\uparrow} + \ket{\uparrow}S^2_-\ket{\uparrow}\ket{\uparrow} + \ket{\uparrow}\ket{\uparrow}S^3_-\ket{\uparrow} \\ 
&= \hbar \ket{\downarrow \uparrow \uparrow} + \hbar \ket{\uparrow \downarrow \uparrow} + \hbar \ket{\uparrow \uparrow \downarrow} \\
\text{Normalize.} &\rightarrow \frac{\hbar}{\sqrt{3}} \left( \ \ket{\downarrow \uparrow \uparrow} + \ket{\uparrow \downarrow \uparrow} + \ket{\uparrow \uparrow \downarrow} \ \right)
\end{align}

The above corresponds to the $\ket{\frac{3}{2}, \frac{1}{2}}$ state. By the same logic we can show that the bottom state, corresponding to $\ket{\downarrow \downarrow \downarrow}$ or $\ket{\frac{3}{2}, -\frac{3}{2}}$ is also an eigenstate of $S^2$ and therefore $S_z$:

\begin{align}
S^T_z\ket{\downarrow \downarrow \downarrow} &= (S^1_z + S^2_z + S^3_z)\ket{\downarrow \downarrow \downarrow} \\
&= -\frac{3\hbar}{2}\ket{\downarrow \downarrow \downarrow}
\end{align}

Applying now a \textit{raising} operator to this eigenstate gives

\begin{align}
S^T_+\ket{\downarrow \downarrow \downarrow} &= (S^1_+ + S^2_+ + S^3_+)\ket{\downarrow \downarrow \downarrow} \\
&= S^1_+\ket{\downarrow}\ket{\downarrow}\ket{\downarrow} + \ket{\downarrow}S^2_+\ket{\downarrow}\ket{\downarrow} + \ket{\downarrow}\ket{\downarrow}S^3_+\ket{\downarrow} \\
&= \hbar \ket{\uparrow \downarrow \downarrow} + \hbar \ket{\downarrow \uparrow \downarrow} + \hbar \ket{\downarrow \downarrow \uparrow} \\
\text{Normalize.} &\rightarrow \frac{\hbar}{3} \left( \ \ket{\uparrow \downarrow \downarrow} + \ket{\downarrow \uparrow \downarrow} + \ket{\downarrow \downarrow \uparrow} \ \right)
\end{align}

I can get at least two more states by just adding a spin-up or spin-down particle to a singlet for two particles:

\begin{align}
\frac{\hbar}{\sqrt{2}}( \ \ket{\uparrow \downarrow \uparrow} + \ket{\downarrow \uparrow \uparrow} \ ) \\
\frac{\hbar}{\sqrt{2}}( \ \ket{\uparrow \downarrow \downarrow} + \ket{\downarrow \uparrow \downarrow} \ )
\end{align}

I get the last two by treating two spin-$\frac{1}{2}$ particles as a single particle of spin 1 and using the $1x\frac{1}{2}$ entry in the Clebsch-Gordan table:

\begin{align}
\frac{\hbar}{\sqrt{6}}( \ 2\ket{\uparrow \uparrow \downarrow} + \ket{\downarrow \uparrow \uparrow} + \ket{\uparrow \downarrow \uparrow} \ ) \\
\frac{\hbar}{\sqrt{6}}( \ 2\ket{\downarrow \downarrow \uparrow} + \ket{\uparrow \downarrow \downarrow} \ket{\downarrow \uparrow \downarrow} \ )
\end{align}

Or, really, just finding one and then flipping all of the spins over.

\section{Problem 4}

We are asked to estimate the energy levels for a quartic potential in one dimension of the form $\lambda x^4$. The Schrodinger equation for this situation reads:

\begin{align}
-\frac{\hbar^2}{2m}\frac{\partial^2 \Psi}{\partial x^2} + \lambda x^4 \Psi = E \Psi
\end{align}

I will use the WKB approximation. Following \textit{Griffiths} equation 8.51, we have that the wavefunctions must match up in the region between the turning points $x_1$ and $x_2$.

\begin{align}
\int_{x_1}^{x_2} p(x) dx = \left(n - \frac{1}{2}\right)\pi \hbar
\end{align}

At a turning point, we must have that the potential energy is equal to the energy:

\begin{align}
E &= \lambda x^4 \\
\frac{E}{\lambda} &= x^4 \\
\left(\frac{E}{\lambda}\right)^{\frac{1}{4}} &= x_{1,2}
\end{align}

The potential is an even function allowing us to double the integral for $p(x)$ and plug in the value we just found for the upper limit:

\begin{align}
2 \int_0^{\left(\frac{E}{\lambda}\right)^{\frac{1}{4}}} \sqrt{2m(E - \lambda x^4)} dx
\end{align}

As it stands this integral is completely impossible. Make the substitution:

\begin{align}
z &= x^4 \\
x &= z^{\frac{1}{4}} \\
dz &= 4x^3 dx \\
&= 4 (z^{\frac{1}{4}})^3 dx \\
&= 4 z^{\frac{3}{4}} dx \\
\rightarrow &= dx = \frac{1}{4}z^{-\frac{3}{4}} dz
\end{align}

We also need to transform the upper limit of integration:

\begin{align}
(\frac{E}{\lambda}^{\frac{1}{4}})^{4} = \frac{E}{\lambda}
\end{align}

So we have

\begin{align}
2 \int_{0}^{\frac{E}{\lambda}} \sqrt{2m(E - \lambda z)}\left[\frac{1}{4}z^{-\frac{3}{4}}\right] dz
\end{align}

Integrating with Mathematica (code attached) I get

\begin{align}
\frac{E^{3/4} \sqrt{2 \pi m} \Gamma \left(\frac{5}{4}\right)}{\sqrt[4]{\lambda } \Gamma \left(\frac{7}{4}\right)}= \left(n - \frac{1}{2}\right)\pi \hbar
\end{align}

Solving this expression for $E$ gives me

\begin{align}
E = \left[\left(n - \frac{1}{2}\right)\pi\hbar \frac{\Gamma(\frac{7}{4})(4 \lambda^{\frac{1}{4}})}{\Gamma(\frac{1}{4})\sqrt{2 \pi m}} \right]^{\frac{4}{3}}
\end{align}

These should be the energy levels for a quartic potential.

\includepdf[pages={1-1}]{quantumfinal.pdf}

%----------------------------------------------------------------------------------------
%	PROBLEM 1
%----------------------------------------------------------------------------------------

\end{document}

### 2016-06-10 Gradschool/quantum/qm_hw1.tex
Added HW 6

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Quantum Mechanics, Homework 6 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

%----------------------------------------------------------------------------------------
%	PROBLEM 1
%----------------------------------------------------------------------------------------

\section{Shankar 12.5.3}

The total angular momentum component operators in the $x$ and $y$ directions are $\mathcal{J}_x$ and $\mathcal{J}_y$ respectively. Thus the expectation values $\braket{\mathcal{J}_x}$ and $\braket{\mathcal{J}_y}$ in the arbitrary angular momentum eigenstate $\ket{lm}$ are given by the following relations:

\begin{align}
\braket{\mathcal{J}_x} &= \braket{lm|\mathcal{J}_x|lm} \\
\braket{\mathcal{J}_y} &= \braket{lm|\mathcal{J}_y|lm}
\end{align}

Using the matrix element definiton of $\mathcal{J}_x$ and $\mathcal{J}_y$ from Shankar equation 12.5.21:

\begin{align}
\mathcal{J}_x &= \frac{\mathcal{J}_+ + \mathcal{J}_-}{2} \\
\mathcal{J}_y &= \frac{\mathcal{J}_+ - \mathcal{J}_-}{2i}
\end{align}

However the same equations also lead to the following result for $\mathcal{J}_x$:

\begin{align}
\braket{j'm'|\mathcal{J}_x|lm} &= \frac{1}{2} \big[\braket{jm|\mathcal{J}_+|jm} + \braket{jm|\mathcal{J}_-|jm} \big] \\
&= \frac{1}{2} \big[ C^+_{jm} \braket{jm|jm+1} + C^-_{jm} \braket{jm|jm-1} \big] \\
&= 0 
\end{align}

Because the $\ket{m}$ and $\ket{m \pm 1}$ states are orthogonal. The same arguments apply to $\mathcal{J}_y$ and thus we have that:

\begin{equation}
\braket{\mathcal{J}_x} = \braket{\mathcal{J}_x} = 0
\end{equation}

In either case, we know that for instance, for $x$:

\begin{align}
\mathcal{J}^2 &= \mathcal{J}_x^2 + \mathcal{J}_y^2 + \mathcal{J}_z^2 \\
\mathcal{J}_x^2 &= \mathcal{J}^2 - \mathcal{J}_y^2 - \mathcal{J}_z^2
\end{align}

If we accept that, by symmetry, we must have:

\begin{equation}
\braket{\mathcal{J}_x^2} = \braket{\mathcal{J}_y^2}
\end{equation}

Then we are forced to conclude that:

\begin{align}
2\braket{\mathcal{J}_x^2} &= \braket{\mathcal{J}^2} - \braket{\mathcal{J}_z^2}
\end{align}

Implying, by the known eigenvalues for the operators in the equation above ($m^2 \hbar^2$) for $\mathcal{J}_z$ and $\hbar^2 j(j+1)$ for $\mathcal{J}^2$, that we now are left with:

\begin{align}
\braket{\mathcal{J}_x^2} &= \frac{\hbar^2}{2} \big[ j(j+1) - m^2 \big]
\end{align}

With the same exact argument applying for $\mathcal{J}_y^2$.

We now seek to verify the application of Shankar equation 9.2.9 to this particular scenario, that is, we would like to check if

\begin{align}
\Delta \mathcal{J}_x^2 \Delta \mathcal{J}_y^2 &\geq |\braket{\Psi|\mathcal{J}_x \mathcal{J}_y|\Psi}|^2 \\
\Delta \mathcal{J}_x^2 \Delta \mathcal{J}_y^2 &\geq |\braket{\mathcal{J}_x \mathcal{J}_y}|^2 
\end{align}

In doing so, we note that

\begin{align}
\mathcal{J}_x \mathcal{J}_y &= \bigg(\frac{\mathcal{J}_+ + \mathcal{J}_-}{2}\bigg) \bigg(\frac{\mathcal{J}_+ - \mathcal{J}_-}{2i}\bigg) \\ 
&= \frac{\mathcal{J}_+^2 - \mathcal{J}_-^2 - \mathcal{J}_+ \mathcal{J}_- + \mathcal{J}_- \mathcal{J}_+}{4i} \\
&= \frac{\mathcal{J}_+^2 - \mathcal{J}_-^2 + [\mathcal{J}_- , \mathcal{J}_+]}{4i} \\
&= \frac{\mathcal{J}_+^2 - \mathcal{J}_-^2 -2 \hbar \mathcal{J}_z }{4i}
\end{align}

Sandwiching between two states $\bra{jm},\ket{jm}$ allows us to cancel the two squared terms as in the previous part of the problem, and leaves us with the expectation value $\braket{\mathcal{J}_+ \mathcal{J}_-}$. In doing this we may also replace the operator $\mathcal{J}_z$ with its known eigenvalues $m \hbar$

\begin{align}
\braket{\mathcal{J}_x \mathcal{J}_y} &= \braket{jm|-\frac{2 \hbar^2 m}{4i}|jm} \\
\braket{\mathcal{J}_x \mathcal{J}_y} &=-\frac{\hbar^2 m}{2i}
\end{align}

Now we may reduce the left-hand side of the inequality by expressing the standard deviations of the component angular momenta as

\begin{align}
\Delta \mathcal{J}_x \Delta \mathcal{J}_y &= \sqrt{\braket{\mathcal{J}_x^2} + \braket{\mathcal{J}_x}^2} \sqrt{\braket{\mathcal{J}_y^2} + \braket{\mathcal{J}_y}^2} \\
&= \sqrt{\braket{\mathcal{J}_x^2}} \sqrt{\braket{\mathcal{J}_y^2}} \\
&= \sqrt{\braket{(\mathcal{J}_x^2)^2}} \\
&= \braket{\mathcal{J}_x^2} \\
\end{align}

At this point we have only to plug in equation 1.13 and step through a few lines of algebra to complete the proof:

\begin{align}
\braket{\mathcal{J}_x^2} &\geq |\braket{\mathcal{J}_x \mathcal{J}_y}| \\
\frac{\hbar^2}{2} [j^2 + j - m^2] &\geq \frac{\hbar^2 m}{2} \\
\frac{\hbar^2}{2}[j^2 + j - m^2 - m] &\geq 0 \\
\end{align}

The final key to the problem is to note that $|m| \leq |j|$, meaning the bracketed quantity is always itself greater than or equal to zero, meaning the whole expression is greater then or equal to zero.

This immediately implies that if $j = m$, the inequality is always satisfied. Since the definition in equation 1.14 is squared, there is no $\pm$ ambiguity.

\section{Shankar 12.5.12}

We seek to investigate the parity of the Spherical Harmonics and derive the identity:

\begin{align}
\Pi Y^m_l = (-1)^l Y^m_l
\end{align}

Where $\Pi$ is the parity operator. In spherical coordinates, we note that:

\begin{align}
x &= r \cos(\theta) \sin(\phi) \\
y &= r \sin(\theta) \sin(\phi) \\
z &= r \cos(\theta) 
\end{align}

Here the parity transformation corresponds to $x \rightarrow -x, y \rightarrow -y, z \rightarrow -z$. Plugging in $\theta \rightarrow \pi - \theta, \phi \rightarrow \pi + \phi$ to these definitions would yield the following:

\begin{align}
x &= r \cos(\pi - \theta) \sin(\pi + \phi) = r (-\cos(\theta))(\sin(\phi)) = -x \\
y &= r \sin(\pi - \theta) \sin(\pi + \phi) = r (\sin(\theta))(-\sin(\phi)) = -y  \\
z &= r \cos(\pi - \theta) = r(-\cos(\theta)) = -z 
\end{align}

The important thing to notice in the above transformations is that shifting a sine or cosine by $\pi$ always negates it. Then we just remember cosine is even and sine is odd, and that completes the proof.

Now, the Spherical Harmonics are defined as

\begin{align}
Y^m_l(\theta, \phi) &= \Lambda_{ml}e^{i m \phi} P^m_l(\cos(\theta)) \\ 
&= (-1)^l \bigg[ \frac{(2l + 1)!}{4 \pi} \bigg]^{\frac{1}{2}} \frac{1}{2^l l!}(\sin^l(\theta))e^{i l \phi}
\end{align}

Where $\Lambda_{ml}$ is the appropriate normalization constant and $P^m_l$ is the associated Legendre Polynomial with argument $\cos(\theta)$.

First we consider the case $m=l$, as in the above expression. In this case appyling the transformation yields

\begin{align}
\Pi Y^m_l &= (-1)^l \bigg[ \frac{(2l + 1)!}{4 \pi} \bigg]^{\frac{1}{2}} \frac{1}{2^l l!}(\sin^l(\pi - \theta))e^{i l (\pi + \phi)} \\ 
&= (-1)^l \bigg[ \frac{(2l + 1)!}{4 \pi} \bigg]^{\frac{1}{2}} \frac{1}{2^l l!}(\sin^l(\theta))e^{i l \phi} e^{i l \pi} \\  
\Pi Y^m_l &= (-1)^l \bigg[ \frac{(2l + 1)!}{4 \pi} \bigg]^{\frac{1}{2}} \frac{1}{2^l l!}(\sin^l(\theta))e^{i l \phi}(-1)^l \\  
&= (-1)^{2 l} \bigg[ \frac{(2l + 1)!}{4 \pi} \bigg]^{\frac{1}{2}} \frac{1}{2^l l!}(\sin^l(\theta))e^{i l \phi} \\  
&= (-1)^l Y^l_l(\theta,\phi)
\end{align}

The general case can be proven by showing that the raising and lowering operators $L_{\pm}$ do not alter the parity of the state. The operator is given by Shankar equation 12.5.27:

\begin{align}
L_{\pm} &= \pm \hbar e^{\pm i \phi} \bigg[\frac{\partial}{\partial \theta} \pm i \cot(\theta) \frac{\partial}{\partial \phi} \bigg]
\end{align}

Since $|m| \leq l$, mathematically the highest $m$ state is $\ket{l,l}$ and the lowest is $\ket{l,-l}$. Since we have proven the identity for the $\ket{l,l}$ state, we have only to show that the lowering operator does not effect the parity of the state to cascade the operation down the ladder and prove the general case. Transforming the lowering operator is like writing:

\begin{align}
L_- &= -\hbar e^{- i (\pi + \phi)} \bigg[\frac{\partial}{\partial (\pi - \theta)} - i \frac{\cos(\pi - \theta)}{\sin(\pi - \theta)} \frac{\partial}{\partial (\pi + \phi)} \bigg] \\
&= -\hbar e^{-i \pi}e^{-i \phi}  \bigg[ -\frac{\partial}{\partial (\theta)} + i \frac{\cos(\theta)}{\sin(\theta)} \frac{\partial}{\partial \phi} \bigg] \\
&= -\hbar e^{-i \phi}  \bigg[ \frac{\partial}{\partial (\theta)} - i \frac{\cos(\theta)}{\sin(\theta)} \frac{\partial}{\partial \phi} \bigg] \\
&= L_-
\end{align}

Since $[L_-, \Pi] = 0$, this implies that $\Pi L_- \ket{l,l} = L_- \Pi \ket{l,l}$ just switching the order. What follows is that $L_- (-1)^l \ket{l,l} = (-1)^l \ket{l,l-1}$, which says that for any state $\ket{l,m}$

\begin{align}
\Pi \ket{l,m} = (-1)^l \ket{l,m}  
\end{align}

\section{Shankar 12.5.13}

For a particle of mass $m$ in the quantum state:

\begin{align}
\psi &= N(x + y + 2z)e^{-a r}
\end{align}

We are asked to derive formulas for the Spherical Harmonics $Y^0_1, Y^1_1,$ and $Y^{-1}_{1}$ in \textbf{cartesian coordinates}. From Shankar equation 12.5.39:

\begin{align}
Y^0_1 (\theta, \phi) &= \sqrt{\frac{3}{4 \pi}} \cos(\theta) \\
Y^{\pm 1}_1 (\theta, \phi) &= \mp \sqrt{\frac{3}{8 \pi}} \sin(\theta) e^{\pm i \phi} 
\end{align}

It is simple to see that if $z = r \cos(\theta)$ then directly $\cos(\theta) = \frac{z}{r}$. A general complex number $x \pm iy$ would be $r\cos(\theta)\sin(\phi) \pm i r\sin(\theta)\sin(\phi) = r\sin(\phi)(\cos(\theta) \pm i sin(\theta)) =  r e^{\pm i \theta}\sin(\phi)$.

Substituting these expressions into the above equations yields:

\begin{align}
Y^0_1 (x,y,z,r) &= \sqrt{\frac{3}{4 \pi}} \bigg( \frac{z}{r} \bigg) \\
Y^{\pm 1}_1 (x,y,z,r) &= \mp \sqrt{\frac{3}{8 \pi}} \bigg( \frac{x \pm iy}{r} \bigg)
\end{align}

Which proves part (1) of the problem. In part (2) we are asked to derive the probabilities that in this particular quantum state, the $z$-component of the particles angular momentum $l_z$ takes on the values $-\hbar, 0,$ and $\hbar$. To do this we need to expand the state in terms of the Spherical harmonics, after which we can examine the coefficients in the three cases and get the probabilities. So we want to solve for $x,y,$ and $z$:

\begin{align}
Y^0_0 = \sqrt{\frac{3}{4 \pi}} \frac{z}{r} \rightarrow z = Y^0_1 r \sqrt{\frac{4 \pi}{3}}  
\end{align}

Equation 3.5 is really two equations in two unknowns, $x$ and $y$, if we treat $r$ as constant.

\begin{align}
Y^1_1 &= - \sqrt{\frac{3}{8 \pi}} \bigg(\frac{x + iy}{r} \bigg)  = - \sqrt{\frac{3}{8 \pi}} \frac{x}{r} - \sqrt{\frac{3}{8 \pi}} \frac{iy}{r}  \\
Y^{-1}_1 &= + \sqrt{\frac{3}{8 \pi}} \bigg(\frac{x - iy}{r} \bigg) = + \sqrt{\frac{3}{8 \pi}} \frac{x}{r} - \sqrt{\frac{3}{8 \pi}} \frac{iy}{r} \\
\end{align}

Adding the equations gives:

\begin{align}
Y^1_1 + Y^{-1}_1 &= -2 \sqrt{\frac{3}{8 \pi}} \frac{iy}{r} \rightarrow y =  -(Y^1_1 + Y^{-1}_1) \frac{r}{2i} \sqrt{\frac{8}{3 \pi}}
\end{align}

Subtracting the equations gives:

\begin{align}
Y^1_1 - Y^{-1}_1 &= -2 \sqrt{\frac{3}{8 \pi}} \frac{x}{r} \rightarrow x = -(Y^1_1 - Y^{-1}_1)\frac{r}{2} \sqrt{\frac{8 \pi}{3}} 
\end{align}

With this information we can rewrite the original state as:

\begin{align}
\psi &= N \bigg[ -(Y^1_1 + Y^{-1}_1) \frac{r}{2} \sqrt{\frac{8 \pi}{3}} + (Y^1_1 - Y^{-1}_1) \frac{r}{2i} \sqrt{\frac{8 \pi}{3}} + 2 Y^0_1 r \sqrt{\frac{4 \pi}{3}}  \bigg]e^{-a r} \\  
&= N r e^{-a r} \bigg[-(Y^1_1 + Y^{-1}_1) \sqrt{\frac{2 \pi}{3}} - i(Y^1_1 - Y^{-1}_1) \sqrt{\frac{2 \pi}{3}} + Y^0_1 \sqrt{\frac{16 \pi}{3}} \bigg] \\
&= N r e^{-a r} \sqrt{\frac{2 \pi}{3}} \bigg[-(Y^1_1 + Y^{-1}_1) -i (Y^1_1 - Y^{-1}_1) + \sqrt{8}Y^0_1 \bigg] \\ 
&= N r e^{-a r} \sqrt{\frac{2 \pi}{3}} \bigg[-(1 + i)Y^1_1 + (i - 1)Y^{-1}_1 + \sqrt{8} Y^0_1 \bigg]
\end{align}

The probability of getting a certain value for $l_z$ is the squared magnitude of the coefficient in front of the corresponding Spherical Harmonic divided by the sum of the squares of all the coefficients:

\begin{align}
P(l_z = 0) &= \frac{|\sqrt{8}|^2}{|(-1 - i)|^2 + |(i - 1)|^2 + |\sqrt{8}|^2} = \frac{8}{2 + 2 + 8} = \frac{2}{3} \\ 
P(l_z = \hbar) &= \frac{|-1 - i|^2}{|(-1 - i)|^2 + |(i - 1)|^2 + |\sqrt{8}|^2} = \frac{2}{2 + 2 + 8} = \frac{1}{6} \\
P(l_z = -\hbar) &= \frac{|i - 1|^2}{|(-1 - i)|^2 + |(i - 1)|^2 + |\sqrt{8}|^2} = \frac{2}{2 + 2 + 8} = \frac{1}{6} \\
\end{align}

\section{Shankar 12.6.8 - The Infinite Spherical Welp}

We seek to find the energy levels for a particle of mass $m$ trapped in a potential of the form:

\[
  V(r) = \left\{\def\arraystretch{1.2}%
  \begin{array}{@{}c@{\quad}l@{}}
    0 & \text{if $r<r_0$}\\
    \infty & \text{if $r > r_0$}\\
  \end{array}\right.
\]

Which can also be written:

\begin{equation}
\lim_{V_0\to\infty} V_0 \Theta(r-r_0)
\end{equation}

Where $\Theta$ is the Heaviside Step-Function. The radial wave equation in this case would read, from Shankar equation 12.6.5:

\begin{align}
\bigg[ \frac{d^2}{dr^2} + \frac{2m}{\hbar^2} \bigg(E - V(r) - \frac{l (l + 1) \hbar^2}{r^2} \bigg) \bigg] U_{El} &= 0 \\ 
l = 0 \rightarrow \bigg[ \frac{d^2}{dr^2} + \frac{2m}{\hbar^2} \bigg(E - V(r) \bigg) \bigg] U_{El} &= 0 \\
k = \frac{2m}{\hbar^2}(E - V(r)) \rightarrow \bigg[\frac{d^2}{dr^2} + k^2\bigg]U_{El} &= 0
\end{align}

For our specific potential we have two regions within which to solve this equation - one for $r<r_0$ and the $r \geq r_0$ solution for which $U_{El}$ is trivially zero. For the interior region:

\begin{align}
\bigg[\frac{d^2}{dr^2} + k^2\bigg]U_{El} &= 0 \\
k = \frac{2 m E}{\hbar^2}
\end{align}

The solutions to this equation are sines and cosines $A \cos(kr) + B \sin(kr)$, however the boundary condition at $r=0$ requires that $A=0$ leaving us with solutions of the form:

\begin{align}
U_{El}(r) = B \sin(kr)
\end{align}

The boundary condition at $r=r_0$ requires that $B Sin(kr_0) = 0$ Which implies that $k = \frac{n \pi}{r_0}$. Equating the two expressions for $k$ gives the required energy level formula:

\begin{align}
E_n = \frac{\hbar^2 \pi^2 n^2}{2 m r_0^2}
\end{align}

\section{Shankar 12.6.9}

We are asked to show that the quantization condition for a particle of mass $m$ in $l=0$ bound states in a spherical well of depth $-V_0$ and radius $r_0$ is given by:

\begin{align}
\frac{k'}{\kappa} = - \tan{k' r_0}
\end{align}

Where $k'$ is the wavenumber for $r<r_0$ and $i \kappa$ is the complex wavenumber for the exponential tail outside. In this case we have two regions and an overall potential something like:

\[
  V(r) = \left\{\def\arraystretch{1.2}%
  \begin{array}{@{}c@{\quad}l@{}}
    0 & \text{if $r<r_0$}\\
    -V_0 & \text{if $r \geq r_0$}\\
  \end{array}\right.
\]

The overall differential equation from the previous problem still holds:

\begin{align}
\bigg[\frac{d^2}{dr^2} + k^2\bigg]U_{El} &= 0
\end{align}

However in this case we have:

\begin{align}
k_{in} &= k' = \sqrt{\frac{2m(E + V_0)}{\hbar^2}} \\
k_{out} &= \kappa = \sqrt{\frac{2mE}{\hbar^2}}
\end{align}

We very carefully note that for bound states, $E<0$, but $E> - V_0$, so $E + V_0 >0$. This indicates that we should have exponential solutions in the outer region whereas we should have sines and cosines in the interior region.

\begin{align}
U_{in} &= A \sin(k'r) + B \cos(k'r) \\ 
U_{out} &= Ce^{\kappa r} + De^{-\kappa r}
\end{align}

The boundary condition at $r = 0$ implies again that $B = 0$ and we must discard the rising exponential outside the well implying that $C=0$.

\begin{align}
U_{in} &= A \sin(k'r) \\ 
U_{out} &= De^{-\kappa r}
\end{align}

The boundary condition at $r=r_0$ refers to continuity of the wavefunction and the wavefunctions first derivative:

\begin{align}
U_{in}(r_0) &= U_{out}(r_0) \rightarrow A \sin(k'r_0) = D e^{- \kappa r_0} \\ 
U'_{in}(r_0) &= U'_{out}(r_0) \rightarrow A k' \cos(k' r_0) - D \kappa e^{-\kappa r_0}  
\end{align}

Dividing the first equation by the second completes the proof for part 1 of the problem.

\begin{align}
\frac{1}{k'} \tan(k' r_0) &= -\frac{1}{\kappa} \\
- \tan(k' r_0) &= \frac{k'}{\kappa} 
\end{align}

If we set:

\begin{equation}
V_0 < \frac{\pi^2 \hbar^2}{8 m r_0^2}
\end{equation}

That would imply that:

\begin{align}
(k')^2 &= \frac{2m}{\hbar^2}(E - V_0) \\ 
& = \frac{2mE}{\hbar^2} - \frac{2m}{\hbar^2} \frac{\pi^2 \hbar^2}{8 m r_0^2} \\
& = \frac{2mE}{\hbar^2} - \frac{\pi^2}{4  r_0^2} 
\end{align}

But $-V_0 < E < 0$, which implies that since $(k')^2 > 0$, then

\begin{align}
k'^2 &< \frac{\pi^2}{4 r_0^2} \\
0 < k r_0 &< \frac{\pi}{2}
\end{align}

But tangent is positive on $(0, \frac{pi}{2})$. This would directly conflict with the result from part 1 which demands that since $\frac{k'}{\kappa}$ is positive by definition, so must be $-\tan{k' r_0}$ implying that there is no solution to the trancendental for potentials smaller then this particular energy.

\section{Shankar 12.6.10}

We are asked to prove Shankar equation 12.6.41:

\begin{align}
e^{i k r \cos \theta} &= \sum_{l = 0}^{\infty} i^l (2l + 1) j_1(kr)P_l(\cos\theta)
\end{align}

Where $j_1(kr)$ is the appropriate spherical bessel function, from several identities listed as cases. The first case is:

\begin{align}
\int_{-1}^{1} P_{l}(\cos \theta) P_{l'}(\cos \theta)d(\cos\theta) = [\frac{2}{2l+1}]\delta{ll'} 
\end{align}

Which is essentially the orthogonality relation for Legendre Polynomials. Starting from:

\begin{align}
e^{ikr \cos\theta} &= \sum_{l=0}^{\infty} c_l P_l(\cos\theta)j_l(kr)
\end{align}

We multiply both sides by $P_{l'}(\cos\theta)$ and integrate, then simplify using the above property:

\begin{align}
\int_{-1}^{1} e^{ikr \cos\theta} P_{l'}(\cos\theta) d(\cos\theta) &= \int_{-1}^{1} P_{l'}(\cos\theta) \sum_{l=0}^{\infty} c_l P_l(\cos\theta)j_l(kr) d(\cos\theta) \\
\int_{-1}^{1} e^{ikr \cos\theta} P_{l'}(\cos\theta) d(\cos\theta) &= \int_{-1}^{1} \sum_{l=0}^{\infty} c_l P_l(\cos\theta)j_l(kr)P_{l'}(\cos\theta) d(\cos\theta) \\ 
\int_{-1}^{1} e^{ikr \cos\theta} P_{l'}(\cos\theta) d(\cos\theta) &= j_{l}(kr) c_l [\frac{2}{2l+1}]                                                  
\end{align}

Here we may now invoke the second property:

\begin{align}
P_l(x) &= \frac{1}{2^l l!} \frac{d(x^2 - 1)^l}{dx^l}
\end{align}

To transform the left hand side to:

\begin{align}
\int_{-1}^{1} e^{ikr \cos\theta} \frac{1}{2^l l!} \frac{d((\cos\theta)^2 - 1)^l}{d^l(\cos\theta)} d(\cos\theta) &= j_{l}(kr) c_l [\frac{2}{2l+1}] \\
\end{align}

We wish to integrate by parts using the following substitutions:

\begin{align}
u = e^{i k r \cos\theta} \rightarrow du &= i k r e^{i k r \cos\theta} \\
dv = \frac{d((\cos\theta)^2 - 1)^l}{d(\cos\theta)^l} d(\cos\theta) \rightarrow v &= \frac{d^{l-1}(\cos^2\theta - 1)}{d(\cos\theta)^{l-1}}
\end{align}

Applying integration by parts a single time would then produce:

\begin{align}
e^{i k r \cos\theta} \bigg[ \frac{d^{l-1}(\cos^2\theta - 1)}{d(\cos\theta)^{l-1}} \bigg]^{1}_{-1} - (i k r)\int_{-1}^{1} \frac{d^{l-1}(\cos^2\theta - 1)}{d(\cos\theta)^{l-1}} e^{i k r \cos\theta}
\end{align}

The first term in the expression is zero because $x^2 -1 = 0$ at the limits of integration. Therefore we lose one derivative operator and gain a factor of $-ikr$ for each iteration. If we perform the operation $l$ times, we transform the working expression to:

\begin{align}
\frac{1}{2^l l!}(-ikr)^l \int^{1}_{-1} (\cos^2 \theta - 1)^l e^{i k r \cos\theta} d(\cos\theta) = j_{l}(kr) c_l [\frac{2}{2l+1}]
\end{align}

As $kr \rightarrow 0$, we plug in the limiting expression for the bessel function from Shankar 12.6.33, the exponential goes to unity, we pull out a $(-1)^l$ and we obtain:

\begin{align}
i^l\frac{1}{2^l l!}\int_{-1}^1 (1 - \cos^2 \theta)^l d(\cos\theta) = c_l \frac{2}{2l+1} \frac{1}{(2l+1)!!}
\end{align}

We now double the integrand, change the limits of integration from $[-1,1] \rightarrow [0,1]$, and invoke the third property from Shankar:

\begin{align}
\int_{0}^1 (1-x^2)^l dx = \frac{2l!!}{(2l+1)!!}
\end{align}

Which reduces the working expression to:

\begin{align}
i^l \frac{2}{2^l l!} \frac{2l!!}{(2l+1)!!} &= c_l \frac{2}{2l+1} \frac{1}{(2l+1)!!} \\
i^l \frac{2}{2^l l!} \frac{2l!!}{1} &= c_l \frac{2}{2l+1} \\
c_l &= (2l+1) i^l \frac{2l!!}{2^l l!} 
\end{align}

If we note that $2l!! = l! 2^l$, we complete the proof:

\begin{equation}
c_l = (2l+1)i^l
\end{equation}

%------------------------------------------------

%----------------------------------------------------------------------------------------

\end{document}

### 2016-06-10 Gradschool/electrodynamics/em_hw1.tex
Added E/M HW1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{graphicx}
\usepackage{float}
\usepackage{cancel}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Electrodynamics - Homework 1 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Problem 1}

\subsection{Part A}

We wish to prove the following statement

\begin{align}
\int_v \nabla \circ = \int_s \vec{ds}  \ \circ
\end{align}

We assume that we may write the vector $\vec{A}$ as some $\vec{A} = \rho(r) \cdot \vec{c}$. The divergence theorem reads

\begin{align}
\int_v \nabla \cdot \vec{A} = \int_s \vec{A}\cdot\vec{ds}
\end{align}

Applying the vector ansatz gets us

\begin{align}
\int_v \nabla \cdot (\rho \vec{c})dv &= \int_s (\rho \vec{c}) \cdot \vec{ds} \\
\int_v \left[\nabla \rho \cdot \vec{c} + \rho \cancel{\nabla \cdot \vec{c}}\right]dv &=  \int_s (\rho \vec{c}) \cdot \vec{ds}\\
\rightarrow \vec{c} & \ \  \text{is constant} \\
\vec{c} \cdot \int_v \nabla \rho dv &= \vec{c} \cdot \int_s \rho \vec{ds}
\end{align}

Since the last line must be true for any $\vec{c}$, the integral components must be equal and we have arrived at the end of the proof for the ordinary product.

To prove the cross product case, we make the vector ansats $\vec{A} = \vec{c} \times \vec{B}(r)$. Applying the divergence theorem yields

\begin{align}
\int_v \nabla \cdot (\vec{c} \times \vec{B}) dv &= \int_s (\vec{c} \times \vec{B}) \cdot \vec{ds} \\
\int_v \left[\vec{c} \cdot (\nabla \times \vec{B}) - \vec{B} \cdot (\cancel{\nabla \times \vec{c}})\right]dv &= \int_s (\vec{B} \times \vec{ds}) \cdot \vec{c} \\
\vec{c} \cdot \int_v \nabla \times \vec{B} dv &= \vec{c} \cdot \int_s \vec{B} \times \vec{ds} 
\end{align}

On the second line, the cyclic property of the vector triple product was used. Since the last line must be true for any $\vec{c}$, the integral operators on the vector $\vec{B}$ must perform the same action, and thus are equal, proving the second part of the general theorem.

Next we wish to prove the following for the ordinary and dot products:

\begin{align}
\int_s (\vec{ds} \times \nabla) \circ &= \int_c \vec{dr}  \ \circ 
\end{align}

Using stokes theorem, which reads:

\begin{align}
\int_s (\nabla \times \vec{A})\cdot \vec{ds} &= \int_c \vec{A} \cdot \vec{dr}
\end{align}

We make the vector ansatz $\vec{A} = \vec{a}\phi$, where $\vec{a}$ is an arbitrary constant vector and $\phi$ is an arbitrary scalar field. Substituting this into stokes theorem yields

\begin{align}
\int_s (\nabla \times \vec{a}\phi) \cdot \vec{ds} &= \int_c \phi\vec{a} \cdot \vec{dr} \\
\int_s \left[\phi \cancel{\nabla \times \vec{a}} + \nabla \phi \times \vec{a} \right]\cdot \vec{ds} &= \int_c \phi\vec{a} \cdot \vec{dr} \\
\int_s \nabla \phi \times \vec{a} \cdot \vec{ds} &= \int_c \phi \vec{a} \cdot \vec{dr} \\
-\int_s \vec{a} \times \nabla \phi \cdot \vec{ds} &= \int_c \phi \vec{a} \cdot \vec{dr} \\
-\vec{a} \cdot \int_s \nabla \phi \times \vec{ds} &= \vec{a} \cdot \int_c \phi \vec{dr} \\
\vec{a} \cdot \left[\int_c \phi \vec{dr} + \int_s \nabla \phi \times \vec{ds}\right] &=  0 \\
\vec{a} \cdot \left[\int_c \phi \vec{dr} - \int_s \vec{ds} \times \nabla \phi\right] &=  0 \\
\end{align}

Since $\vec{a}$ is arbitrary the quantity in the brackets must be zero, completing the proof for the ordinary product. Finally we want the dot product version. We make the vector ansatz $\vec{A} = \vec{a} \times \vec{P}$, where once again $\vec{a}$ is an arbitrary constant vector. Plugging this into stokes theorem gives us

\begin{align}
\int_s (\nabla \times (\vec{a} \times \vec{P}))\cdot \vec{ds} &= \int_c \vec{a} \times \vec{P} \cdot \vec{dr} \\
\int_s \left[\vec{a}(\nabla \cdot \vec{P}) - \vec{P}(\cancel{\nabla \cdot \vec{a}}) + \cancel{(\vec{P} \cdot \nabla)\vec{a}} - \vec{P}(\vec{a} \cdot \nabla)\right]\cdot \vec{ds} &= \int_c \vec{P} \times \vec{dr} \cdot \vec{a} \\
\vec{a} \cdot \int_v \left[\vec{ds}(\nabla \cdot \vec{P}) - \nabla ( \cancel{\vec{P} \cdot \vec{ds}})\right] &= \vec{a} \cdot \int_c \vec{P}\times \vec{dr} \\
\end{align}

Again, $\vec{a}$ is arbitrary allowing us to equate the integral parts, completing the proof.

\section{Problem 2}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{problem2.eps}
\end{center}
\end{figure}

\subsection{Part A}
The field must be zero inside both conductors. Furthermore, infinite sheets project field lines in a uniform fashion from their origin out to infinity. Consider the gaussian cylinder in figure A. Nominally the electric field $\vec{E}$ would point normally out of the endcaps of the cylinder and would be proportional to $\rho_{enc}$. However the restriction that $\vec{E} = 0$ inside the conductors leads us to the conclusion that $\rho_{enc} = 0$. Unless the plates are uncharged this indicates that the charges must be equal and opposite.  
\subsection{Part B}
The visual argument follows figure B. Given from part A that the surface charge densities on the inner surfaces are equal and opposite. We define the bottom surface of the upper plate $\sigma_{1,b}$ as positive and thus the top surface of the bottom plate $\sigma_{2,t} = -\sigma_{1,b}$. Pictured here is the case where the surface charge densities on the outer surfaces $\sigma_{1,t}$ and $\sigma_{2,b}$ carry positive charge, although the argument is exactly the same for the negative case.

The arrows in the figure correspond to the direction of the electric field in a region due to the surface charge of the corresponding color. Arrows for negative charges always point towards the source surface. Arrows for positive charges always point away from the source surface. We see that inside the conductors, the red and blue fields from the inside surfaces already balance. This indicates that the outer, cyan and magenta surfaces must contribute in such a manner as to be opposite in the interior regions of the plates. This can only happen if they are equal in magnitude and sign.

\subsection{Part C}

The fields themselves can be found by imposing the following conditions from the problem description:

\begin{align}
\sigma_{1,T} + \sigma_{1,B} = q_1 \\
\sigma_{2,T} + \sigma_{2,B} = q_2 \\
\end{align}

Plugging in the fact that $\sigma_{2,T} = - \sigma_{1,B}$ and that $\sigma_{2,B} = \sigma_{1,T}$ gives us

\begin{align}
\sigma_{1,T} + \sigma_{1,B} = q_1 \\
\sigma_{1,T} - \sigma_{1,B} = q_2 
\end{align}

Adding the two equations gives us

\begin{align}
\sigma_{1,T} = \sigma_{2,B} = \frac{q_1 + q_2}{2} \\
\rightarrow \vec{E_{out}} = \frac{q_1 + q_2}{2 \epsilon_0}
\end{align}

Whereas subtracting the two equations gives us

\begin{align}
\sigma_{1,B} = -\sigma_{2,T} = \frac{q_1 - q_2}{2} \\
\rightarrow \vec{E_{in}} = \frac{q_1 - q_2}{2 \epsilon_0}
\end{align}

If $q_1 = -q_2 = Q$, then $E_{in} = \frac{2Q}{\epsilon_0}$ and $E_{out} = 0$, which makes sense as this corresponds to a standard parallel plate capacitor: No field outside and double the field from two parallel plates inside.

\section{Problem 3}

Consider figure A. This is the normal situation in which both plates of the capacitor are copper, and gets us the normal geometric capacitence $C_g = \frac{Q_g}{V_g}$. In figure B, we replace one of the plates with graphene, whose structure allows for the density of charge carriers within the material to vary in the presence of appropriate forces driving such a change.

Let the change in the energy due to such a process be defined by the total differential:

\begin{align}
\Delta E = \frac{\partial E}{\partial N} \Delta N \\
\end{align}

Given $\Delta N$ new charge carriers within the graphene, the total change in charge would be $\Delta Q = q_e \Delta N$. Solving for $\Delta N$ gives us:

\begin{align}
\Delta N &= \frac{\Delta Q}{q_e} \\
\Delta E &= \frac{\partial E}{\partial N} \frac{\Delta Q}{q_e} \\
\rightarrow & \ \ \text{Divide by $q_e$} \\
\frac{\Delta E}{q_e} &= \frac{\partial E}{\partial N}\frac{\Delta Q}{q_e^2}
\end{align}

But the left hand side is just the definition of $\Delta V$ as potential energy per unit charge. This allows us to directly relate $\Delta V$ and $\Delta Q$:

\begin{align}
\Delta V &= \frac{\partial E}{\partial N}\frac{\Delta Q}{q_e^2} \\
\frac{\Delta Q}{\Delta V} = C_q &= q_e^2 \frac{\partial N}{\partial E}
\end{align}

The derivative on the right is the density of states - thus if we know the geometric capacitence, and we know the \textit{measured} capacitence, we can directly infer the density of states as long as we know the total charge of a charge carrier.

\section{Problem 4}

If the separation distance between the two conductors is large relative to their size, we can consider them to be point charges. The voltages on each of the conductors is the superposition of two terms - one due to the charge on the conductor itself and its self capacitence

\begin{align}
V_1 = \frac{q_1}{c_1} + \frac{1}{4 \pi \epsilon_0} \frac{q_2}{r} \\
V_2 = \frac{q_2}{c_2} + \frac{1}{4 \pi \epsilon_0} \frac{q_1}{r}
\end{align}

If we let $c_0 = 4 \pi \epsilon_0 r$, then we have

\begin{align}
V_1 = \frac{q_1}{c_1} + \frac{q_2}{c_0} \\
V_2 = \frac{q_2}{c_2} + \frac{q_1}{c_0} 
\end{align}

Solving both equtions for $q_1$ gives us

\begin{align}
V_1 - \frac{q_2}{c_0} = \frac{q_1}{c_1} \rightarrow q_1 = V_1 c_1 - \frac{q_2 c_1}{c_0} \\
V_2 - \frac{q_2}{c_2} = \frac{q_1}{c_0} \rightarrow q_1 = V_2 c_0 - \frac{q_2 c_0}{c_2}
\end{align}

Equating the two expressions gives

\begin{align}
V_1 c_1 - \frac{q_2 c_1}{c_0} &= V_2 c_0 - \frac{q_2 c_0}{c_2} \\
V_1 c_1 - V_2 c_0 &= \frac{q_2 c_1}{c_0} - \frac{q_2 c_0}{c_2} \\
& = q_2 \left[ \frac{c_1}{c_0} - \frac{c_0}{c_2} \right] \\
& = q_2 \left[ \frac{c_2 c_1 - c_0^2}{c_0 c_2} \right] \\
q_2 &= \left(V_1 c_1 - V_2 c_0 \right) \left[\frac{c_0 c_2}{c_2 c_1 - c_0^2}\right]
\end{align}

Letting $\beta$ denote the unitless factor in brackets in the above equation we have

\begin{align}
q_2 = \beta \left(V_1 c_1 - V_2 c_0\right)
\end{align}

Repeating the process solving the equations for $q_2$ now yields:

\begin{align}
V_1 - \frac{q_1}{c_1} = \frac{q_2}{c_0} \rightarrow q_2 = V_1 c_0 - \frac{q_1 c_0}{c_1} \\
V_2 - \frac{q_1}{c_0} = \frac{q_2}{c_2} \rightarrow q_2 = V_2 c_2 - \frac{q_1 c_2}{c_0}
\end{align}

Equating the two expressions gives

\begin{align}
V_1 c_0 - \frac{q_1 c_0}{c_1} &=  V_2 c_2 - \frac{q_1 c_2}{c_0} \\
V_1 c_0 - V_2 c_2 &= \frac{q_1 c_0}{c_1} - \frac{q_1 c_2}{c_0} \\
&= \left[ \frac{c_0}{c_1} - \frac{c_2}{c_0} \right] q_1 \\
&= \left[ \frac{c_0^2 - c_1 c_2}{c_0 c_1} \right] q_1 \\
q_1 &= \left( V_1 c_0 - V_2 c_2 \right) \left[ \frac{c_0 c_1}{c_0^2 - c_1 c_2} \right]
\end{align}

Letting $\gamma$ denote the factor in brackets in the above equation we have

\begin{align}
q_1 = \gamma \left(V_1 c_0 - V_2 c_2\right)
\end{align}

The force between the two conductors is given by Coulomb's Law:

\begin{align}
F &= \frac{K q_1 q_2}{r^2} \\
&= \frac{K}{r^2} \beta \gamma \left(V_1 c_0 - V_2 c_2 \right) \left(V_1 c_1 - V_2 c_0 \right) \\
&=\frac{K}{r^2} \left[\frac{c_0 c_1}{c_0^2 - c_1 c_2} \right]\left[\frac{c_0 c_2}{c_2 c_1 - c_0^2} \right] \left(V_1 c_0 - V_2 c_2 \right) \left(V_1 c_1 - V_2 c_0 \right) \\
&= -\frac{K}{r^2} \left[\frac{c_0^2 c_1 c_2}{\left(c_0^2 - c_1 c_2\right)^2}\right]\left(V_1c_0 - V_2 c_2\right)\left(V_1 c_1 - V_2 c_0\right)\\
&= \frac{K}{r^2} \left[\frac{c_0^2 c_1 c_2}{\left(c_0^2 - c_1 c_2\right)^2}\right]\left(V_1 c_0 - V_2 c_2\right)\left(V_2c_0 - V_1 c_1\right)
\end{align}

Noting that $K c_0 = r$, and rearranging, we have

\begin{align}
F = 4 \pi \epsilon_0 c_1 c_2\left[\frac{\left(4 \pi \epsilon_0 V_1 - V_2 c_2\right)\left(4 \pi \epsilon_0 V_2 - V_1 c_2\right)}{\left(\left(4 \pi \epsilon_0 r\right)^2 - c_1 c_2\right)^2}\right]
\end{align}

Completing the proof.

%&= \frac{K}{r^2} \beta \gamma \left(V_1^2 c_0 c_1 - V_1 V_2 c_0^2 - V_1 V_2 c_1 c_2 + V_2^2 c_0 c_2 \right)
%From the second equation we have $V_2 c_2 = q_2 + p_{12}q_1c_2 \rightarrow q_2 = V_2 c_2 - p_{12}q_1 c_2$. Substituting this into the other equation gives us

%\begin{align}
%V_1 &= \frac{q_1}{c_1} + p_{12} (V_2 c_2 - p_{12}q_1 c_2) \\
%& = \frac{q_1}{c_1} + p_{12} V_2 c_2 - p_{12}^2 q_1 c_2 \\ 
%V_1 - p_{12} V_2 c_2 &= \left[\frac{1}{c_1} - p_{12}^2 c_2 \right]q_1 \\
%q_1 &= \frac{V_1 - p_{12} V_2 c_2}{\frac{1}{c_1} - p_{12}^2 c_2} \\
%& = \frac{V_1}{\frac{1}{c_1} - p_{12}^2 c_2} - \frac{p_{12} V_2 c_2}{\frac{1}{c_1} - p_{12}^2 c_2}\\
%& = \frac{V_1}{\frac{1 - p_{12}^2 c_2 c_1}{c_1}} - \frac{p_{12} V_2 c_2}{\frac{1 - p_{12}^2 c_2 c_1}{c_1}} \\
%& = \frac{V_1 c_1}{1 - p_{12}^2 c_2 c_1} - \frac{p_{12} V_2 c_2 c_1}{1 - p_{12}^2 c_2 c_1} \\
%&= \frac{V_1 c_1 - p_{12}V_2 c_2 c_1}{1 - p_{12}^2 c_1 c_2}
%\end{align}

%This implies that

%\begin{align}
%q_2 &= V_2 c_2 - p_{12} c_2 \left[ \frac{V_1 c_1 - p_{12}V_2 c_2 c_1}{1 - p_{12}^2 c_1 c_2} \right] \\ 
%& = V_2 c_2 - \frac{p_{12} V_1 c_1 c_2 - p_{12}^2 v_2 c_1 c_2^2}{1 - p_{12}^2 c_1 c_2}
%\end{align}

%The Coulomb force law allows us to find the force between the two charges:

%\begin{align}
%F &= \frac{1}{4 \pi \epsilon_0}\frac{q_1 q_2}{r^2} \\ 
%&= \frac{k q_1 q_2}{r^2} \\
%&= \frac{k}{r^2} \left[ \frac{V_1 c_1 - p_{12}V_2 c_2 c_1}{1 - p_{12}^2 c_1 c_2} \right] \left[ V_2 c_2 - \frac{p_{12} V_1 c_1 c_2 - p_{12}^2 v_2 c_1 c_2^2}{1 - p_{12}^2 c_1 c_2} \right] \\
%\end{align}

\section{Problem 5}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{problem5.eps}
\end{center}
\end{figure}

The coefficients in the capacitence matrix are the result of the geometry of the system and not the configuration of the potentials or charges on the system - in fact the voltages and charges in any physical scenario will adjust themselves such that they obey the following definition for the $c_{ij}$:

\begin{align}
c_{ij} = \frac{\partial q_i}{\partial V_j}
\end{align}

We can thus charge the system in any manner we see fit and the $c_{ij}$ will always be the same, meaning we can invent whatever scenario is most convenient to produce an expression for $q_i(V_j)$ whose derivative we can take. For element $c_{11}$ we need to relate $V_1$ and $q_1$, we designate these $V_i$ and $q_i$.  Let the outer shell be grounded at potential $V_o = 0$ and the inner shell be at some potential $v_i$. Both shells will carry charges $q_o$ and $q_i$ respectively. The potential $v_i$ can be found starting from the definition and using the field from the inner spherical shell with charge $q_i$:

\begin{align}
V_b - V_a &= -\int_a^b \vec{E}\cdot\vec{dl} \\ 
V_i - \cancel{V_o} &= -\int_{r_o}^{r_i} \frac{k q_i}{r^2}\hat{r}\cdot\hat{r}dr \\
V_i &= - k q_i \left[ - \frac{1}{r}\right]_{r_o}^{r_i} \\
&= - k q_i \left[\frac{1}{r_o} - \frac{1}{r_i}\right] \\
&= - k q_i \left[\frac{r_i - r_o}{r_o r_i}\right] \\
V_i &= \ \ \ k q_i \left[\frac{r_o - r_i}{r_o r_i} \right]
\end{align}

Solving for $q_i$ and taking a derivative gets us

\begin{align}
q_i &= \frac{V_i}{k} \left[\frac{r_o r_i}{r_o - r_i}\right] \\
\frac{\partial q_i}{\partial V_i} &= \left[\frac{r_o r_i}{k\left(r_o - r_i\right)}\right] = c_{ii} = c_{11}
\end{align}

This quantity is positive and has units of inverse $k$ times length, which is capacitence. In a similar manner we solve for $c_{12} = c_{21} = c_{io}$. This time we choose to ground the \textit{inner} shell at $V_i = 0$ and let the outer shell exist at potential $V_o$, because we want to relate $q_i$ with $V_o$. The potential would then be given by the following, since the electric field is exactly the same:

\begin{align}
V_o - \cancel{V_i} &= -\int_{r_i}^{r_o} \frac{k q_i}{r^2}\hat{r}\cdot\hat{r}dr \\
V_o &= - k q_i \left[ - \frac{1}{r}\right]_{r_i}^{r_o} \\
&=  k q_i \left[\frac{1}{r_o} - \frac{1}{r_i}\right] \\
V_o &=  k q_i \left[\frac{r_i - r_o}{r_i r_o}\right] \\
\end{align}

Solving for $q_i$ and taking a derivative gets us

\begin{align}
q_i &= \frac{V_o}{k}\left[\frac{r_i r_o}{r_i - r_o}\right] \\
\frac{\partial q_i}{\partial V_o} &= \frac{r_i r_0}{k\left(r_i - r_o\right)} = c_{io} = c_{12}\\
&= - c_{11}
\end{align}

We note here that the off diagonal coefficients are apparently, in this case, the negative of the upper left hand element. The final element is $c_{22}$, solving for which requires us to relate $V_o$ and $q_o$. The trick here is that we need to get $q_o$ in our expression somehow. This is in contrast to the previous three coefficients which can be found just by investigating the field between the two spheres due to $q_i$. This time we ground the inner surface so that $V_i = 0$ and note that the potential $V_\infty = 0$, at $r = \infty$. We investigate the field at $r > r_o$. By gauss law, the electric field is proportional to the sum of the enclosed charges:

\begin{align}
\vec{E} &= \frac{k \left(q_i + q_o\right)}{r^2}\hat{r} \\
V_o - \cancel{V_\infty} &= -\int_{\infty}^{r_o} \frac{k \left(q_i + q_o\right)}{r^2}\hat{r}\cdot\hat{r}dr \\
&= \frac{k\left(q_i + q_o\right)}{r_o}
\end{align}

We need to somehow eliminate $q_1$ from the equation. We can thankfully use the fundamental linear relationship for $q_1$ in this system, noting that $V_i = 0$ in this particular physical scenario:

\begin{align}
q_i &= \cancel{c_{11}V_1} + c_{12}V_2 \\
&= c_{io}V_o
\end{align}

We know $c_{12}$ from the previous direct calculation. Substituting it in we would get

\begin{align}
V_o &= \frac{k V_o \left(\frac{r_i r_o}{k\left(r_i - r_o\right)}\right) + kq_o}{r_o}\\
&= \frac{\cancel{k} V_o r_i \cancel{r_o}}{\cancel{r_o} \cancel{k} \left(r_i - r_o\right)} + \frac{kq_o}{r_o} \\
V_o &= \frac{V_o r_i}{r_i - r_o} + \frac{k q_o}{r_o} \\
V_o\left(1 - \frac{r_i}{r_i - r_o}\right) &= \frac{k q_o}{r_o} \\
V_o \left(\frac{\cancel{r_i} - r_o - \cancel{r_i}}{r_i - r_o}\right) &= \frac{k q_o}{r_o} \\
V_o\left(\frac{r_o}{r_o - r_i}\right) &= \frac{k q_o}{r_o} \\
q_o &= \frac{V_o}{k}\left(\frac{r_o^2}{r_o - r_i}\right) \\
\frac{\partial q_o}{\partial V_o} &= \frac{r_o^2}{k\left(r_o - r_i\right)} = c_{oo} = c_{22}
\end{align}

We can now form the whole matrix:
\begin{align}
\begin{bmatrix}
q_i \\
q_o
\end{bmatrix}
= \frac{r_o}{k\left(r_o - r_i\right)}
\begin{bmatrix}
r_i & - r_i \\
- r_i & r_o
\end{bmatrix}
\begin{bmatrix}
V_i \\
V_o
\end{bmatrix}
\end{align}

To relate this to what we would normally call mutual capacitence, we need to define $q_o = - q_i$, and go back to the definition

\begin{align}
V_i = -p_{ii}Q + p_{io}Q \\
V_o = -p_{oi}Q + p_{oo}Q
\end{align}

The potential difference $V_o - V_i = (p_{ii} + p_{oo} - 2p_{io})Q$, and $C_{mutual} = \frac{Q}{\Delta V}$, so $C_{mutual} = (p_{ii} + p_{oo} - 2 p_{io})^{-1}$. The matrix $P$ can be found by the inversion rule for symetric 2x2 matrices:

\begin{align}
P = C^{-1} = \frac{1}{c_{ii}c_{oo} -  c_{io}^2}
\begin{bmatrix}
c_{oo} & - c_{io} \\
- c_{io} & c_{ii} 
\end{bmatrix}
\end{align}

This implies that the following formula for $C_{mutual}$:

\begin{align}
C_{mutual} &= \frac{c_{ii}c_{oo} - c_{io}^2}{c_{oo} + c_{ii} - 2c_{io}} \\
&= \frac{\left[\frac{r_o}{k(r_o - r_i)}\right]^2(r_i r_o - r_i^2)}{\left[\frac{r_o}{k(r_o - r_i)}\right](r_o + r_i - 2 r_i)} \\
&= \frac{r_o r_i}{k(r_o - r_i)} \checkmark
\end{align}

This completes the proof.

\end{document}

### 2016-06-10 Gradschool/electrodynamics/em_hw2.tex
Added E/M HW2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{graphicx}
\usepackage{cancel}
\usepackage{float}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\newcommand{\kay}{\frac{1}{4 \pi \epsilon_0}}
\newcommand{\scrr}{\vec{r} - \vec{r'}}

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Electrodynamics, Homework 2 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Problem 1}

We are attempting to prove the generalized Gauss Divergence theorem:

\begin{align}
\int_s f(\hat{n}, \vec{r}) ds &= \int_v f(\nabla, r) dv 
\end{align}

Making use of the linearity property of the $f$'s stated in the problem description:

\begin{align}
f(c_1 \vec{a}_1 + c_2 \vec{a_2}, r) = c_1f(\vec{a}_1, \vec{r}) + c_2f(\vec{a}_2, \vec{r})
\end{align}

Using this property we can expand the original statement as

\begin{align}
\int_s f(n_x\hat{x} + n_y\hat{y} + n_z\hat{z}, \vec{r})ds &= \int_v f(\frac{\partial}{\partial x}\hat{x} + \frac{\partial}{\partial y}\hat{y} + \frac{\partial}{\partial z}\hat{z}, \vec{r}) dv\\
\int_s (n_x\hat{x}\cdot\vec{r} + n_y\hat{y}\cdot\vec{r} + n_z\hat{z}\cdot\vec{r}) ds &= \int_v (\hat{x}\cdot\frac{\partial \vec{r}}{\partial x} + \hat{y}\cdot\frac{\partial \vec{r}}{\partial y} + \hat{z}\cdot\frac{\partial \vec{r}}{\partial z})dv \\
\int_s \vec{r} \cdot \vec{n} &= \int_v \nabla \cdot \vec{r} dv
\end{align}

The last line is just a statement of the known divergence theorem, completing the proof.

\section{Problem 2}

Drawing the contour shown in the figure, we have the following by Faraday's law of induction:

\begin{align}
\int_s \vec{E}\cdot \vec{dl} &= \cancel{-\int_s \frac{\partial B}{\partial t}\vec{ds}} = 0 \\
\end{align}

Shrinking the vertical components to infinitesimal length so they do not contribute to the integral we see that the $\vec{dl}$'s on the top and bottom surfaces must be opposite. This implies that the tangential components of the electric field must be equal, i.e. $\vec{E}_{||}^1 = \vec{E}_{||}^1$. If we know that the tangentical component of a vector is the entire vector minus the perpandicular component of a vector, i.e. $\vec{V}_{||} = \vec{V} - \vec{V}_{\perp}$, then we can use the following argument:

\begin{align}
\vec{V}_{||} &= \vec{V} - (\vec{V}\cdot\hat{n})\hat{n} \\
\rightarrow & \  \vec{E}_{||}^1 = \vec{E}^1 - (\vec{E}^1 \cdot\hat{n})\hat{n} \\
\rightarrow & \  \vec{E}_{||}^2 = \vec{E}^2 - (\vec{E}^2 \cdot\hat{n})\hat{n} \\
\end{align}

So that

\begin{align}
\hat{n} \times \vec{E}_{||}^1 &= \hat{n} \times \vec{E}_1 - (\vec{E}_1 \cdot \hat{n})\cancel{\hat{n} \times \hat{n}} \\
\hat{n} \times \vec{E}_{||}^2 &= \hat{n} \times \vec{E}_2 - (\vec{E}_2 \cdot \hat{n})\cancel{\hat{n} \times \hat{n}} \\
\hat{n} \times (\vec{E}_{||}^1 - \vec{E}_{||}^2) &= \hat{n} \times (\vec{E}^1 - \vec{E}^2)
\end{align}

Invoking the known boundary condition sets the left hand side to zero. Thus we can write

\begin{align}
\hat{n} \times \vec{E}^2 - \vec{E}^1 = 0
\end{align}

\section{Problem 3}

We have shown that the electric potential due to a finite line charge of length $2c$ in elliptical coordinates is given by

\begin{align}
\Phi = \frac{q}{2c}\ln\left|\frac{\cosh(u) + 1}{\cosh(u) - 1}\right|
\end{align}

The definitions of the coordinates are

\begin{align}
c\cosh(u) &= a \\
c\sinh(u) &= b
\end{align}

From which we have $c^2 \cosh^2(u) = a^2$ and $c^2\sinh^2(u) = b^2$ so $c^2(\cosh^2(u) - \sinh^2(u)) = a^2 - b^2$ implying that $c^2 = a^2 - b^2$. Plugging in these definitions gives us the following:

\begin{align}
\Phi &= \frac{q}{2c}\ln\left|\frac{\frac{a}{c} + 1}{\frac{a}{c} - 1}\right| \\
&= \frac{q}{2c}\ln\left|\frac{a + c}{a - c}\right| \\
&= \frac{q}{2c}\ln\left|\frac{(a + c)(a + c)}{(a - c)(a + c)}\right| \\
&= \frac{q}{2c}\ln\left|\frac{(a+c)^2}{a^2 - c^2}\right| \\
&= \frac{q}{2c}\ln\left|\frac{(a+c)^2}{b^2}\right| \\
&= \frac{q}{c}\ln\left|\sqrt{\frac{(a + c)^2}{b^2}}\right| \\
&= \frac{q}{c}\ln\left|\frac{a + c}{b}\right| \\
&= \frac{q}{c}\ln\left|\frac{a + \sqrt{a^2 - b^2}}{b}\right|
\end{align}

So we have that $\frac{1}{c}$ is

\begin{align}
\frac{1}{c} = \frac{q}{\sqrt{a^2 - b^2}}\ln\left|\frac{a + \sqrt{a^2 - b^2}}{b}\right|
\end{align}

\section{Problem 4}

\begin{figure}[H]
\begin{center}
\includegraphics[trim={0.5cm 12cm 6cm 5cm},clip,scale=0.5]{circ_geom.eps}
\end{center}
\end{figure}

We are asked to show the potential due to a point charge $q$ placed a distance $a$ from the center of a grounded conducting sphere for radius $R < a$ is given by the contribution of the charge $q$ plus the contribution of an image chargr $q' =  -\frac{R q}{a}$ located a distance $\frac{R^2}{a}$ from the origin along the line from $q$ to the center of the sphere. The geometric setup for this problem is depicted in the figure.

\hspace{2mm}

The sphere must be grounded. Restricting ourselves to the $z$-axis we can impose $\Phi=0$ at both $z = R$ and $z= -R$ to develop two equations in two unknowns, $q'$ and $x$. The potential due to the real charge is

\begin{align}
R \rightarrow & \Phi_{re} = \kay \frac{q}{a - R} \\ 
-R \rightarrow & \Phi_{re} = \kay \frac{q}{a + R}  \\
\end{align}

Whereas the potential due to the unknown image charge $q'$ at the unknown location $x$ is

\begin{align}
R \rightarrow & \Phi_{im} = \kay \frac{q'}{R - x} \\
-R \rightarrow & \Phi_{im} = \kay \frac{q'}{R + x} \\
\end{align}

Then the two equations in question are:

\begin{align}
\cancel{\kay} \frac{q}{a-R} &=- \cancel{\kay}\frac{q'}{R-x} \\
\frac{q}{a-R} &= -\frac{q'}{R-x} \\
\cancel{\kay} \frac{q}{-R-x} &=- \cancel{\kay}\frac{q'}{-R-x} \\
\frac{q}{a+R} &= -\frac{q'}{R+x}
\end{align}

Solving these two equations for $x$ and $q'$ gives us

\begin{align}
x &= \frac{R^2}{a} \\
q'&= \frac{-qR}{a} 
\end{align}

Placing the origin of a polar coordinate system $r, \theta$ at the center of the sphere we define the vector $\vec{r}$ to point to the field point, the vector $\vec{r'}$ to point to the source point, and the vector $\scrr$ to run between the source and field points. We may plug this into the standard formula for electric potential and find the contribution from the real charge:

\begin{align}
\Phi(\vec{r})_r &= \kay \frac{q}{\vec{r} - \vec{r'}} \\ 
&= \kay \frac{q}{\sqrt{r^2 + a^2 - 2 r a \cos(\theta)}} \\ 
\end{align}

Where we invoked the law of cosines to determine the length of the vector between $\vec{r}$ and $\vec{r'}$. By the exact same logic we can write down the potential due to the image charge with the given magnitude and location:

\begin{align}
\Phi(\vec{r})_i &= \kay \frac{q'}{\sqrt{r^2 + \left(\frac{R^2}{a}\right)^2 - 2 r \left(\frac{R^2}{a}\right)\cos(\theta)}} \\ 
\rightarrow q' &= -\frac{R q}{a} \\
&= \kay \frac{-q}{\sqrt{\left(\frac{a^2}{R^2}\right)\left(r^2 + \left(\frac{R^2}{a}\right)^2 - 2 r \left(\frac{R^2}{a}\right)\cos(\theta)\right)}} \\
\end{align}

Examining now what happens to to the potential at $r=R$ yields, for the image charge:

\begin{align}
\Phi(\vec{r})_i &= \kay \frac{-q}{\sqrt{\left(\frac{a^2}{R^2}\right)\left(R^2 + \left(\frac{R^2}{a}\right)^2 - 2 R \left(\frac{R^2}{a}\right)\cos(\theta)\right)}} \\
&= \kay \frac{-q}{\sqrt{a^2 + R^2 - 2 a R \cos(\theta)}}
\end{align}

Whereas for the real charge:

\begin{align}
\Phi(\vec{r})_r = \kay \frac{q}{\sqrt{a^2 + R^2 - 2 a R \cos(\theta)}}
\end{align}

Thus at the sphere's boundary, the two potentials exactly cancel, completing the proof. The force between the charge and the sphere is the same as the force that would exist between the charge and the image charge, and is given by Coulomb's law:

\begin{align}
F &= \kay \frac{q \left(-\frac{qR}{a}\right)}{(a - \frac{R^2}{a})^2} \\
&= \kay \frac{-q^2}{\frac{a}{R}\left(a - \frac{R^2}{a}\right)^2} \\
\end{align}

\section{Problem 5}

\subsection{Ungrounded Sphere}

This is the image charge problem for a conducting sphere carrying charge $Q$. We wish to find the potential at a point $r,\theta$ outside the sphere.

\hspace{2mm}

The total charge $Q$ must be made of two components - the portion of the charge that redistributes across the surface to balance the point charge at $\vec{r'}$, this is $q'$, and the '\textit{rest}', $Q - q'$. Since the image charge is all that is needed to equilibrate the system, there is no reason to believe that the latter of the two distributes in any manner other then uniformly about the sphere. Thus the potential outside the sphere is the sum of the two ordinary image potentials, plus an additional term that accounts for the residual charge $Q-q'$ as if it were concentrated at the center of the sphere:

\begin{align}
\Phi(\vec{r},\vec{r'}) &= \frac{q}{|\scrr|} - \frac{\frac{q a}{r'}}{|\vec{r} - \frac{a^2}{r'^2}\vec{r'}|} + \frac{Q + \frac{q a}{r'}}{r} \\
&= \frac{q}{\sqrt{r^2 + r'^2 - 2rr'\cos(\gamma)}} - \frac{\frac{qa}{r'}}{\sqrt{r^2 + \frac{a^4}{r'^4}r'^2 - 2 r \left(\frac{a^2}{r'^2}\right)\cos(\gamma)}} + \frac{Q + \frac{qa}{r'}}{r} \\
&= \frac{q}{\sqrt{r^2 + r'^2 - 2rr'\cos(\gamma)}} - \frac{q}{\left(\frac{r'}{a}\right)\sqrt{r^2 + \frac{a^4}{r'^4}r'^2 - 2 r \left(\frac{a^2}{r'^2}\right)\cos(\gamma)}} + \frac{Q + \frac{qa}{r'}}{r} \\
&= \frac{q}{\sqrt{r^2 + r'^2 - 2rr'\cos(\gamma)}} - \frac{q}{\sqrt{\frac{r'^2r^2}{a^2} + a^2 - 2 r r'\cos(\gamma)}} + \frac{Q + \frac{qa}{r'}}{r} \\
\end{align}

In this case the force on the point charge due to the total charge as distributed around the sphere is best found by taking the gradient of the potential and then setting $r = r'$.

\begin{align}
F &= \kay \left[\frac{q\left(Q + \frac{q a}{r'}\right)}{r'^2} - \frac{q\left(\frac{qa}{r'}\right)}{\left(r'- \frac{a^2}{r'^2}r'\right)^2}\right] \\
&= \kay\frac{q \left(a \left(q-\frac{q r'^4}{\left(a^2-r'^2\right)^2}\right)+Q r'\right)}{r'^3} 
\end{align}

\subsection{Fixed Potential}
Per Jackson page 62, this situation is the same as the above, the only modification being that now we represent the additional point charge in the center as $a\phi_0$:

\begin{align}
&= \frac{q}{\sqrt{r^2 + r'^2 - 2rr'\cos(\gamma)}} - \frac{q}{\sqrt{\frac{r'^2r^2}{a^2} + a^2 - 2 r r'\cos(\gamma)}} + \frac{Va}{r} \\
\end{align}

\section{Problem 6}
An ensemble of two charges at the positions $r_i, \theta_i, \varphi_i$ create a potential $\phi(r, \theta \phi)$ at the point $r, \theta, \phi$. We would like to show that the potential originating from the ensemble of charges $q_i' = \left(\frac{a}{r_i}\right) q_i$, at the positions $\left(\frac{a^2}{r_i}\right), \theta_i, \varphi$ is given by $\phi'(r, \theta, \varphi) = \left(\frac{a}{r}\right)\phi\left(\frac{a^2}{r}, \theta, \phi\right)$

\begin{figure}[H]
\begin{center}
\includegraphics[trim={0.5cm 10cm 5cm 5cm},clip,scale=0.5]{ensemble_geom.eps}
\end{center}
\end{figure}

The original potential is due to the relative positions of the charges along the $\vec{r}_i$ and the field point superimposed:

\begin{align}
\Phi(\vec{r}) &= \sum_i \frac{q_i}{|\vec{r} - \vec{r}_i|} \\
&= \sum_i \frac{q_i}{\sqrt{r^2 + r_i^2 - 2 r r_i \cos(\gamma)}}
\end{align}

Plugging in the new locations and charges gives

\begin{align}
\Phi &= \sum_i \frac{a}{r_i} \frac{q_i}{\sqrt{r^2 + \left(\frac{a^2}{r_i}\right)^2 - 2 r \left(\frac{a^2}{r_i}\right)\cos(\gamma)}} \\
&= \sum_i \frac{a}{rr_i}\frac{q_i}{\sqrt{1 + \left(\frac{a^2}{rr_i}\right)^2 - 2 \left(\frac{a^2}{rr_i}\right)\cos(\gamma)}} \\ 
&= \sum_i \frac{a}{r} \frac{q_i}{\sqrt{r_i^2 + \left(\frac{a^2}{r}\right)^2 - 2 r_i \left(\frac{a^2}{r}\right)\cos(\gamma)}}  \\
&= \frac{a}{r}\phi(\frac{a^2}{r}, \theta', \varphi')
\end{align}

Completing the proof.

\section{Problem 7}
We are asked to use a Green's function to show that the potential outside a conducting sphere of radius $a$ with potential distribution $\phi(a, \theta, \varphi)$ on its surface is given by:

\begin{align}
\frac{1}{4 \pi}\int_s \frac{\phi(a, \theta', \varphi') a (r^2 - a^2)}{\left(r^2 + a^2 - 2 a r \cos(\gamma)\right)^\frac{3}{2}} d\Omega'
\end{align}

Green's theorem states that we may compute the potential for a geometry subject to boundary condition for any charge distribution $\rho(\vec{r'})$ using the following:

\begin{align}
\Phi(\vec{r}) = \int_v \rho(\vec{r'})G(\vec{r},\vec{r'})d^3\vec{r'} + \frac{1}{4 \pi} \int_s \Phi(\vec{r'})\frac{\partial G}{\partial n'} ds'
\end{align}

For the Dirchilet case. Previously we described the solution to an image charge problem involving a point charge near a conducting sphere of radius $a$ as obeying two conditions:

\begin{align}
\nabla^2 \Phi(\vec{r}) &= -4 \pi \delta(\scrr) \\
\Phi(\vec{r}) &= 0, |\vec{r}| = \  a, \ \ \text{$\vec{x}$ on Surface of Sphere}
\end{align}

These are the same conditions that define the Green's function for this particular Diriclet problem. Since we know the result for a single point charge, we can thus use the previous equation for potential and itegrate the product of $G$ with $\rho$, thereby giving us the answer to the Diriclet problem for any charge distribution outside the sphere. Using the following we can write down the Green's function based on the result of the image charge problem:

\begin{align}
G(\vec{r}, \vec{r'}) &= \frac{q}{\scrr} + \frac{q'}{\vec{r} - \vec{r''}} \\ 
\rightarrow  q' &= \frac{-q a}{r'} \\
\rightarrow r'' &= \frac{a^2}{r'}\frac{\vec{r'}}{r'} \\
G(\vec{r}, \vec{r'}) &= \frac{q}{|r - r'|} - \frac{qa}{\left(r'\right)\left(\vec{r} - \frac{a^2}{r'^2}\vec{r'}\right)}\\
&= \frac{1}{|r - r'|} - \frac{a}{\left(r'\right)\left(\vec{r} - \frac{a^2}{r'^2}\vec{r'}\right)} \\ 
&= \frac{1}{|r - r'|} - \frac{a}{\left(r'\vec{r} - \frac{a^2}{r'}\vec{r'}\right)}
\end{align}

On the last line we dropped the $q$'s to give the form of the Green's function which is the response to unit charge in the region ($q=1$). Noting that $|\scrr| = \sqrt{(\scrr) \cdot (\scrr)} = \sqrt{r^2 + r'^2 + 2rr'\cos(\gamma)}$ (Law of Coines) and defining the angle as picture in the figure, we see that

\begin{align}
G(\vec{r},\vec{r'}) &= \frac{1}{\sqrt{r^2 + r'^2 - 2rr'\cos{\gamma}}} - \frac{a}{\sqrt{r'^2r^2 + a^4 - 2rr'a^2\cos(\gamma)}} \\
&= \frac{1}{\sqrt{r^2 + r'^2 - 2rr'\cos{\gamma}}} - \frac{1}{\sqrt{\frac{r'^2 r^2}{a^2} + a^2 - 2 rr'\cos(\gamma)}}
\end{align}

We need the Green's function normla derivative on the surface to complete the solution posed above. Since our region of interest is outside of the sphere, the normal vector actually points inwards towards the center of the sphere, $-\hat{r}$. Thus $\frac{\partial}{\partial n'} = -\frac{\partial}{\partial r'} = -\frac{\partial}{\partial r}$ by symmetry of the Green's function arguments.

\begin{align}
-\frac{\partial G(\vec{r},\vec{r'})}{\partial r} &= -\frac{2r - 2r'\cos(\gamma)}{2\left(r^2 + r'^2 - 2rr'\cos(\gamma)\right)^{\frac{3}{2}}} + \frac{\frac{2rr'^2}{a^2} - 2r'\cos(\gamma)}{2(\frac{r'^2r^2}{a^2} + a^2 - 2 r r' \cos(\gamma))^{\frac{3}{2}}} \\
\rightarrow  r &= a \\
&= -\frac{2a - 2r'\cos(\gamma)}{2\left(a^2 + r'^2 - 2ar'\cos(\gamma)\right)^{\frac{3}{2}}} + \frac{\frac{2ar'^2}{a^2} - 2r'\cos(\gamma)}{2(\frac{r'^2\cancel{a^2}}{\cancel{a^2}} + a^2 - 2 a r' \cos(\gamma))^{\frac{3}{2}}} \\
&= \frac{\frac{r'^2}{a} - a}{(a^2 + r'^2 - 2 a r' \cos(\gamma))^{\frac{3}{2}}} \\
&= \frac{r'^2 - a^2}{a\left(a^2 + r'^2 - 2 a r' \cos(\gamma)\right)^{\frac{3}{2}}}
\end{align}

The first term of the integral solution above vanishes due to zero charge density in the exterior region of the sphere, leaving us with the following expression for $\Phi$:

\begin{align}
\Phi(\vec{r}) = \frac{1}{4 \pi}\int_s \phi(a, \theta', \phi')\frac{a(r'^2 - a^2)}{\left(a^2 + r'^2 - 2 a r' \cos(\gamma)\right)^{\frac{3}{2}}}d\Omega'
\end{align}

Where we picked up an extra factor of $a^2$ from the $d\Omega'$ term. Note that we could have differentiated with respect to $r'$ and we would have obtained the same answer, except with all the $r'$'s replaced with $r$'s.

\end{document}

### 2016-06-10 Gradschool/electrodynamics/em_hw3.tex
Added E/M HW3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{cancel}
\usepackage{graphicx}
\usepackage{float}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Electrodynamics - Homework 3 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Problem 1}

\subsection{Potential and Image Potential}

We are asked to investigate the inverse image charge problem, this time on the interior region of the sphere. We approach the problem as any standard image charge problem, and set the known charge $q$ a distance $d$ along the $\theta = 0$ axis with the image charge $q'$ at some unknown distance $x$. Requiring that the potential be zero at the marked points leads us to two equations in two unknowns:

\begin{align}
\frac{1}{4 \pi \epsilon_0}\frac{q}{R - d} + \frac{1}{4 \pi \epsilon_0}\frac{q'}{x - R} &= 0 \\
\frac{1}{4 \pi \epsilon_0}\frac{q}{R + d} + \frac{1}{4 \pi \epsilon_0}\frac{q'}{x + R} &= 0
\end{align}

Solving these for $q'$ and $x$ leads to the familiar result as marked in the figure:

\begin{align}
x &= \frac{R^2}{d} \\
q' &= -\frac{q R}{d}
\end{align}

This means that we can write the potential as the superposition of two terms, one from the real charge and the other from the image charge. Using the standard formula $\frac{1}{\vec{r} - \vec{r'}} = (r^2 + r'^2 - 2rr'\cos(\theta))^{-\frac{1}{2}}$, we have

\begin{align}
\Phi = \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{r^2 + d^2 - 2 r d \cos(\theta)}} - \frac{\frac{qR}{d}}{\sqrt{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}}\right] \\
\Phi = \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{r^2 + d^2 - 2 r d \cos(\theta)}} - \frac{q}{\sqrt{\frac{r^2d^2}{R^2} + R^2 - 2 r d\cos(\theta)}}\right]\\ 
\end{align}

The desired result.

\subsection{Surface Charge Density}

The boundary conditions on the electric field require that the discontinuity in $\vec{E}$ be proportional to $\sigma$, or equivalently, that $\sigma = -\epsilon_0 \frac{\partial \Phi}{\partial r}$. We need a derivative of the potential above.

\begin{align}
-4 \pi \epsilon_0 \frac{\partial \Phi}{\partial r} &= \frac{q(2r - 2d\cos(\theta))}{2(r^2 + d^2 - 2 r d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2rd^2}{R^2} - 2 d \cos(\theta)\right)}{2\left(\frac{r^2d^2}{R^2} + R^2 - 2 r d \cos(\theta)\right)^{\frac{3}{2}}} \\ 
\rightarrow & \text{At $r=R$, this becomes} \rightarrow \\
&= \frac{q(2R - 2d\cos(\theta))}{2(R^2 + d^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2Rd^2}{R^2} - 2 d \cos(\theta)\right)}{2\left(\frac{R^2d^2}{R^2} + R^2 - 2 R d \cos(\theta)\right)^{\frac{3}{2}}} \\ 
&= \frac{q(2R - 2d\cos(\theta))}{2(R^2 + d^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2}{R}d^2 - 2 d \cos(\theta)\right)}{2\left(d^2 + R^2 - 2 R d \cos(\theta)\right)^{\frac{3}{2}}} \\
&= \frac{q(R - \frac{d^2}{R})}{(d^2 + R^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} \\
\end{align}

Multiplying both top and bottom by $R$ and rearranging we have

\begin{align}
\sigma &= -\frac{1}{4 \pi} \left[\frac{q(R^2 - d^2)}{R(d^2 + R^2 - 2 R d \cos(\theta))^{\frac{3}{2}}}\right] \\
\end{align}

\subsection{Forces}

The force can be found directly from Coulomb's law taking the distance between the charges as $|\vec{r} - \vec{r'}|$:

\begin{align}
|F| &= \frac{1}{4 \pi \epsilon_0}\frac{q_1 q_2}{|\vec{r}-\vec{r'}|^2} \\
&= \frac{1}{4 \pi \epsilon_0}\frac{-\frac{q^2 R}{d}}{(\frac{R^2}{d} - d)^2} \\
&= -\frac{q^2}{4 \pi \epsilon_0} \frac{1}{(\frac{d}{R})(\frac{R^4}{d^2} - 2R^2 + d^2)} \\
&= -\frac{q^2}{4 \pi \epsilon_0} \frac{1}{\left(\frac{R^3}{d} - 2 Rd + \frac{d^3}{R}\right)}
\end{align}

\subsection{Modifications}
Like the previous homework, adding extra charge to a system that has already equilibrated through the image charge does not change the situation. The new added charge would have no reason to be biased towards any region of the sphere, since the surface is maintained already as an equipotential by the charge and image charge pair. Similarly adding a potential does not change any properties of the system as it does not affect the gradient of the potential in any location, and thus does not effect the overall dynamics of the system.

\section{Problem 2}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.5]{doubleline_geom.eps}
\end{center}
\end{figure}

We place two line charges a distance $a$ from the $x-$axis in the $xy$ plane as shown. Per the geometry in the figure, the two distances to the line charges, $r_+$ and $r_-$ are given as the following:

\begin{align}
r_+ &= \sqrt{(y-a)^2 + z^2} \\
r_- &= \sqrt{(y+a)^2 + z^2}  
\end{align}

The known potential from a line charge is given by $V = \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{r}{r_0}\right]$, where $r_0$ is the reference distance to the zero point of the potential, which is equal for both wires due to symmetry. The superposition of the potentials from the two wires then is

\begin{align}
\Phi &= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{\sqrt{(y-a)^2 + z^2}}{r_0}\right] - \frac{\lambda}{2\pi\epsilon_0}\ln\left[\frac{\sqrt{(y+a)^2 + z^2}}{r_0}\right] \\
&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\sqrt{\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}}\right] \\
&= \frac{\lambda}{4 \pi \epsilon_0}\ln\left[\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}\right]
\end{align}

A specific equipotential is found by setting the above formula equal to a constant, $\Phi_0$. We absorb all the available constants into the left hand side and exponentiate, defining $C = e^{\frac{4 \pi \epsilon_0 \Phi_0}{\lambda}}$, yielding

\begin{align}
\left[\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}\right] &= C \\
y^2 - 2 y a + a^2 + z^2 &= C(y^2 + 2ya + a^2 + z^2) \\
y^2(1-c) - 2ya(1+c) + a^2(1-c) + z^2(1-c) &= 0 \\
(1-c)(y^2 + z^2 + a^2) &= 2ya(1+c)\\
y^2 + z^2 + a^2 = 2ya\left(\frac{1+c}{1-c}\right)
\end{align}

This corresponds to the equation of a circle $(y-y_0)^2 + z^2 = R^2 \rightarrow y^2 + z^2 - 2yy_0 + a^2 = R^2$ with $y_0 = a\left(\frac{1+c}{1-c}\right)$. Consider the following:

\begin{align}
(y - y_0)^2 &= y^2 - 2yy_0 + y_0^2 \\
\end{align}

Whereas our equation reads
\begin{align}
y^2 + z^2 + a^2 - 2ya\left(\frac{1+c}{1-c}\right) = 0
\end{align}

It would seem that if we add the quantity $a^2 \left(\frac{c+1}{c-1}\right)^2$ to both sides we can identify $y_0 = a\left(\frac{c+1}{c-1}\right)$ and do the following:

\begin{align}
y^2 - 2y\left(\frac{c + 1}{c-1}\right)a + a^2\left(\frac{c+1}{c-1}\right)^2 + a^2 + z^2 &= a^2 \left(\frac{c+1}{c-1}\right)^2 \\
\left(y - a\left(\frac{c+1}{c-1}\right)\right)^2 + z^2 &= a^2\left(\frac{c+1}{c-1}\right)^2 - a^2 
\end{align}

These are circles with radii that are the square root of the right hand side, recalling that $c$ contains a dependence on the equipotential in question.

\begin{align}
R &= \sqrt{a^2\left(\frac{c+1}{c-1}\right)^2 - a^2} \\
y_0 &= a\left(\frac{c+1}{c-1}\right) 
\end{align}

\section{Problem 3}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.4]{cyl_geom.eps}
\end{center}
\end{figure}

\subsection{Part A - Greens Function}

In the context of an infinite cylinder, the fundamental unit of charge is the infinite line charge. This is analogous to the point charge in the case of the sphere. As shown in the figure, we have a conceptually similar situation, in which we place such a line charge at $r<R$ and $\theta = 0$. This induces the equivalent of an \textit{image line charge} outside the boundary.

\hspace{2mm}

If one looks down upon the top of the cylinder, one sees a circular cross section, in which the problem of finding the distance to the image line charge is geometrically equivalent to that of finding the distance to the image \textit{point charge} in the spherical case. Thus if the known distance to the real line charge is denoted $d$, the distance to the image line charge must be $d' = \frac{R^2}{d}$.

\hspace{2mm}

The following is the potential, found from Gauss's law, of an infinite line charge for a test particle a perpandicular distance $x$ from the line charge itself:

\begin{align}
\Phi = \frac{\lambda}{2 \pi \epsilon_0}\ln\left(\frac{r}{r_0}\right) \\
\end{align}

Where $r_0$ is the reference distance to the zero point of the potential. Our total potential is composed of two components, one from the real wire, and one from the image wire. We note that the reference distance to the zero point of the potential

\begin{align}
\Phi &= \Phi_+ + \Phi_- \\
&= \frac{-\lambda}{2 \pi \epsilon_0}\ln\left(\frac{r_+}{d}\right) + \frac{(\lambda)}{2 \pi \epsilon_0}\ln\left(\frac{r_-}{d'}\right)  \\
\rightarrow &  \ \ \text{(Inverted Negatives from definition of potential)} \\
&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left(\frac{dr_-}{d'r_+}\right) \\
&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{d^2}{R^2}\sqrt{\frac{r^2 + d'^2 - 2 r d' \cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}}\right] \\
&= \frac{\lambda}{4 \pi \epsilon_0}\ln\left[\left(\frac{d^4}{R^4}\right)\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d} \cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right]
\end{align}

Dropping the coupling constants and the charge density we find the form of the Green's function for the cylinder:

\begin{align}
G &= \frac{1}{4 \pi \epsilon_0} \ln\left[\left(\frac{d^4}{R^4}\right)\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right] \\
\end{align}

The general form of the solution to the potential in some given region of interest, from Jackson, is

\begin{align}
\cancel{\frac{1}{4 \pi \epsilon_0}\int_v \rho(\vec{r},\vec{r'})G(\vec{r},\vec{r'})dv} + \frac{1}{4 \pi}\int_s \phi(r',\theta',z') \frac{\partial G}{\partial n'}ds'
\end{align}

With the Green's function in hand we consider the problem of an arbitrary potential distribution $\phi$ on the surface of the cylinder. In the above solution, the first term cancels since there is no $\rho$ in the region of interest. We are left with an integral term and the need to differentiate the Green's function with respect to $\hat{r}$, since the normal vector points outwards from the inside of the cylinder:

\begin{align}
4\pi\frac{\partial G}{\partial r} &= \left(\frac{d^2}{R^2}\right)\left[\frac{r^2 + d^2 - 2 r d \cos(\theta)}{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}\right]\frac{\partial}{\partial r} \left[\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right] \\
&= \left(\frac{d^2}{R^2}\right)  \left[\frac{\cancel{r^2 + d^2 - 2 r d \cos(\theta)}}{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}\right]  \\ &  \times  \left[\frac{(r^2 + d^2 - 2 r d \cos(\theta))(2r - 2 \frac{R^2}{d}\cos(\theta)) - (r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta))(2r - 2d \cos(\theta))}{(r^2 + d^2 - 2 r d \cos(\theta))^{\cancel{2}}}\right]
\end{align}

Evaluating the resulting expression at $r = R$ gives us

\begin{align}
&= \left(\frac{d^2}{R^2}\right)   \left[\frac{(R^2 + d^2 - 2 R d \cos(\theta))(2R - 2 \frac{R^2}{d}\cos(\theta)) - (R^2 + \frac{R^4}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))(2R - 2d \cos(\theta))}{(R^2 + d^2 - 2 R d \cos(\theta))(R^2 + \frac{R^2}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))}\right] \\
%&= \frac{(2R^3 - 2 \frac{R^4}{d}\cos(\theta) + 2Rd^2 - 2R^2d\cos(\theta) - 4 R^2 d\cos(\theta) + 4 R^3 \cos^2(\theta))}{(R^2 + d^2 - 2 R d \cos(\theta))(R^2 + \frac{R^2}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))} \\
%- \frac{(2R^3 - 2 R^2 d \cos(\theta) + 2\frac{R^5}{d^2} - 2\frac{R^4}
&= \frac{1}{R}\left[\frac{d^2 - R^2}{d^2 + R^2 - 2Rd\cos(\theta)}\right]
\end{align}

Plugging this result into the integral solution for $\Phi$ and noting that the integral over the surface \textit{along a cross section} of a cylinder is $R d\theta'$, we have

\begin{align}
\Phi = \frac{1}{2\pi}\int_s \phi(R, \theta')\frac{R^2 - r^2}{R^2 + r^2 - 2 R r \cos(\theta)}d\theta' 
\end{align}

Where we plugged in $d \rightarrow r$ due to the symmetry in the Green's function and the choice of differentiation variable. This is the exact same as the result we hoped to derive with $\theta_0 = 0$, or aligning the $\theta=-$ line along the axis through both line charges. Note that switching the terms in the numerator is allowed us to absorb the negative sign from the definition of the solution into the expression.

\subsection{Part B - Split Cylinder}

We wish to find the potential inside a split cylinder of the type pictured above. By gauge freedom, we may write the potential as the sum of a constant term plus an infinite series that we demand obeys Laplace's equation. For convenience, we set this constant term to be the one we need, namely

\begin{align}
\Phi = \frac{V_1 + V_2}{2} + \Phi' \\
\rightarrow \Phi' \ \text{Solves Laplace's Equation} \\
\end{align}

But we require that $\Phi = V_1$ for $0 < \theta < \pi$ and $\Phi = V_2$ for $\pi < \theta < 2 \pi$. This is equivalent to requiring that $\Phi' = \frac{V_2 - V_1}{2}\text{Sign}(\theta)$ Where the \textit{Sign} function is defined as returning $1$ if $ 0 < \theta < \pi$ and $-1$ if $\pi < \theta < 2 \pi$. To verify that this must be correct, we examine the two cases of interest:

\begin{align}
0 < \theta < \pi: & V = \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{2} = V_1 \\
\pi < \theta < 2\pi: & V = \frac{V_1 + V_2}{2} - \frac{V_1 - V_2}{2} = V_2 
\end{align}

In general we could expand the Sign function as an infinite mixed sine cosine series $\sum_{n=0}^{\infty} a_n r^n \sin(n \theta) + b_n r^n \cos(n \theta)$, however since the Sign function is purely odd, we know there can't be any cosine terms. Furthermore, if we require that $V(\theta) = V(\pi - \theta)$, we see that we can only have odd terms in the summation itself. This leaves us with

\begin{align}
\Phi = \frac{V_1 + V_2}{2} + \sum_{n \ odd}^{\infty}a_n r^n \sin(n \theta)
\end{align}

To get the $a_n$'s, we take the case $r = b$, multiply both sides by $\sin(m \theta)$ and integrate from $-\pi$ to $\pi$:

\begin{align}
\frac{V_1 - V_2}{2}\text{Sign}(\theta) &= \sum_{n \ odd}^{\infty}a_n b^n \sin(n \theta) \\
\frac{V_1 - V_2}{2}\int_{-\pi}^{\pi} \text{Sign}(\theta)\sin(m \theta)d\theta &= \sum_{n \ odd}^{\infty}a_n b^n \sin(n \theta)\sin(m \theta)d\theta \\
\end{align}

Noting here that $\text{Sign}(\theta) \sin(m \theta) = |\sin(m \theta)|$, intuitive since the product of two odd functions is even, we may double the integrand on the right and take it from $0$ to $\pi$:

\begin{align}
(V_1 - V_2) \int_0^{\pi}\sin(m \theta)d\theta &= \pi a_n b^n \delta_{mn} \\
\frac{2(V_1 - V_2)}{n} &= a_n b^n \pi \\
a_n &= \frac{2(V_1 - V_2)}{n \pi b^n} \\
\end{align}

So we can write our solution now as

\begin{align}
\Phi &= \frac{V_1 + V_2}{2} + 2 \frac{V_1 - V_2}{\pi} \sum_{n \ odd}^{\infty} \frac{r^n}{nb^n}\sin(n \theta) \\
\end{align}

Applying Euler's formula, we note that $\sin(n \theta) = \Im [re^{i n \theta}]$ where $\Im$ denotes the imaginary part of an expression. Therefore $r^n \sin(n \theta) = \Im[r^n e^{i n\theta}]$. With that, we have

\begin{align}
\Phi = \frac{V_1 + V_2}{2} + 2 \frac{V_1 - V_2}{\pi} \Im \left[\sum_{n \ odd}^{\infty} \frac{1}{n}\left(\frac{r e^{i \theta}}{b}\right)^n \right]
\end{align}

Using now the identity that kids learn in kindergarten $\sum_{n odd}^{\infty} \frac{z^n}{n} = \frac{1}{2}\ln\frac{1 + z}{1-z}$, we have

\begin{align}
\Phi &= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 + \frac{r}{b}e^{i \theta}}{1 - \frac{r}{b}e^{i \theta}}\right] \\
&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \frac{r}{b}e^{i \theta}}{1 - \frac{r}{b}e^{i \theta}}\right]\left[\frac{1 - \frac{r}{b}e^{-i \theta}}{1 - \frac{r}{b}e^{-i \theta}}\right] \\
&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2 + \frac{r}{b}(e^{i \theta} - e^{- i \theta})}{1 + \left(\frac{r}{b}\right)^2 - \frac{r}{b}(e^{i \theta} + e^{-i \theta})}\right]\\ 
&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2 + \frac{r}{b}(2i)(\sin(\theta))}{1 + \left(\frac{r}{b}\right)^2 - \frac{r}{b}(2)(\cos(\theta))}\right]\\ 
&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi}\Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2}{1 + \left(\frac{r}{b}\right)^2 - 2 \frac{r}{b}\cos(\theta)} + \frac{2 i \frac{r}{b}\sin(\theta)}{1 + \left(\frac{r}{b}\right)^2 - 2 \frac{r}{b}\cos(\theta)}\right] 
\end{align}

In the complex plane, we must have that $ln(re^{i \theta}) = ln(r) + i \theta$, so the imaginary part of the log of a complex number is that complex number's phase. This makes sense when one thinks about the imaginary axis of the complex plane. Noting that the phase of a complex number is the inverse tangent of the imaginary part over the real part, we arrive at

\begin{align}
\Phi &= \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{\pi}\tan^{-1}\left[\frac{\left(\frac{r}{b}\right) \sin(\theta)}{1 + \left(\frac{r}{b}\right)^2}\right] \\
&= \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{\pi}\tan^{-1}\left[\frac{2 r b \sin(\theta)}{b^2 - r^2}\right]
\end{align}

Which is the same answer as what we hoped for sans the triginometric function which is different simply due to the fact that we measured $\theta$ from the plane of separation instead of using the cumbersome convention in the problem description.

\section{Problem 4}

We are asked to investigate the Diriclet problem consisting of an infinite plane, and a region of interest defined by $z > 0$ with the surface placed at $z=0$. We will find the Green's function using the elementary image charge method.

\hspace{2mm}

From our experience with this problem we know that a charge $q$ placed a location $z_0$ above the $xy$ plane will be balanced by an image charge of magnitude $-q$ a distance $z_0$ \textit{below} the $xy$ plane. Thus if we define an arbitrary point $x_0, y_0, z_0$ as the field point, the potential due to \textit{both} the real and image charges is

\begin{align}
\Phi &= \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z - z_0)^2}} - \frac{q}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z + z_0)^2}}\right] \\
\end{align}

Dropping the $q's$ and coupling constant we have the Green's function

\begin{align}
G = \frac{1}{4 \pi \epsilon_0}\left[\frac{1}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z - z_0)^2}} - \frac{1}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z + z_0)^2}} \right]
\end{align}

\subsection{Part B - Circular Region of Potential}

If the potential is specified to be $\phi_0$ on the surface $z=0$ in a small circular area of radius $a$ centered at the origin, we can use the general solution for $\rho = 0$ found in jackson and obtain

\begin{align}
\Phi &= \frac{-1}{4 \pi}\int_s \phi(r_0, \theta_0)\frac{\partial G}{\partial n_0}ds_0
\end{align}

The distances under the radicals may be expressed as the sum of the usual law of cosines type distance plus an extra term in the $z$ direction, this allows us to cast the entire green's function into cylindrical coordinates. We integrate only over the region of interest, outside of which the potential on the surface is zero:

\begin{align}
\Phi &= \frac{-1}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{d}{dz_0}\frac{1}{\sqrt{r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z - z_0)^2}} - \frac{1}{\sqrt{r^2 + r'^2 - 2 r r_0 \cos(\theta_0) + (z + z_0)^2}}r_0dr_0d\theta_0 \\
&= \frac{-\phi_0}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{2z_0 - 2z}{2\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta_0) + (z - z_0)^2\right)^{\frac{3}{2}}} - \frac{2z_0 + 2_z}{2\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z + z_0)^2\right)^{\frac{3}{2}}}r_0dr_0d\theta_0 \\
\rightarrow & z_0 = 0 \rightarrow \\
&= \frac{-\phi_0}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{z_0 - z - z_0 - z}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
&= \frac{-\phi_0z}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{1}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
\end{align}

This is the simplest form of the result we seek.

\subsection{Part C - Potential Along the $Z$-Axis}

If $x$ and $y$ are zero, then $r$ must be zero. This simplifies the result from the previous subsection as

\begin{align}
&= \frac{-\phi_0z}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{1}{\left(r_0^2 + z^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
&= \frac{-\phi_0 z}{4 \pi}\int_0^a \frac{r_0}{(r_0^2 + z^2)^{\frac{3}{2}}}dr_0
\end{align}

Let $u = r_0^2 + z^2$ and $du = 2r_0dr_0$ so $dr_0 = \frac{du}{2r_0}$.

\begin{align}
\Phi &= \frac{-\phi_0 z}{4 \pi}\int_{z^2}^{a^2 + z^2}\frac{\cancel{r_0}}{u^{\frac{3}{2}}}\frac{du}{2\cancel{r_0}} \\
%&= \frac{\phi_0 z}{\epsilon_0} \left(\frac{1}{2}u^{-\frac{1}{2}}\right)\bigg|_{u=z^2}^{u=z^2+a^2} \\
&= \frac{\phi_0 z}{4 \pi} \left(\frac{1}{\sqrt{u}}\right)\bigg|_{u=z^2}^{u=z^2 + a^2} \\
&= \frac{\phi_0 z}{4 \pi} \left[\frac{1}{\sqrt{z^2 + a^2}} - \frac{1}{\sqrt{z^2}}\right] \\
&= \frac{\phi_0 z}{4 \pi} \left[ 1 - \frac{z}{\sqrt{z^2 + a^2}}\right]\\
\end{align}

As required.

\subsection{Expansion}

Starting from
\begin{align}
&= \frac{-\phi_0z}{4 \pi} \int_0^a \int_0^{2 \pi} \frac{r_0dr_0d\theta_0}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}}  \\
\end{align}

We divide both numerator and denominator by a factor of $(r^2 + z^2)^{\frac{3}{2}}$ and rearrange:

\begin{align}
\frac{-\phi_0 z}{4 \pi} \int_0^a \int_0^{2\pi}\left(1 + \frac{r_0^2 -2 r r_0 \cos(\theta_0)}{r^2 + z^2}\right)^{\frac{3}{2}}
\end{align}

Using $(1 + x)^n \sim 1 + nx + \frac{n(n-1)}{2}x^2...$ we carry the expansion to two terms as follows:

\begin{align}
\frac{-\phi_0 z}{4 \pi}\int_0^a \int_0^{2 \pi}\left[1 - \frac{3}{2(r^2 + z^2)}(r_0^2 - 2rr_0\cos(\theta_0)) + \frac{15}{8(r^2 + z^2)^2}(r_0^2 - 2rr_0\cos(\theta_0))^2\right]r_0 dr_0 d\theta_0 \\
\end{align}

Integrating term by term we have

The first term, without the constants, or prefactor, is

\begin{align}
&= \int_0^a \int_0^{2 \pi}d\theta_0r_0dr_0 = \pi a^2 \\
\end{align}

The second term is

\begin{align}
&=\frac{1}{(r^2 + z^2)} \int_0^a \int_0^{2 \pi} \left(-\frac{3}{2}\right)(r_0^2 - 2rr_0^2\cos(\theta))d\theta_0dr_0 \\
&= \frac{1}{(r^2 + z^2)} \left[\int_0^a \int_0^{2 \pi}r_0^3 d\theta_0 dr_0 - \int_0^a \int_0^{2 \pi}2rr_0^2\cos(\theta_0)d\theta_0 dr_0\right] \\
&= \frac{1}{(r^2 + z^2)} \left[\frac{\pi a^4}{2} - \frac{2 r a^3}{3}(\sin(\theta_0 - 2 \pi) - \sin(\theta_0))\right] \\
&= -\frac{3}{2}\frac{1}{(r^2 + z^2)}\frac{\pi a^4}{2} \\
&= -\frac{3\pi a^4}{4(r^2 + z^2)}
\end{align}

The last integral is

\begin{align}
&=\frac{15}{8} \frac{1}{(r^2 + z^2)^2} \int_0^a \int_0^{2 \pi}\left[r_0^5 + 4 r^2 r_0^3\cos^2(\theta_0) - 4 r r_0^4\cos(\theta_0)\right]d\theta_0dr_0 \\
&= \frac{15}{8} \frac{1}{(r^2 + z^2)^2}\bigg[\int_0^a \int_0^{2 \pi}r_0^5 d\theta_0 dr_0 + \int_0^a \int_0^{2 \pi} 4 r^2 r_0^3 \cos^2(\theta_0)d\theta_0dr_0  \\ &  - \int_0^a \int_0^{2 \pi} 4 r r_0^4 \cos(\theta_0)d\theta_0 dr_0\bigg] \\
&= \frac{15}{8}\frac{1}{(z^2 + r^2)^2}\left[\frac{\pi a^6}{3} + \pi r^2 a^4\right] \\
&= \frac{5a^2 \pi (3r^2 a^2 + a^4)}{8 (r^2 + z^2)^2}
\end{align}

Reinserting the prefactors and constants yields

\begin{align}
\Phi = \frac{-\phi_0  a^2}{4} \frac{z}{(z^2 + r^2)^{\frac{3}{2}}}\left[1 - \frac{3 a^2}{4(r^2 + z^2)} + \frac{5(3 r^2 a^2 + a^4)}{8(r^2 + z^2)^2}\right]
\end{align}

%Considering the figure, we see that the total vector to the field point from the origin is made up of a law of cosines like component in the plane plus a vector that accounts for the $z$ component. Thus to transform to cylinderical coordinates we must write the Green's function as

%\begin{align}
%G = \frac{1}{4 \pi \epsilon_)}\left[\frac{1}{
%\end{align}

\section{Problem 5}

We begin with separation of variables assuming a solution of the form $\Phi  = R(r)\Theta(\theta)Z(z)$.The Laplacian in cylindrical coordinates is given by $\frac{1}{r}\frac{\partial}{\partial r}\left[r\frac{\partial}{\partial r}\right] + \frac{1}{r^2}\frac{\partial^2}{\partial \theta^2} + \frac{\partial^2}{\partial z^2}$. Acting upon $\Phi$ with this operator and setting the result equals to zero gives us

\begin{align}
Z\Theta\frac{1}{r} \frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{RZ}{r^2}\frac{\partial^2 \Theta}{\partial \theta^2} + R\Theta\frac{\partial^2 Z}{\partial z^2} &= 0 \\
\frac{1}{rR}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{r^2 \Theta}\frac{\partial^2 \Theta}{\partial \theta^2} + \frac{1}{Z}\frac{\partial^2 Z}{\partial z^2} &= 0 \\
\end{align}

The only way that two functions of separate variables may be equal for all values of those variables are if they are both equal to a constant - the \textit{same} constant.

\begin{align}
\frac{1}{rR}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{r^2 \Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= -k^2 \\ 
\frac{1}{Z}\frac{\partial^2 Z}{\partial z^2} &= k^2
\end{align}

The equation for $Z$ implies that $Z$ is proportional to real exponentials - or linear combinations of $sinh$ and $cosh$.

\begin{align}
Z \sim A\sinh(kz) + B\cosh(kz)
\end{align}

The other equation reads

\begin{align}
\frac{r}{R} \frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= -k^2r^2 \\
\frac{r}{R}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + k^2 r^2 = -\frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} \\
\end{align}

A similar argument concerning constant functions allows us to write the two equations:

\begin{align}
\frac{r}{R}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + k^2r^2 &= \mu^2 \\
\frac{r}{R}\left[\frac{\partial R}{\partial r} + r\frac{\partial^2 R}{\partial r^2}\right] + k^2 r^2 &= \mu^2 \\
\frac{r}{R}\frac{\partial R}{\partial r} + \frac{r^2}{R}\frac{\partial^2 R}{\partial r^2} + k^2 r^2 &= \mu^2R \\
r^2 \frac{\partial^2 R}{\partial r^2} + r \frac{\partial R}{\partial r} + k^2 r^2 R = \mu^2R \\
r^2 \frac{\partial^2 R}{\partial r^2} + r \frac{\partial R}{\partial r} + [k^2 r^2 - \mu^2]R &= 0 \\
-\frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= \mu^2 \\
\end{align}

The equation in $R$ can be identified as Bessel's equation of order $\mu$ with argument $kr$. The $\Theta$ equation solutions are oscillatory.

\begin{align}
R &\sim J_{\mu}(kr) + \cancel{Y_\mu(kr)} \\
\Theta &\sim \cancel{\sin(\mu\theta)}+ \cos(\mu\theta) \\
\end{align}

We may exclude Neumann functions as they diverge within the region of interest $r=0$. Furthermore, since $\phi_0$ is a constant on the sides of the cylinder we must have that $\Theta'' = \mu^2\Theta = 0$. Unless $\Theta = 0$, this requires that $\mu = 0$, which in turns restricts us to cosines in $\Theta$ and a cosine term equal to unity. The boundary condition at $z=0$ implies that the cosh term is also zero. With this information we have narrowed the solution down to something like

\begin{align}
\Phi \sim CJ_0(kr)\sinh(kz)
\end{align}

We take the top cap at $z=L$ as the one with the constant potential. To use the final boundary condition we need to lower the global potential by $\phi_0$ and impose the condition that $\phi = 0$ at $z=L$, or equivalently, that $\sinh(k(z-L)) = 0$. Additionally, $J_0(ka) = 0$, which restricts $k$ to be $\frac{x_{0n}}{a}$ where $x_{0n}$ is the $n'th$ zero of the zeroth order bessel function.

\begin{align}
\Phi &= \sum_{n=0}^{\infty} a_n J_0(\frac{x_{0n}}{a}r)\sinh(x_{0n}\frac{(z-L)}{a}) \\ 
\text{at $z=0$} \rightarrow \phi_0 &=  \sum_{n=0}^{\infty}a_nJ_0(\frac{x_{0n}}{a}r)\sinh(-x_{0n}\frac{L}{a})\\
\end{align}

To find the coefficients we need to take a Fourier-Bessel series by multiplying both sides by $rJ_0(\frac{x_{0n'}}{a}$ and integrate from $0$ to $a$:

\begin{align}
\int_0^a \phi_0 rJ_0\left(\frac{x_{0n'}}{a}r\right)dr &= \sum_{n=0}^{\infty}a_n \sinh(-x_{0n}\frac{L}{a}) \int_0^a r J_0\left(\frac{x_{0n}}{a}r\right) J_0\left(\frac{x_{0n'}}{a}r\right)dr \\
\int_0^a \phi_0 rJ_0\left(\frac{x_{0n'}}{a}r\right)dr &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2\delta_{nn'} \ \rightarrow \ \sinh \ \text{an odd function.} \\ 
\end{align}

The orthogonality of bessel functions was used on the right hand side. What we would like to do next is get the left hand side into a form that is suitable for an integral identity. Let $u = \frac{x_{0n}}{a}r$. Then $r = \frac{a}{x_{0n}}u$ and  $du = \frac{x_{0n}}{a}dr$. Furthermore, This allows us to rewrite the current equality as

\begin{align}
\phi_0\int_0^a \cancel{\frac{a}{x_{0n}}}uJ_0(u)\cancel{\frac{x_{0n}}{a}}du &= \sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2 \\
\phi_0\int_0^a uJ_0(u)du &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2 \\
\rightarrow \  \ \text{Using an Integral identity}& \rightarrow \\
\phi_0\frac{\cancel{a^2}}{x_{0n}}\cancel{J_1(x_{0n})} &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{\cancel{a^2}}{2}[J_1(x_{0n})]^{\cancel{2}}\\
a_n &= 2\frac{\phi_0 \ \text{cosech}(x_{0n}\frac{L}{a})}{x_{0n}J_1(x_{0n})}
\end{align}

The solution is thus as follows

\begin{align}
\Phi(r,z) = \phi_0 + 2\sum_{n=1}^{\infty}\frac{\phi_0 \ \text{cosech}(x_{0n}\frac{L}{a})}{x_{0n}J_1(x_{0n})}\sinh(x_{0n}\frac{(z-L)}{a})J_0\left(\frac{x_{0n}}{a}r\right)
\end{align}

\end{document}

### 2016-06-10 Gradschool/math/MathMethodsFinal.tex
Added Math Final

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{cancel}
\usepackage{pdfpages}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Methods of Theoretical Physics Final Exam \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Q1 Part 1}

We are asked to solve the following initial value problem:

\begin{align}
y'' + 4y' + 4y = 0
\end{align}

Using the initial conditions $y(0) = 1$, and $y'(0) = 3$. The equation relates $y$ to its derivatives in a manner that seems to state that $y$ and its derivatives have the same dimension - this suggests exponential solutions of the form

\begin{align}
y &= e^{\alpha x} \\
y' &= \alpha e^{\alpha x} \\
y'' &= \alpha^2 e^{\alpha x}
\end{align}

Plugging these expressions into equation 1.1 gives us:

\begin{align}
\alpha^2 e^{\alpha x} + 4 \alpha e^{\alpha x} + 4 e^{\alpha x} = 0
\end{align}

Cancelling the exponential behavior yields a quadratic equation in $\alpha$:

\begin{align}
\alpha^2 + 4 \alpha + 4 &= 0 \\
(\alpha + 2)^2 &= 0
\end{align}

There is a repeated root at $\alpha = -2$. This indicates that we have a general solution of the form

\begin{align}
y(x) = c_1 e^{-2x} + c_2 x e^{-2x}
\end{align}

With two undetermined constants $c_1$ and $c_2$ that must be set by the initial conditions. To figure them out we need to go get two equations in two unknowns from the derivatives of our general solution:

\begin{align}
y' &= e^{-2 x} (-2 c_1 -2 c_2x + c_2) \\
y'' &= 4 e^{-2x} (c_1 + c_2(x-1))
\end{align}

So we know the following must hold at $x=0$:

\begin{align}
y(0) &= c_1 = 1\\ 
y'(0) &= c_2 - 2c_1 = 3 \\
\rightarrow c_2 &= 5
\end{align}

Meaning we can write down the solution to the initial value problem in full:

\begin{align}
y(x) = e^{-2x} + 5x e^{-2x}
\end{align}

This solution was verified by Mathematica. \checkmark

\section{Q1 Part 2}
We are asked to find a general solution for $x > 0$ for the differential equation:

\begin{align}
\frac{1}{x}\frac{dy}{dx} - \frac{2}{x} y = x \cos(x)
\end{align}

We begin by multiplying through by $x$ to cast the ODE into the standard form of Arfken equation 7.11.

\begin{align}
\frac{dy}{dx} - 2y = x^2 \cos(x)
\end{align}

With $P(x) = -2$ and $Q(x) = x^2 \cos(x)$. Then, from Arfken equation 7.15, we find the Integrating Factor $\alpha(x)$:

\begin{align}
\alpha(x) &= e^{\int^{x} P(x) dx} \\
&= e^{-\int^{x} 2 dx} \\
&= e^{- 2x}
\end{align}

With $\alpha$ in hand, we directly apply Arfken equation 7.16, which states that

\begin{align}
y(x) &= \frac{1}{\alpha(x)} \left[ \int^{x} \alpha(x) Q(x) dx + C \right] \\
&= e^{2x} \left[ \int^{x} e^{-2x}x^2\cos(x) + C \right]
\end{align}

The integral evaluates to

\begin{align}
\frac{1}{125} e^{-2 x} \left[\left(25 x^2+40 x+22\right) \sin(x) -2 \left(25 x^2+15 x+2\right) \cos(x) \right]
\end{align}

Multiplying by the $e^{2x}$ out front takes the total exponent to $0$ so we may cancel the exponential off the solution completely, leaving us with the solution

\begin{align}
y(x) =  \frac{1}{125} \left[\left(25 x^2+40 x+22\right) \sin(x) -2 \left(25 x^2+15 x+2\right) \cos(x) \right]
\end{align}

This solution has been verified by Mathematica. \checkmark

\section{Q1 Part 3}

We are asked to consider the differential equation

\begin{align}
2y - 6x + \left(3x - \frac{4 x^2}{y}\right)\frac{dy}{dx} = 0
\end{align}

We begin by casting the ODE into the standard form of Arfken equation 7.6.

\begin{align}
P(x,y)dx + Q(x,y)dy &= 0 \\ 
(2y - 6x)dx + \left(3x - \frac{4x^2}{y}\right)dy &= 0
\end{align}

Meaning we must have that $P(x,y) = 2y - 6x$ and $Q(x,y) = 3x - \frac{4x^2}{y}$. To constitute an exact equation, we must have that

\begin{align}
\frac{\partial P(x,y)}{\partial y} = \frac{\partial Q(x,y)}{\partial x}
\end{align}

And currently we have $\frac{\partial P(x,y)}{\partial y} = 2$ and $\frac{\partial Q(x,y)}{\partial x} = 3 - \frac{8x}{y}$. So, our equation is not exact. However, when we multiply the entire equation by the integrating factor $xy^2$:

\begin{align}
x y^2 (2y - 6x)dx + x y^2 \left(3x - \frac{4x^2}{y}\right)dy &= 0 \\
(2 x y^3 - 6 x^2 y^2)dx + \left(3 x^2 y^2 - 4 x^3 y \right)dy &= 0
\end{align}

In this modified situation, we have $P'(x,y) = (2xy^3 - 6x^2y^2)$ and $Q'(x,y) = (3x^2 y^2 - 4 x^3 y)$. Note that the primes do not denote differentiation. Merely that these are different functions. Then we have $\frac{\partial P'(x,y)}{\partial y} = (6 x y^2 - 12 x^2 y)$ and  $\frac{\partial Q'(x,y)}{\partial x} = (6 x y^2 - 12 x^2 y)$. \checkmark

\hspace{2mm}
We now seek to actually solve the equation. We expect that in the exact equation case, both $P'$ and $Q'$ can be written as derivatives of the solution

\begin{align}
P'(x,y) &= \frac{\partial \Psi}{\partial x} \\
Q'(x,y) &= \frac{\partial \Psi}{\partial y} 
\end{align}

It doesn't really matter which function we choose to integrate for this next step, I'm picking the $x$ derivative. Thus

\begin{align}
\Psi &= \int \left( 2 x y^3 - 6 x^2 y^2 \right)dx \\
&= \int 2xy^3 dx - \int 6x^2 y^2 dx \\
&= x^2 y^3 - 2x^3 y^2 + C(y)
\end{align}

Where the "constant" of integration is really a function of $y$, as any such function will cancel when differentiated with respect to $x$. So we go ahead and differentiate again with respect to the \textit{other} variable, $y$.

\begin{align}
3x^2 y^2 - 4 x^3 y + C'(y)
\end{align}

And set this equal to $Q'(x,y)$

\begin{align}
\cancel{3x^2 y^2} - \cancel{4x^3 y} + C'(y) &= \cancel{3x^2 y^2}- \cancel{4 x^3 y} \\
\rightarrow C'(y) &= 0 \\
C(y) &= K
\end{align}

Where $K$ is the constant of integration. We may now write the full solution to the original differential equation:

\begin{align}
\Psi = x^2 y^3 - 2 x^3 y^2 + K
\end{align}

\section{Q1 Part 4}
We are asked to find a general solution, on the interval $-\frac{\pi}{2}, \frac{\pi}{2}$, to the differential equation:

\begin{equation}
\frac{d^2 y}{dx^2} + y = \tan(x)
\end{equation}

We will use the method of variation of parameters which begins by finding the solutions to the homogenous equation:

\begin{align}
\frac{d^2 y}{dx^2} + y = 0
\end{align}

The solutions to this equation are simple triginometric functions $y_1(x) = \sin(x)$, $y_2(x) = \cos(x)$. They have simple derivatives $y'_1(x) = \cos(x)$, $y'_2(x) = -\sin(x)$.

\hspace{2mm}

This means that the simeltaneous equations for Variation of Parameters read:

\begin{align}
y_1 u'_1 + y_2 u'_2 &= 0 \\
y'_1 u'_1 + y'_2 u'_2 &= F(x)
\end{align}

Where $F(x)$ is the function corresponding to the source term in the differential equation.

\begin{align}
\sin(x)u'_1(x) + \cos(x)u'_2(x) &= 0 \\
\cos(x)u'_1(x) - \sin(x)u'_2(x) &= \tan(x)
\end{align}

%Squaring both equations:

%\begin{align}
%\sin^2(x)u'^2_1(x) + 2 \sin(x) \cos(x) u'_1(x) u'_2(x) + \cos^2 u'^2_2(x) &= 0 \\
%\cos^2(x)u'^2_1(x) - 2 \sin(x) \cos(x) u'_1(x) u'_2(x) + \sin^2 u'^2_2(x) &= \tan^2(x)
%\end{align}

%Adding now both equations, the cross terms cancel, and we end up with

%\begin{equation}
%u'^2_1(x) + u'^2_2(x) = \tan^2(x)
%\end{equation}

%But from the first equation we have that $u'_2(x) = - \tan(x)u'_1(x)$. Substituting this into the second equation, we have that $\cos(x)u'_1(x) - \sin(x)(-\tan(x))u'_1(x) = \tan(x)$

These two algebraic equations for the derivatives of the functions $u_i$ may be solved to yield the following values:

\begin{align}
u'_1(x) &= \sin(x) \\
u'_2(x) &= -\sin(x)\tan(x)
\end{align}

The integrals are thus

\begin{align}
u_1(x) &= \int \sin(x) dx = -\cos(x) + c \\
u_2(x) &= -\int \sin(x) \tan(x) = \ln |\sec(x) + \tan(x)| - \sin(x) + c
\end{align}

Variation of Parameters started out by assuming we can write our particular solution as these two functions $u_i$ times our homogenous solutions:

\begin{align}
y_p(x) &= u_1(x)y_1(x) + u_2(x)y_2(x) \\
&= [-\cos(x)][\sin(x)] + [-\ln |\sec(x) + \tan(x)| + \sin(x)][\cos(x)]\\
&= \cancel{\cos(x)\sin(x)} - \cancel{\cos(x)\sin(x)} - \cos(x)\ln|\sec(x) + \tan(x)| \\
&= - \cos(x)\ln|\sec(x) + \tan(x)|
\end{align}

Meaning we can write the \textit{most general} solution to the differential equation as the sum of the two homogenous solutions plus the particular solution thanks to the source term:

\begin{align}
y(x) = c_1 \sin(x) + c_2 \cos(x) - \cos(x)\ln|\sec(x) + \tan(x)|
\end{align}

Completing the problem.

\section{Q1 Part 5}

Given that $\cos(x)$ is a solution, we are asked to find the general solution to the equation

\begin{align}
\sin(x) y'' - 2 \cos(x) y' - \sin(x) y = 0
\end{align}

On the interval $0 < x < \pi$. We remind ourselves that given an ODE solution we may construct a second linearly independent solution with the Wronskian double integral (Arfken equation 7.67):

\begin{align}
y_2(x) = y_1(x) \int^{x} \frac{e^{-\int^{s} P(t) dt}}{y_1(s)^2} ds
\end{align}

The variables are initially confusing but we notice it is just set up so that the integral in the exponential is a function of $s$ and so the whole integral is a function of $x$.

\hspace{2mm}

We rewrite the ODE in standard form:

\begin{align}
y'' - 2\cot(x) y' - y = 0
\end{align}

Here $P(x)$ is the coefficient on the $y'$ term:

\begin{align}
P(x) = - 2 \cot(x)
\end{align}

Taking first the integral inside the exponential function, we have a negative sign from the sign of hte integral and another one from the $P(x)$ making the integral \textit{positive}. Doing it the other way gets you the complete wrong answer

\begin{align}
2\int \cot(x) dx & \\
\rightarrow 2 \ln (\sin(x)) & \\
\rightarrow  \ln (\sin(x)^{2}) & \\
\end{align}

So our Wronskian double integral reads:

\begin{align}
y_2(x) &= \cos(x) \int \frac{e^{\ln(\sin^2(x))}}{\cos^2(x)} dx \\
&= \cos(x) \int \frac{\sin^2(x)}{\cos^2(x)} dx \\
&= \cos(x) \int \tan^2(x) dx \\
&= \cos(x) (\tan(x) - x) \\ 
&= \sin(x) - x\cos(x)
\end{align}

Thus the most general solution to this homogenous differential equation looks like:

\begin{align}
y(x) = c_1 \cos(x) + c_2 \left[\sin(x) - x \cos(x) \right]
\end{align}

\section{Q1 Part 6}

We are asked to find a power series expansion for the second order differential equation

\begin{align}
(1 + x^2) y'' - y' + y = 0 
\end{align}

If we cast the ODE into standard form, we would just get

\begin{align}
y'' - \frac{1}{1 + x^2} y' + \frac{1}{1+x^2}y = 0
\end{align}

Since we are expanding around the point $x=0$, we check this equation for singularities indicating the necessity of the Frobenius method. Finding none, we assume a power series solution of the form

\begin{align}
y(x) &= \sum_{n=0}^{\infty} a_n x^n \\
y'(x) &= \sum_{n=1}^{\infty} a_n n x^{n-1} \\
y''(x) &= \sum_{n=2}^{\infty} a_n n(n-1) x^{n-2}
\end{align}

We lose an index on the bottom each time we take a derivative, because the original series is something like $a_0 + a_1 x + a_2 x^2 ...$, so the first derivative cancels the first term, the second gets rid of the second term, and so on, and so forth.

\hspace{2mm}

Substituting this into the original differential equation gives us

\begin{align}
\sum_{n=2}^{\infty} a_n n(n-1) x^{n-2} - x^2 \sum_{n=2}^{\infty} a_n n(n-1) x^{n-2} - \sum_{n=1}^{\infty} a_n n x^{n-1} + \sum_{n=0}^{\infty} a_n x^n &= 0\\
\sum_{n=2}^{\infty} a_n n(n-1) x^{n-2} - \sum_{n=2}^{\infty} a_n n(n-1) x^{n} - \sum_{n=1}^{\infty} a_n n x^{n-1} + \sum_{n=0}^{\infty} a_n x^n &= 0\\
\end{align}

%We want to get everything under one summation so we can say something about the coefficients $a_n$. Do do this we need all the summations to start at the same point, and ideally we need all the powers of $x$ to be the same as well.

%\hspace{2mm}

%To the first summation, we apply the transformation $m = n - 2 \rightarrow n = m + 2$:

%\begin{align}
%\sum_{n=2}^{\infty} a_n n (n-1) x^{n-2} \rightarrow \sum_{m + 2 = 2}^{\infty} a_{m+2} (m + 2)(m + 2 - 1)x^{m + 2 - 2} \rightarrow \sum_{m = 0}^{\infty} a_{m+2}(m+2)(m+1)x^m
%\end{align}

%To the third summation, we apply the transformation $m = n - 1 \rightarrow n = m + 1$:

%\begin{align}
%\sum_{n=1}^{\infty} a_n n x^{n-1} \rightarrow \sum_{m + 1 = 1}^{\infty} a_{m+1}(m+1) x^{m + 1 - 1} \rightarrow \sum_{m=0}^{\infty} a_{m+1}(m+1)x^m
%\end{align}

%As for the second summation, we might as well write it as a sum from $0$ anyways, seeing as how the $n=0$ and $n=1$ are zero anyways. Thus we should have

%\begin{align}
%\sum_{n=0}^{\infty} a_{n+2} \left[(n+2)(n+1)\right] - a_n n(n-1)x^n - a_{n+1} (n+1)x^n + a_n x^n &=0 \\
%\sum_{n=0}^{\infty} a_{n+2} \left[(n+2)(n+1)\right] 
%\end{align}

There are some ways that we could rearrange this into one power series with a single expression for all the coefficients $a_n$. But since we are only asked to evaluate the first few terms, we'll use the following process.

\hspace{2mm}

Sometimes, in an infinite sum, two terms that occur at different values of $n$ may cancel each other. But in a power series, we can never cancel a power of $x$ with a different power of $x$, they just behave fundamentally differently and don't cancel. So we know individually that the coefficients on each power of $x$ must themselves all vanish.

\hspace{2mm}

We might as well set all of the lower indices of the summations to zero, the terms will cancel for all the appropriate values of $n$.

\hspace{2mm}

The $x^0$ terms:

\begin{align}
a_2(2(2-1))x^0 - \cancel{a_0(0(0-1))x^0} &- a_1(1)x^0 + a_0x^0 = 2a_2 + a_1 +  a_0 = 0 \\
&\rightarrow a_2 = -\frac{a_1 + a_0}{2}
\end{align}

The $x^1$ terms:

\begin{align}
&a_3(3(3-1))x^1 - \cancel{a_1(1(1-1))x^1} - a_2(2)x^1 + a_1 x^1 = 6a_3 - 2 a_2 + a_1 = 0 \\
&\rightarrow a_3 = \frac{2a_2 - a_1}{6} = \frac{2 (-\frac{a_1 + a_0}{2}) - a_1}{6} = \frac{a_0 - 2 a_1}{6}
\end{align}

The $x^2$ terms:

\begin{align}
&a_4(4(4-1))x^2 - a_2(2(2-1))x^2 - a_3(3)x^2 + a_2 x^2 = 12a_4 - \cancel{a_2} - 3a_3 + \cancel{a_2} = 0 \\
&\rightarrow a_4 = \frac{a_3}{4} = \frac{a_0 - 2a_1}{24} 
\end{align}

%The $x^3$ terms:

%\begin{align}
%&a_5(5(5-1))x^3 - a_3(3(3-1))x^3 - a_4(4)x^3 + a_3x^3 = 20a_5 - 6a_3 - 4a_4 + a_3 = 0 \\
%&\rightarrow 20a_5 - 5a_3 - 4a_4 = 0  
%\end{align}

Since this is a second order differential equation, we know there must be two undetermined coefficients. These are $a_0$ and $a_1$. Since we have information about the coefficients of powers up through $x^4$ from demanding each power individually goes to zero, we can write the power series expansion as follows:

\begin{align}
y(x) = a_0 + a_1 x - \frac{a_1 + a_0}{2} x^2 + \frac{a_0 - 2a_1}{6}x^3 + \frac{a_0 - 2a_1}{24}x^4 ....
\end{align}

We can theoretically iterate this process ad infinitum and get all the terms. But we will stop here.

\hspace{2mm}

Fuch's Theorem can give us some information about the radius of convergence of the series solutions. It states that the radius of convergence of the power series solution to a differential equation is at least as large as the smaller of the two radii of convergence of $P(x)$ and $Q(x)$ from the standard form equation. Here $P(x) = -\frac{1}{1 + x^2}$ and $Q(x) = \frac{1}{1+x^2}$. Both of these have, with opposite sign, the power series expansion:

\begin{align}
\sum_{n=0}^{\infty}(-1)^n x^{2n} \rightarrow R.O.C. = 1
\end{align}

So we know by Fuch's Theorem that the power series solution to our particular differential equation must have a radius of convergence $R>1$.

\section{Q2 Part 1}

We are asked to investigate the full asymptotic behavior of the integral

\begin{align}
\int_0^1 e^{i x t^2} dt, \ \ x \rightarrow + \infty
\end{align}

My solution follows notes from Bender and Orszag, \textbf{\textit{Advanced Mathematics for Scientists and Engineers, Asymptotic Methods and Perturbation Theory, Ch.6}}. The method of Steepest Descent concerns integrals of the form

\begin{align}
I(\lambda) = \int_{C} e^{\lambda P(t)}Q(t) dt
\end{align}

We recognize the "constant" $\lambda = x$ meaning our function $P(t) = i t^2$ and $Q(t) = 1$. The critical points of $P(t)$ occur when $\frac{dP(t)}{dt} = 0$ or when $2 i t = 0 \rightarrow t=0$.

\hspace{2mm}

The objective is to use Cauchy's theorem to deform the contour $0 < t < 1$, which is along the real axis of the complex plane, into some kind of contour of constant complex phase. We also want it to pass through the critical point so we are looking for a contour of constant complex phase that passes through $t=0$.

Letting

\begin{align}
t &= u + iv \\
t^2 &= u^2 + 2 i u v - v^2 \\
P(t) &= i(u^2 - v^2) - 2 u v
\end{align}

But at $t = 0$ we must have that the imaginary part of $P(t)$ equals zero. This is equivalent to requiring $u = \pm v$ on the contour in question. If we take the contours for which $u = -v$, then we would have that the real part,

\begin{align}
\Re{P(t)} &= 2v^2 \\
|e^{xP(t)}| &= e^{2xv^2} 
\end{align}

Since $t = u + iv$ and $u = -v$ we have that $t = (-v + iv) = (i - 1)v$. But as $t \rightarrow \infty$ the real part of $P(t)$ blows up. Since the modified exponential doesn't have a maximum (only a minimum, in fact) on the chosen contour we can't use Laplace's method. However the other choice, $u = v$, gives us the opposite case

\begin{align}
\Re{P(t)} &= -2v^2 \\
|e^{xP(t)}| &= e^{-2xv^2} 
\end{align}

Except this time we pick up an extra negative sign from the definition of $t$ as a complex number and we get $t = (i + 1)v$. Now things are inverted and the exponential actually has a maximum at $t=0$.

\hspace{2mm}

Now we need to apply similar analysis to the other endpoint of the integral. Here the imaginary part o f $P(t)$ must be equal to $1$. So we have $u^2 - v^2 = 1 \rightarrow u = \sqrt{1 + v^2}$. Plugging this back into the defintion of $t$ gives us the second required contour, $t = \sqrt{1 + v^2} + iv$. Our results are:

\begin{align}
&C_1: t = (1+i)v \\
&C_3: t = \sqrt{1 + v^2} + iv
\end{align}

We need to deform the original path along the real axis into the contours $C_1 + C_3$. Both contours have constant complex magnitude, but since both of the contours have \textit{different} constant complex magnitudes ($0$ vs $1$) it's impossible to deform the original path into both of them at once. Thus we need some contour somewhere out in the complex plane that bridges between the two paths, $C_2$. If $V$ is the magnitude of the complex part of the contour, let's bridge the contours with a horizontal line joining the points $V + iV$ and $\sqrt{1 + V^2} + iV$ somewhere very far off in the complex plane. As $V\rightarrow \infty$, the contribution from $C_2$ vanishes anyways.

\hspace{2mm}

Thus we can write the original integral as follows:

\begin{align}
I(x) = \int_{C_1} e^{i x t^2} + \int_{C_3} e^{i x t^2} dt
\end{align}

We wish to evaluate the first integral exactly. Along the contour, $t = (1 + i)v$, $dt = (1 + i)dv$. We substitute this in and set the bounds of the integral to be $0 \rightarrow \infty$, which is what our contour has to be for us to have assumed that the contribution from $C_2 = 0$

\begin{align}
\int_{C_1} &= e^{i x ((1 + i)v^2)}(1 + i) dv \\
&= (1 + i) \int_{0}^{\infty} e^{i x (2 i v^2)} dv \\
&= (1 + i) \int_{0}^{\infty} e^{- 2 x v^2} dv \\
&= (1 + i) \frac{1}{2}\sqrt{\frac{\pi}{2x}} \\
\end{align}

The factor $1 + i$ is like drawing a $45^\circ$ right triangle in the first quadrant of the complex plane and so we can convert this factor to the corresponding complex exponential:

\begin{align}
1 + i = e^{\frac{i \pi}{4}}
\end{align}

Giving us the result that the integral along $C_1$ is equal to $e^{\frac{i \pi}{4}} \frac{1}{2} \sqrt{\frac{\pi}{2x}}$.

\hspace{2mm}

When we apply the same arguments to the integral along $C_3$, we have $t = \sqrt{1 + v^2} + iv$. Plugging this in gives

\begin{align}
&=\int_{C_3}e^{ix (\sqrt{1 + v^2} + iv)^2} dt \\
&\rightarrow \int_{C_3}e^{ix (1 + 2 i v \sqrt{1 + v^2})}dt \\
&\rightarrow \int_{C_3}e^{ix - 2xv \sqrt{1 + v^2}} dt \\
&\rightarrow \int_{C_3}e^{ix}e^{-2xv \sqrt{1 + v^2}} dt
\end{align}

If we make the change of variables $it^2 \rightarrow  i - s$, for $s = 2v \sqrt{1 + v^2}$ we can rewrite the integral as

\begin{align}
e^{ix} \int_{C_3} e^{-sx} dv
\end{align}

To rewrite $dt$ in terms of $ds$ we notice that $t^2 = 1 + is \rightarrow t = \sqrt{is + 1}$ So $dt = \frac{1}{2} (1 + is)^{- \frac{1}{2}}(i) = \frac{1}{2\sqrt{1 + is}}$. Then we can write

\begin{align}
ie^{ix} \int_{C_3} \frac{e^{-sx}}{2 \sqrt{1 + is}} ds
\end{align}

This is now of a form appropriate for use of Watson's Lemma. Using the Taylor Expansion:

\begin{align}
(1 + is)^{-\frac{1}{2}} &= \sum_{n=0}^{\infty}(-is)^n \frac{\Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) n!} \\  
(1 + is)^{-\frac{1}{2}} &= \sum_{n=0}^{\infty}(-i)^n(s)^n \frac{\Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) n!}   
\end{align}

This is of the form $\sum_{n=1}^{\infty} a_n (-is)^n$. Watson's Lemma applies to integrals of the form

\begin{align}
\int e^{-xs}f(s)ds
\end{align}

And states that if the function $f(s)$ has asymptotic expansion $f(s) \sim s^{\alpha} \sum_{n=0}^{\infty} a_n s^{\beta n}$, then the integral $I(x)$ has the asymptotic behavior

\begin{align}
I(x) \sim \sum_{n=0}^{\infty} \frac{a_n \Gamma( \alpha + \beta n + 1)}{x^{\alpha + \beta n + 1}}
\end{align}

In our case we have the Taylor expansion for $f(s)$, which is exact, so surely $f(s)$ is asymptotic to said series as well. From equation 7.26, we identify that $\alpha = 0$ since the series  has no power of $s$ out front. Likewise since the power of $s$ inside the series is just $n$ we identify that $\beta = 1$. Finally we identify the coefficient $n$ dependence as $a_n = \frac{(-i)^n \Gamma(n + \frac{1}{2}}{\Gamma(\frac{1}{2}) n!}$.

Plugging into the formula for Watson's Lemma, we have:

\begin{align}
\int_{C_3} e^{i x t^2} dt &\sim \frac{1}{2}i e^{ix} \sum_{n=0}^{\infty} \frac{(-i)^n \Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) n!} \frac{\Gamma(0 + n + 1)}{x^{0 + n + 1}} \\
&\sim \frac{1}{2}i e^{ix} \sum_{n=0}^{\infty} \frac{(-i)^n \Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) \cancel{n!}} \frac{\cancel{\Gamma(n + 1)}}{x^{n + 1}} \\
&\sim \frac{1}{2}i e^{ix} \sum_{n=0}^{\infty} (-i)^n \frac{\Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) x^{n+1}}
\end{align}

Combining with the first integral over $C_1$, we have the full asymptotic behavior:

\begin{align}
e^{\frac{i \pi}{4}} \frac{1}{2} \sqrt{\frac{\pi}{2x}} \frac{1}{2}i e^{ix} + \frac{1}{2} i e^{ix} \sum_{n=0}^{\infty} (-i)^n \frac{\Gamma(n + \frac{1}{2})}{\Gamma(\frac{1}{2}) x^{n+1}}
\end{align}

\includepdf[pages={1-2}]{FinalMath.pdf}

%----------------------------------------------------------------------------------------
%	PROBLEM 1
%----------------------------------------------------------------------------------------

\end{document}

### 2016-06-10 Gradschool/electrodynamics/em_hw4.tex
Aded E/M HW4

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}
\usepackage{float}
\usepackage{cancel}
\usepackage{graphicx}

\newcommand{\partd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\partdd}[2]{\frac{\partial^2 #1}{\partial #2^2}}
%\newcommand{\coth}{\text{coth}}
\newcommand{\csch}{\text{csch}}
\newcommand{\kay}{\frac{1}{4 \pi \epsilon_0}}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Electrodynamics - Homework 4 \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Problem 1}

We are asked to investigate the Diriclet problem in prolate spheroidal coordinates via separation of variables. Like any similar problem involving curvilinear coordinates our first task is to cast the Laplacian into the appropriate form so that we can write Laplace's equation:

\begin{align}
\nabla^2 \Phi = 0
\end{align}

The general form of the Laplacian in curvilinear coordinates is given by

\begin{align}
\nabla \cdot \nabla = \nabla^2 = \frac{1}{h_1 h_2 h_3}\left[\partd{}{u_1}\left(\frac{h_2 h_3}{h_1}\partd{}{u_2}\right) + \partd{}{u_2}\left(\frac{h_3h_1}{h_2}\partd{}{u_2}\right) + \partd{}{u_3}\left(\frac{h_1h_2}{h_3}\partd{}{u_3}\right)\right]
\end{align}

Where the $h_i$ are the scale factors and run over all the coordinates relevant to the space - in the case of prolate spheroidal coordinates, these are $\alpha, \beta$, and $\phi$. These parameters are given in cartesian coordinates as

\begin{align}
x &= c \sinh(\alpha)\sin(\beta)\cos(\phi) \\
y &= c \sinh(\alpha)\sin(\beta)\sin(\phi) \\
z &= c\cosh(\alpha)\cos(\beta)
\end{align}

Whereas the scale factors themselves are given by

\begin{align}
h_i &= \left|\partd{\vec{r}}{q_i} \right| \\
&= \sqrt{\partd{\vec{r}}{q_i} \cdot \partd{\vec{r}}{q_i}} \\
\end{align}

For $\vec{r} = x\hat{x} + y\hat{y} + z\hat{z}$. We need to plug in our coordinate transformations to the components of our position vector and then take derivatives with respect to those coordinates to get the scale factors. The position vector looks like

\begin{align}
\vec{r} &= c\sinh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\sinh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\cosh(\alpha)\cos(\beta)\hat{z} \\
\partd{\vec{r}}{\alpha} &= c\cosh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\cosh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\sinh(\alpha)\cos(\beta)\hat{z}\\
\end{align}

Dotting this with itself gives

\begin{align}
\partd{\vec{r}}{\alpha} \cdot \partd{\vec{r}}{\alpha} &= c^2 \cosh^2(\alpha)\sin^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
&= c^2\cosh^2(\alpha)\sin^2(\beta)(\cos^2(\phi) + \sin^2(\phi)) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
&= c^2\left[\cosh^2(\alpha)\sin^2(\beta) + \sinh^2(\alpha)\cos^2(\beta)\right] \\
&= c^2\left[(1 + \sinh^2(\alpha))\sin^2(\beta) + \sinh^2(\alpha)(1 - \sin^2(\beta))\right] \\
&= c^2\left[\cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sinh^2(\alpha)\right]\\
h_{\alpha} &= \sqrt{\sin^2(\beta) + \sinh^2(\alpha)} 
\end{align}

Similarly for the other coordinates

\begin{align}
\partd{\vec{r}}{\beta}\cdot \partd{\vec{r}}{\beta}&= c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\ 
&= c^2\sinh^2(\alpha)\cos^2(\beta) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\
&= c^2\left[\sinh^2(\alpha)(1-\sin^2(\beta)) + (\sinh^2(\alpha) + 1)\sin^2(\beta)\right]\\
&= c^2\left[\sinh^2(\alpha) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta)\right] \\
h_{\beta} &= \sqrt{c^2\sinh^2(\alpha) +\sin^2(\beta)} 
\end{align}

\begin{align}
\partd{\vec{r}}{\phi} \cdot \partd{\vec{r}}{\phi} &= c^2\sinh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + \sinh^2(\alpha)\sin^2(\beta)\cos^2(\phi) \\
&= c^2\sinh^2(\alpha)\sin^2(\beta)(\sin^2(\phi) + \cos^2(\phi)) \\
&= c^2\sinh^2(\alpha)\sin^2(\beta) \\
h_{\phi} &= \sqrt{c^2\sinh^2(\alpha)\sin^2(\beta)}
\end{align}

Since $h_1 = h_2$ the definition of the Laplacian reduces in this case to

\begin{align}
\frac{1}{h_1h_2h_3}\left[\partd{}{\alpha}h_3\partd{}{\alpha} + \partd{}{\beta}h_3\partd{}{\beta} + \partd{}{\phi}\frac{h_1h_2}{h_3}\partd{}{\phi}\right] \\
\end{align}

The term in the brackets simplifies as

\begin{align}
&=\partd{}{\alpha}(\sinh(\alpha)\sin(\beta))\partd{}{\alpha}\partd{}{\alpha} + \partd{}{\beta}(\sinh(\alpha)\sin(\beta))\partd{}{\beta} + \partd{}{\phi}\left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partd{}{\phi} \\
&= \cosh(\alpha)\sin(\beta)\partd{}{\alpha} + \sinh(\alpha)\sin(\beta)\partdd{}{\alpha} + \sinh(\alpha)\cos(\beta)\partd{}{\beta} + \sinh(\alpha)\sin(\beta)\partdd{}{\beta} ... \\
&+ \left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partdd{}{\phi}
\end{align}

Dividing out the $\sinh(\alpha)\sin(\beta)$ term from the prefactor gets us the final Laplacian:

\begin{align}
\nabla^2 = \frac{1}{c^2(\sin^2\beta)+\sinh^2(\alpha)}\left[\coth(\alpha)\partd{}{\alpha} + \cot(\beta)\partd{}{\beta}+ \partdd{}{\alpha} + \partdd{}{\beta} + (\csch(\alpha) + \csc(\beta))\partdd{}{\phi}\right]
\end{align}

We are given that the problem has azimuthal symmetry, so the ($\phi$) derivatives are zero. This leads us to assume a solution of the form $\Phi = A(\alpha)B(\beta)$. Acting our Laplacian, sans the $\phi$ derivative part, on our ansatz gets us the following

\begin{align}
\partdd{A}{\alpha}B + \partdd{B}{\beta}A + B\coth(\alpha)\partd{A}{\alpha} + A\cot(\beta)\partd{B}{\beta} &= 0 \\
\frac{1}{A}\partdd{A}{\alpha} + \frac{1}{B}\partdd{B}{\beta} + \frac{\coth(\alpha)}{A}\partd{A}{\alpha} + \frac{\cot(\beta)}{B}\partd{B}{\beta} &= 0
\end{align}

As usual the only way two functions of separate varibles can be equal for all values of their respective arguments are if they are both equal to the same constant. In anticipation for transforming our result into Legendre's differential equation we select $l(l+1)$ for our constant and arrive at the following two second order ODEs as our solutions:

\begin{align}
\partdd{A}{\alpha} + \coth(\alpha)\partd{A}{\alpha} - Al(l+1) &= 0 \\
\partdd{B}{\beta} + \cot(\beta)\partd{B}{\beta} + Bl(l+1) &= 0
\end{align}

Starting with the equation for $A$, we would like to transform our derivatives from $\partd{}{\alpha}$ to $\partd{}{\cosh(\alpha)}$. We can write $\partd{}{\alpha} = \partd{\cosh(\alpha)}{\alpha}\partd{}{\cosh(\alpha)} = \sinh(\alpha)\partd{}{\cosh(\alpha)}$. Taking another derivative of this expression via product rule, we get $\partd{}{\alpha}\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh(\alpha)\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{}{\cosh(\alpha)}$. With this information the equation for $A$ transforms as

\begin{align}
\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \coth(\alpha)\sinh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \cosh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
\sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + 2 \cosh(\alpha)\partd{A}{\alpha} - Al(l+1)) = 0 \\
-(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} + 2\cosh(\alpha)\partd{A}{\alpha}-Al(l+1) = 0 \\
(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} - 2\cosh(\alpha)\partd{A}{\alpha} + Al(l+1) = 0
\end{align}

Which is Legendre's equation for $A(\alpha)$. As for the $B$ equation, we want the argument of the resulting Legendre polynomials to be $\cos(\beta)$. So we note that $\partd{}{\beta} = \partd{\cos(\beta)}{\beta}\partd{}{\cos(\beta)} = -\sin(\beta)\partd{}{\cos(\beta)}$ and $\partdd{}{\beta} = \partd{}{\beta}\left[-\sin(\beta)\partd{}{\cos(\beta)}\right] = -\cos(\beta)\partd{}{\cos(\beta)} + \sin^2(\beta)\partdd{}{\cos(\beta)}$. Acting the revised operators on $B$ yields

\begin{align}
\sin^2(\beta)\partdd{B}{\cos(\beta)} - \cos(\beta)\partd{B}{\cos(\beta)} - \cot(\beta)\sin(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0 \\
(1-\cos^2(\beta))\partdd{B}{\cos(\beta)} - 2 \cos(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0
\end{align}

Which is again Legendre's equation for $B(\beta)$. Thus we can represent our solution as

\begin{align}
\Phi = \sum_n A_n P_n(\cosh(\alpha)P_n(\cos(\beta))
\end{align}

Which is the desired result.

\section{Problem 2}

We want to the scalar monopole moment, vector dipole moment, and tensor quadrupole moment for the two charge distributions pictured above, for both cartesian and spherical coordinates. The moments themselves are defined as follows:

\begin{align}
M &= \kay \int_v \rho(\vec{r'})d^3\vec{r'} = \ \text{Total Charge}  \\
\vec{P} &= \kay \int_v \vec{r'}\rho(\vec{r})d^3\vec{r'} = \ \text{Dipole Moment Vector} \\
Q_{ij} &= \kay \int_v (3x_ix_j - r_i^2\delta_{ij})\rho(\vec{r'})d^3(r') = \ \text{Quadrupole Moment Tensor Components}
\end{align}

We note that the total charge of the system sums to zero. Thus we have no monopole moment for the charge distribution. As for the dipole moment, the relevant formula in the discrete case would have to be

\begin{align}
\vec{P}_i &= \sum_{n=1}^3 r_i^{(n)}q_i^{(n)} \ \text{\textit{Component i} of the dipole vector} \\
\end{align}

\subsection{Distribution 1}

Labelling the charges as shown in the figure, the dipole moment  would come out to be $qa\hat{z} - qa\hat{z} -\cancel{2q(0)} = 0$, when summing over all the charges. Thus the dipole moment of the first charge distribution is also zero. The discrete quadrupole formula written as a summation over a discrete charge distribution must be

\begin{align}
Q_{ij} = \sum_{n=1}^3 (3x_i^{(n)}x_j^{(n)} - (r_i^{(n)})^2\delta_{ij})q^{(n)}
\end{align}

For the charge distribution in question we can expand this formula as follows:

\begin{align}
Q_{ij} &= (3x_i^{(1)}x_j^{(1)} - (r_i^{(1)})^2\delta_{ij})q^{(1)} + (3x_i^{(2)}x_j^{(2)} - (r_i^{(2)})^2\delta_{ij})q^{(2)} + (3x_i^{(3)}x_j^{(3)} - (r_i^{(3)})^2\delta_{ij})q^{(3)} \\
\end{align}

Let's look at the off-diagonal case first. If $i \neq j$, then obviously all the $\delta_{ij}$ cancel. Furthermore, when we notice that the charges lie only on the $z$ axis, we find that since \textit{at least one} of the coordinates ($i$ or $j$) must \textit{not} be $z$, all of the other products cancel too. Thus we conclude that all of the off-diagonal elements in this tensor are zero. This leaves us with three to calculate:

\begin{align}
Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)} \\
&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\ 
&= -2a^2q \\
Q_{22} &= (3x_2^{(1)}x_2^{(1)} - (r_2^{(1)})^2)q^{(1)} + (3x_2^{(2)}x_2^{(2)} - (r_2^{(2)})^2)q^{(2)} + (3x_2^{(3)}x_2^{(3)} - (r_2^{(3)})^2)q^{(3)} \\
&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\
&= -2a^2q \\
Q_{33} &= (3x_3^{(1)}x_3^{(1)} - (r_3^{(1)})^2)q^{(1)} + (3x_3^{(2)}x_3^{(2)} - (r_3^{(2)})^2)q^{(2)} + (3x_3^{(3)}x_3^{(3)} - (r_3^{(3)})^2)q^{(3)} \\
&= (3a^2 - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (3a^2 - a^2)q \\
&= 4a^2q
\end{align}

Thus, the quadrupole moment tensor, which we note is traceless, is

\begin{align}
Q = 2a^2
\begin{bmatrix}
-1 & 0 & 0 \\
0 & -1 & 0 \\
0 & 0 & 2
\end{bmatrix}
\end{align}

To translate to spherical coordinates, we make use of Jackson equations 4.4 through 4.6. Since the total charge and dipole moments are known to be zero, the $l=0$ and $l=1$ $q$'s are all zero. Furthermore, if all the off diagonal terms of $Q$ are zero, $q_{21}$ is automatically zero. Even more, $q_{22} = \text{const.} * (Q_{11} - \cancel{iQ_{12}} - Q_{22})$ is zero because $Q_{11} = Q_{22}$. This leaves only the $q_{20}$ moment to calculate.

\begin{align}
q_{20} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}Q_{33} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}(4a^2) = \sqrt{\frac{5}{\pi}}a^2
\end{align}

As the only nonzero $q_{lm}$.

\subsection{Distribution 2}

We note that as before the total charge over the system is zero. So we must have a monopole moment that vanishes. The components of the dipole moment vector $\vec{P}$ would be as follows:

\begin{align}
P_x &= r_x^{(1)}q^{(1)} + r_x^{(2)}q^{(2)} + r_x^{(3)}q^{(3)} + r_x^{(4)}q^{(4)} \\
&= a(-q) + \cancel{0(q)} -a(-q) + \cancel{a(0)} = 0  \\
P_y &= r_y^{(1)}q^{(1)} + r_y^{(2)}q^{(2)} + r_y^{(3)}q^{(3)} + r_y^{(4)}q^{(4)} \\
&= \cancel{(0)-q} -a(q) +\cancel{(0)(-q)} -a(-q) = 0 \\
P_z &= 0+0+0+0 = 0
\end{align}

So we conclude, explicitly, that $\vec{P} = 0$ for this distribution as well. The components of the Quadrupole tensor can be found through the same process as above. For this case, we note that any and all terms involving $z$ are automatically zero because the charges are in the $xy$ plane. That elminiates 5 of 9 components that we need to calculate. Furthermore, since we know that $Q_{33}$ is zero, we have the additonal constraint that $Q_{22} = - Q_{11}$ for the tensor to remain traceless. This leaves us with only two components to calculate:

\begin{align}
Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)}  ... \\
&+ (3x_1^{(4)}x_1^{(4)} - (r_1^{(4)})^2)q^{(4)}\\
&= -(3 a^2 - a^2)(-q) + (3(0)(0) - a^2)(q) + (3 a^2 - a^2)(-q) - (3(0)(0) - a^2)q \\
&= -6a^2 \\
Q_{12} &= (3x_1^{(1)}x_2^{(1)})q^{(1)} + (3x_1^{(2)}x_2^{(2)})q^{(2)} + (3x_1^{(3)}x_2^{(3)})q^{(3)}+(3x_1^{(4)}x_2^{(4)})q^{(4)} \\
&= \cancel{(3(a)(0))q} + \cancel{(3(0)(a))q} + \cancel{(3(0)(a))q} + \cancel{(3(a)(0))q} = 0 
\end{align}

Which in retrospect was obvious because no individual charge has both an $x$ and $y$ component. The quadrupole moment tensor is, then,

\begin{align}
Q = -6a^2
\begin{bmatrix}
-1 & 0 & 0\\
0 &1 & 0 \\
0 & 0 & 0 
\end{bmatrix}
\end{align}

In spherical coordinates, using the same logic as above, no monopole or dipole moments means $q_{1m}$ and $q_{00}$ vanish. This time, $q_{20}$ also vanishes because $Q_{33}=0$. The only nonzero $q_{lm}$ is, then:

\begin{align}
q_{22} =  \frac{1}{12}\sqrt{\frac{15}{2\pi}}(Q_{11} - Q_{22}) = \frac{1}{12}\sqrt{\frac{15}{2\pi}}(8a^2) = \frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2
\end{align}

\subsection{Far Field Potential}

Using Jackson Equation 4.1 and plugging in our results from above, we have, for the first source distribution,

\begin{align}
\Phi &= \kay \sum_{l=0}^{\infty}\sum_{m=-l}^l \frac{4\pi}{2l + 1}q_{lm}\frac{Y_{lm}(\theta,\phi)}{r^{l+1}} \\
&=
\begin{cases}
\frac{1}{\cancel{4 \pi} \epsilon_0} \sqrt{\frac{5}{\pi}}a^2 \frac{\cancel{4\pi}}{5}\frac{1}{4}\sqrt{\frac{5}{\pi}}\left[3\cos^2(\theta)-1\right]r^{-3} = \frac{a^2}{4 \pi}\left[3\cos^2(\theta)-1\right]r^{-3}, & \ \text{Distribution 1} \\ 
\frac{1}{4 \pi \epsilon_0}\frac{4 \pi}{5}\frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2\left[\frac{1}{4}\sqrt{\frac{15}{2\pi}}\right]\sin^2(\theta)e^{2i\phi}r^{-3} = \frac{a^2}{\epsilon_0 \pi}\sin^2(\theta)e^{2i\phi}r^{-3}, & \ \text{Distribution 2} 
\end{cases}
\end{align}

\section{Problem 3}

We are given a "localized" charge distribution in spherical coordinates with the following form:

\begin{align}
\rho(\vec{r}) = \frac{1}{64 \pi}r^2 e^{-r}\sin^2(\theta)
\end{align}

\subsection{Multipole Moments}

The formula for multipole moments is given in Jackson as

\begin{align}
q_{lm} &= \int_v Y^*_{lm}(\theta',\phi')r'^l\rho(\vec{r'})d^3\vec{r'} 
\end{align}

Right off the bat we make two observations that simplify our task. Firstly, the problem is azimuthally symmetric. Thus we can write our spherical harmonics, per Jackson equation 3.57, as

\begin{align}
Y_{l0}(\theta) = \sqrt{\frac{2l + 1}{4 \pi}}P_l(\cos(\theta)) 
\end{align}

Secondly, we note that since the charge distribution has \textit{at most} a $sin^2(\theta)$ term, we are guaranteed that we will not have multipole moments higher then $l=2$, as those contain Legendre polynomails that carry $\sin^3(\theta)$ terms and higher. This leaves us with no more then 6 coefficients to evaluate, since the $-m$ versions are related to the $+m$ versions via $q_{l, -m} = (-1)^m q^*_{lm}$. The general formula for the $q_{l0}$ then is as follows, plugging in our particular charge distribution:

\begin{align}
q_{l0} &= \int_v \sqrt{\frac{2l + 1}{4\pi}}P_l(\cos(\theta'))r'^l*\left[\frac{1}{64\pi}r'^2 e^{-r'}\sin^2(\theta')\right]r'^2\sin(\theta')dr'd\theta'd\phi' \\
&= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{64\pi} \int_{r=0}^{r=\infty}\int_{\theta=0}^{\theta=\pi}\int_{\phi=0}^{\phi=2\pi}r'^{l+4}e^{-r}\sin^3(\theta')P_l(\cos(\theta'))d\phi'd\theta'dr'
\end{align}

$\phi$ integrates out and we pick up a factor of $2 \pi$. The $\theta$ and $r$ integrals are independent and thus factorize.

\begin{align}
q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_0^{\pi}\sin^3(\theta')P_l(\cos(\theta')d\theta'\right]
\end{align}

The $\theta$ integral can be accomplished via change of variables $x=\cos(\theta)$. Then $dx = -\sin(\theta)d\theta$ and $d\theta = -\frac{dx}{\sin(\theta)}$. Noting that $\sin^3(\theta) = \sin^2(\theta)\sin(\theta) = \sin(\theta)[1-\cos^2(\theta)]$, we know that the factor $\sin^3(\theta)d\theta = (1 - x^2)dx$. Finally, the limits of integration change as $x = \cos(0) \rightarrow 1$  and $x = \cos(\pi) \rightarrow -1$.

\begin{align}
q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_{-1}^{1}(1-x^2)P_l(x)dx\right]
\end{align}

Splitting the $x$ integral up into two pieces, one sees that the first piece,

\begin{align}
\int_{-1}^{1}P_l(x)dx
\end{align}

Can be written as 
\begin{align}
\int_{-1}^{1}P_l(x)P_0(x)dx
\end{align}

Since $P_0(x) = 1$. Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$, Since $P_0(x) = 1$. Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$. This means that the integral above is zero unless $l=0$, in which case it just equals $2$. The second integral is more annoying and requires Jackson equation 3.32:

\begin{align}
\int_{-1}^1x^2P_l(x)P_n(x)dx = \begin{cases}
\frac{2(n+1)(n+2)}{(2n+1)(2n+1)(2n+3)}, & l=n+2\\
\frac{2(2n^2 + 2n - 1)}{(2n-1)(2n+1)(2n+5)}, & l=n\\
\ \ \ \ \ \ \ \ \ \  0, & l = \text{Anything Else}
\end{cases}
\end{align}

In our case $n=0$ and we are concerned with the $l=0$ and $l=2$ cases. We have thus concluded that the $\theta$ portion of the integral, and thus the entire integral, and thus the multipole moment, is zero for all values of $l$ except for $l=0$ and $l=2$. Simplifying the $\theta$ integral for the two nonzero cases we get

\begin{align}
\int_{\theta} = 
\begin{cases}
2 - \int_{-1}^1 x^2 P_0(x)P_l(x)dx = 2 - \frac{2}{3} = \frac{4}{3}, & l=0 \\
-\int_{-1}^1 x^2 P_0(x)P_2(x)dx = -\frac{4}{15}, & l=2  
\end{cases}
\end{align}

This leaves us with the integral over $r$ which we recognize as the definition of the factorial function for the exponent of $r$:

\begin{align}
\int_0^{\infty}r^n e^{-r}dr \equiv n!
\end{align}

So we have

\begin{align}
\int_r = 
\begin{cases}
\int_0^{\infty}r^4e^{-r}dr = 4! = 24, & l=0 \\
\int_0^{\infty}r^6e^{-r}dr = 6! = 720, & l=2
\end{cases}
\end{align}

Combining everything together we have:

\begin{align}
q_{l0} = 
\begin{cases}
\sqrt{\frac{1}{4\pi}}\frac{1}{32}\left(\frac{4}{3}\right)(24) = \sqrt{\frac{1}{4\pi}}, & l=0 \\
\sqrt{\frac{4 + 1}{4\pi}}\frac{1}{32}\left(-\frac{4}{15}\right)(720) = -3\sqrt{\frac{5}{\pi}}, & l=2
\end{cases}
\end{align}

This means that if the general formula for the multipole expansion is

\begin{align}
\Phi = \kay \sum_{l=0}^{\infty}\sum_{m=-l}^{l}\frac{4\pi}{2l + 1}q_{lm}Y_{lm}(\theta,\phi)r^{-(l+1)}
\end{align}

Then we would have

\begin{align}
\Phi &= \kay \left[\cancel{\frac{4\pi}{1}}\cancel{\sqrt{\frac{1}{4 \pi}}}P_0(\cos(\theta))\cancel{\sqrt{\frac{1}{4\pi}}}r^{-1} + \cancel{\frac{4\pi}{5}}\left(-3\cancel{\sqrt{\frac{5}{\pi}}}\right)\cancel{\frac{1}{4}\sqrt{\frac{5}{\pi}}}P_2(\cos(\theta))r^{-3}  \right] \\
&= \kay \left[\frac{1}{r} - \frac{3}{2r^3}(3\cos^2(\theta) - 1)\right]
\end{align}

As our final expression. In the far-field, we can probably neglect the $r^{-3}$ term, which would give us just the potential of a point charge, $\Phi = \kay \frac{1}{r}$ as our answer.

\section{Problem 4}
By calculating the moments up to quadrupolar order, we wish to show tha the charge distribution $\rho(\vec{r}) = -(\vec{P}\cdot \nabla)\delta(\vec{r})$ represents an elementary dipole with dipole moment $\vec{P}$ located at the origin, given the hint that this is most easily accomplished in cartesian coordinates.

\subsection{Monopole Moment}

Directly integrating the charge distribution over space gives us an integral which we can expand as follows:

\begin{align}
&= - \int_v (\vec{P}\cdot \nabla)\delta(\vec{r})d^3(\vec{r'}) \\
&= -\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[P_x \partd{}{x} + P_y \partd{}{y} + P_z \partd{}{z}\right] \delta (x) \delta (y) \delta (z) dx dy dz
\end{align}

Picking out just the $x$ term, we have

\begin{align}
&= \int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}\delta(z)dz \int_{-\infty}^{\infty}P_x\partd{}{x}\delta(x)dx  \\
&= \int_{-\infty}^{\infty}P_x\partd{\delta(x)}{x}dx
\end{align}

Integrating by parts with $u = P_x, du = P'_x dx, dv = \delta'(x)dx, v=\delta(x)$, we have

\begin{align}
\cancel{P_x \delta(x)\bigg|_{-\infty}^{\infty}} - \int_{-\infty}^{\infty}\delta(x)P'_x(x)dx
\end{align}

The first term vanishes at the boundaries because $\delta(x) = 0$ everywhere except $x=0$. The second, integral term vanishes as long as $\vec{P} = $ const. This implies that the $x$ integral is zero, and applying the same logic to the other portions of the total integral, we get a monopole term that vanishes completely, as is expected for a pure dipole.

\subsection{Dipole Moment}

This is an integral of the form

\begin{align}
-\int_v \vec{r}\left[\vec{P} \cdot \nabla\right] d^3(\vec{r})
\end{align}

Which can be expanded explicitly in cartesian coordinates as

\begin{align}
-\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[x\hat{x} +y\hat{y} + z\hat{z}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
\end{align}

From this we can see that the integral breaks up into two types of terms:

\begin{enumerate}
\item \textit{Uniform} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_x\partd{}{x}\delta(x)\delta(y)\delta(z) dxdydz$
\item \textit{Mixed} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_y\partd{}{y}\delta(x)\delta(y)\delta(z)dxdydz$ 
\end{enumerate}

Examining the uniform case in $x$, we integrate by parts with $u=xP_x \hat{x}, du=P_xdx\hat{x}, v=\delta'(x)dx, v = \delta(x)$ and including the negative out front, obtain

\begin{align}
\cancel{-xP_x\delta(x)\bigg|_{-\infty}^{\infty}} + \int_{-\infty}^{\infty}P_x\delta(x)dx\hat{x}
\end{align}

Which produces the value of $P_x\hat{x}$ from the term on the right hand side. However, the \textit{mixed} integral cases look like

\begin{align}
\int_{-\infty}^{\infty}x\hat{x}\delta(x)dx\int_{-\infty}^{\infty}\delta(z)dz\cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} = 0
\end{align}

The last integral factor on the right hand side is zero, because it is of the exact same form as the monopole integral above. Applying the same logic to the other 7 integrals, we end up with simply $P_x \hat{x} + P_y \hat{y} + P_z\hat{z}$, which corresponds to a pure dipole of magnitude $P$ located at the origin.

\subsection{Quadrupole Moment}

The integral for the quadrupole moment tensor components, given in Jackson, is

\begin{align}
Q_{ij} = \int_v \left[3x_ix_j - r^2\delta_{ij}\right]\rho(x)d^3(x)
\end{align}

Inserting our charge distribution yields the integral

\begin{align}
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\left[3x_ix_j - r^2 \delta_{ij}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
\end{align}

Let's examine the first term in brackets. For $i=j$, we have, for instance $3x^2 - x^2 - y^2 - z^2 = 2x^2 - y^2 - z^2$. For $i\neq j$, we have terms that look like just $xy$. The first term of the diagonal case reads

\begin{align}
-2\int_{-\infty}^{\infty}\delta(x)dx\int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}x^2P_x\delta'(x)dx
\end{align}

Integrating the $x-$factor by parts with $u=x^2P_x, du=2xP_x dx, dv=\delta'(x)dx, v=\delta(x)$ the boundary term cancels as usual and the resulting integral is

\begin{align}
-2\int_{-\infty}^{\infty} \delta(x)xP_xdx 
\end{align}

Which vanishes, because the integrand $x|_{x=0} = 0$ at the location of the delta function. The other terms in the $xx$ case are integrals of the form

\begin{align}
-2\int_{-\infty}^{\infty}x^2\delta(x)dx \int_{-\infty}^{\infty}\delta(z)dz \cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} 
\end{align}

The last factor on the right vanishes as it is another monopole-like integral already shown to be zero. The same construction of course works for the other diagonal elements. That just leaves the off diagonal case, which consists of integrals like

\begin{align}
3\int_{-\infty}^{\infty}xP_x\delta'(x)\cancel{\int_{-\infty}^{\infty}y\delta(y)dy}\int_{-\infty}^{\infty}\delta(z)dz
\end{align}

and

\begin{align}
3\int_{-\infty}^{\infty}x\delta{x}dx\int_{-\infty}^{\infty}y\delta(y)dy\int_{-\infty}^{\infty}P_z\partd{}{z}\delta{z}dz
\end{align}

In the latter case, \textit{all three} integral factors are zero, making the three out of nine cases that look like this one \textit{exceptionally} zero.

\end{document}

### 2016-08-10 Gradschool/thesis/detAMR.tex
initial commit of thesis stuff

--- 2016-06-10 Gradschool/electrodynamics/em_hw4.tex
@@
 \usepackage{braket}
-\usepackage{float}
-\usepackage{cancel}
-\usepackage{graphicx}
-
-\newcommand{\partd}[2]{\frac{\partial #1}{\partial #2}}
-\newcommand{\partdd}[2]{\frac{\partial^2 #1}{\partial #2^2}}
-%\newcommand{\coth}{\text{coth}}
-\newcommand{\csch}{\text{csch}}
-\newcommand{\kay}{\frac{1}{4 \pi \epsilon_0}}
 
@@
 \horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge Electrodynamics - Homework 4 \\ % The assignment title
+\huge Detector AM Response Vectorization \\ % The assignment title
 \horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
@@
 
-\section{Problem 1}
-
-We are asked to investigate the Diriclet problem in prolate spheroidal coordinates via separation of variables.
-Like any similar problem involving curvilinear coordinates our first task is to cast the Laplacian into the appropriate form so that we can write Laplace's equation:
+\section{ComputeDetAMResponse}
+The output of this function are the components of what will become a complex number, $F_{+}$ and $F_{\times}$. These are long sums of different terms that, based on the current code, look like this:
 
 \begin{align}
-\nabla^2 \Phi = 0
+F_{+} &= [R_{0,0}X_0 + R_{0,1}X_1 + R_{0,2}X_2]X_0 - [R_{0,0}Y_0 + R_{0,1}Y_1 + R_{0,2}Y_2]Y_0 \\
+&+ [R_{1,0}X_0 + R_{1,1}X_1 + R_{1,2}X_2]X_1 - [R_{1,0}Y_0 + R_{1,1}Y_1 + R_{1,2}Y_2]Y_1 \\
+&+ [R_{2,0}X_0 + R_{2,1}X_1 + R_{2,2}X_2]X_2 - [R_{2,0}Y_0 + R_{2,1}Y_1 + R_{2,2}Y_2]Y_2
 \end{align}
 
-The general form of the Laplacian in curvilinear coordinates is given by
+Collecting all the positive terms of this expression gives
+\begin{align}
+F_{+} &= R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\
+&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\ 
+&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 
+\end{align}
+At which point we see that the same result is computable as an outer product of the vector $\vec{X}$ using
 
 \begin{align}
-\nabla \cdot \nabla = \nabla^2 = \frac{1}{h_1 h_2 h_3}\left[\partd{}{u_1}\left(\frac{h_2 h_3}{h_1}\partd{}{u_2}\right) + \partd{}{u_2}\left(\frac{h_3h_1}{h_2}\partd{}{u_2}\right) + \partd{}{u_3}\left(\frac{h_1h_2}{h_3}\partd{}{u_3}\right)\right]
+F_{+} &= \vec{X}\mathbf{R}\vec{X} - \vec{Y}\mathbf{R}\vec{Y}
 \end{align}
 
-Where the $h_i$ are the scale factors and run over all the coordinates relevant to the space - in the case of prolate spheroidal coordinates, these are $\alpha, \beta$, and $\phi$.
-These parameters are given in cartesian coordinates as
+And the complex part of the gravitational wave is thus
 
 \begin{align}
-x &= c \sinh(\alpha)\sin(\beta)\cos(\phi) \\
-y &= c \sinh(\alpha)\sin(\beta)\sin(\phi) \\
-z &= c\cosh(\alpha)\cos(\beta)
+F_{\times} &= \vec{X}\mathbf{R}\vec{Y} + \vec{Y}\mathbf{R}\vec{X}
 \end{align}
 
-Whereas the scale factors themselves are given by
+One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
+Thus we need a function that takes as input a \textit{list} vectors (or \textit{vector}) of vectors and produces a vector with the right components as output.
+To that end we define the tensor $X^{i}_{j}$ where
 
 \begin{align}
-h_i &= \left|\partd{\vec{r}}{q_i} \right| \\
-&= \sqrt{\partd{\vec{r}}{q_i} \cdot \partd{\vec{r}}{q_i}} \\
-\end{align}
-
-For $\vec{r} = x\hat{x} + y\hat{y} + z\hat{z}$.
-We need to plug in our coordinate transformations to the components of our position vector and then take derivatives with respect to those coordinates to get the scale factors.
-The position vector looks like
-
-\begin{align}
-\vec{r} &= c\sinh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\sinh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\cosh(\alpha)\cos(\beta)\hat{z} \\
-\partd{\vec{r}}{\alpha} &= c\cosh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\cosh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\sinh(\alpha)\cos(\beta)\hat{z}\\
-\end{align}
-
-Dotting this with itself gives
-
-\begin{align}
-\partd{\vec{r}}{\alpha} \cdot \partd{\vec{r}}{\alpha} &= c^2 \cosh^2(\alpha)\sin^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
-&= c^2\cosh^2(\alpha)\sin^2(\beta)(\cos^2(\phi) + \sin^2(\phi)) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
-&= c^2\left[\cosh^2(\alpha)\sin^2(\beta) + \sinh^2(\alpha)\cos^2(\beta)\right] \\
-&= c^2\left[(1 + \sinh^2(\alpha))\sin^2(\beta) + \sinh^2(\alpha)(1 - \sin^2(\beta))\right] \\
-&= c^2\left[\cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sinh^2(\alpha)\right]\\
-h_{\alpha} &= \sqrt{\sin^2(\beta) + \sinh^2(\alpha)} 
-\end{align}
-
-Similarly for the other coordinates
-
-\begin{align}
-\partd{\vec{r}}{\beta}\cdot \partd{\vec{r}}{\beta}&= c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\ 
-&= c^2\sinh^2(\alpha)\cos^2(\beta) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\
-&= c^2\left[\sinh^2(\alpha)(1-\sin^2(\beta)) + (\sinh^2(\alpha) + 1)\sin^2(\beta)\right]\\
-&= c^2\left[\sinh^2(\alpha) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta)\right] \\
-h_{\beta} &= \sqrt{c^2\sinh^2(\alpha) +\sin^2(\beta)} 
-\end{align}
-
-\begin{align}
-\partd{\vec{r}}{\phi} \cdot \partd{\vec{r}}{\phi} &= c^2\sinh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + \sinh^2(\alpha)\sin^2(\beta)\cos^2(\phi) \\
-&= c^2\sinh^2(\alpha)\sin^2(\beta)(\sin^2(\phi) + \cos^2(\phi)) \\
-&= c^2\sinh^2(\alpha)\sin^2(\beta) \\
-h_{\phi} &= \sqrt{c^2\sinh^2(\alpha)\sin^2(\beta)}
-\end{align}
-
-Since $h_1 = h_2$ the definition of the Laplacian reduces in this case to
-
-\begin{align}
-\frac{1}{h_1h_2h_3}\left[\partd{}{\alpha}h_3\partd{}{\alpha} + \partd{}{\beta}h_3\partd{}{\beta} + \partd{}{\phi}\frac{h_1h_2}{h_3}\partd{}{\phi}\right] \\
-\end{align}
-
-The term in the brackets simplifies as
-
-\begin{align}
-&=\partd{}{\alpha}(\sinh(\alpha)\sin(\beta))\partd{}{\alpha}\partd{}{\alpha} + \partd{}{\beta}(\sinh(\alpha)\sin(\beta))\partd{}{\beta} + \partd{}{\phi}\left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partd{}{\phi} \\
-&= \cosh(\alpha)\sin(\beta)\partd{}{\alpha} + \sinh(\alpha)\sin(\beta)\partdd{}{\alpha} + \sinh(\alpha)\cos(\beta)\partd{}{\beta} + \sinh(\alpha)\sin(\beta)\partdd{}{\beta} ... \\
-&+ \left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partdd{}{\phi}
-\end{align}
-
-Dividing out the $\sinh(\alpha)\sin(\beta)$ term from the prefactor gets us the final Laplacian:
-
-\begin{align}
-\nabla^2 = \frac{1}{c^2(\sin^2\beta)+\sinh^2(\alpha)}\left[\coth(\alpha)\partd{}{\alpha} + \cot(\beta)\partd{}{\beta}+ \partdd{}{\alpha} + \partdd{}{\beta} + (\csch(\alpha) + \csc(\beta))\partdd{}{\phi}\right]
-\end{align}
-
-We are given that the problem has azimuthal symmetry, so the ($\phi$) derivatives are zero.
-This leads us to assume a solution of the form $\Phi = A(\alpha)B(\beta)$.
-Acting our Laplacian, sans the $\phi$ derivative part, on our ansatz gets us the following
-
-\begin{align}
-\partdd{A}{\alpha}B + \partdd{B}{\beta}A + B\coth(\alpha)\partd{A}{\alpha} + A\cot(\beta)\partd{B}{\beta} &= 0 \\
-\frac{1}{A}\partdd{A}{\alpha} + \frac{1}{B}\partdd{B}{\beta} + \frac{\coth(\alpha)}{A}\partd{A}{\alpha} + \frac{\cot(\beta)}{B}\partd{B}{\beta} &= 0
-\end{align}
-
-As usual the only way two functions of separate varibles can be equal for all values of their respective arguments are if they are both equal to the same constant.
-In anticipation for transforming our result into Legendre's differential equation we select $l(l+1)$ for our constant and arrive at the following two second order ODEs as our solutions:
-
-\begin{align}
-\partdd{A}{\alpha} + \coth(\alpha)\partd{A}{\alpha} - Al(l+1) &= 0 \\
-\partdd{B}{\beta} + \cot(\beta)\partd{B}{\beta} + Bl(l+1) &= 0
-\end{align}
-
-Starting with the equation for $A$, we would like to transform our derivatives from $\partd{}{\alpha}$ to $\partd{}{\cosh(\alpha)}$.
-We can write $\partd{}{\alpha} = \partd{\cosh(\alpha)}{\alpha}\partd{}{\cosh(\alpha)} = \sinh(\alpha)\partd{}{\cosh(\alpha)}$.
-Taking another derivative of this expression via product rule, we get $\partd{}{\alpha}\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh(\alpha)\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{}{\cosh(\alpha)}$.
-With this information the equation for $A$ transforms as
-
-\begin{align}
-\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \coth(\alpha)\sinh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
-\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \cosh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
-\sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + 2 \cosh(\alpha)\partd{A}{\alpha} - Al(l+1)) = 0 \\
--(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} + 2\cosh(\alpha)\partd{A}{\alpha}-Al(l+1) = 0 \\
-(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} - 2\cosh(\alpha)\partd{A}{\alpha} + Al(l+1) = 0
-\end{align}
-
-Which is Legendre's equation for $A(\alpha)$.
-As for the $B$ equation, we want the argument of the resulting Legendre polynomials to be $\cos(\beta)$.
-So we note that $\partd{}{\beta} = \partd{\cos(\beta)}{\beta}\partd{}{\cos(\beta)} = -\sin(\beta)\partd{}{\cos(\beta)}$ and $\partdd{}{\beta} = \partd{}{\beta}\left[-\sin(\beta)\partd{}{\cos(\beta)}\right] = -\cos(\beta)\partd{}{\cos(\beta)} + \sin^2(\beta)\partdd{}{\cos(\beta)}$.
-Acting the revised operators on $B$ yields
-
-\begin{align}
-\sin^2(\beta)\partdd{B}{\cos(\beta)} - \cos(\beta)\partd{B}{\cos(\beta)} - \cot(\beta)\sin(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0 \\
-(1-\cos^2(\beta))\partdd{B}{\cos(\beta)} - 2 \cos(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0
-\end{align}
-
-Which is again Legendre's equation for $B(\beta)$. Thus we can represent our solution as
-
-\begin{align}
-\Phi = \sum_n A_n P_n(\cosh(\alpha)P_n(\cos(\beta))
-\end{align}
-
-Which is the desired result.
-
-\section{Problem 2}
-
-We want to the scalar monopole moment, vector dipole moment, and tensor quadrupole moment for the two charge distributions pictured above, for both cartesian and spherical coordinates.
-The moments themselves are defined as follows:
-
-\begin{align}
-M &= \kay \int_v \rho(\vec{r'})d^3\vec{r'} = \ \text{Total Charge}  \\
-\vec{P} &= \kay \int_v \vec{r'}\rho(\vec{r})d^3\vec{r'} = \ \text{Dipole Moment Vector} \\
-Q_{ij} &= \kay \int_v (3x_ix_j - r_i^2\delta_{ij})\rho(\vec{r'})d^3(r') = \ \text{Quadrupole Moment Tensor Components}
-\end{align}
-
-We note that the total charge of the system sums to zero.
-Thus we have no monopole moment for the charge distribution.
-As for the dipole moment, the relevant formula in the discrete case would have to be
-
-\begin{align}
-\vec{P}_i &= \sum_{n=1}^3 r_i^{(n)}q_i^{(n)} \ \text{\textit{Component i} of the dipole vector} \\
-\end{align}
-
-\subsection{Distribution 1}
-
-Labelling the charges as shown in the figure, the dipole moment  would come out to be $qa\hat{z} - qa\hat{z} -\cancel{2q(0)} = 0$, when summing over all the charges.
-Thus the dipole moment of the first charge distribution is also zero.
-The discrete quadrupole formula written as a summation over a discrete charge distribution must be
-
-\begin{align}
-Q_{ij} = \sum_{n=1}^3 (3x_i^{(n)}x_j^{(n)} - (r_i^{(n)})^2\delta_{ij})q^{(n)}
-\end{align}
-
-For the charge distribution in question we can expand this formula as follows:
-
-\begin{align}
-Q_{ij} &= (3x_i^{(1)}x_j^{(1)} - (r_i^{(1)})^2\delta_{ij})q^{(1)} + (3x_i^{(2)}x_j^{(2)} - (r_i^{(2)})^2\delta_{ij})q^{(2)} + (3x_i^{(3)}x_j^{(3)} - (r_i^{(3)})^2\delta_{ij})q^{(3)} \\
-\end{align}
-
-Let's look at the off-diagonal case first.
-If $i \neq j$, then obviously all the $\delta_{ij}$ cancel.
-Furthermore, when we notice that the charges lie only on the $z$ axis, we find that since \textit{at least one} of the coordinates ($i$ or $j$) must \textit{not} be $z$, all of the other products cancel too.
-Thus we conclude that all of the off-diagonal elements in this tensor are zero.
-This leaves us with three to calculate:
-
-\begin{align}
-Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)} \\
-&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\ 
-&= -2a^2q \\
-Q_{22} &= (3x_2^{(1)}x_2^{(1)} - (r_2^{(1)})^2)q^{(1)} + (3x_2^{(2)}x_2^{(2)} - (r_2^{(2)})^2)q^{(2)} + (3x_2^{(3)}x_2^{(3)} - (r_2^{(3)})^2)q^{(3)} \\
-&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\
-&= -2a^2q \\
-Q_{33} &= (3x_3^{(1)}x_3^{(1)} - (r_3^{(1)})^2)q^{(1)} + (3x_3^{(2)}x_3^{(2)} - (r_3^{(2)})^2)q^{(2)} + (3x_3^{(3)}x_3^{(3)} - (r_3^{(3)})^2)q^{(3)} \\
-&= (3a^2 - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (3a^2 - a^2)q \\
-&= 4a^2q
-\end{align}
-
-Thus, the quadrupole moment tensor, which we note is traceless, is
-
-\begin{align}
-Q = 2a^2
+X^i &= 
 \begin{bmatrix}
--1 & 0 & 0 \\
-0 & -1 & 0 \\
-0 & 0 & 2
+X^{i}_0 \\
+X^{i}_1 \\
+X^{i}_2
 \end{bmatrix}
@@
 
-To translate to spherical coordinates, we make use of Jackson equations 4.4 through 4.6.
-Since the total charge and dipole moments are known to be zero, the $l=0$ and $l=1$ $q$'s are all zero.
-Furthermore, if all the off diagonal terms of $Q$ are zero, $q_{21}$ is automatically zero.
-Even more, $q_{22} = \text{const.} * (Q_{11} - \cancel{iQ_{12}} - Q_{22})$ is zero because $Q_{11} = Q_{22}$.
-This leaves only the $q_{20}$ moment to calculate.
+As well as the tensor $R^{i}_{jk}$ where
 
 \begin{align}
-q_{20} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}Q_{33} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}(4a^2) = \sqrt{\frac{5}{\pi}}a^2
-\end{align}
-
-As the only nonzero $q_{lm}$.
-
-\subsection{Distribution 2}
-
-We note that as before the total charge over the system is zero. So we must have a monopole moment that vanishes. The components of the dipole moment vector $\vec{P}$ would be as follows:
-
-\begin{align}
-P_x &= r_x^{(1)}q^{(1)} + r_x^{(2)}q^{(2)} + r_x^{(3)}q^{(3)} + r_x^{(4)}q^{(4)} \\
-&= a(-q) + \cancel{0(q)} -a(-q) + \cancel{a(0)} = 0  \\
-P_y &= r_y^{(1)}q^{(1)} + r_y^{(2)}q^{(2)} + r_y^{(3)}q^{(3)} + r_y^{(4)}q^{(4)} \\
-&= \cancel{(0)-q} -a(q) +\cancel{(0)(-q)} -a(-q) = 0 \\
-P_z &= 0+0+0+0 = 0
-\end{align}
-
-So we conclude, explicitly, that $\vec{P} = 0$ for this distribution as well.
-The components of the Quadrupole tensor can be found through the same process as above.
-For this case, we note that any and all terms involving $z$ are automatically zero because the charges are in the $xy$ plane.
-That elminiates 5 of 9 components that we need to calculate.
-Furthermore, since we know that $Q_{33}$ is zero, we have the additonal constraint that $Q_{22} = - Q_{11}$ for the tensor to remain traceless.
-This leaves us with only two components to calculate:
-
-\begin{align}
-Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)}  ... \\
-&+ (3x_1^{(4)}x_1^{(4)} - (r_1^{(4)})^2)q^{(4)}\\
-&= -(3 a^2 - a^2)(-q) + (3(0)(0) - a^2)(q) + (3 a^2 - a^2)(-q) - (3(0)(0) - a^2)q \\
-&= -6a^2 \\
-Q_{12} &= (3x_1^{(1)}x_2^{(1)})q^{(1)} + (3x_1^{(2)}x_2^{(2)})q^{(2)} + (3x_1^{(3)}x_2^{(3)})q^{(3)}+(3x_1^{(4)}x_2^{(4)})q^{(4)} \\
-&= \cancel{(3(a)(0))q} + \cancel{(3(0)(a))q} + \cancel{(3(0)(a))q} + \cancel{(3(a)(0))q} = 0 
-\end{align}
-
-Which in retrospect was obvious because no individual charge has both an $x$ and $y$ component. The quadrupole moment tensor is, then,
-
-\begin{align}
-Q = -6a^2
+R^{i} &= 
 \begin{bmatrix}
--1 & 0 & 0\\
-0 &1 & 0 \\
-0 & 0 & 0 
+R^{i}_{00} & R^{i}_{01} & R^{i}_{02} \\ 
+R^{i}_{10} & R^{i}_{11} & R^{i}_{12} \\ 
+R^{i}_{20} & R^{i}_{21} & R^{i}_{22}  
 \end{bmatrix}
@@
 
-In spherical coordinates, using the same logic as above, no monopole or dipole moments means $q_{1m}$ and $q_{00}$ vanish.
-This time, $q_{20}$ also vanishes because $Q_{33}=0$.
-The only nonzero $q_{lm}$ is, then:
+The tensor $X$ is like a stack of all the different possible $\vec{X}$ coming out of the page.
+The Tensor $R$ is like $n$ copies of the matrix $\mathbf{R}$ stacked on top of each other.
+In this way the desired vector is obtainable with the tensor contraction
 
 \begin{align}
-q_{22} =  \frac{1}{12}\sqrt{\frac{15}{2\pi}}(Q_{11} - Q_{22}) = \frac{1}{12}\sqrt{\frac{15}{2\pi}}(8a^2) = \frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2
+F^{i}_{+} &= X^{lm}R^{i}_{lj}X^{j}_{m} - Y^{lm}R^{i}_{lj}Y^{j}_{m} \\ 
+F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
 \end{align}
 
-\subsection{Far Field Potential}
-
-Using Jackson Equation 4.1 and plugging in our results from above, we have, for the first source distribution,
-
-\begin{align}
-\Phi &= \kay \sum_{l=0}^{\infty}\sum_{m=-l}^l \frac{4\pi}{2l + 1}q_{lm}\frac{Y_{lm}(\theta,\phi)}{r^{l+1}} \\
-&=
-\begin{cases}
-\frac{1}{\cancel{4 \pi} \epsilon_0} \sqrt{\frac{5}{\pi}}a^2 \frac{\cancel{4\pi}}{5}\frac{1}{4}\sqrt{\frac{5}{\pi}}\left[3\cos^2(\theta)-1\right]r^{-3} = \frac{a^2}{4 \pi}\left[3\cos^2(\theta)-1\right]r^{-3}, & \ \text{Distribution 1} \\
-\frac{1}{4 \pi \epsilon_0}\frac{4 \pi}{5}\frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2\left[\frac{1}{4}\sqrt{\frac{15}{2\pi}}\right]\sin^2(\theta)e^{2i\phi}r^{-3} = \frac{a^2}{\epsilon_0 \pi}\sin^2(\theta)e^{2i\phi}r^{-3}, & \ \text{Distribution 2}
-\end{cases}
-\end{align}
-
-\section{Problem 3}
-
-We are given a "localized" charge distribution in spherical coordinates with the following form:
-
-\begin{align}
-\rho(\vec{r}) = \frac{1}{64 \pi}r^2 e^{-r}\sin^2(\theta)
-\end{align}
-
-\subsection{Multipole Moments}
-
-The formula for multipole moments is given in Jackson as
-
-\begin{align}
-q_{lm} &= \int_v Y^*_{lm}(\theta',\phi')r'^l\rho(\vec{r'})d^3\vec{r'} 
-\end{align}
-
-Right off the bat we make two observations that simplify our task. Firstly, the problem is azimuthally symmetric. Thus we can write our spherical harmonics, per Jackson equation 3.57, as
-
-\begin{align}
-Y_{l0}(\theta) = \sqrt{\frac{2l + 1}{4 \pi}}P_l(\cos(\theta)) 
-\end{align}
-
-Secondly, we note that since the charge distribution has \textit{at most} a $sin^2(\theta)$ term, we are guaranteed that we will not have multipole moments higher then $l=2$, as those contain Legendre polynomails that carry $\sin^3(\theta)$ terms and higher.
-This leaves us with no more then 6 coefficients to evaluate, since the $-m$ versions are related to the $+m$ versions via $q_{l, -m} = (-1)^m q^*_{lm}$.
-The general formula for the $q_{l0}$ then is as follows, plugging in our particular charge distribution:
-
-\begin{align}
-q_{l0} &= \int_v \sqrt{\frac{2l + 1}{4\pi}}P_l(\cos(\theta'))r'^l*\left[\frac{1}{64\pi}r'^2 e^{-r'}\sin^2(\theta')\right]r'^2\sin(\theta')dr'd\theta'd\phi' \\
-&= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{64\pi} \int_{r=0}^{r=\infty}\int_{\theta=0}^{\theta=\pi}\int_{\phi=0}^{\phi=2\pi}r'^{l+4}e^{-r}\sin^3(\theta')P_l(\cos(\theta'))d\phi'd\theta'dr'
-\end{align}
-
-$\phi$ integrates out and we pick up a factor of $2 \pi$. The $\theta$ and $r$ integrals are independent and thus factorize.
-
-\begin{align}
-q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_0^{\pi}\sin^3(\theta')P_l(\cos(\theta')d\theta'\right]
-\end{align}
-
-The $\theta$ integral can be accomplished via change of variables $x=\cos(\theta)$.
-Then $dx = -\sin(\theta)d\theta$ and $d\theta = -\frac{dx}{\sin(\theta)}$.
-Noting that $\sin^3(\theta) = \sin^2(\theta)\sin(\theta) = \sin(\theta)[1-\cos^2(\theta)]$, we know that the factor $\sin^3(\theta)d\theta = (1 - x^2)dx$.
-Finally, the limits of integration change as $x = \cos(0) \rightarrow 1$  and $x = \cos(\pi) \rightarrow -1$.
-
-\begin{align}
-q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_{-1}^{1}(1-x^2)P_l(x)dx\right]
-\end{align}
-
-Splitting the $x$ integral up into two pieces, one sees that the first piece,
-
-\begin{align}
-\int_{-1}^{1}P_l(x)dx
-\end{align}
-
-Can be written as 
-\begin{align}
-\int_{-1}^{1}P_l(x)P_0(x)dx
-\end{align}
-
-Since $P_0(x) = 1$.
-Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$, Since $P_0(x) = 1$.
-Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$.
-This means that the integral above is zero unless $l=0$, in which case it just equals $2$.
-The second integral is more annoying and requires Jackson equation 3.32:
-
-\begin{align}
-\int_{-1}^1x^2P_l(x)P_n(x)dx = \begin{cases}
-\frac{2(n+1)(n+2)}{(2n+1)(2n+1)(2n+3)}, & l=n+2\\
-\frac{2(2n^2 + 2n - 1)}{(2n-1)(2n+1)(2n+5)}, & l=n\\
-\ \ \ \ \ \ \ \ \ \  0, & l = \text{Anything Else}
-\end{cases}
-\end{align}
-
-In our case $n=0$ and we are concerned with the $l=0$ and $l=2$ cases.
-We have thus concluded that the $\theta$ portion of the integral, and thus the entire integral, and thus the multipole moment, is zero for all values of $l$ except for $l=0$ and $l=2$.
-Simplifying the $\theta$ integral for the two nonzero cases we get
-
-\begin{align}
-\int_{\theta} = 
-\begin{cases}
-2 - \int_{-1}^1 x^2 P_0(x)P_l(x)dx = 2 - \frac{2}{3} = \frac{4}{3}, & l=0 \\
--\int_{-1}^1 x^2 P_0(x)P_2(x)dx = -\frac{4}{15}, & l=2  
-\end{cases}
-\end{align}
-
-This leaves us with the integral over $r$ which we recognize as the definition of the factorial function for the exponent of $r$:
-
-\begin{align}
-\int_0^{\infty}r^n e^{-r}dr \equiv n!
-\end{align}
-
-So we have
-
-\begin{align}
-\int_r = 
-\begin{cases}
-\int_0^{\infty}r^4e^{-r}dr = 4! = 24, & l=0 \\
-\int_0^{\infty}r^6e^{-r}dr = 6! = 720, & l=2
-\end{cases}
-\end{align}
-
-Combining everything together we have:
-
-\begin{align}
-q_{l0} = 
-\begin{cases}
-\sqrt{\frac{1}{4\pi}}\frac{1}{32}\left(\frac{4}{3}\right)(24) = \sqrt{\frac{1}{4\pi}}, & l=0 \\
-\sqrt{\frac{4 + 1}{4\pi}}\frac{1}{32}\left(-\frac{4}{15}\right)(720) = -3\sqrt{\frac{5}{\pi}}, & l=2
-\end{cases}
-\end{align}
-
-This means that if the general formula for the multipole expansion is
-
-\begin{align}
-\Phi = \kay \sum_{l=0}^{\infty}\sum_{m=-l}^{l}\frac{4\pi}{2l + 1}q_{lm}Y_{lm}(\theta,\phi)r^{-(l+1)}
-\end{align}
-
-Then we would have
-
-\begin{align}
-\Phi &= \kay \left[\cancel{\frac{4\pi}{1}}\cancel{\sqrt{\frac{1}{4 \pi}}}P_0(\cos(\theta))\cancel{\sqrt{\frac{1}{4\pi}}}r^{-1} + \cancel{\frac{4\pi}{5}}\left(-3\cancel{\sqrt{\frac{5}{\pi}}}\right)\cancel{\frac{1}{4}\sqrt{\frac{5}{\pi}}}P_2(\cos(\theta))r^{-3}  \right] \\
-&= \kay \left[\frac{1}{r} - \frac{3}{2r^3}(3\cos^2(\theta) - 1)\right]
-\end{align}
-
-As our final expression. In the far-field, we can probably neglect the $r^{-3}$ term, which would give us just the potential of a point charge, $\Phi = \kay \frac{1}{r}$ as our answer.
-
-\section{Problem 4}
-By calculating the moments up to quadrupolar order, we wish to show tha the charge distribution $\rho(\vec{r}) = -(\vec{P}\cdot \nabla)\delta(\vec{r})$ represents an elementary dipole with dipole moment $\vec{P}$ located at the origin, given the hint that this is most easily accomplished in cartesian coordinates.
-
-\subsection{Monopole Moment}
-
-Directly integrating the charge distribution over space gives us an integral which we can expand as follows:
-
-\begin{align}
-&= - \int_v (\vec{P}\cdot \nabla)\delta(\vec{r})d^3(\vec{r'}) \\
-&= -\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[P_x \partd{}{x} + P_y \partd{}{y} + P_z \partd{}{z}\right] \delta (x) \delta (y) \delta (z) dx dy dz
-\end{align}
-
-Picking out just the $x$ term, we have
-
-\begin{align}
-&= \int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}\delta(z)dz \int_{-\infty}^{\infty}P_x\partd{}{x}\delta(x)dx  \\
-&= \int_{-\infty}^{\infty}P_x\partd{\delta(x)}{x}dx
-\end{align}
-
-Integrating by parts with $u = P_x, du = P'_x dx, dv = \delta'(x)dx, v=\delta(x)$, we have
-
-\begin{align}
-\cancel{P_x \delta(x)\bigg|_{-\infty}^{\infty}} - \int_{-\infty}^{\infty}\delta(x)P'_x(x)dx
-\end{align}
-
-The first term vanishes at the boundaries because $\delta(x) = 0$ everywhere except $x=0$.
-The second, integral term vanishes as long as $\vec{P} = $ const.
-This implies that the $x$ integral is zero, and applying the same logic to the other portions of the total integral, we get a monopole term that vanishes completely, as is expected for a pure dipole.
-
-\subsection{Dipole Moment}
-
-This is an integral of the form
-
-\begin{align}
--\int_v \vec{r}\left[\vec{P} \cdot \nabla\right] d^3(\vec{r})
-\end{align}
-
-Which can be expanded explicitly in cartesian coordinates as
-
-\begin{align}
--\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[x\hat{x} +y\hat{y} + z\hat{z}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
-\end{align}
-
-From this we can see that the integral breaks up into two types of terms:
-
-\begin{enumerate}
-\item \textit{Uniform} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_x\partd{}{x}\delta(x)\delta(y)\delta(z) dxdydz$
-\item \textit{Mixed} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_y\partd{}{y}\delta(x)\delta(y)\delta(z)dxdydz$ 
-\end{enumerate}
-
-Examining the uniform case in $x$, we integrate by parts with $u=xP_x \hat{x}, du=P_xdx\hat{x}, v=\delta'(x)dx, v = \delta(x)$ and including the negative out front, obtain
-
-\begin{align}
-\cancel{-xP_x\delta(x)\bigg|_{-\infty}^{\infty}} + \int_{-\infty}^{\infty}P_x\delta(x)dx\hat{x}
-\end{align}
-
-Which produces the value of $P_x\hat{x}$ from the term on the right hand side. However, the \textit{mixed} integral cases look like
-
-\begin{align}
-\int_{-\infty}^{\infty}x\hat{x}\delta(x)dx\int_{-\infty}^{\infty}\delta(z)dz\cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} = 0
-\end{align}
-
-The last integral factor on the right hand side is zero, because it is of the exact same form as the monopole integral above.
-Applying the same logic to the other 7 integrals, we end up with simply $P_x \hat{x} + P_y \hat{y} + P_z\hat{z}$, which corresponds to a pure dipole of magnitude $P$ located at the origin.
-
-\subsection{Quadrupole Moment}
-
-The integral for the quadrupole moment tensor components, given in Jackson, is
-
-\begin{align}
-Q_{ij} = \int_v \left[3x_ix_j - r^2\delta_{ij}\right]\rho(x)d^3(x)
-\end{align}
-
-Inserting our charge distribution yields the integral
-
-\begin{align}
-\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\left[3x_ix_j - r^2 \delta_{ij}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
-\end{align}
-
-Let's examine the first term in brackets.
-For $i=j$, we have, for instance $3x^2 - x^2 - y^2 - z^2 = 2x^2 - y^2 - z^2$.
-For $i\neq j$, we have terms that look like just $xy$.
-The first term of the diagonal case reads
-
-\begin{align}
--2\int_{-\infty}^{\infty}\delta(x)dx\int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}x^2P_x\delta'(x)dx
-\end{align}
-
-Integrating the $x-$factor by parts with $u=x^2P_x, du=2xP_x dx, dv=\delta'(x)dx, v=\delta(x)$ the boundary term cancels as usual and the resulting integral is
-
-\begin{align}
--2\int_{-\infty}^{\infty} \delta(x)xP_xdx 
-\end{align}
-
-Which vanishes, because the integrand $x|_{x=0} = 0$ at the location of the delta function. The other terms in the $xx$ case are integrals of the form
-
-\begin{align}
--2\int_{-\infty}^{\infty}x^2\delta(x)dx \int_{-\infty}^{\infty}\delta(z)dz \cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} 
-\end{align}
-
-The last factor on the right vanishes as it is another monopole-like integral already shown to be zero.
-The same construction of course works for the other diagonal elements.
-That just leaves the off diagonal case, which consists of integrals like
-
-\begin{align}
-3\int_{-\infty}^{\infty}xP_x\delta'(x)\cancel{\int_{-\infty}^{\infty}y\delta(y)dy}\int_{-\infty}^{\infty}\delta(z)dz
-\end{align}
-
-and
-
-\begin{align}
-3\int_{-\infty}^{\infty}x\delta{x}dx\int_{-\infty}^{\infty}y\delta(y)dy\int_{-\infty}^{\infty}P_z\partd{}{z}\delta{z}dz
-\end{align}
-
-In the latter case, \textit{all three} integral factors are zero, making the three out of nine cases that look like this one \textit{exceptionally} zero.
-
 \end{document}

### 2016-08-10 Gradschool/thesis/fourier.tex
initial commit of thesis stuff

--- 2016-06-10 Gradschool/electrodynamics/em_hw4.tex
@@
 \horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge Electrodynamics - Homework 4 \\ % The assignment title
+\huge Fourier Transforms, U and V Matrices \\ % The assignment title
 \horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
@@
 
-\section{Problem 1}
-
-We are asked to investigate the Diriclet problem in prolate spheroidal coordinates via separation of variables.
-Like any similar problem involving curvilinear coordinates our first task is to cast the Laplacian into the appropriate form so that we can write Laplace's equation:
+The fourier transform $\hat{h}(\omega)$ of a real or complex function of time $h(t)$ is defined as
 
 \begin{align}
-\nabla^2 \Phi = 0
+\hat{h}(\omega) &\equiv \int_{-\infty}^{\infty} h(t)e^{-i \omega t} dt
 \end{align}
 
-The general form of the Laplacian in curvilinear coordinates is given by
+For purely real $h(t)$ the following relation holds. Note that the complex conjugation and reversal operations commute:
 
 \begin{align}
-\nabla \cdot \nabla = \nabla^2 = \frac{1}{h_1 h_2 h_3}\left[\partd{}{u_1}\left(\frac{h_2 h_3}{h_1}\partd{}{u_2}\right) + \partd{}{u_2}\left(\frac{h_3h_1}{h_2}\partd{}{u_2}\right) + \partd{}{u_3}\left(\frac{h_1h_2}{h_3}\partd{}{u_3}\right)\right]
+[\hat{h}(\omega)]^{*} &= \int_{-\infty}^{\infty}h(t)e^{i \omega t}dt \\
+\rightarrow [\hat{h}(-\omega)]^{*} &= \int_{-\infty}^{\infty}h(t)e^{-i \omega t}dt \\
+\rightarrow [\hat{h}(-\omega)]^{*} &= \hat{h}(\omega)
 \end{align}
 
-Where the $h_i$ are the scale factors and run over all the coordinates relevant to the space - in the case of prolate spheroidal coordinates, these are $\alpha, \beta$, and $\phi$.
-These parameters are given in cartesian coordinates as
+This is Hermitian symmetry. In the case of a complex function of time $h(t) \equiv h_{+}(t) - i h_{\times}(t)$ the same sequence of operations yields
 
 \begin{align}
-x &= c \sinh(\alpha)\sin(\beta)\cos(\phi) \\
-y &= c \sinh(\alpha)\sin(\beta)\sin(\phi) \\
-z &= c\cosh(\alpha)\cos(\beta)
+\hat{h}(\omega) &= \int_{-\infty}^{\infty}[h_{+}(t) - i h_{\times}(t)] e^{-i \omega t} dt \\
+\rightarrow [\hat{h}(\omega)]^{*} &= \int_{-\infty}^{\infty}[h_{+}(t) + i h_{\times}(t)]e^{i \omega t}dt \\
+\rightarrow [\hat{h}(-\omega)]^{*} &= \int_{-\infty}^{\infty}[h_{+}(t) + i h_{\times}(t)]e^{-i \omega t}dt \\
+\rightarrow [\hat{h}(-\omega)]^{*} &= \widehat{h^{*}}(\omega)
 \end{align}
 
-Whereas the scale factors themselves are given by
+The right hand side of the last formula is the \textit{fourier transform of the complex conugate of}$h(t)$. The "Complex conjugate in time" operation is defined as
 
 \begin{align}
-h_i &= \left|\partd{\vec{r}}{q_i} \right| \\
-&= \sqrt{\partd{\vec{r}}{q_i} \cdot \partd{\vec{r}}{q_i}} \\
+\mathcal{I}\hat{h}(\omega) &= \int_{-\infty}^{\infty}h^{*}(t)e^{-i \omega t}dt 
 \end{align}
 
-For $\vec{r} = x\hat{x} + y\hat{y} + z\hat{z}$.
-We need to plug in our coordinate transformations to the components of our position vector and then take derivatives with respect to those coordinates to get the scale factors.
-The position vector looks like
+This is also the fourier transform of the couplex conjugate of the time domain function $h(t)$, and therefore equivalent, for any complex function, to:
 
 \begin{align}
-\vec{r} &= c\sinh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\sinh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\cosh(\alpha)\cos(\beta)\hat{z} \\
-\partd{\vec{r}}{\alpha} &= c\cosh(\alpha)\sin(\beta)\cos(\phi)\hat{x} + c\cosh(\alpha)\sin(\beta)\sin(\phi)\hat{y} + c\sinh(\alpha)\cos(\beta)\hat{z}\\
+\mathcal{I}\hat{h}(\omega) = [\hat{h}(-\omega)]^{*}
 \end{align}
 
-Dotting this with itself gives
-
-\begin{align}
-\partd{\vec{r}}{\alpha} \cdot \partd{\vec{r}}{\alpha} &= c^2 \cosh^2(\alpha)\sin^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
-&= c^2\cosh^2(\alpha)\sin^2(\beta)(\cos^2(\phi) + \sin^2(\phi)) + c^2 \sinh^2(\alpha)\cos^2(\beta) \\
-&= c^2\left[\cosh^2(\alpha)\sin^2(\beta) + \sinh^2(\alpha)\cos^2(\beta)\right] \\
-&= c^2\left[(1 + \sinh^2(\alpha))\sin^2(\beta) + \sinh^2(\alpha)(1 - \sin^2(\beta))\right] \\
-&= c^2\left[\cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sinh^2(\alpha)\right]\\
-h_{\alpha} &= \sqrt{\sin^2(\beta) + \sinh^2(\alpha)} 
-\end{align}
-
-Similarly for the other coordinates
-
-\begin{align}
-\partd{\vec{r}}{\beta}\cdot \partd{\vec{r}}{\beta}&= c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2\sinh^2(\alpha)\cos^2(\beta)\cos^2(\phi) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\ 
-&= c^2\sinh^2(\alpha)\cos^2(\beta) + c^2 \cosh^2(\alpha)\sin^2(\beta) \\
-&= c^2\left[\sinh^2(\alpha)(1-\sin^2(\beta)) + (\sinh^2(\alpha) + 1)\sin^2(\beta)\right]\\
-&= c^2\left[\sinh^2(\alpha) - \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \cancel{\sinh^2(\alpha)\sin^2(\beta)} + \sin^2(\beta)\right] \\
-h_{\beta} &= \sqrt{c^2\sinh^2(\alpha) +\sin^2(\beta)} 
-\end{align}
-
-\begin{align}
-\partd{\vec{r}}{\phi} \cdot \partd{\vec{r}}{\phi} &= c^2\sinh^2(\alpha)\sin^2(\beta)\sin^2(\phi) + \sinh^2(\alpha)\sin^2(\beta)\cos^2(\phi) \\
-&= c^2\sinh^2(\alpha)\sin^2(\beta)(\sin^2(\phi) + \cos^2(\phi)) \\
-&= c^2\sinh^2(\alpha)\sin^2(\beta) \\
-h_{\phi} &= \sqrt{c^2\sinh^2(\alpha)\sin^2(\beta)}
-\end{align}
-
-Since $h_1 = h_2$ the definition of the Laplacian reduces in this case to
-
-\begin{align}
-\frac{1}{h_1h_2h_3}\left[\partd{}{\alpha}h_3\partd{}{\alpha} + \partd{}{\beta}h_3\partd{}{\beta} + \partd{}{\phi}\frac{h_1h_2}{h_3}\partd{}{\phi}\right] \\
-\end{align}
-
-The term in the brackets simplifies as
-
-\begin{align}
-&=\partd{}{\alpha}(\sinh(\alpha)\sin(\beta))\partd{}{\alpha}\partd{}{\alpha} + \partd{}{\beta}(\sinh(\alpha)\sin(\beta))\partd{}{\beta} + \partd{}{\phi}\left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partd{}{\phi} \\
-&= \cosh(\alpha)\sin(\beta)\partd{}{\alpha} + \sinh(\alpha)\sin(\beta)\partdd{}{\alpha} + \sinh(\alpha)\cos(\beta)\partd{}{\beta} + \sinh(\alpha)\sin(\beta)\partdd{}{\beta} ... \\
-&+ \left(\frac{c^2(\sin^2(\beta)+\sinh^2(\alpha))}{c\sinh(\alpha)\sin(\beta)}\right)\partdd{}{\phi}
-\end{align}
-
-Dividing out the $\sinh(\alpha)\sin(\beta)$ term from the prefactor gets us the final Laplacian:
-
-\begin{align}
-\nabla^2 = \frac{1}{c^2(\sin^2\beta)+\sinh^2(\alpha)}\left[\coth(\alpha)\partd{}{\alpha} + \cot(\beta)\partd{}{\beta}+ \partdd{}{\alpha} + \partdd{}{\beta} + (\csch(\alpha) + \csc(\beta))\partdd{}{\phi}\right]
-\end{align}
-
-We are given that the problem has azimuthal symmetry, so the ($\phi$) derivatives are zero.
-This leads us to assume a solution of the form $\Phi = A(\alpha)B(\beta)$.
-Acting our Laplacian, sans the $\phi$ derivative part, on our ansatz gets us the following
-
-\begin{align}
-\partdd{A}{\alpha}B + \partdd{B}{\beta}A + B\coth(\alpha)\partd{A}{\alpha} + A\cot(\beta)\partd{B}{\beta} &= 0 \\
-\frac{1}{A}\partdd{A}{\alpha} + \frac{1}{B}\partdd{B}{\beta} + \frac{\coth(\alpha)}{A}\partd{A}{\alpha} + \frac{\cot(\beta)}{B}\partd{B}{\beta} &= 0
-\end{align}
-
-As usual the only way two functions of separate varibles can be equal for all values of their respective arguments are if they are both equal to the same constant.
-In anticipation for transforming our result into Legendre's differential equation we select $l(l+1)$ for our constant and arrive at the following two second order ODEs as our solutions:
-
-\begin{align}
-\partdd{A}{\alpha} + \coth(\alpha)\partd{A}{\alpha} - Al(l+1) &= 0 \\
-\partdd{B}{\beta} + \cot(\beta)\partd{B}{\beta} + Bl(l+1) &= 0
-\end{align}
-
-Starting with the equation for $A$, we would like to transform our derivatives from $\partd{}{\alpha}$ to $\partd{}{\cosh(\alpha)}$.
-We can write $\partd{}{\alpha} = \partd{\cosh(\alpha)}{\alpha}\partd{}{\cosh(\alpha)} = \sinh(\alpha)\partd{}{\cosh(\alpha)}$.
-Taking another derivative of this expression via product rule, we get $\partd{}{\alpha}\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh(\alpha)\left[\sinh(\alpha)\partd{}{\cosh(\alpha)}\partd{}{\cosh(\alpha)}\right] = \cosh(\alpha)\partd{}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{}{\cosh(\alpha)}$.
-With this information the equation for $A$ transforms as
-
-\begin{align}
-\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \coth(\alpha)\sinh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
-\cosh(\alpha)\partd{A}{\cosh(\alpha)} + \sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + \cosh(\alpha)\partd{A}{\cosh(\alpha)} - Al(l+1) = 0 \\
-\sinh^2(\alpha)\partdd{A}{\cosh(\alpha)} + 2 \cosh(\alpha)\partd{A}{\alpha} - Al(l+1)) = 0 \\
--(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} + 2\cosh(\alpha)\partd{A}{\alpha}-Al(l+1) = 0 \\
-(1-\cosh^2(\alpha))\partdd{A}{\cosh(\alpha)} - 2\cosh(\alpha)\partd{A}{\alpha} + Al(l+1) = 0
-\end{align}
-
-Which is Legendre's equation for $A(\alpha)$.
-As for the $B$ equation, we want the argument of the resulting Legendre polynomials to be $\cos(\beta)$.
-So we note that $\partd{}{\beta} = \partd{\cos(\beta)}{\beta}\partd{}{\cos(\beta)} = -\sin(\beta)\partd{}{\cos(\beta)}$ and $\partdd{}{\beta} = \partd{}{\beta}\left[-\sin(\beta)\partd{}{\cos(\beta)}\right] = -\cos(\beta)\partd{}{\cos(\beta)} + \sin^2(\beta)\partdd{}{\cos(\beta)}$.
-Acting the revised operators on $B$ yields
-
-\begin{align}
-\sin^2(\beta)\partdd{B}{\cos(\beta)} - \cos(\beta)\partd{B}{\cos(\beta)} - \cot(\beta)\sin(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0 \\
-(1-\cos^2(\beta))\partdd{B}{\cos(\beta)} - 2 \cos(\beta)\partd{B}{\cos(\beta)} + Bl(l+1) &= 0
-\end{align}
-
-Which is again Legendre's equation for $B(\beta)$. Thus we can represent our solution as
-
-\begin{align}
-\Phi = \sum_n A_n P_n(\cosh(\alpha)P_n(\cos(\beta))
-\end{align}
-
-Which is the desired result.
-
-\section{Problem 2}
-
-We want to the scalar monopole moment, vector dipole moment, and tensor quadrupole moment for the two charge distributions pictured above, for both cartesian and spherical coordinates.
-The moments themselves are defined as follows:
-
-\begin{align}
-M &= \kay \int_v \rho(\vec{r'})d^3\vec{r'} = \ \text{Total Charge}  \\
-\vec{P} &= \kay \int_v \vec{r'}\rho(\vec{r})d^3\vec{r'} = \ \text{Dipole Moment Vector} \\
-Q_{ij} &= \kay \int_v (3x_ix_j - r_i^2\delta_{ij})\rho(\vec{r'})d^3(r') = \ \text{Quadrupole Moment Tensor Components}
-\end{align}
-
-We note that the total charge of the system sums to zero.
-Thus we have no monopole moment for the charge distribution.
-As for the dipole moment, the relevant formula in the discrete case would have to be
-
-\begin{align}
-\vec{P}_i &= \sum_{n=1}^3 r_i^{(n)}q_i^{(n)} \ \text{\textit{Component i} of the dipole vector} \\
-\end{align}
-
-\subsection{Distribution 1}
-
-Labelling the charges as shown in the figure, the dipole moment  would come out to be $qa\hat{z} - qa\hat{z} -\cancel{2q(0)} = 0$, when summing over all the charges.
-Thus the dipole moment of the first charge distribution is also zero.
-The discrete quadrupole formula written as a summation over a discrete charge distribution must be
-
-\begin{align}
-Q_{ij} = \sum_{n=1}^3 (3x_i^{(n)}x_j^{(n)} - (r_i^{(n)})^2\delta_{ij})q^{(n)}
-\end{align}
-
-For the charge distribution in question we can expand this formula as follows:
-
-\begin{align}
-Q_{ij} &= (3x_i^{(1)}x_j^{(1)} - (r_i^{(1)})^2\delta_{ij})q^{(1)} + (3x_i^{(2)}x_j^{(2)} - (r_i^{(2)})^2\delta_{ij})q^{(2)} + (3x_i^{(3)}x_j^{(3)} - (r_i^{(3)})^2\delta_{ij})q^{(3)} \\
-\end{align}
-
-Let's look at the off-diagonal case first.
-If $i \neq j$, then obviously all the $\delta_{ij}$ cancel.
-Furthermore, when we notice that the charges lie only on the $z$ axis, we find that since \textit{at least one} of the coordinates ($i$ or $j$) must \textit{not} be $z$, all of the other products cancel too.
-Thus we conclude that all of the off-diagonal elements in this tensor are zero.
-This leaves us with three to calculate:
-
-\begin{align}
-Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)} \\
-&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\ 
-&= -2a^2q \\
-Q_{22} &= (3x_2^{(1)}x_2^{(1)} - (r_2^{(1)})^2)q^{(1)} + (3x_2^{(2)}x_2^{(2)} - (r_2^{(2)})^2)q^{(2)} + (3x_2^{(3)}x_2^{(3)} - (r_2^{(3)})^2)q^{(3)} \\
-&= (\cancel{3(0)(0)} - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (\cancel{3(0)(0)} - a^2)q \\
-&= -2a^2q \\
-Q_{33} &= (3x_3^{(1)}x_3^{(1)} - (r_3^{(1)})^2)q^{(1)} + (3x_3^{(2)}x_3^{(2)} - (r_3^{(2)})^2)q^{(2)} + (3x_3^{(3)}x_3^{(3)} - (r_3^{(3)})^2)q^{(3)} \\
-&= (3a^2 - a^2)q - 2\cancel{(3(0)(0) - 0^2)q} + (3a^2 - a^2)q \\
-&= 4a^2q
-\end{align}
-
-Thus, the quadrupole moment tensor, which we note is traceless, is
-
-\begin{align}
-Q = 2a^2
-\begin{bmatrix}
--1 & 0 & 0 \\
-0 & -1 & 0 \\
-0 & 0 & 2
-\end{bmatrix}
-\end{align}
-
-To translate to spherical coordinates, we make use of Jackson equations 4.4 through 4.6.
-Since the total charge and dipole moments are known to be zero, the $l=0$ and $l=1$ $q$'s are all zero.
-Furthermore, if all the off diagonal terms of $Q$ are zero, $q_{21}$ is automatically zero.
-Even more, $q_{22} = \text{const.} * (Q_{11} - \cancel{iQ_{12}} - Q_{22})$ is zero because $Q_{11} = Q_{22}$.
-This leaves only the $q_{20}$ moment to calculate.
-
-\begin{align}
-q_{20} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}Q_{33} = \frac{1}{2}\sqrt{\frac{5}{4\pi}}(4a^2) = \sqrt{\frac{5}{\pi}}a^2
-\end{align}
-
-As the only nonzero $q_{lm}$.
-
-\subsection{Distribution 2}
-
-We note that as before the total charge over the system is zero. So we must have a monopole moment that vanishes. The components of the dipole moment vector $\vec{P}$ would be as follows:
-
-\begin{align}
-P_x &= r_x^{(1)}q^{(1)} + r_x^{(2)}q^{(2)} + r_x^{(3)}q^{(3)} + r_x^{(4)}q^{(4)} \\
-&= a(-q) + \cancel{0(q)} -a(-q) + \cancel{a(0)} = 0  \\
-P_y &= r_y^{(1)}q^{(1)} + r_y^{(2)}q^{(2)} + r_y^{(3)}q^{(3)} + r_y^{(4)}q^{(4)} \\
-&= \cancel{(0)-q} -a(q) +\cancel{(0)(-q)} -a(-q) = 0 \\
-P_z &= 0+0+0+0 = 0
-\end{align}
-
-So we conclude, explicitly, that $\vec{P} = 0$ for this distribution as well.
-The components of the Quadrupole tensor can be found through the same process as above.
-For this case, we note that any and all terms involving $z$ are automatically zero because the charges are in the $xy$ plane.
-That elminiates 5 of 9 components that we need to calculate.
-Furthermore, since we know that $Q_{33}$ is zero, we have the additonal constraint that $Q_{22} = - Q_{11}$ for the tensor to remain traceless.
-This leaves us with only two components to calculate:
-
-\begin{align}
-Q_{11} &= (3x_1^{(1)}x_1^{(1)} - (r_1^{(1)})^2)q^{(1)} + (3x_1^{(2)}x_1^{(2)} - (r_1^{(2)})^2)q^{(2)} + (3x_1^{(3)}x_1^{(3)} - (r_1^{(3)})^2)q^{(3)}  ... \\
-&+ (3x_1^{(4)}x_1^{(4)} - (r_1^{(4)})^2)q^{(4)}\\
-&= -(3 a^2 - a^2)(-q) + (3(0)(0) - a^2)(q) + (3 a^2 - a^2)(-q) - (3(0)(0) - a^2)q \\
-&= -6a^2 \\
-Q_{12} &= (3x_1^{(1)}x_2^{(1)})q^{(1)} + (3x_1^{(2)}x_2^{(2)})q^{(2)} + (3x_1^{(3)}x_2^{(3)})q^{(3)}+(3x_1^{(4)}x_2^{(4)})q^{(4)} \\
-&= \cancel{(3(a)(0))q} + \cancel{(3(0)(a))q} + \cancel{(3(0)(a))q} + \cancel{(3(a)(0))q} = 0 
-\end{align}
-
-Which in retrospect was obvious because no individual charge has both an $x$ and $y$ component. The quadrupole moment tensor is, then,
-
-\begin{align}
-Q = -6a^2
-\begin{bmatrix}
--1 & 0 & 0\\
-0 &1 & 0 \\
-0 & 0 & 0 
-\end{bmatrix}
-\end{align}
-
-In spherical coordinates, using the same logic as above, no monopole or dipole moments means $q_{1m}$ and $q_{00}$ vanish.
-This time, $q_{20}$ also vanishes because $Q_{33}=0$.
-The only nonzero $q_{lm}$ is, then:
-
-\begin{align}
-q_{22} =  \frac{1}{12}\sqrt{\frac{15}{2\pi}}(Q_{11} - Q_{22}) = \frac{1}{12}\sqrt{\frac{15}{2\pi}}(8a^2) = \frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2
-\end{align}
-
-\subsection{Far Field Potential}
-
-Using Jackson Equation 4.1 and plugging in our results from above, we have, for the first source distribution,
-
-\begin{align}
-\Phi &= \kay \sum_{l=0}^{\infty}\sum_{m=-l}^l \frac{4\pi}{2l + 1}q_{lm}\frac{Y_{lm}(\theta,\phi)}{r^{l+1}} \\
-&=
-\begin{cases}
-\frac{1}{\cancel{4 \pi} \epsilon_0} \sqrt{\frac{5}{\pi}}a^2 \frac{\cancel{4\pi}}{5}\frac{1}{4}\sqrt{\frac{5}{\pi}}\left[3\cos^2(\theta)-1\right]r^{-3} = \frac{a^2}{4 \pi}\left[3\cos^2(\theta)-1\right]r^{-3}, & \ \text{Distribution 1} \\
-\frac{1}{4 \pi \epsilon_0}\frac{4 \pi}{5}\frac{2}{3}\sqrt{\frac{15}{2\pi}}a^2\left[\frac{1}{4}\sqrt{\frac{15}{2\pi}}\right]\sin^2(\theta)e^{2i\phi}r^{-3} = \frac{a^2}{\epsilon_0 \pi}\sin^2(\theta)e^{2i\phi}r^{-3}, & \ \text{Distribution 2}
-\end{cases}
-\end{align}
-
-\section{Problem 3}
-
-We are given a "localized" charge distribution in spherical coordinates with the following form:
-
-\begin{align}
-\rho(\vec{r}) = \frac{1}{64 \pi}r^2 e^{-r}\sin^2(\theta)
-\end{align}
-
-\subsection{Multipole Moments}
-
-The formula for multipole moments is given in Jackson as
-
-\begin{align}
-q_{lm} &= \int_v Y^*_{lm}(\theta',\phi')r'^l\rho(\vec{r'})d^3\vec{r'} 
-\end{align}
-
-Right off the bat we make two observations that simplify our task. Firstly, the problem is azimuthally symmetric. Thus we can write our spherical harmonics, per Jackson equation 3.57, as
-
-\begin{align}
-Y_{l0}(\theta) = \sqrt{\frac{2l + 1}{4 \pi}}P_l(\cos(\theta)) 
-\end{align}
-
-Secondly, we note that since the charge distribution has \textit{at most} a $sin^2(\theta)$ term, we are guaranteed that we will not have multipole moments higher then $l=2$, as those contain Legendre polynomails that carry $\sin^3(\theta)$ terms and higher.
-This leaves us with no more then 6 coefficients to evaluate, since the $-m$ versions are related to the $+m$ versions via $q_{l, -m} = (-1)^m q^*_{lm}$.
-The general formula for the $q_{l0}$ then is as follows, plugging in our particular charge distribution:
-
-\begin{align}
-q_{l0} &= \int_v \sqrt{\frac{2l + 1}{4\pi}}P_l(\cos(\theta'))r'^l*\left[\frac{1}{64\pi}r'^2 e^{-r'}\sin^2(\theta')\right]r'^2\sin(\theta')dr'd\theta'd\phi' \\
-&= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{64\pi} \int_{r=0}^{r=\infty}\int_{\theta=0}^{\theta=\pi}\int_{\phi=0}^{\phi=2\pi}r'^{l+4}e^{-r}\sin^3(\theta')P_l(\cos(\theta'))d\phi'd\theta'dr'
-\end{align}
-
-$\phi$ integrates out and we pick up a factor of $2 \pi$. The $\theta$ and $r$ integrals are independent and thus factorize.
-
-\begin{align}
-q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_0^{\pi}\sin^3(\theta')P_l(\cos(\theta')d\theta'\right]
-\end{align}
-
-The $\theta$ integral can be accomplished via change of variables $x=\cos(\theta)$.
-Then $dx = -\sin(\theta)d\theta$ and $d\theta = -\frac{dx}{\sin(\theta)}$.
-Noting that $\sin^3(\theta) = \sin^2(\theta)\sin(\theta) = \sin(\theta)[1-\cos^2(\theta)]$, we know that the factor $\sin^3(\theta)d\theta = (1 - x^2)dx$.
-Finally, the limits of integration change as $x = \cos(0) \rightarrow 1$  and $x = \cos(\pi) \rightarrow -1$.
-
-\begin{align}
-q_{l0} &= \sqrt{\frac{2l + 1}{4 \pi}}\frac{1}{32}\left[\int_0^\infty r'^{l+4}e^{-r}dr\right]\left[\int_{-1}^{1}(1-x^2)P_l(x)dx\right]
-\end{align}
-
-Splitting the $x$ integral up into two pieces, one sees that the first piece,
-
-\begin{align}
-\int_{-1}^{1}P_l(x)dx
-\end{align}
-
-Can be written as 
-\begin{align}
-\int_{-1}^{1}P_l(x)P_0(x)dx
-\end{align}
-
-Since $P_0(x) = 1$.
-Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$, Since $P_0(x) = 1$.
-Then, by orthogonality of the Legendre polynomials, we have that $\int_{-1}^1 P_l(x)P_n(x)dx = \frac{2}{2l+1}\delta_{nl}$.
-This means that the integral above is zero unless $l=0$, in which case it just equals $2$.
-The second integral is more annoying and requires Jackson equation 3.32:
-
-\begin{align}
-\int_{-1}^1x^2P_l(x)P_n(x)dx = \begin{cases}
-\frac{2(n+1)(n+2)}{(2n+1)(2n+1)(2n+3)}, & l=n+2\\
-\frac{2(2n^2 + 2n - 1)}{(2n-1)(2n+1)(2n+5)}, & l=n\\
-\ \ \ \ \ \ \ \ \ \  0, & l = \text{Anything Else}
-\end{cases}
-\end{align}
-
-In our case $n=0$ and we are concerned with the $l=0$ and $l=2$ cases.
-We have thus concluded that the $\theta$ portion of the integral, and thus the entire integral, and thus the multipole moment, is zero for all values of $l$ except for $l=0$ and $l=2$.
-Simplifying the $\theta$ integral for the two nonzero cases we get
-
-\begin{align}
-\int_{\theta} = 
-\begin{cases}
-2 - \int_{-1}^1 x^2 P_0(x)P_l(x)dx = 2 - \frac{2}{3} = \frac{4}{3}, & l=0 \\
--\int_{-1}^1 x^2 P_0(x)P_2(x)dx = -\frac{4}{15}, & l=2  
-\end{cases}
-\end{align}
-
-This leaves us with the integral over $r$ which we recognize as the definition of the factorial function for the exponent of $r$:
-
-\begin{align}
-\int_0^{\infty}r^n e^{-r}dr \equiv n!
-\end{align}
-
-So we have
-
-\begin{align}
-\int_r = 
-\begin{cases}
-\int_0^{\infty}r^4e^{-r}dr = 4! = 24, & l=0 \\
-\int_0^{\infty}r^6e^{-r}dr = 6! = 720, & l=2
-\end{cases}
-\end{align}
-
-Combining everything together we have:
-
-\begin{align}
-q_{l0} = 
-\begin{cases}
-\sqrt{\frac{1}{4\pi}}\frac{1}{32}\left(\frac{4}{3}\right)(24) = \sqrt{\frac{1}{4\pi}}, & l=0 \\
-\sqrt{\frac{4 + 1}{4\pi}}\frac{1}{32}\left(-\frac{4}{15}\right)(720) = -3\sqrt{\frac{5}{\pi}}, & l=2
-\end{cases}
-\end{align}
-
-This means that if the general formula for the multipole expansion is
-
-\begin{align}
-\Phi = \kay \sum_{l=0}^{\infty}\sum_{m=-l}^{l}\frac{4\pi}{2l + 1}q_{lm}Y_{lm}(\theta,\phi)r^{-(l+1)}
-\end{align}
-
-Then we would have
-
-\begin{align}
-\Phi &= \kay \left[\cancel{\frac{4\pi}{1}}\cancel{\sqrt{\frac{1}{4 \pi}}}P_0(\cos(\theta))\cancel{\sqrt{\frac{1}{4\pi}}}r^{-1} + \cancel{\frac{4\pi}{5}}\left(-3\cancel{\sqrt{\frac{5}{\pi}}}\right)\cancel{\frac{1}{4}\sqrt{\frac{5}{\pi}}}P_2(\cos(\theta))r^{-3}  \right] \\
-&= \kay \left[\frac{1}{r} - \frac{3}{2r^3}(3\cos^2(\theta) - 1)\right]
-\end{align}
-
-As our final expression. In the far-field, we can probably neglect the $r^{-3}$ term, which would give us just the potential of a point charge, $\Phi = \kay \frac{1}{r}$ as our answer.
-
-\section{Problem 4}
-By calculating the moments up to quadrupolar order, we wish to show tha the charge distribution $\rho(\vec{r}) = -(\vec{P}\cdot \nabla)\delta(\vec{r})$ represents an elementary dipole with dipole moment $\vec{P}$ located at the origin, given the hint that this is most easily accomplished in cartesian coordinates.
-
-\subsection{Monopole Moment}
-
-Directly integrating the charge distribution over space gives us an integral which we can expand as follows:
-
-\begin{align}
-&= - \int_v (\vec{P}\cdot \nabla)\delta(\vec{r})d^3(\vec{r'}) \\
-&= -\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[P_x \partd{}{x} + P_y \partd{}{y} + P_z \partd{}{z}\right] \delta (x) \delta (y) \delta (z) dx dy dz
-\end{align}
-
-Picking out just the $x$ term, we have
-
-\begin{align}
-&= \int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}\delta(z)dz \int_{-\infty}^{\infty}P_x\partd{}{x}\delta(x)dx  \\
-&= \int_{-\infty}^{\infty}P_x\partd{\delta(x)}{x}dx
-\end{align}
-
-Integrating by parts with $u = P_x, du = P'_x dx, dv = \delta'(x)dx, v=\delta(x)$, we have
-
-\begin{align}
-\cancel{P_x \delta(x)\bigg|_{-\infty}^{\infty}} - \int_{-\infty}^{\infty}\delta(x)P'_x(x)dx
-\end{align}
-
-The first term vanishes at the boundaries because $\delta(x) = 0$ everywhere except $x=0$.
-The second, integral term vanishes as long as $\vec{P} = $ const.
-This implies that the $x$ integral is zero, and applying the same logic to the other portions of the total integral, we get a monopole term that vanishes completely, as is expected for a pure dipole.
-
-\subsection{Dipole Moment}
-
-This is an integral of the form
-
-\begin{align}
--\int_v \vec{r}\left[\vec{P} \cdot \nabla\right] d^3(\vec{r})
-\end{align}
-
-Which can be expanded explicitly in cartesian coordinates as
-
-\begin{align}
--\int_{x=-\infty}^{x=\infty} \int_{y=-\infty}^{y=\infty} \int_{z=-\infty}^{z=\infty}\left[x\hat{x} +y\hat{y} + z\hat{z}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
-\end{align}
-
-From this we can see that the integral breaks up into two types of terms:
-
-\begin{enumerate}
-\item \textit{Uniform} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_x\partd{}{x}\delta(x)\delta(y)\delta(z) dxdydz$
-\item \textit{Mixed} types of the form $\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\hat{x}P_y\partd{}{y}\delta(x)\delta(y)\delta(z)dxdydz$ 
-\end{enumerate}
-
-Examining the uniform case in $x$, we integrate by parts with $u=xP_x \hat{x}, du=P_xdx\hat{x}, v=\delta'(x)dx, v = \delta(x)$ and including the negative out front, obtain
-
-\begin{align}
-\cancel{-xP_x\delta(x)\bigg|_{-\infty}^{\infty}} + \int_{-\infty}^{\infty}P_x\delta(x)dx\hat{x}
-\end{align}
-
-Which produces the value of $P_x\hat{x}$ from the term on the right hand side. However, the \textit{mixed} integral cases look like
-
-\begin{align}
-\int_{-\infty}^{\infty}x\hat{x}\delta(x)dx\int_{-\infty}^{\infty}\delta(z)dz\cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} = 0
-\end{align}
-
-The last integral factor on the right hand side is zero, because it is of the exact same form as the monopole integral above.
-Applying the same logic to the other 7 integrals, we end up with simply $P_x \hat{x} + P_y \hat{y} + P_z\hat{z}$, which corresponds to a pure dipole of magnitude $P$ located at the origin.
-
-\subsection{Quadrupole Moment}
-
-The integral for the quadrupole moment tensor components, given in Jackson, is
-
-\begin{align}
-Q_{ij} = \int_v \left[3x_ix_j - r^2\delta_{ij}\right]\rho(x)d^3(x)
-\end{align}
-
-Inserting our charge distribution yields the integral
-
-\begin{align}
-\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}\left[3x_ix_j - r^2 \delta_{ij}\right]\left[P_x\partd{}{x} + P_y\partd{}{y} + P_z\partd{}{z}\right]\delta(x)\delta(y)\delta(z)dxdydz
-\end{align}
-
-Let's examine the first term in brackets.
-For $i=j$, we have, for instance $3x^2 - x^2 - y^2 - z^2 = 2x^2 - y^2 - z^2$.
-For $i\neq j$, we have terms that look like just $xy$.
-The first term of the diagonal case reads
-
-\begin{align}
--2\int_{-\infty}^{\infty}\delta(x)dx\int_{-\infty}^{\infty}\delta(y)dy\int_{-\infty}^{\infty}x^2P_x\delta'(x)dx
-\end{align}
-
-Integrating the $x-$factor by parts with $u=x^2P_x, du=2xP_x dx, dv=\delta'(x)dx, v=\delta(x)$ the boundary term cancels as usual and the resulting integral is
-
-\begin{align}
--2\int_{-\infty}^{\infty} \delta(x)xP_xdx 
-\end{align}
-
-Which vanishes, because the integrand $x|_{x=0} = 0$ at the location of the delta function. The other terms in the $xx$ case are integrals of the form
-
-\begin{align}
--2\int_{-\infty}^{\infty}x^2\delta(x)dx \int_{-\infty}^{\infty}\delta(z)dz \cancel{\int_{-\infty}^{\infty}P_y\partd{}{y}\delta(y)dy} 
-\end{align}
-
-The last factor on the right vanishes as it is another monopole-like integral already shown to be zero.
-The same construction of course works for the other diagonal elements.
-That just leaves the off diagonal case, which consists of integrals like
-
-\begin{align}
-3\int_{-\infty}^{\infty}xP_x\delta'(x)\cancel{\int_{-\infty}^{\infty}y\delta(y)dy}\int_{-\infty}^{\infty}\delta(z)dz
-\end{align}
-
-and
-
-\begin{align}
-3\int_{-\infty}^{\infty}x\delta{x}dx\int_{-\infty}^{\infty}y\delta(y)dy\int_{-\infty}^{\infty}P_z\partd{}{z}\delta{z}dz
-\end{align}
-
-In the latter case, \textit{all three} integral factors are zero, making the three out of nine cases that look like this one \textit{exceptionally} zero.
+This is true for any arbitrary function that can be written $h(t) = h_{+}(t) + i h_{\times}(t)$.
 

### 2016-08-10 Gradschool/thesis/msthesis.tex
initial commit of thesis stuff

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Masters Thesis \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

\section{Background and Motivation}

Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations. The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.

\subsection{Pieces of the Einstein Equation}
\subsubsection{Reimann Tensor}
The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime. The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.

The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space. Here we motivate the concept with a visual explanation of parallel transport. Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \textbf{FIGURE HERE}. If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$. This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$. Let this vector be denoted $\vec{d}$. Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector. From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above. Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop. It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question. More remarkable however is the simplicity of the formal definition within this picture:

\begin{align}
R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
\end{align}

It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero. Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold. Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation. Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}

\begin{align}
R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
\end{align}

If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic. Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics. It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation

\begin{align}
\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
\end{align}

%----------------------------------------------------------------------------------------

\end{document}

### 2016-08-10 Gradschool/thesis/vec_sdll.tex
initial commit of thesis stuff

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Short Sectioned Assignment
% LaTeX Template
% Version 1.0 (5/5/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Frits Wenneker (http://www.howtotex.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size

\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\usepackage[adobe-utopia]{mathdesign}
%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
\usepackage[english]{babel} % English language/hyphenation
\usepackage{amsmath,amsfonts,amsthm} % Math packages

\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template

\usepackage{listings}

\usepackage{sectsty} % Allows customizing section commands

\usepackage{braket}

\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps

\usepackage{fancyhdr} % Custom headers and footers
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{13.6pt} % Customize the height of the header

\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)

\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge Single Detector Log Likelihood \\ % The assignment title
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Brandon B. Miller} % Your name

\date{\normalsize\today} % Today's date or a custom date

\begin{document}

\maketitle % Print the title

%----------------------------------------------------------------------------------------
%	PROBLEM 1
%----------------------------------------------------------------------------------------
\subsection{Vectorization}
\renewcommand*{\arraystretch}{0.5}
Equation 24 from Arxiv $15502.05370v1.pdf$ for the network log likelihood reads

\begin{equation}
\begin{align}
\ln{\mathcal{L}} = &\frac{D_{ref}}{D}Re \sum_{k}\sum_{(l,m)}\left[F_k Y_{lm}\right]^{*}Q_{k,lm} \\ 
& - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] \\
 & - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right]
\end{align}
\end{equation}

Consider a single detector, thus dropping the sum over $k$. The first term is of the form $\vec{A}\cdot\vec{B} = \sum_{i=0}^{d}A_iB_i$, so if $Q_{k,(l,m)}$ were a simple vector, we could write it as

\begin{equation}
- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2} F*\vec{Y}\cdot\vec{Q}
\end{equation}

However the $Q_{k,(l,m)}$ are actually harmonic mode time series and not single values. We desire a vector whose values are the likelihoods at each point in the time series.

\newpage
Consider the case where we have only the $(2,-2), (2,0)$ and $(2,2)$ modes. If we write all the mode time series $Q^0, Q^1, Q^2...$ as the columns of a matrix , then the desired result is obtained with

\begin{equation}
F*
\begin{bmatrix}
Q^0_{2,-2} & Q^0_{2,+0} & Q^0_{2,+2} \\
Q^1_{2,-2} & Q^1_{2,+0} & Q^1_{2,+2} \\ 
Q^2_{2,-2} & Q^2_{2,+0} & Q^2_{2,+2} \\
\vdots & \vdots & \vdots
\end{bmatrix}
\begin{bmatrix}

\left(Y_{2,-2}\right)\\
\hspace{0mm} \\
\left(Y_{2,+0}\right) \\
\hspace{0mm} \\
\left(Y_{2,+2}\right) \\

\end{bmatrix}
=
\begin{bmatrix}
Q^0_{2,-2} Y_{2,-2} + Q^0_{2,-2}Y_{2,+0} + Q^0_{2,+2}Y_{2,+2} \\
Q^1_{2,-2} Y_{2,-2} + Q^1_{2,-2}Y_{2,+0} + Q^1_{2,+2}Y_{2,+2} \\
Q^2_{2,-2} Y_{2,-2} + Q^2_{2,-2}Y_{2,+0} + Q^2_{2,+2}Y_{2,+2} \\
\vdots
\end{bmatrix}

\end{equation}

With $\vec{Y}$ and $\mathbf{Q}$ defined as the matrix and vector above respectively, we have for the first term

\begin{equation}
\frac{D_{ref}}{D}Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right]
\end{equation}

The second term is a sum once over all the possible combinations of $(l,m), (l',m')$ pairs using the $U_{(l,m),(l',m')}$ cross terms. Its result is a scalar quantity made up of terms like

\begin{align}
&Y_{2,-2}^{*}Y_{2,-2}U_{(2,-2),(2,-2)} + Y_{2,-2}^{*}Y_{2,+0}U_{(2,-2),(2,+0)} + Y_{2,-2}^{*}Y_{2,+2}U_{(2,-2),(2,+2)} \\ 
 +  \ &Y_{2,+0}^{*}Y_{2,-2}U_{(2,+0),(2,-2)} + Y_{2,+0}^{*}Y_{2,+0}U_{(2,+0),(2,+0)} + Y_{2,+0}^{*}Y_{2,+2}U_{(2,+0),(2,+2)} \\ 
+ \  &Y_{2,+2}^{*}Y_{2,-2}U_{(2,+2),(2,-2)} + Y_{2,+2}^{*}Y_{2,+0}U_{(2,+2),(2,+0)} + Y_{2,+2}^{*}Y_{2,+2}U_{(2,+2),(2,+2)}
\end{align}

If we pack the $U_{(l,m),(l',m')}$ into the matrix $\mathbf{U}$ as defined below then the following set of matrix operations produces the same sum

\begin{align*}
\begin{bmatrix}
Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
\end{bmatrix}
\begin{bmatrix}
U_{(2,-2),(2,-2)} &  U_{(2,-2),(2,+0)} &  U_{(2,-2),(2,+2)} \\
U_{(2,+0),(2,-2)} &  U_{(2,+0),(2,+0)} &  U_{(2,+0),(2,+2)} \\
U_{(2,+2),(2,-2)} &  U_{(2,+2),(2,+0)} &  U_{(2,+2),(2,+2)}
\end{bmatrix}
\begin{bmatrix}
Y_{2,-2} \\
Y_{2,+0} \\
Y_{2,+2}
\end{bmatrix}
\end{align*}

because when you multiply $\mathbf{U}$ into $\vec{Y}$ this simplifies to

\begin{align*}
\begin{bmatrix}
Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
\end{bmatrix}
\begin{bmatrix}
U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} 
\end{bmatrix}
\end{align*}

Which becomes the desired scalar. This allows us to write the second term as

\begin{align}
- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] =  - \left[\frac{D_{ref}}{2D}\right]^{2} |F^2|\vec{Y}^{*}\mathbf{U}\vec{Y}
\end{align}

We must always set up the spherical harmonic vectors based on the value of $m$ and the cross terms in row-major form based first on $m_2$ and then on $m_1$. If we organize the matrix $\mathbf{V}$ in the same way then the same set of steps will lead us to conclude that

\begin{align}
- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2}Re \left[F^2 \vec{Y}\mathbf{V}\vec{Y} \right]
\end{align}

Combining the results the single detector log likelihood is

\begin{align}
\ln{\mathcal{L}} = \frac{D_{ref}}{D}\Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right] - \left[\frac{D_{ref}}{2D}\right]^{2}\left[|F|^2 \vec{Y}^{*}\mathbf{U}\vec{Y} - \Re\left(F^2 \vec{Y}\mathbf{V}\vec{Y}\right) \right]
\end{align}

\subsection{Implementation}

As it stands the data is organized into dictionaries keyed by tuples corresponding to modes or pairs of modes. I wrote a very basic function that takes one of these dictionaries as argument and produces matrices organized as described in the previous section. I do not think it requires a huge amount of attention since it just needs to run once at startup.

\vspace{10mm}

I have written only the most basic implementation for this that I can think of:

\vspace{10mm}

\begin{lstlisting}[language=Python]

def sdll_matrix(rholm_vals, crossTermsU, crossTermsV, Ylms, F, dist):
 #Compute single detector log likelihood with vectorized operations
 invDistMpc = distRef/dist
 Fstar = np.conj(F)

 Ylms_conj = np.conj(Ylms)
 term1 = Fstar*np.dot(rholm_vals, Ylms_conj)
 term2 = np.dot(Fstar*F*Ylms_conj, np.dot(crossTermsU, Ylms))
 term3 = np.real(np.dot(F*F*Ylms, np.dot(crossTermsV, Ylms)))
 return invDistMpc*np.real(term1) - 0.25*invDistMpc**2 * np.real((term2 + term3))

\end{lstlisting}

\end{document}

### 2016-08-16 Gradschool/thesis/msthesis.tex
working on linearized gravity section

@@
 \usepackage{braket}
+\usepackage{cancel}
 
@@
 \horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge Masters Thesis \\ % The assignment title
+\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
 \horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
@@
 
-\author{Brandon B. Miller} % Your name
+\author{Brandon B. Miller, Masters Thesis} % Your name
 
@@
 \maketitle % Print the title
-
+\newpage
 \section{Background and Motivation}
@@
 
+The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
+
+\begin{align}
+R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
+\end{align}
+
 If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
@@
 
+\section{Linearized Gravity}
+
+In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation
+
+\begin{align}
+g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
+\end{align}
+
+Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources.
+The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
+If we tried to use the full form of equation \textbf{EQUATION}, we would have
+
+\begin{align}
+g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
+&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
+&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
+\end{align}
+
+The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity.
+We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
+
+\begin{align}
+h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
+\end{align}
+
+The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
+
+\begin{align}
+\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
+&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
+\end{align}
+
+Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
+Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero.
+This leaves us with three:
+
+\begin{align}
+\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
+&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
+\end{align}
+
+Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself.
+It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate.
+The same assumption allows for simplification of the Reimann and Ricci tensors:
+
+\begin{align}
+R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
+R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
+\end{align}
+
+Plugging in equation \textbf{EQUATION} we have
+
+\begin{align}
+R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
+&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
+&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
+\end{align}
+
+The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$.
+The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$.
+The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:
+
+\begin{align}
+R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
+\end{align}
+
+Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is
+
+\begin{align}
+R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
+\end{align}
+
+Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity
+
+\begin{align}
+V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
+\end{align}
+
+With which the first two terms of the linearized Ricci tensor can be reorganized as
+
+\begin{align}
+\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
+&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
+\end{align}
+
+Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields
+
+\begin{align}
+R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
+\end{align}
+
+The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with
+
+\begin{align}
+R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
+\end{align}
+
+Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
+
+\begin{align}
+\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
+\end{align}
+
 %----------------------------------------------------------------------------------------

### 2016-08-18 Gradschool/thesis/msthesis.tex
expanded implementation section

@@
 
+\section{Implementation}
+The main weight of the computational work described above falls under an expensive Monte Carlo integration of the evidence as part of the Bayesian posterior.
+Zooming in it quickly becomes apparent that the main challenge to be overcome is efficient evaluation of the function to be integrated, in this case the factored likelihood.
+Prior to this work, the factored likelihood is was computed with a series of looping structures that individually compiled all the terms of the equation in an independent and serial manner.
+The process took substantial advantage of many high level operations available within Python to match up the correct terms across various harmonic mode time series, spherical harmonics, and antenna patterns.
+Thanks to clever optimization and gratuitous use of fast numerical Python libraries such as NumPy, the code was able to compute approximately $10^3$ likelihood evaluations on the order of seconds, the limiting factor being the serial nature of loops and the inherent lethargy of high level languages such as Python.
+Although NumPy commonly passes target data to compiled routines written in faster languages such as C, some portions of the computation (in particular the marginalization over time) benefit little from this capability and form a bottleneck for efficient evaluation of the likelihood.
+Furthermore the original implementation heavily relied upon older structures within the existing LIGO Algorithms Library (LAL), a set of C routines bound to Python through Swig, to perform many intermediate calculations, such as computing spherical harmonics.
+The goal of the first phase of this work was thorough vectorization of the process used to build the terms of the likelihood utilizing the BLAS subroutines available through NumPy as often as possible, as well as reducing dependency on LAL.
+To that end, the components of the factored likelihod were reworked.
+
+\subsection{Complex Antenna Factor}
+The complex antenna factor is a complex number that modulates the gravitational wave signal as a function of incoming direction.
+This is necessary due to the non isotropic sensitivity of the LIGO interferometers themselves.
+$F_{+}$ and $F_{\times}$.
+
+\begin{align}
+F_{+} &= [R_{0,0}X_0 + R_{0,1}X_1 + R_{0,2}X_2]X_0 - [R_{0,0}Y_0 + R_{0,1}Y_1 + R_{0,2}Y_2]Y_0 \\
+&+ [R_{1,0}X_0 + R_{1,1}X_1 + R_{1,2}X_2]X_1 - [R_{1,0}Y_0 + R_{1,1}Y_1 + R_{1,2}Y_2]Y_1 \\
+&+ [R_{2,0}X_0 + R_{2,1}X_1 + R_{2,2}X_2]X_2 - [R_{2,0}Y_0 + R_{2,1}Y_1 + R_{2,2}Y_2]Y_2
+\end{align}
+
+Collecting all the positive terms of this expression gives
+\begin{align}
+F_{+} &= R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\
+&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\ 
+&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 
+\end{align}
+At which point we see that the same result is computable as an outer product of the vector $\vec{X}$ using
+
+\begin{align}
+F_{+} &= \vec{X}\mathbf{R}\vec{X} - \vec{Y}\mathbf{R}\vec{Y}
+\end{align}
+
+And the complex part of the gravitational wave is thus
+
+\begin{align}
+F_{\times} &= \vec{X}\mathbf{R}\vec{Y} + \vec{Y}\mathbf{R}\vec{X}
+\end{align}
+
+One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
+Thus we need a function that takes as input a \textit{list} vectors (or \textit{vector}) of vectors and produces a vector with the right components as output.
+To that end we define the tensor $X^{i}_{j}$ where
+
+\begin{align}
+X^i &= 
+\begin{bmatrix}
+X^{i}_0 \\
+X^{i}_1 \\
+X^{i}_2
+\end{bmatrix}
+\end{align}
+
+As well as the tensor $R^{i}_{jk}$ where
+
+\begin{align}
+R^{i} &= 
+\begin{bmatrix}
+R^{i}_{00} & R^{i}_{01} & R^{i}_{02} \\ 
+R^{i}_{10} & R^{i}_{11} & R^{i}_{12} \\ 
+R^{i}_{20} & R^{i}_{21} & R^{i}_{22}  
+\end{bmatrix}
+\end{align}
+
+The tensor $X$ is like a stack of all the different possible $\vec{X}$ coming out of the page.
+The Tensor $R$ is like $n$ copies of the matrix $\mathbf{R}$ stacked on top of each other.
+In this way the desired vector is obtainable with the tensor contraction
+
+\begin{align}
+F^{i}_{+} &= X^{lm}R^{i}_{lj}X^{j}_{m} - Y^{lm}R^{i}_{lj}Y^{j}_{m} \\ 
+F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
+\end{align}
+
+\subsection{Vectorized Single Detector Log Likelihood}
+
+With the spherical harmonics and antenna factor in hand the following set of operations yield the factored likelihood for a single detector, the network likelihood being a simple sum over all of the detectors.
+
+Equation 24 from Arxiv $15502.05370v1.pdf$ for the network log likelihood reads
+
+\begin{align}
+\ln{\mathcal{L}} = &\frac{D_{ref}}{D}Re \sum_{k}\sum_{(l,m)}\left[F_k Y_{lm}\right]^{*}Q_{k,lm} \\ 
+& - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] \\
+ & - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right]
+\end{align}
+
+Consider a single detector, thus dropping the sum over $k$. The first term is of the form $\vec{A}\cdot\vec{B} = \sum_{i=0}^{d}A_iB_i$, so if $Q_{k,(l,m)}$ were a simple vector, we could write it as
+
+\begin{equation}
+- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2} F*\vec{Y}\cdot\vec{Q}
+\end{equation}
+
+However the $Q_{k,(l,m)}$ are actually harmonic mode time series and not single values. We desire a vector whose values are the likelihoods at each point in the time series.
+
+Consider the case where we have only the $(2,-2), (2,0)$ and $(2,2)$ modes.
+If we write all the mode time series $Q^0, Q^1, Q^2...$ as the columns of a matrix , then the desired result is obtained with
+
+\begin{align}
+F*
+\begin{bmatrix}
+Q^0_{2,-2} & Q^0_{2,+0} & Q^0_{2,+2} \\
+Q^1_{2,-2} & Q^1_{2,+0} & Q^1_{2,+2} \\ 
+Q^2_{2,-2} & Q^2_{2,+0} & Q^2_{2,+2} \\
+\vdots & \vdots & \vdots
+\end{bmatrix}
+\begin{bmatrix}
+\left(Y_{2,-2}\right)\\
+\hspace{0mm} \\
+\left(Y_{2,+0}\right) \\
+\hspace{0mm} \\
+\left(Y_{2,+2}\right) \\
+\end{bmatrix}
+=
+\begin{bmatrix}
+Q^0_{2,-2} Y_{2,-2} + Q^0_{2,-2}Y_{2,+0} + Q^0_{2,+2}Y_{2,+2} \\
+Q^1_{2,-2} Y_{2,-2} + Q^1_{2,-2}Y_{2,+0} + Q^1_{2,+2}Y_{2,+2} \\
+Q^2_{2,-2} Y_{2,-2} + Q^2_{2,-2}Y_{2,+0} + Q^2_{2,+2}Y_{2,+2} \\
+\vdots
+\end{bmatrix}
+\end{align}
+
+With $\vec{Y}$ and $\mathbf{Q}$ defined as the matrix and vector above respectively, we have for the first term
+
+\begin{equation}
+\frac{D_{ref}}{D}Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right]
+\end{equation}
+
+The second term is a sum once over all the possible combinations of $(l,m), (l',m')$ pairs using the $U_{(l,m),(l',m')}$ cross terms. Its result is a scalar quantity made up of terms like
+
+\begin{align}
+&Y_{2,-2}^{*}Y_{2,-2}U_{(2,-2),(2,-2)} + Y_{2,-2}^{*}Y_{2,+0}U_{(2,-2),(2,+0)} + Y_{2,-2}^{*}Y_{2,+2}U_{(2,-2),(2,+2)} \\ 
+ +  \ &Y_{2,+0}^{*}Y_{2,-2}U_{(2,+0),(2,-2)} + Y_{2,+0}^{*}Y_{2,+0}U_{(2,+0),(2,+0)} + Y_{2,+0}^{*}Y_{2,+2}U_{(2,+0),(2,+2)} \\ 
++ \  &Y_{2,+2}^{*}Y_{2,-2}U_{(2,+2),(2,-2)} + Y_{2,+2}^{*}Y_{2,+0}U_{(2,+2),(2,+0)} + Y_{2,+2}^{*}Y_{2,+2}U_{(2,+2),(2,+2)}
+\end{align}
+
+If we pack the $U_{(l,m),(l',m')}$ into the matrix $\mathbf{U}$ as defined below then the following set of matrix operations produces the same sum
+
+\begin{align*}
+\begin{bmatrix}
+Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
+\end{bmatrix}
+\begin{bmatrix}
+U_{(2,-2),(2,-2)} &  U_{(2,-2),(2,+0)} &  U_{(2,-2),(2,+2)} \\
+U_{(2,+0),(2,-2)} &  U_{(2,+0),(2,+0)} &  U_{(2,+0),(2,+2)} \\
+U_{(2,+2),(2,-2)} &  U_{(2,+2),(2,+0)} &  U_{(2,+2),(2,+2)}
+\end{bmatrix}
+\begin{bmatrix}
+Y_{2,-2} \\
+Y_{2,+0} \\
+Y_{2,+2}
+\end{bmatrix}
+\end{align*}
+
+because when you multiply $\mathbf{U}$ into $\vec{Y}$ this simplifies to
+
+\begin{align*}
+\begin{bmatrix}
+Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
+\end{bmatrix}
+\begin{bmatrix}
+U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
+U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
+U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} 
+\end{bmatrix}
+\end{align*}
+
+Which becomes the desired scalar. This allows us to write the second term as
+
+\begin{align}
+- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] =  - \left[\frac{D_{ref}}{2D}\right]^{2} |F^2|\vec{Y}^{*}\mathbf{U}\vec{Y}
+\end{align}
+
+We must always set up the spherical harmonic vectors based on the value of $m$ and the cross terms in row-major form based first on $m_2$ and then on $m_1$.
+If we organize the matrix $\mathbf{V}$ in the same way then the same set of steps will lead us to conclude that
+
+\begin{align}
+- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2}Re \left[F^2 \vec{Y}\mathbf{V}\vec{Y} \right]
+\end{align}
+
+Combining the results the single detector log likelihood is
+
+\begin{align}
+\ln{\mathcal{L}} = \frac{D_{ref}}{D}\Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right] - \left[\frac{D_{ref}}{2D}\right]^{2}\left[|F|^2 \vec{Y}^{*}\mathbf{U}\vec{Y} - \Re\left(F^2 \vec{Y}\mathbf{V}\vec{Y}\right) \right]
+\end{align}
+
+\subsection{GPU Implementation}
+Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures.
+This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency.
+Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level.
+A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.
+
 %----------------------------------------------------------------------------------------

### 2016-08-19 Gradschool/thesis/msthesis.tex
expanded cuda section

@@
 
+$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics.
+However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner.
+Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e.
+$dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed.
+Thus we have
+
+\begin{align}
+h_{\mu \nu} = h_{\nu \mu}
+\end{align}
+
+Which, for a sixteen component tensor, makes six of the components redundant.
+This leaves us with ten.
+Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation.
+This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively.
+It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form
+
+\begin{align}
+\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
+V &\rightarrow V -\frac{\partial \Psi}{\partial t}
+\end{align}
+
+The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
+Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
+
+\begin{align}
+\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
+\rightarrow \partial^{\mu}A_{\mu} &= 0
+\end{align}
+
+It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden!
+While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
+It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}.
+We do not have to state \textit{what it is}, as long as we know that it exists.
+A proof that there legitimately is such a gauge function exists but is out of the scope of this outline.
+There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
+In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
+
+\begin{align}
+x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
+\end{align}
+
+In particular, we can show that there exists a coordinate transformation such that
+
+\begin{align}
+\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
+\end{align}
+
+\textbf{PUT PROOF HERE}
+
+\begin{align}
+\text{PUT PROOF HERE}
+\end{align}
+
+Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply
+
+\begin{align}
+\Box h_{\mu \nu} = 0
+\end{align}
+
+This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
+Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}.
+In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
+This leaves us with six.
+
+\section{Bayesian Statistics}
+We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
+Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \textbf{FIGURE}.
+Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
+We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$.
+A perpandicular cross section would simply be $P(y|x)$.
+We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$.
+Thus we have that
+
+\begin{align}
+P(x|y)P(y) &= P(y|x)P(x) \\
+P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
+\end{align}
+
+Which is Bayes' theorem.
+The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome.
+Reading $\neg x$ as "not x", we have
+
+\begin{align}
+P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
+&= \sum_{k=1}^K P(x|y_k)P(y_k) 
+\end{align}
+
+ Assuming a signal is present, the measured strain in a gravitational wave interferometer $d(t)$ can be decomposed into time dependent strain and noise components
+
+\begin{align}
+d(t) = h(t) + n(t)
+\end{align}
+
+then the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
+
+\begin{align}
+P[\vec{\mu}|d(t)] = \frac{P[d(t)|\vec{\mu}]P(\vec{\mu})}{P[d(t)]}
+\end{align}
+
+Extending equation \textbf{EQUATION} to the case of a continuous state space we have for the denominator
+
+\begin{align}
+P[d(t)] &= \int_{\Omega}P[d(t)|\vec{\mu}]P(\vec{\mu})d\vec{\mu}
+\end{align}
+
+Where $\Omega$ indicates integration over the entire probabilistic state space.
+
+\subsection{Harmonic Mode Decomposition}
+
+For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:
+
+\begin{align}
+h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
+\end{align}
+
+Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
+
+\begin{align}
+\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
+\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
+&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
+\end{align}
+
+Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,
+
+\begin{align}
+\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
+&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
+&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
+\end{align}
+
+The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters.
+The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
+Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.
+
 \section{Implementation}
@@
 A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.
+Reaching the limit of the sample throughput with conventional serial python libraries, our investigation turned to GPU based acceleration as the only means forward.
+The following is the result of this investigation.
+
+\subsection{Hardware Limitations and Logical Programming}
+GPUs work significantly differently from CPUs and must be paired with one to operate, they serve only as accelerators for complex operations and not as standalone units.
+Data must be passed back and forth from host memory on the node to RAM placed directly on the card itself through expensive transfers via the PCI bus.
+This is known as device global memory.
+While the onboard RAM on bleeding edge cards exceeds 12GB, memory transfers are the dominant cost of many GPU programs and must be minimized.
+Once the data exists on the device, it is accessed and manipulated by anywhere between nine and fifteen logical cores called Streaming Multiprocessors.
+These cores operate concurrently and contain a smaller granularity of parallel processing units that \textit{may} operate concurrently depending on the individual resources required by the requested operations.
+Programming in CUDA involves launching groups of threads called \textit{grids} that are further subdivided into \textit{blocks}.
+The blocks themselves are indexed by local variables that are rapidly accessible by all the contained threads, and the threads themselves all have access to indices placed within registers that exist for the lifetime of the thread in question.
+Both the blocks and and grids can be up to three dimensional and thus each block and thread may have access to up to three registers containing block indices and three containing thread indices.
+In addition thread blocks have access to an extremely high speed shared memory space where ideally most of the computations are performed, however it is limited in size to \textbf{SIZE} and is not designed to hold the entire target dataset.
+One major design goal for CUDA programs is implementing a process which copies portions of memory from the global space to the shared, processes it, and copies it back, while using the thread and block indices to blanket the dataset with threads and control which threads access what data.
+The current CUDA API allows for the logical launch of \textbf{HOW MANY} blocks with a maximum of 1024 threads per block, for a total of \textit{HOW MANY} possible "concurrent" threads.
+It is important to note that these threads are not \textit{physically} concurrent.
+Indeed, if each of the logical threads containing six integers in registers were to exist simultaneously, the GPU would be storing \textbf{HOW MANY} registers at once!
+Similarly perposterous amounts of shared memory would have to exist.
+In actuality, the shared memory and registers are located on the logical cores themselves, and blocks thread blocks line up to be processed.
+How many are processed at once is heavily dependent on the resources requested by the blocks themselves.
+Defining too many local variables in a kernel function can lead to \textit{register pressure} and decreased \textit{occupancy} of the multiprocessors on the GPU, a similar effect can come from requesting too much shared memory.
+Furthermore, the blocks are not guaranteed to execute in any particular order.
+While the compiler is responsible for sorting out how many blocks can be processed at a time as well as the order, the programmer is responsible for performance tuning with regard to optimizing block occupancy.
+This is often a late stage design consideration, but also leads to some of the greatest returns after algorithmic optimization has been completed.
+
+\subsection{PyCuda} 
+PyCuda is a flexible library for GPU scripting that includes a broad spectrum of GPU based array operations built to mimic the usage of NumPy arrays within a program.
+It allows for relatively easy access to the Nvidia CUDA API directly from within Python, and many of its operations are directly compatible with NumPy arrays with careful casting to the correct data types.
+Certainly one of the most challenging aspects of incorporating GPU code into a higher level language such as Python is the interface layer.
+As it stands much of the widely used LAL software consists of C code bound to Python through complex Swig interfaces, and while this is a perfectly valid way to access faster libraries, it can often be deceptively difficult to work with as many of the resulting Python objects contain read-only data structures masquerading as dynamic Python objects.
+This sometimes yields unpredictable behavior and it was decided this should be avoided as a way to interface with CUDA C.
+PyCuda is a very elegant solution to this problem as it seamlessly bridges this gap using the native GPUArray class, and a set of functions that create GPUArray copies of NumPy arrays that are built initially on the node CPU hosts.
+With a smaller group of developers, the built-in GPUArray class contained within PyCuda does not yet support \textit{all} of the rich spectrum of high level functions available within the NDArray NumPy class.
+Perhaps more importantly however, PyCuda allows for custom CUDA kernel functions to be written as pure CUDA C directly within special containers called source modules, that are contained within the main body of the Python code.
+PyCuda actually calls the real Nvidia compiler at runtime and links the resulting temporary executable to the main program.
+One may call these functions with NumPy arrays as arguments, as if they were normal Python functions.
+This is perhaps the most important feature of PyCuda as it allows the developer to tune the customization of the code to whatever level is desired.
+Those intermediate operations that may be handled with the built in methods of the GPUArray are performed as such.
+More complex operations that are not supported can be custom built with source modules.
+If desired, the entire program may consist of CUDA C with no more then basic startup routines written in actual Python meant to instantiate the original data and pass it down to the GPU.
+This particular work used mostly custom kernels, for the simple reason that applying the same matrix operation to a contiguous block of memory containing multiple slightly different copies constitutes tensor operations similar in nature to the results of section \textbf{SECTION}, for which there is no general support for in PyCuda.
 

### 2016-08-22 Gradschool/thesis/msthesis.tex
updated bayesian statistics section

@@
 \usepackage{cancel}
+\usepackage{graphicx}
+\usepackage{wrapfig}
+\usepackage[mathscr]{euscript}
 
@@
 
+\begin{figure}[h!]
+\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
+\caption{A visual depiction of the Reimann Curvature tensor.
+Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors.
+The orange vector $\vec{d}$ points between the tips.
+The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
+\label{fig:sphere}
+\end{figure}
+
 The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space.
 Here we motivate the concept with a visual explanation of parallel transport.
-Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \textbf{FIGURE HERE}.
+Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}.
 If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$.
@@
 We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
-Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \textbf{FIGURE}.
+Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
 Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
@@
 
-\begin{align}
+\begin{figure}[h!]
+\center
+\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
+\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
+\label{fig:sphere}
+\end{figure}
+
+\begin{align} \label{eq:bayes}
 P(x|y)P(y) &= P(y|x)P(x) \\
@@
 
-\begin{align}
+\begin{align}\label{eq:continuous}
 P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
-&= \sum_{k=1}^K P(x|y_k)P(y_k) 
+&= \sum_{k=1}^K P(y|x_k)P(x_k) 
 \end{align}
 
- Assuming a signal is present, the measured strain in a gravitational wave interferometer $d(t)$ can be decomposed into time dependent strain and noise components
+\clearpage 
+This allows us to rewrite the posterior probability as
 
 \begin{align}
-d(t) = h(t) + n(t)
+P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
 \end{align}
 
-then the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
+Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has
+
+\begin{align} \label{eq:lratio}
+P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
+\end{align}
+
+Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression.
+The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur.
+In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio.
+This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.
+
+\subsection{Signal Processing}
+Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:
 
 \begin{align}
-P[\vec{\mu}|d(t)] = \frac{P[d(t)|\vec{\mu}]P(\vec{\mu})}{P[d(t)]}
+P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
 \end{align}
 
-Extending equation \textbf{EQUATION} to the case of a continuous state space we have for the denominator
+If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as
 
 \begin{align}
-P[d(t)] &= \int_{\Omega}P[d(t)|\vec{\mu}]P(\vec{\mu})d\vec{\mu}
+P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
+\end{align}
+
+For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
+This is
+
+\begin{align}
+\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
+\end{align}
+
+In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$.
+If there is no signal present in the data, then the data and noise are equal, meaning
+
+\begin{align}
+d(t) &= n(t) \\
+P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
+\end{align}
+
+Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:
+
+\begin{align}
+d(t) &= n(t) + h(t) \\
+p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
+\end{align}
+
+So the likelihood ratio for gravitational wave signal processing is exactly equal to
+
+\begin{align}
+\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
+\end{align}
+
+By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
+
+\begin{align}\label{eq:parampos}
+P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
+\end{align}
+
+Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator
+
+\begin{align}
+P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
 \end{align}
@@
 Where $\Omega$ indicates integration over the entire probabilistic state space.
+Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us
+
+\begin{align}\label{eq:bayeslikely}
+P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
+\end{align}
+
+Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters.
+For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source.
+We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively.
+The denominator of equation \ref{eq:bayeslikely} is thus factorizable as
+
+\begin{align}
+\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta}
+\end{align}
+
+Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
+Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
+These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm.
+Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
+A new and more embarassingly parallel scheme is needed.
 
@@
 
-%----------------------------------------------------------------------------------------
+\subsection{General Design Considerations}
+Computing the time marginalized likelihood for a single sample involves processing the likelihood for each point in a signal time series, multiplied again by the number of associated harmonic modes.
+As the number of samples selected by the integrator as an adaptation chunk increases, the global memory required for the program to execute becomes dominated by these time series.
+Consequently the code is built upon managing and manipulating a large contiguous block of memory whose rows contain the harmonic mode time series for each sample, this lends itself naturally to using a two dimensional grid of one dimensional blocks as the launch configuration for most kernel functions.
+The one dimensional blocks lie conceptually along the rows and may line up if the number of time series samples exceeds the maximum threds per block.
+The grid extends downwards in the $y$-direction and will generally assign a row of blocks to each sample, or in some cases blanket the memory entirely in threads.
+It is the nature of CUDA programs that thread blocks operate most efficiently when the total number of threads per block is a power of two.
+Aside from the microprocessor architecture optimizing for these cases, memory reads by groups of thirty two threads (known as a \textit{Warp}) are coalesced by the compiler and recieve the value to be read as a broadcast.
+This massively boosts throughput if used correctly and it is thus advantageous in most cases to cast the problem dimensions into some workable multiple of the maximum threads per block on the hardware.
+For most Nvidia cards this is 1024.
+It is substantially easier to pad the time series with zeros out to the next greatest multiple of this number then to design specialized cleanup blocks whose sole purpose is to process the remainder.
+In the worse case this corresponds to holding \textbf{NUMBER} extra zeros in memory, which can be a painful amount of unecessary overhead in some scenarios.
+These factors come together to make memory management the main bottleneck for the entire algorithm, and a main area of focus for further development of this method.
+
+\subsection{Spherical Harmonics}
+
+The spin-weighted spherical harmonics are precomputed at startup.
+There is no analytic formula that returns the spin-weighted spherical harmonic for arbitrary quantum numbers $l, m$ and $s$.
+The most efficient way of retrieving them closely follows the method used by the current, serial LAL functions, which implement a lookup table that contains the explicit formulas.
+This is often sufficient as most waveform templates require only the $l=2$ modes, and thus only the corresponding spherical harmonics.
+Indeed, the LSC has yet to require support for $s \neq 2$, so the corresponding CUDA function does not attempt to extend this capability.
+
+\begin{wrapfig}
+\hspace{-1cm}
+\includegraphics[trim={0cm, 2cm, 2cm, 2cm}, clip]{spherical_harmonics.eps}
+\caption{Computation of Spherical Harmonics, $l=2$.}
+\label{fig:spharms}
+\end{wrapfig}
+
+Instead, it takes as argument an ordered list of $m$ values and executes the following set of steps to return a row-major ordered block of results whose rows correspond to values of $m$: Launch a one dimensional grid of one dimensional blocks, with one thread per sample.
+Compute the answer for this value of $m$ and place the result in global memory.
+Increment the local thread index by the number of samples and repeat.
+As long as it loops through the $m$ values in order, the results will be placed correctly, as per figure \ref{fig:spharms}.
+As the extrinsic parameter values used in computation are accessed relatively few times, this function does not use shared memory.
+Regardless as a standalone function it is able to compute several billion results per second, so little further optimization is necessary.
+
+\clearpage
+
+\subsection{Complex Antenna Factor}
+The antenna factor requires somewhat less fancy footwork then the spherical harmonics, as there is only one per sample and not an arbirary multiple depending on the number of selected modes (which are not necessarily all the values between $-l$ and $l$.
+The position vectors corresponding to the extrinsic parameters are contracted with the detector response tensor in a thread-wise fashion: one thread for each sample.
+We avoid using BLAS based routines to perform the contraction due to the small size of the matrices involved, startup routines for BLAS might take longer then a simple loop over nine elements of the detector tensor.
+We again use a one dimensional grid of one dimensional blocks, however since the position vectors contain commonly accessed elements, we choose to store these independently in device shared memory as well as the detector tensor itself in even higher speed device constant memory.
+This makes the antenna factor computation one of the most efficient steps of the process.
+
+\subsection{Main Routine}
+
+Due to the variation in extrinsic parameters over the samples, it is necessary for each to manipulate a separate group of copies of the harmonic mode time series.
+The number of samples may be large, so the main block of memory intended to hold these time series is handed down to the GPU initialized to zeros and a GPU function is used to expanded the time series into a vertial stack of copies, per figure \ref{fig:expandrho}.
+To streamline the marginalization process this block of memory is padded with zeros out to the nearest multiple of $T_{max}$, which is generally either 512 or 1024 depending on the GPU architecture.
+
+\begin{wrapfig}
+\vspace{10mm}
+\includegraphics[trim={0cm, 13cm, 0cm, 3.0cm}, clip, scale=0.65]{expand_rhots.eps}
+\label{fig:expandrho}
+\caption{Expansion of time series. One thread per relevant item in resulting memory block. Zeros are padded out to the nearest multiple of $T_{max}$.
+\end{wrapfig}
+
+\clearpage
+
+The result of this expansion of the time series into GPU memory is the desired vertical stack of time series copies.
+
+The next step is to transform these time series into term one of equation \textbf{EQUATION}.
+This involves combining the antenna factor (for which there is one complex number per sample) with the spin-weighted spherical harmonics (for which there are $N_m$ complex numbers per sample) into a single quantity and inserting them into the correct positions within the time series block.
+It is more efficient to complete the combination of the antenna factor and spherical harmonics prior to multiplying them into the time series, and this action is easily completed in an elementwise fasion using the built in machinery of PyCuda.
+It is steps like these for which PyCuda is an essential tool for performing what would otherwise be complex GPU operations with a single line of code.
+
+\begin{wrapfigure}{l}{0.6\textwidth}
+\hspace{-5cm}
+\vspace{-1cm}
+  \begin{center}
+    \includegraphics[trim={4cm, 13cm, 2cm, 3cm}, clip, width=0.75\textwidth]{FYrho.eps}
+  \end{center}
+  \caption{Insertion of the antenna factor-spherical harmonic pairs. }
+\end{wrapfigure}
+
+This step coupled with the previous is all that is necessary to build the first term of the likelihood for all of the samples on the GPU, the rest of the work is simply a series of summing operations.
+These must be handled somewhat carefully due to the possibility of race conditions and queued memory accesses, both of which can hinder or dangerously skew the results of a computation.
+The main issue stems from the fact that there is no guarantee of the order of execution of threadblocks within a program.
+This means that any program that depends upon memory reads and writes occuring in a specific order is prone to producting incorrect results.
+For this reason it is important to design the summation process in a way that does not exhibit this behavior: each thread should only read and write to memory locations unique to it.
+Note that each of the functions described thus far follows this paradigm.
+Queued memory are serialized by the CUDA compiler.
+This means that memory reads to the same location by multiple threads take place one by one.
+While not as dangerous as a race condition, it is a nontrivial performance consideration.
+Note that again the process thus far has been designed to avoid these situations.
+This is particularly important for justifying the nature of the next series of steps.
+
+\clearpage
+
+\begin{wrapfigure}{l}{0.6\textwidth}
+\vspace{-0.78cm}
+\begin{center}
+\includegraphics[trim={1.67cm, 10cm, 0, 3cm}, clip, scale=0.5]{sumrhots.eps}
+\caption{Downwards summation of the harmonic modes to form term one of the factored likelihood for a group of samples.}
+\end{center}
+\end{wrapfigure}
+
+The time series are summed downwards within memory by single threads.
+One row of threads is launched per sample, this leaves $N_m$ values for each thread to collect in a summation that resides in the bottom row.
+The memory locations accessed and written to by the individual threads in the grid are unique to that thread, avoiding races and queued memory accesses.
+The result are rows that are the same as those that would have been formed taking the product $\mathbf{Q}(F\vec{Y})$, this is the first term of the factored likelihood.
+It remains only to build the second term and subtract it from the rows of this matrix to complete the computation.
+
+\subsection{Marginalization Over Time}
+
+\begin{wrapfigure}{l}{0.6\textwidth}
+\begin{center}
+\includegraphics[scale=0.7]{onedmarg.eps}
+\end{center}
+\end{wrapfigure}
+
+%---------------------------------------------------------------------------------------
 

### 2016-08-25 Gradschool/thesis/msthesis.tex
still working

@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Short Sectioned Assignment
-% LaTeX Template
-% Version 1.0 (5/5/12)
-%
-% This template has been downloaded from:
-% http://www.LaTeXTemplates.com
-%
-% Original author:
-% Frits Wenneker (http://www.howtotex.com)
-%
-% License:
-% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%----------------------------------------------------------------------------------------
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%----------------------------------------------------------------------------------------
-
 \documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
-
 \usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
@@
 \usepackage{amsmath,amsfonts,amsthm} % Math packages
-
 \usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
-
 \usepackage{sectsty} % Allows customizing section commands
@@
 
-\begin{align}
+\begin{align}\label{eq:flteinst}
 \Box h_{\mu \nu} = 0
@@
 This leaves us with six.
+
+In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read
+
+\begin{align}
+\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
+-\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
+\end{align}
+
+This is the general form of the time dependent wave equation whose solutions take the form
+
+\begin{align}
+f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
+\end{align}
+
+Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector.
+Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have
+
+\begin{align}
+\text{PUT EQ HERE}
+\end{align}
 
@@
 &= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
-&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
+&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h^*_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y^*_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
 \end{align}
+
+Now, for any complex number $z = a + ib$, we have that $z*z + \Re(z^2) = 2z*z$, so $z*z = \frac{1}{2}(z*z + \Re z^2)$. Then the double sum over $l,m$ pairs becomes
 
@@
 \subsection{Complex Antenna Factor}
-The complex antenna factor is a complex number that modulates the gravitational wave signal as a function of incoming direction.
-This is necessary due to the non isotropic sensitivity of the LIGO interferometers themselves.
-$F_{+}$ and $F_{\times}$.
+Gravitational wave interferometers by nature possess nonisotropic sensitivity patterns that are a function of the specific geometry and orientation of the detector.
+As a result any incoming signal is modulated not only by directional effects \textbf{RA, DEC} relative to the celestial coordinate grid, but also temporal effects that are dependent on the orientation of the earth at the signal arrival time.
+Following closely the treatment from \textbf{CITATION} we note that as described in section \textbf{SECTION}, there are six degrees of freedom carried by gravitational waves.
+Four of these may be eliminated by a change of basis leaving us with only two spatial tensor polarizations $\epsilon^k_{ij}$ which can be expressed as
 
 \begin{align}
-F_{+} &= [R_{0,0}X_0 + R_{0,1}X_1 + R_{0,2}X_2]X_0 - [R_{0,0}Y_0 + R_{0,1}Y_1 + R_{0,2}Y_2]Y_0 \\
-&+ [R_{1,0}X_0 + R_{1,1}X_1 + R_{1,2}X_2]X_1 - [R_{1,0}Y_0 + R_{1,1}Y_1 + R_{1,2}Y_2]Y_1 \\
-&+ [R_{2,0}X_0 + R_{2,1}X_1 + R_{2,2}X_2]X_2 - [R_{2,0}Y_0 + R_{2,1}Y_1 + R_{2,2}Y_2]Y_2
+\epsilon^+_{ij} = 
+\begin{bmatrix}
+1 & 0 & 0 \\
+0 & -1 & 0 \\
+0 & 0 & 0
+\end{bmatrix}
+, \ \ e^{\times}_{ij} =
+\begin{bmatrix}
+0 & 1 & 0 \\
+1 & 0 & 0 \\
+0 & 0 & 0
+\end{bmatrix}
 \end{align}
 
-Collecting all the positive terms of this expression gives
-\begin{align}
-F_{+} &= R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\ 
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 
-\end{align}
-At which point we see that the same result is computable as an outer product of the vector $\vec{X}$ using
+These matrices are geometrically consistent with the orthogonal warping of ring like structures of test particles that are commonly used to visualize the effect of a passing gravitational wave: $\epsilon^+_{i,j}$ tends to grow vectors in the direction of the unit basis vectors whereas $\epsilon^{\times}_{ij}$ tends to shear them at an angle $\frac{\pi}{4}$ offset from the same basis.
+Moreover, there is an additional effect from the polarization angle $\psi$ of the incoming wave relative to the detector basis.
+Manifestation of a gravitational wave as measurable, real valued detector strain $h(t)$ is given by a the contraction of an object known as the detector tensor with these polarization tensors.
 
 \begin{align}
-F_{+} &= \vec{X}\mathbf{R}\vec{X} - \vec{Y}\mathbf{R}\vec{Y}
+\mathbf{R^{ij}} = \frac{1}{2}(v^i v^j - u^i u^j)
 \end{align}
 
-And the complex part of the gravitational wave is thus
+With the various polarization tensors to form the linear combination
 
 \begin{align}
-F_{\times} &= \vec{X}\mathbf{R}\vec{Y} + \vec{Y}\mathbf{R}\vec{X}
+h(t) = h_{+}(t)R^{ij}_e^+_{ij} + h_{\times}(t)R^{ij}e^{\times}_{ij}
 \end{align}
 
+The products $R^{ij}_e^+_{ij}$ and $R^{ij}e^{\times}_{ij}$ are known as the \textit{antenna factors}, $F_{+}$ and $F_{\times}$.
+As with the gravitational wave strain itself, we represent these as a single complex number $F_{+} - i F_{\times}$, in keeping with the formalism of section \textbf{SECTION}.
+It can be shown that in the wave basis, the scalar in question can be expressed as
+
+\begin{align}
+F_+ - i F_{\times} &= \vec{X}^T\mathbf{R}\vec{X} - \vec{Y}^T\mathbf{R}\vec{Y} - i(\vec{X}^T\mathbf{R}\vec{Y} + \vec{Y}^T\mathbf{R}\vec{X})
+\end{align}
+
+Where $\vec{X}$ and $\vec{Y}$ are the wave-frame axes. These can be redefined in terms of the working parameters as
+
+\begin{align}
+\text{PUT DEFINITIONS HERE}
+\end{align}
+
+Our objective is to produce this quantity for a group of samples, all at once, as a coherent vectorized operation.
 One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
@@
 
-\clearpage
-
 \subsection{Complex Antenna Factor}
@@
 \end{wrapfig}
-
-\clearpage
 
@@
 \subsection{Marginalization Over Time}
+Having integrated over the extrinsic parameters numerically, it remains to marginalize over the entire time series to produce a single scalar representing the likelihood associated with a sample.
+While a simplistic operation in theory special attention was taken to optimize the summation process.
+This is due to the "width" of the data to be processed.
+The time series themselves can be thousands of entries long depending on the signal sample rate.
+Accelerating the preparation of the time series, as demonstrated in the previous steps, can be performed with brute force by flooding the calculation with threads, parallel reduction on a GPU requires finesse.
+In fact, it was found that this particular step was the main bottleneck in the serial code, thus it was expected that here the greatest returns would be achieved.
+The objective is to keep as many threads busy as possible to maintain throughput, however as we will demonstrate, the best that can be achieved without excessively complicated schemes is to start with $\frac{N_t}{2}$ threads and recursively cut the working group in half, a total of $\log_2(N_t)$ times, as per figure \ref{fig:recursive}.
 
 \begin{wrapfigure}{l}{0.6\textwidth}
+\label{fig:recursive}
+\caption{Sequential addressing for parallel reduction.}
 \begin{center}
-\includegraphics[scale=0.7]{onedmarg.eps}
+\includegraphics[trim={0, 13cm, 0, 0}, clip, scale=0.5]{onedmarg.eps}
 \end{center}
 \end{wrapfigure}
+
+\begin{wrapfigure}{h}{0.6\textwidth}
+\label{fig:folding}
+\caption{"Folding" of the time series block for time marginalization}
+\begin{center}
+\includegraphics[trim={0, 13cm, 0, 0}, clip, scale=0.5]{marg_fold.eps}
+\end{center}
+\end{wrapfig}
+
+\section{Results and Prospects for Low Latency Electromagnetic Followup}
+\begin{figure}[H!]
+\hspace{-5.0cm}
+\includegraphics[scale=0.57]{speedup.eps}
+\end{figure}
+
+The dawn of gravitational wave astronomy in the 2015-2016 era has touched off a revolution in our ability to observe and classify astronomical phenomena.
+As with any entirely novel method of gleaning information from distant sources, development of methods to process and decipher such data has been explosive.
+Pure gravitational wave astrophysics has exciting prospects for the future, but the possibility of advancing other classical forms of astrophysics through their interplay with these newer methods is equally compelling.
+Significant research \textbf{CITATION} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers.
+A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration klonovae.
+Such early warning could provide valuable insights as to the central engine of such gamma ray bursts and confirm recent models of these processes.
+However, electromagnetic observational resources are limited and expensive to operate and such it is imperative that any communication to such resources in the context of EM followup of GW triggers must be rapid and accurate to avoid wasting time and money.
+To that end it is desireable to develop existing parameter estimation codes to such a point that they can produces fast and correct results to reliably indicate potential sources of interest while minimizing false positives.
+The main goal of this work is to push this envelope and reduce the end-to-end runtime of parallel PE codes (specifically RapidPE) for the express purpose of low latency electromagnetic follow-up.
+
+\subsection{Gamma Ray Bursts}
+Gamma ray bursts (GRBs) are short duration transient events thought to occur during or shortly after the merger of compact binary systems, specifically those which include a single black hole and one neutron star.
+The presence of strong radiative emission from the poles of black holes has been confirmed by \textbf{CITATION}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs.
+The process is depicted in figure \textbf{FIGURE} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole.
+The radiation eminated from the poles of the BH impacts this disk and generates gamma rays, this is known as an internal shock.
+As the resulting cloud of matter ploughs into the surrounding interstellar medium, it slows, leaving a lower density region in its wake.
+Subsequent waves of matter may propagate through this region at a higher velocity then the initial blast wave, catching up and ramming into the back of the initial wave, producing further shocks.
+It is not yet explicitly confirmed that this is the exact process by which GRBs are generated.
+It is generally accepted that to confirm this hypothesis we would need to observe direct coincidence between a GW trigger localized to some region of the sky and a GRB from the same direction.
+These phenomena have been demonstrated in numerical simulations by \textbf{CITATION} and are highly dependent on initial system parameters.
+In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors \textbf{CITATION}.
+We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
+We use rate predictions from \textbf{CITATION} to restrict the mass range to be studied to fall between \textbf{SOME RANGE}.
+These bounds corespond to a chirp mass range of \textbf{WHAT RANGE}.
+From this we can determine the expected time that the signal will spend in the Advanced LIGO band.
+In figure \textbf{FIGURE}, we have visualized this range in \textbf{SOME PLANE}.
+Using the results of our code we find that anadditional \textbf{SOME PERCENTAGE} of expected NSBH sytems expected to be picked up by Advanced LIGO are now candidates for joint GW-EM astronomy.
+
+\begin{figure}[H!]
+\hspace{-0.5cm}
+\includegraphics[scale=0.35]{gamray.png}
+\end{figure}
+
+\subsection{Kilonovae}
+Compact binary mergers of neutron stars are thought to produce transient electromagnetic counterparts known as \textit{Kilonovae}.
+These smaller class of novae are dimmer and lower energy then regular novae or supernovae, and are hypothesized to occur due to the radiative emission of remnant matter displaced by the merger dynamics.
+It has been shown that a substantial fraction of elements heavier then Iron-56 are formed through r-process nucleosynthesis in the hearts of large stars, during which nuclei are subjected to high neutron flux as well as high electron density filling quantum states up to the Fermi energy.
+When this occurs $\beta^-$ decay of neutrons into protons is blocked \textit{CITATION} resulting in accumulation of large amounts of neutrons within the nuclei themselves.
+These nuclei later decay into heavy elements.
+These conditions are present during the merger of neutron stars or the tidal disruption of a neutron star by a black hole.
+Models of kilonovae have been successful in reproducing faint EM transients using r-process emission as a primary mechanism for radiation.
+Coincident measurements of such transients with localized gravitational waves is expected to be a smoking gun for confirming this model of kilonova genesis, thus it is important to assess the capability of modern parameter estimation to process signals from candidate systems and produce meaningful information useful for kilonova followup.
+Since merger ejecta from NSBH mergers is not subject to relativistic beaming, kilonovae emission is expected to be relatively isotropic.
+Assuming a random and uniform distribution of black hole spin directions there is thus an increased probability of detecting kilonovae of \textbf{SOLID ANGLE OVER 4 PI} relative to axially emissive GRBs.
+This translates to a better chance of finding a kilonova as a merger signature.
+Moreover, direct observation of kilonovae combined with numerical simulations can provide degeneracy-breaking information about the original progenitor through careful measurements of redshift compared to simulated spectra.
+The timescales characteristic of kilonova signatures are derived from numerical simulations, and most suggest a peak in the emission intensity roughly a day after merger.
+That said, the same simulations predict a steep rise in luminosity in the period just before the peak.
+For this reason crucial information about the astrophysics of kilonovae may be rapidly lost, and it is important to collect it as rapidly as possible.
+Figure \textbf{SOME FIGURE} depicts the timescales associated with kilonovae from \textbf{CITATION} as well as current PE methods prospects of convergence prior to certain features of the kilonova light curve.
 

### 2016-08-25 Gradschool/thesis/msthesis.tex
added memory concerns

@@
 
+\subsection{Memory Considerations}
+Here we provide an overview of the memory resources required by the contributing structures involved in the likelihood calculation.
+This is a bookkeeping step necessary to select the correct sample array sizes given available hardware and tune the implementation to achieve optimal performance.
+The samples themselves are double precision arrays of length $N_s$.
+For $P$ params this equates to $8PN_s$ bytes of memory.
+The first computation in the routine is the generation of spherical harmonics, given a maximum value of $l$.
+Since there are $2l+1$ possible values of $m$ for each $l$, there are
+
+\begin{align}
+\sum_{l=0}^{L_{max}} (2l + 1) 
+\end{align}
+
+Total values of $m$ under a certain $L_{max}$.
+The partial sums are $S_L = (L_{max} + 1)^2$, so there are up to $16N_s(L_{max}+1)^2$ bytes of memory associated the spherical harmonics for a given value of $L_{max}$, requiring an extra factor of two to store complex numbers in double precision.
+Finally, we access to the complex conjugate of the spherical harmonics, this adds another factor of two bringing the total contribution to $32N_s(L_{max}+1)^2$ bytes.
+The antenna factor is also a complex number, but there is only a single value for each sample, adding only $16N_s$ bytes to the total memory usage, with an additional $16N_s$ for the conjugate.
+This is amounts to an additional $32N_s$ bytes.
+By far the greatest memory concern is the large time series block inside which the main calculation occurs.
+This block requires a row for each sample, multiplied by the number of modes for a total of $N_m N_s$ rows.
+Each of these rows holds $N_t$ elements that could, in the worst case, need to be padded by up to $T_{max} - 1$ zeros to enable proper time marginalization.
+Each element of this block is a double precision complex number.
+This equates to $16N_m N_s (N_t + T_{max} - 1)$ bytes of memory.
+An unfortunate shortcoming of PyCuda is that some simple operations can not be performed in place - one such operation is taking the real part of an array.
+This is due to the recasting of data types on the GPU itself.
+This means that for a brief portion of the calculation, an additional $50\%$ memory is needed raising the requirement to $24N_m N_s (N_t + T_{max} - 1)$ at peak for the main block.
+The $U$ and $V$ crossterms are complex numbers that require $16N_s$ bytes each, for a total of $32N_s$.
+The second term of the likelihood, which is a combination of these, is a real number of the same length, adding $8N_s$ bytes to the total.
+The sum of these contributions is the total number of bytes $B$ and is given by
+
+\begin{align}
+N_s(8P + 32(L_{max} + 1)^2 +  24N_m(N_t + T_{max} - 1) + 72)
+\end{align}
+
+Below we have tabulated $B$ for some values of $N_s$ to be used as reference when running the code.
+
 \section{Results and Prospects for Low Latency Electromagnetic Followup}
@@
 
+\begin{figure}[H!]
+\hspace{-6cm}
+\includegraphics[scale=0.5]{EMFollow.eps}
+\end{figure}
+
 %---------------------------------------------------------------------------------------

### 2016-08-26 Gradschool/thesis/final/chap1.tex
added MIT template

--- 2016-08-22 Gradschool/thesis/msthesis.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Short Sectioned Assignment
-% LaTeX Template
-% Version 1.0 (5/5/12)
-%
-% This template has been downloaded from:
-% http://www.LaTeXTemplates.com
-%
-% Original author:
-% Frits Wenneker (http://www.howtotex.com)
-%
-% License:
-% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%----------------------------------------------------------------------------------------
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%----------------------------------------------------------------------------------------
-
-\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
-
-\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
-\usepackage[adobe-utopia]{mathdesign}
-%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
-\usepackage[english]{babel} % English language/hyphenation
-\usepackage{amsmath,amsfonts,amsthm} % Math packages
-
-\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
-
-\usepackage{sectsty} % Allows customizing section commands
-
-\usepackage{braket}
-\usepackage{cancel}
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage[mathscr]{euscript}
-
-\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps
-
-\usepackage{fancyhdr} % Custom headers and footers
-\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
-\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
-\fancyfoot[L]{} % Empty left footer
-\fancyfoot[C]{} % Empty center footer
-\fancyfoot[R]{\thepage} % Page numbering for right footer
-\renewcommand{\headrulewidth}{0pt} % Remove header underlines
-\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
-\setlength{\headheight}{13.6pt} % Customize the height of the header
-
-\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-
-\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text
-
-%----------------------------------------------------------------------------------------
-%	TITLE SECTION
-%----------------------------------------------------------------------------------------
-
-\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height
-
-\title{	
-\normalfont \normalsize 
-\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
-\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
-\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
-}
-
-\author{Brandon B. Miller, Masters Thesis} % Your name
-
-\date{\normalsize\today} % Today's date or a custom date
-
-\begin{document}
-
-\maketitle % Print the title
-\newpage
-\section{Background and Motivation}
-
+%% This is an example first chapter.  You should put chapter/appendix that you
+%% write into a separate file, and add a line \include{yourfilename} to
+%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
+%% You can process specific files by typing their names in at the 
+%% \files=
+%% prompt when you run the file main.tex through LaTeX.
+\chapter{Linearized Gravity}
 Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
@@
 Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}
+Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel symbols defined in \nameref{Appendix C}.
 
@@
 If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
+Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{\rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
 It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
@@
 
-\begin{align}
+\begin{align}\label{eq:metpluspert}
 g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
@@
 The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \textbf{EQUATION}, we would have
+If we tried to use the full form of equation \ref{eq:metpluspert}, we would have
 
@@
 
-The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity.
+The second two terms being second order in $h$, the peturbation, are neglected in linearized gravity.
 We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
@@
 
-\begin{align}
+\begin{align}\label{eq:linchrist}
 \Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
@@
 
-Plugging in equation \textbf{EQUATION} we have
+Plugging in equation \ref{eq:linchrist} we have
 
@@
 &= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
-&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
+&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]\label{eq:linricci}
 \end{align}
@@
 
-\begin{align}
+\begin{align}\label{eq:lorentzgauge}
 V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
@@
 
-Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields
+Which upon substitution into equation \ref{eq:linricci} for $R_{\mu \nu}$ yields
 
@@
 
-Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
+Setting the right hand side of this equation equal to zero gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
 
@@
 
-\begin{align}
+\begin{align}\label{eq:emtrans}
 \vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
@@
 
-The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
+The function $\Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
 Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
 
-\begin{align}
+\begin{align}\label{eq:emgauge}
 \nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
@@
 While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
-It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}.
+It actually is there inside the $V$ and $\vec{A}$ written in equation \ref{eq:emgauge}.
 We do not have to state \textit{what it is}, as long as we know that it exists.
@@
 There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
-In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
+In the same sense as equation \ref{eq:emtrans}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
 
@@
 This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
-Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}.
+Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \ref{eq:lorentzgauge}.
 In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
@@
 
-\section{Bayesian Statistics}
-We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
-Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
-Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
-We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$.
-A perpandicular cross section would simply be $P(y|x)$.
-We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$.
-Thus we have that
+In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read
 
-\begin{figure}[h!]
-\center
-\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
-\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
-\label{fig:sphere}
-\end{figure}
-
-\begin{align} \label{eq:bayes}
-P(x|y)P(y) &= P(y|x)P(x) \\
-P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
+\begin{align}
+\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
+-\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
 \end{align}
 
-Which is Bayes' theorem.
-The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome.
-Reading $\neg x$ as "not x", we have
+This is the general form of the time dependent wave equation whose solutions take the form
 
-\begin{align}\label{eq:continuous}
-P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
-&= \sum_{k=1}^K P(y|x_k)P(x_k) 
+\begin{align}
+f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
 \end{align}
 
-\clearpage 
-This allows us to rewrite the posterior probability as
+Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector.
+Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have
 
 \begin{align}
-P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
+\text{PUT EQ HERE}
 \end{align}
-
-Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has
-
-\begin{align} \label{eq:lratio}
-P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
-\end{align}
-
-Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression.
-The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur.
-In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio.
-This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.
-
-\subsection{Signal Processing}
-Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:
-
-\begin{align}
-P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
-\end{align}
-
-If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as
-
-\begin{align}
-P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
-\end{align}
-
-For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
-This is
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
-\end{align}
-
-In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$.
-If there is no signal present in the data, then the data and noise are equal, meaning
-
-\begin{align}
-d(t) &= n(t) \\
-P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
-\end{align}
-
-Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:
-
-\begin{align}
-d(t) &= n(t) + h(t) \\
-p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
-\end{align}
-
-So the likelihood ratio for gravitational wave signal processing is exactly equal to
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
-\end{align}
-
-By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
-
-\begin{align}\label{eq:parampos}
-P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
-\end{align}
-
-Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator
-
-\begin{align}
-P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
-\end{align}
-
-Where $\Omega$ indicates integration over the entire probabilistic state space.
-Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us
-
-\begin{align}\label{eq:bayeslikely}
-P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
-\end{align}
-
-Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters.
-For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source.
-We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively.
-The denominator of equation \ref{eq:bayeslikely} is thus factorizable as
-
-\begin{align}
-\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta}
-\end{align}
-
-Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
-Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
-These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm.
-Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
-A new and more embarassingly parallel scheme is needed.
-
-\subsection{Harmonic Mode Decomposition}
-
-For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:
-
-\begin{align}
-h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
-\end{align}
-
-Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
-
-\begin{align}
-\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
-\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
-&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
-\end{align}
-
-Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,
-
-\begin{align}
-\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
-&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
-&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
-\end{align}
-
-The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters.
-The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
-Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.
-
-\section{Implementation}
-The main weight of the computational work described above falls under an expensive Monte Carlo integration of the evidence as part of the Bayesian posterior.
-Zooming in it quickly becomes apparent that the main challenge to be overcome is efficient evaluation of the function to be integrated, in this case the factored likelihood.
-Prior to this work, the factored likelihood is was computed with a series of looping structures that individually compiled all the terms of the equation in an independent and serial manner.
-The process took substantial advantage of many high level operations available within Python to match up the correct terms across various harmonic mode time series, spherical harmonics, and antenna patterns.
-Thanks to clever optimization and gratuitous use of fast numerical Python libraries such as NumPy, the code was able to compute approximately $10^3$ likelihood evaluations on the order of seconds, the limiting factor being the serial nature of loops and the inherent lethargy of high level languages such as Python.
-Although NumPy commonly passes target data to compiled routines written in faster languages such as C, some portions of the computation (in particular the marginalization over time) benefit little from this capability and form a bottleneck for efficient evaluation of the likelihood.
-Furthermore the original implementation heavily relied upon older structures within the existing LIGO Algorithms Library (LAL), a set of C routines bound to Python through Swig, to perform many intermediate calculations, such as computing spherical harmonics.
-The goal of the first phase of this work was thorough vectorization of the process used to build the terms of the likelihood utilizing the BLAS subroutines available through NumPy as often as possible, as well as reducing dependency on LAL.
-To that end, the components of the factored likelihod were reworked.
-
-\subsection{Complex Antenna Factor}
-The complex antenna factor is a complex number that modulates the gravitational wave signal as a function of incoming direction.
-This is necessary due to the non isotropic sensitivity of the LIGO interferometers themselves.
-$F_{+}$ and $F_{\times}$.
-
-\begin{align}
-F_{+} &= [R_{0,0}X_0 + R_{0,1}X_1 + R_{0,2}X_2]X_0 - [R_{0,0}Y_0 + R_{0,1}Y_1 + R_{0,2}Y_2]Y_0 \\
-&+ [R_{1,0}X_0 + R_{1,1}X_1 + R_{1,2}X_2]X_1 - [R_{1,0}Y_0 + R_{1,1}Y_1 + R_{1,2}Y_2]Y_1 \\
-&+ [R_{2,0}X_0 + R_{2,1}X_1 + R_{2,2}X_2]X_2 - [R_{2,0}Y_0 + R_{2,1}Y_1 + R_{2,2}Y_2]Y_2
-\end{align}
-
-Collecting all the positive terms of this expression gives
-\begin{align}
-F_{+} &= R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\ 
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 
-\end{align}
-At which point we see that the same result is computable as an outer product of the vector $\vec{X}$ using
-
-\begin{align}
-F_{+} &= \vec{X}\mathbf{R}\vec{X} - \vec{Y}\mathbf{R}\vec{Y}
-\end{align}
-
-And the complex part of the gravitational wave is thus
-
-\begin{align}
-F_{\times} &= \vec{X}\mathbf{R}\vec{Y} + \vec{Y}\mathbf{R}\vec{X}
-\end{align}
-
-One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
-Thus we need a function that takes as input a \textit{list} vectors (or \textit{vector}) of vectors and produces a vector with the right components as output.
-To that end we define the tensor $X^{i}_{j}$ where
-
-\begin{align}
-X^i &= 
-\begin{bmatrix}
-X^{i}_0 \\
-X^{i}_1 \\
-X^{i}_2
-\end{bmatrix}
-\end{align}
-
-As well as the tensor $R^{i}_{jk}$ where
-
-\begin{align}
-R^{i} &= 
-\begin{bmatrix}
-R^{i}_{00} & R^{i}_{01} & R^{i}_{02} \\ 
-R^{i}_{10} & R^{i}_{11} & R^{i}_{12} \\ 
-R^{i}_{20} & R^{i}_{21} & R^{i}_{22}  
-\end{bmatrix}
-\end{align}
-
-The tensor $X$ is like a stack of all the different possible $\vec{X}$ coming out of the page.
-The Tensor $R$ is like $n$ copies of the matrix $\mathbf{R}$ stacked on top of each other.
-In this way the desired vector is obtainable with the tensor contraction
-
-\begin{align}
-F^{i}_{+} &= X^{lm}R^{i}_{lj}X^{j}_{m} - Y^{lm}R^{i}_{lj}Y^{j}_{m} \\ 
-F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
-\end{align}
-
-\subsection{Vectorized Single Detector Log Likelihood}
-
-With the spherical harmonics and antenna factor in hand the following set of operations yield the factored likelihood for a single detector, the network likelihood being a simple sum over all of the detectors.
-
-Equation 24 from Arxiv $15502.05370v1.pdf$ for the network log likelihood reads
-
-\begin{align}
-\ln{\mathcal{L}} = &\frac{D_{ref}}{D}Re \sum_{k}\sum_{(l,m)}\left[F_k Y_{lm}\right]^{*}Q_{k,lm} \\ 
-& - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] \\
- & - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right]
-\end{align}
-
-Consider a single detector, thus dropping the sum over $k$. The first term is of the form $\vec{A}\cdot\vec{B} = \sum_{i=0}^{d}A_iB_i$, so if $Q_{k,(l,m)}$ were a simple vector, we could write it as
-
-\begin{equation}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2} F*\vec{Y}\cdot\vec{Q}
-\end{equation}
-
-However the $Q_{k,(l,m)}$ are actually harmonic mode time series and not single values. We desire a vector whose values are the likelihoods at each point in the time series.
-
-Consider the case where we have only the $(2,-2), (2,0)$ and $(2,2)$ modes.
-If we write all the mode time series $Q^0, Q^1, Q^2...$ as the columns of a matrix , then the desired result is obtained with
-
-\begin{align}
-F*
-\begin{bmatrix}
-Q^0_{2,-2} & Q^0_{2,+0} & Q^0_{2,+2} \\
-Q^1_{2,-2} & Q^1_{2,+0} & Q^1_{2,+2} \\ 
-Q^2_{2,-2} & Q^2_{2,+0} & Q^2_{2,+2} \\
-\vdots & \vdots & \vdots
-\end{bmatrix}
-\begin{bmatrix}
-\left(Y_{2,-2}\right)\\
-\hspace{0mm} \\
-\left(Y_{2,+0}\right) \\
-\hspace{0mm} \\
-\left(Y_{2,+2}\right) \\
-\end{bmatrix}
-=
-\begin{bmatrix}
-Q^0_{2,-2} Y_{2,-2} + Q^0_{2,-2}Y_{2,+0} + Q^0_{2,+2}Y_{2,+2} \\
-Q^1_{2,-2} Y_{2,-2} + Q^1_{2,-2}Y_{2,+0} + Q^1_{2,+2}Y_{2,+2} \\
-Q^2_{2,-2} Y_{2,-2} + Q^2_{2,-2}Y_{2,+0} + Q^2_{2,+2}Y_{2,+2} \\
-\vdots
-\end{bmatrix}
-\end{align}
-
-With $\vec{Y}$ and $\mathbf{Q}$ defined as the matrix and vector above respectively, we have for the first term
-
-\begin{equation}
-\frac{D_{ref}}{D}Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right]
-\end{equation}
-
-The second term is a sum once over all the possible combinations of $(l,m), (l',m')$ pairs using the $U_{(l,m),(l',m')}$ cross terms. Its result is a scalar quantity made up of terms like
-
-\begin{align}
-&Y_{2,-2}^{*}Y_{2,-2}U_{(2,-2),(2,-2)} + Y_{2,-2}^{*}Y_{2,+0}U_{(2,-2),(2,+0)} + Y_{2,-2}^{*}Y_{2,+2}U_{(2,-2),(2,+2)} \\ 
- +  \ &Y_{2,+0}^{*}Y_{2,-2}U_{(2,+0),(2,-2)} + Y_{2,+0}^{*}Y_{2,+0}U_{(2,+0),(2,+0)} + Y_{2,+0}^{*}Y_{2,+2}U_{(2,+0),(2,+2)} \\ 
-+ \  &Y_{2,+2}^{*}Y_{2,-2}U_{(2,+2),(2,-2)} + Y_{2,+2}^{*}Y_{2,+0}U_{(2,+2),(2,+0)} + Y_{2,+2}^{*}Y_{2,+2}U_{(2,+2),(2,+2)}
-\end{align}
-
-If we pack the $U_{(l,m),(l',m')}$ into the matrix $\mathbf{U}$ as defined below then the following set of matrix operations produces the same sum
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} &  U_{(2,-2),(2,+0)} &  U_{(2,-2),(2,+2)} \\
-U_{(2,+0),(2,-2)} &  U_{(2,+0),(2,+0)} &  U_{(2,+0),(2,+2)} \\
-U_{(2,+2),(2,-2)} &  U_{(2,+2),(2,+0)} &  U_{(2,+2),(2,+2)}
-\end{bmatrix}
-\begin{bmatrix}
-Y_{2,-2} \\
-Y_{2,+0} \\
-Y_{2,+2}
-\end{bmatrix}
-\end{align*}
-
-because when you multiply $\mathbf{U}$ into $\vec{Y}$ this simplifies to
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} 
-\end{bmatrix}
-\end{align*}
-
-Which becomes the desired scalar. This allows us to write the second term as
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] =  - \left[\frac{D_{ref}}{2D}\right]^{2} |F^2|\vec{Y}^{*}\mathbf{U}\vec{Y}
-\end{align}
-
-We must always set up the spherical harmonic vectors based on the value of $m$ and the cross terms in row-major form based first on $m_2$ and then on $m_1$.
-If we organize the matrix $\mathbf{V}$ in the same way then the same set of steps will lead us to conclude that
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2}Re \left[F^2 \vec{Y}\mathbf{V}\vec{Y} \right]
-\end{align}
-
-Combining the results the single detector log likelihood is
-
-\begin{align}
-\ln{\mathcal{L}} = \frac{D_{ref}}{D}\Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right] - \left[\frac{D_{ref}}{2D}\right]^{2}\left[|F|^2 \vec{Y}^{*}\mathbf{U}\vec{Y} - \Re\left(F^2 \vec{Y}\mathbf{V}\vec{Y}\right) \right]
-\end{align}
-
-\subsection{GPU Implementation}
-Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures.
-This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency.
-Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level.
-A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.
-Reaching the limit of the sample throughput with conventional serial python libraries, our investigation turned to GPU based acceleration as the only means forward.
-The following is the result of this investigation.
-
-\subsection{Hardware Limitations and Logical Programming}
-GPUs work significantly differently from CPUs and must be paired with one to operate, they serve only as accelerators for complex operations and not as standalone units.
-Data must be passed back and forth from host memory on the node to RAM placed directly on the card itself through expensive transfers via the PCI bus.
-This is known as device global memory.
-While the onboard RAM on bleeding edge cards exceeds 12GB, memory transfers are the dominant cost of many GPU programs and must be minimized.
-Once the data exists on the device, it is accessed and manipulated by anywhere between nine and fifteen logical cores called Streaming Multiprocessors.
-These cores operate concurrently and contain a smaller granularity of parallel processing units that \textit{may} operate concurrently depending on the individual resources required by the requested operations.
-Programming in CUDA involves launching groups of threads called \textit{grids} that are further subdivided into \textit{blocks}.
-The blocks themselves are indexed by local variables that are rapidly accessible by all the contained threads, and the threads themselves all have access to indices placed within registers that exist for the lifetime of the thread in question.
-Both the blocks and and grids can be up to three dimensional and thus each block and thread may have access to up to three registers containing block indices and three containing thread indices.
-In addition thread blocks have access to an extremely high speed shared memory space where ideally most of the computations are performed, however it is limited in size to \textbf{SIZE} and is not designed to hold the entire target dataset.
-One major design goal for CUDA programs is implementing a process which copies portions of memory from the global space to the shared, processes it, and copies it back, while using the thread and block indices to blanket the dataset with threads and control which threads access what data.
-The current CUDA API allows for the logical launch of \textbf{HOW MANY} blocks with a maximum of 1024 threads per block, for a total of \textit{HOW MANY} possible "concurrent" threads.
-It is important to note that these threads are not \textit{physically} concurrent.
-Indeed, if each of the logical threads containing six integers in registers were to exist simultaneously, the GPU would be storing \textbf{HOW MANY} registers at once!
-Similarly perposterous amounts of shared memory would have to exist.
-In actuality, the shared memory and registers are located on the logical cores themselves, and blocks thread blocks line up to be processed.
-How many are processed at once is heavily dependent on the resources requested by the blocks themselves.
-Defining too many local variables in a kernel function can lead to \textit{register pressure} and decreased \textit{occupancy} of the multiprocessors on the GPU, a similar effect can come from requesting too much shared memory.
-Furthermore, the blocks are not guaranteed to execute in any particular order.
-While the compiler is responsible for sorting out how many blocks can be processed at a time as well as the order, the programmer is responsible for performance tuning with regard to optimizing block occupancy.
-This is often a late stage design consideration, but also leads to some of the greatest returns after algorithmic optimization has been completed.
-
-\subsection{PyCuda} 
-PyCuda is a flexible library for GPU scripting that includes a broad spectrum of GPU based array operations built to mimic the usage of NumPy arrays within a program.
-It allows for relatively easy access to the Nvidia CUDA API directly from within Python, and many of its operations are directly compatible with NumPy arrays with careful casting to the correct data types.
-Certainly one of the most challenging aspects of incorporating GPU code into a higher level language such as Python is the interface layer.
-As it stands much of the widely used LAL software consists of C code bound to Python through complex Swig interfaces, and while this is a perfectly valid way to access faster libraries, it can often be deceptively difficult to work with as many of the resulting Python objects contain read-only data structures masquerading as dynamic Python objects.
-This sometimes yields unpredictable behavior and it was decided this should be avoided as a way to interface with CUDA C.
-PyCuda is a very elegant solution to this problem as it seamlessly bridges this gap using the native GPUArray class, and a set of functions that create GPUArray copies of NumPy arrays that are built initially on the node CPU hosts.
-With a smaller group of developers, the built-in GPUArray class contained within PyCuda does not yet support \textit{all} of the rich spectrum of high level functions available within the NDArray NumPy class.
-Perhaps more importantly however, PyCuda allows for custom CUDA kernel functions to be written as pure CUDA C directly within special containers called source modules, that are contained within the main body of the Python code.
-PyCuda actually calls the real Nvidia compiler at runtime and links the resulting temporary executable to the main program.
-One may call these functions with NumPy arrays as arguments, as if they were normal Python functions.
-This is perhaps the most important feature of PyCuda as it allows the developer to tune the customization of the code to whatever level is desired.
-Those intermediate operations that may be handled with the built in methods of the GPUArray are performed as such.
-More complex operations that are not supported can be custom built with source modules.
-If desired, the entire program may consist of CUDA C with no more then basic startup routines written in actual Python meant to instantiate the original data and pass it down to the GPU.
-This particular work used mostly custom kernels, for the simple reason that applying the same matrix operation to a contiguous block of memory containing multiple slightly different copies constitutes tensor operations similar in nature to the results of section \textbf{SECTION}, for which there is no general support for in PyCuda.
-
-\subsection{General Design Considerations}
-Computing the time marginalized likelihood for a single sample involves processing the likelihood for each point in a signal time series, multiplied again by the number of associated harmonic modes.
-As the number of samples selected by the integrator as an adaptation chunk increases, the global memory required for the program to execute becomes dominated by these time series.
-Consequently the code is built upon managing and manipulating a large contiguous block of memory whose rows contain the harmonic mode time series for each sample, this lends itself naturally to using a two dimensional grid of one dimensional blocks as the launch configuration for most kernel functions.
-The one dimensional blocks lie conceptually along the rows and may line up if the number of time series samples exceeds the maximum threds per block.
-The grid extends downwards in the $y$-direction and will generally assign a row of blocks to each sample, or in some cases blanket the memory entirely in threads.
-It is the nature of CUDA programs that thread blocks operate most efficiently when the total number of threads per block is a power of two.
-Aside from the microprocessor architecture optimizing for these cases, memory reads by groups of thirty two threads (known as a \textit{Warp}) are coalesced by the compiler and recieve the value to be read as a broadcast.
-This massively boosts throughput if used correctly and it is thus advantageous in most cases to cast the problem dimensions into some workable multiple of the maximum threads per block on the hardware.
-For most Nvidia cards this is 1024.
-It is substantially easier to pad the time series with zeros out to the next greatest multiple of this number then to design specialized cleanup blocks whose sole purpose is to process the remainder.
-In the worse case this corresponds to holding \textbf{NUMBER} extra zeros in memory, which can be a painful amount of unecessary overhead in some scenarios.
-These factors come together to make memory management the main bottleneck for the entire algorithm, and a main area of focus for further development of this method.
-
-\subsection{Spherical Harmonics}
-
-The spin-weighted spherical harmonics are precomputed at startup.
-There is no analytic formula that returns the spin-weighted spherical harmonic for arbitrary quantum numbers $l, m$ and $s$.
-The most efficient way of retrieving them closely follows the method used by the current, serial LAL functions, which implement a lookup table that contains the explicit formulas.
-This is often sufficient as most waveform templates require only the $l=2$ modes, and thus only the corresponding spherical harmonics.
-Indeed, the LSC has yet to require support for $s \neq 2$, so the corresponding CUDA function does not attempt to extend this capability.
-
-\begin{wrapfig}
-\hspace{-1cm}
-\includegraphics[trim={0cm, 2cm, 2cm, 2cm}, clip]{spherical_harmonics.eps}
-\caption{Computation of Spherical Harmonics, $l=2$.}
-\label{fig:spharms}
-\end{wrapfig}
-
-Instead, it takes as argument an ordered list of $m$ values and executes the following set of steps to return a row-major ordered block of results whose rows correspond to values of $m$: Launch a one dimensional grid of one dimensional blocks, with one thread per sample.
-Compute the answer for this value of $m$ and place the result in global memory.
-Increment the local thread index by the number of samples and repeat.
-As long as it loops through the $m$ values in order, the results will be placed correctly, as per figure \ref{fig:spharms}.
-As the extrinsic parameter values used in computation are accessed relatively few times, this function does not use shared memory.
-Regardless as a standalone function it is able to compute several billion results per second, so little further optimization is necessary.
-
-\clearpage
-
-\subsection{Complex Antenna Factor}
-The antenna factor requires somewhat less fancy footwork then the spherical harmonics, as there is only one per sample and not an arbirary multiple depending on the number of selected modes (which are not necessarily all the values between $-l$ and $l$.
-The position vectors corresponding to the extrinsic parameters are contracted with the detector response tensor in a thread-wise fashion: one thread for each sample.
-We avoid using BLAS based routines to perform the contraction due to the small size of the matrices involved, startup routines for BLAS might take longer then a simple loop over nine elements of the detector tensor.
-We again use a one dimensional grid of one dimensional blocks, however since the position vectors contain commonly accessed elements, we choose to store these independently in device shared memory as well as the detector tensor itself in even higher speed device constant memory.
-This makes the antenna factor computation one of the most efficient steps of the process.
-
-\subsection{Main Routine}
-
-Due to the variation in extrinsic parameters over the samples, it is necessary for each to manipulate a separate group of copies of the harmonic mode time series.
-The number of samples may be large, so the main block of memory intended to hold these time series is handed down to the GPU initialized to zeros and a GPU function is used to expanded the time series into a vertial stack of copies, per figure \ref{fig:expandrho}.
-To streamline the marginalization process this block of memory is padded with zeros out to the nearest multiple of $T_{max}$, which is generally either 512 or 1024 depending on the GPU architecture.
-
-\begin{wrapfig}
-\vspace{10mm}
-\includegraphics[trim={0cm, 13cm, 0cm, 3.0cm}, clip, scale=0.65]{expand_rhots.eps}
-\label{fig:expandrho}
-\caption{Expansion of time series. One thread per relevant item in resulting memory block. Zeros are padded out to the nearest multiple of $T_{max}$.
-\end{wrapfig}
-
-\clearpage
-
-The result of this expansion of the time series into GPU memory is the desired vertical stack of time series copies.
-
-The next step is to transform these time series into term one of equation \textbf{EQUATION}.
-This involves combining the antenna factor (for which there is one complex number per sample) with the spin-weighted spherical harmonics (for which there are $N_m$ complex numbers per sample) into a single quantity and inserting them into the correct positions within the time series block.
-It is more efficient to complete the combination of the antenna factor and spherical harmonics prior to multiplying them into the time series, and this action is easily completed in an elementwise fasion using the built in machinery of PyCuda.
-It is steps like these for which PyCuda is an essential tool for performing what would otherwise be complex GPU operations with a single line of code.
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\hspace{-5cm}
-\vspace{-1cm}
-  \begin{center}
-    \includegraphics[trim={4cm, 13cm, 2cm, 3cm}, clip, width=0.75\textwidth]{FYrho.eps}
-  \end{center}
-  \caption{Insertion of the antenna factor-spherical harmonic pairs. }
-\end{wrapfigure}
-
-This step coupled with the previous is all that is necessary to build the first term of the likelihood for all of the samples on the GPU, the rest of the work is simply a series of summing operations.
-These must be handled somewhat carefully due to the possibility of race conditions and queued memory accesses, both of which can hinder or dangerously skew the results of a computation.
-The main issue stems from the fact that there is no guarantee of the order of execution of threadblocks within a program.
-This means that any program that depends upon memory reads and writes occuring in a specific order is prone to producting incorrect results.
-For this reason it is important to design the summation process in a way that does not exhibit this behavior: each thread should only read and write to memory locations unique to it.
-Note that each of the functions described thus far follows this paradigm.
-Queued memory are serialized by the CUDA compiler.
-This means that memory reads to the same location by multiple threads take place one by one.
-While not as dangerous as a race condition, it is a nontrivial performance consideration.
-Note that again the process thus far has been designed to avoid these situations.
-This is particularly important for justifying the nature of the next series of steps.
-
-\clearpage
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\vspace{-0.78cm}
-\begin{center}
-\includegraphics[trim={1.67cm, 10cm, 0, 3cm}, clip, scale=0.5]{sumrhots.eps}
-\caption{Downwards summation of the harmonic modes to form term one of the factored likelihood for a group of samples.}
-\end{center}
-\end{wrapfigure}
-
-The time series are summed downwards within memory by single threads.
-One row of threads is launched per sample, this leaves $N_m$ values for each thread to collect in a summation that resides in the bottom row.
-The memory locations accessed and written to by the individual threads in the grid are unique to that thread, avoiding races and queued memory accesses.
-The result are rows that are the same as those that would have been formed taking the product $\mathbf{Q}(F\vec{Y})$, this is the first term of the factored likelihood.
-It remains only to build the second term and subtract it from the rows of this matrix to complete the computation.
-
-\subsection{Marginalization Over Time}
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\begin{center}
-\includegraphics[scale=0.7]{onedmarg.eps}
-\end{center}
-\end{wrapfigure}
-
-%---------------------------------------------------------------------------------------
-
-\end{document}

### 2016-08-26 Gradschool/thesis/final/chap4.tex
added MIT template

--- 2016-08-25 Gradschool/thesis/msthesis.tex
@@
-\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
-\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
-\usepackage[adobe-utopia]{mathdesign}
-%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
-\usepackage[english]{babel} % English language/hyphenation
-\usepackage{amsmath,amsfonts,amsthm} % Math packages
-\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
-\usepackage{sectsty} % Allows customizing section commands
-
-\usepackage{braket}
-\usepackage{cancel}
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage[mathscr]{euscript}
-
-\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps
-
-\usepackage{fancyhdr} % Custom headers and footers
-\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
-\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
-\fancyfoot[L]{} % Empty left footer
-\fancyfoot[C]{} % Empty center footer
-\fancyfoot[R]{\thepage} % Page numbering for right footer
-\renewcommand{\headrulewidth}{0pt} % Remove header underlines
-\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
-\setlength{\headheight}{13.6pt} % Customize the height of the header
-
-\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-
-\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text
-
-%----------------------------------------------------------------------------------------
-%	TITLE SECTION
-%----------------------------------------------------------------------------------------
-
-\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height
-
-\title{	
-\normalfont \normalsize 
-\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
-\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
-\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
-}
-
-\author{Brandon B. Miller, Masters Thesis} % Your name
-
-\date{\normalsize\today} % Today's date or a custom date
-
-\begin{document}
-
-\maketitle % Print the title
-\newpage
-\section{Background and Motivation}
-
-Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
-The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.
-
-\subsection{Pieces of the Einstein Equation}
-\subsubsection{Reimann Tensor}
-The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
-The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.
-
-\begin{figure}[h!]
-\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
-\caption{A visual depiction of the Reimann Curvature tensor.
-Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors.
-The orange vector $\vec{d}$ points between the tips.
-The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
-\label{fig:sphere}
-\end{figure}
-
-The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space.
-Here we motivate the concept with a visual explanation of parallel transport.
-Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}.
-If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$.
-This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$.
-Let this vector be denoted $\vec{d}$.
-Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector.
-From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
-Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above.
-Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop.
-It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question.
-More remarkable however is the simplicity of the formal definition within this picture:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
-\end{align}
-
-It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
-\end{align}
-
-The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
-\end{align}
-
-If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
-It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
-
-\begin{align}
-\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
-\end{align}
-
-\section{Linearized Gravity}
-
-In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation
-
-\begin{align}
-g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
-\end{align}
-
-Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources.
-The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \textbf{EQUATION}, we would have
-
-\begin{align}
-g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
-&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
-&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
-\end{align}
-
-The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity.
-We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
-
-\begin{align}
-h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
-\end{align}
-
-The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
-&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
-\end{align}
-
-Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
-Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero.
-This leaves us with three:
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
-&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
-\end{align}
-
-Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself.
-It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate.
-The same assumption allows for simplification of the Reimann and Ricci tensors:
-
-\begin{align}
-R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
-R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
-\end{align}
-
-Plugging in equation \textbf{EQUATION} we have
-
-\begin{align}
-R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
-&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
-&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
-\end{align}
-
-The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$.
-The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$.
-The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:
-
-\begin{align}
-R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
-\end{align}
-
-Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is
-
-\begin{align}
-R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
-\end{align}
-
-Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity
-
-\begin{align}
-V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
-\end{align}
-
-With which the first two terms of the linearized Ricci tensor can be reorganized as
-
-\begin{align}
-\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
-&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
-\end{align}
-
-Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields
-
-\begin{align}
-R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
-\end{align}
-
-The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with
-
-\begin{align}
-R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
-\end{align}
-
-Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
-
-\begin{align}
-\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
-\end{align}
-
-$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics.
-However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner.
-Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e.
-$dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed.
-Thus we have
-
-\begin{align}
-h_{\mu \nu} = h_{\nu \mu}
-\end{align}
-
-Which, for a sixteen component tensor, makes six of the components redundant.
-This leaves us with ten.
-Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation.
-This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively.
-It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form
-
-\begin{align}
-\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
-V &\rightarrow V -\frac{\partial \Psi}{\partial t}
-\end{align}
-
-The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
-Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
-
-\begin{align}
-\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
-\rightarrow \partial^{\mu}A_{\mu} &= 0
-\end{align}
-
-It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden!
-While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
-It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}.
-We do not have to state \textit{what it is}, as long as we know that it exists.
-A proof that there legitimately is such a gauge function exists but is out of the scope of this outline.
-There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
-In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
-
-\begin{align}
-x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
-\end{align}
-
-In particular, we can show that there exists a coordinate transformation such that
-
-\begin{align}
-\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
-\end{align}
-
-\textbf{PUT PROOF HERE}
-
-\begin{align}
-\text{PUT PROOF HERE}
-\end{align}
-
-Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply
-
-\begin{align}\label{eq:flteinst}
-\Box h_{\mu \nu} = 0
-\end{align}
-
-This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
-Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}.
-In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
-This leaves us with six.
-
-In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read
-
-\begin{align}
-\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
--\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
-\end{align}
-
-This is the general form of the time dependent wave equation whose solutions take the form
-
-\begin{align}
-f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
-\end{align}
-
-Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector.
-Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have
-
-\begin{align}
-\text{PUT EQ HERE}
-\end{align}
-
-\section{Bayesian Statistics}
-We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
-Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
-Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
-We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$.
-A perpandicular cross section would simply be $P(y|x)$.
-We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$.
-Thus we have that
-
-\begin{figure}[h!]
-\center
-\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
-\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
-\label{fig:sphere}
-\end{figure}
-
-\begin{align} \label{eq:bayes}
-P(x|y)P(y) &= P(y|x)P(x) \\
-P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
-\end{align}
-
-Which is Bayes' theorem.
-The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome.
-Reading $\neg x$ as "not x", we have
-
-\begin{align}\label{eq:continuous}
-P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
-&= \sum_{k=1}^K P(y|x_k)P(x_k) 
-\end{align}
-
-\clearpage 
-This allows us to rewrite the posterior probability as
-
-\begin{align}
-P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
-\end{align}
-
-Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has
-
-\begin{align} \label{eq:lratio}
-P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
-\end{align}
-
-Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression.
-The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur.
-In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio.
-This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.
-
-\subsection{Signal Processing}
-Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:
-
-\begin{align}
-P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
-\end{align}
-
-If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as
-
-\begin{align}
-P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
-\end{align}
-
-For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
-This is
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
-\end{align}
-
-In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$.
-If there is no signal present in the data, then the data and noise are equal, meaning
-
-\begin{align}
-d(t) &= n(t) \\
-P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
-\end{align}
-
-Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:
-
-\begin{align}
-d(t) &= n(t) + h(t) \\
-p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
-\end{align}
-
-So the likelihood ratio for gravitational wave signal processing is exactly equal to
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
-\end{align}
-
-By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
-
-\begin{align}\label{eq:parampos}
-P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
-\end{align}
-
-Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator
-
-\begin{align}
-P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
-\end{align}
-
-Where $\Omega$ indicates integration over the entire probabilistic state space.
-Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us
-
-\begin{align}\label{eq:bayeslikely}
-P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
-\end{align}
-
-Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters.
-For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source.
-We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively.
-The denominator of equation \ref{eq:bayeslikely} is thus factorizable as
-
-\begin{align}
-\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta}
-\end{align}
-
-Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
-Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
-These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm.
-Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
-A new and more embarassingly parallel scheme is needed.
-
-\subsection{Harmonic Mode Decomposition}
-
-For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:
-
-\begin{align}
-h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
-\end{align}
-
-Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
-
-\begin{align}
-\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
-\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
-&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
-\end{align}
-
-Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,
-
-\begin{align}
-\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
-&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
-&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h^*_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y^*_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
-\end{align}
-
-Now, for any complex number $z = a + ib$, we have that $z*z + \Re(z^2) = 2z*z$, so $z*z = \frac{1}{2}(z*z + \Re z^2)$. Then the double sum over $l,m$ pairs becomes
-
-The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters.
-The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
-Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.
-
-\section{Implementation}
-The main weight of the computational work described above falls under an expensive Monte Carlo integration of the evidence as part of the Bayesian posterior.
-Zooming in it quickly becomes apparent that the main challenge to be overcome is efficient evaluation of the function to be integrated, in this case the factored likelihood.
-Prior to this work, the factored likelihood is was computed with a series of looping structures that individually compiled all the terms of the equation in an independent and serial manner.
-The process took substantial advantage of many high level operations available within Python to match up the correct terms across various harmonic mode time series, spherical harmonics, and antenna patterns.
-Thanks to clever optimization and gratuitous use of fast numerical Python libraries such as NumPy, the code was able to compute approximately $10^3$ likelihood evaluations on the order of seconds, the limiting factor being the serial nature of loops and the inherent lethargy of high level languages such as Python.
-Although NumPy commonly passes target data to compiled routines written in faster languages such as C, some portions of the computation (in particular the marginalization over time) benefit little from this capability and form a bottleneck for efficient evaluation of the likelihood.
-Furthermore the original implementation heavily relied upon older structures within the existing LIGO Algorithms Library (LAL), a set of C routines bound to Python through Swig, to perform many intermediate calculations, such as computing spherical harmonics.
-The goal of the first phase of this work was thorough vectorization of the process used to build the terms of the likelihood utilizing the BLAS subroutines available through NumPy as often as possible, as well as reducing dependency on LAL.
-To that end, the components of the factored likelihod were reworked.
-
-\subsection{Complex Antenna Factor}
-Gravitational wave interferometers by nature possess nonisotropic sensitivity patterns that are a function of the specific geometry and orientation of the detector.
-As a result any incoming signal is modulated not only by directional effects \textbf{RA, DEC} relative to the celestial coordinate grid, but also temporal effects that are dependent on the orientation of the earth at the signal arrival time.
-Following closely the treatment from \textbf{CITATION} we note that as described in section \textbf{SECTION}, there are six degrees of freedom carried by gravitational waves.
-Four of these may be eliminated by a change of basis leaving us with only two spatial tensor polarizations $\epsilon^k_{ij}$ which can be expressed as
-
-\begin{align}
-\epsilon^+_{ij} = 
-\begin{bmatrix}
-1 & 0 & 0 \\
-0 & -1 & 0 \\
-0 & 0 & 0
-\end{bmatrix}
-, \ \ e^{\times}_{ij} =
-\begin{bmatrix}
-0 & 1 & 0 \\
-1 & 0 & 0 \\
-0 & 0 & 0
-\end{bmatrix}
-\end{align}
-
-These matrices are geometrically consistent with the orthogonal warping of ring like structures of test particles that are commonly used to visualize the effect of a passing gravitational wave: $\epsilon^+_{i,j}$ tends to grow vectors in the direction of the unit basis vectors whereas $\epsilon^{\times}_{ij}$ tends to shear them at an angle $\frac{\pi}{4}$ offset from the same basis.
-Moreover, there is an additional effect from the polarization angle $\psi$ of the incoming wave relative to the detector basis.
-Manifestation of a gravitational wave as measurable, real valued detector strain $h(t)$ is given by a the contraction of an object known as the detector tensor with these polarization tensors.
-
-\begin{align}
-\mathbf{R^{ij}} = \frac{1}{2}(v^i v^j - u^i u^j)
-\end{align}
-
-With the various polarization tensors to form the linear combination
-
-\begin{align}
-h(t) = h_{+}(t)R^{ij}_e^+_{ij} + h_{\times}(t)R^{ij}e^{\times}_{ij}
-\end{align}
-
-The products $R^{ij}_e^+_{ij}$ and $R^{ij}e^{\times}_{ij}$ are known as the \textit{antenna factors}, $F_{+}$ and $F_{\times}$.
-As with the gravitational wave strain itself, we represent these as a single complex number $F_{+} - i F_{\times}$, in keeping with the formalism of section \textbf{SECTION}.
-It can be shown that in the wave basis, the scalar in question can be expressed as
-
-\begin{align}
-F_+ - i F_{\times} &= \vec{X}^T\mathbf{R}\vec{X} - \vec{Y}^T\mathbf{R}\vec{Y} - i(\vec{X}^T\mathbf{R}\vec{Y} + \vec{Y}^T\mathbf{R}\vec{X})
-\end{align}
-
-Where $\vec{X}$ and $\vec{Y}$ are the wave-frame axes. These can be redefined in terms of the working parameters as
-
-\begin{align}
-\text{PUT DEFINITIONS HERE}
-\end{align}
-
-Our objective is to produce this quantity for a group of samples, all at once, as a coherent vectorized operation.
-One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
-Thus we need a function that takes as input a \textit{list} vectors (or \textit{vector}) of vectors and produces a vector with the right components as output.
-To that end we define the tensor $X^{i}_{j}$ where
-
-\begin{align}
-X^i &= 
-\begin{bmatrix}
-X^{i}_0 \\
-X^{i}_1 \\
-X^{i}_2
-\end{bmatrix}
-\end{align}
-
-As well as the tensor $R^{i}_{jk}$ where
-
-\begin{align}
-R^{i} &= 
-\begin{bmatrix}
-R^{i}_{00} & R^{i}_{01} & R^{i}_{02} \\ 
-R^{i}_{10} & R^{i}_{11} & R^{i}_{12} \\ 
-R^{i}_{20} & R^{i}_{21} & R^{i}_{22}  
-\end{bmatrix}
-\end{align}
-
-The tensor $X$ is like a stack of all the different possible $\vec{X}$ coming out of the page.
-The Tensor $R$ is like $n$ copies of the matrix $\mathbf{R}$ stacked on top of each other.
-In this way the desired vector is obtainable with the tensor contraction
-
-\begin{align}
-F^{i}_{+} &= X^{lm}R^{i}_{lj}X^{j}_{m} - Y^{lm}R^{i}_{lj}Y^{j}_{m} \\ 
-F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
-\end{align}
-
-\subsection{Vectorized Single Detector Log Likelihood}
-
-With the spherical harmonics and antenna factor in hand the following set of operations yield the factored likelihood for a single detector, the network likelihood being a simple sum over all of the detectors.
-
-Equation 24 from Arxiv $15502.05370v1.pdf$ for the network log likelihood reads
-
-\begin{align}
-\ln{\mathcal{L}} = &\frac{D_{ref}}{D}Re \sum_{k}\sum_{(l,m)}\left[F_k Y_{lm}\right]^{*}Q_{k,lm} \\ 
-& - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] \\
- & - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right]
-\end{align}
-
-Consider a single detector, thus dropping the sum over $k$. The first term is of the form $\vec{A}\cdot\vec{B} = \sum_{i=0}^{d}A_iB_i$, so if $Q_{k,(l,m)}$ were a simple vector, we could write it as
-
-\begin{equation}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2} F*\vec{Y}\cdot\vec{Q}
-\end{equation}
-
-However the $Q_{k,(l,m)}$ are actually harmonic mode time series and not single values. We desire a vector whose values are the likelihoods at each point in the time series.
-
-Consider the case where we have only the $(2,-2), (2,0)$ and $(2,2)$ modes.
-If we write all the mode time series $Q^0, Q^1, Q^2...$ as the columns of a matrix , then the desired result is obtained with
-
-\begin{align}
-F*
-\begin{bmatrix}
-Q^0_{2,-2} & Q^0_{2,+0} & Q^0_{2,+2} \\
-Q^1_{2,-2} & Q^1_{2,+0} & Q^1_{2,+2} \\ 
-Q^2_{2,-2} & Q^2_{2,+0} & Q^2_{2,+2} \\
-\vdots & \vdots & \vdots
-\end{bmatrix}
-\begin{bmatrix}
-\left(Y_{2,-2}\right)\\
-\hspace{0mm} \\
-\left(Y_{2,+0}\right) \\
-\hspace{0mm} \\
-\left(Y_{2,+2}\right) \\
-\end{bmatrix}
-=
-\begin{bmatrix}
-Q^0_{2,-2} Y_{2,-2} + Q^0_{2,-2}Y_{2,+0} + Q^0_{2,+2}Y_{2,+2} \\
-Q^1_{2,-2} Y_{2,-2} + Q^1_{2,-2}Y_{2,+0} + Q^1_{2,+2}Y_{2,+2} \\
-Q^2_{2,-2} Y_{2,-2} + Q^2_{2,-2}Y_{2,+0} + Q^2_{2,+2}Y_{2,+2} \\
-\vdots
-\end{bmatrix}
-\end{align}
-
-With $\vec{Y}$ and $\mathbf{Q}$ defined as the matrix and vector above respectively, we have for the first term
-
-\begin{equation}
-\frac{D_{ref}}{D}Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right]
-\end{equation}
-
-The second term is a sum once over all the possible combinations of $(l,m), (l',m')$ pairs using the $U_{(l,m),(l',m')}$ cross terms. Its result is a scalar quantity made up of terms like
-
-\begin{align}
-&Y_{2,-2}^{*}Y_{2,-2}U_{(2,-2),(2,-2)} + Y_{2,-2}^{*}Y_{2,+0}U_{(2,-2),(2,+0)} + Y_{2,-2}^{*}Y_{2,+2}U_{(2,-2),(2,+2)} \\ 
- +  \ &Y_{2,+0}^{*}Y_{2,-2}U_{(2,+0),(2,-2)} + Y_{2,+0}^{*}Y_{2,+0}U_{(2,+0),(2,+0)} + Y_{2,+0}^{*}Y_{2,+2}U_{(2,+0),(2,+2)} \\ 
-+ \  &Y_{2,+2}^{*}Y_{2,-2}U_{(2,+2),(2,-2)} + Y_{2,+2}^{*}Y_{2,+0}U_{(2,+2),(2,+0)} + Y_{2,+2}^{*}Y_{2,+2}U_{(2,+2),(2,+2)}
-\end{align}
-
-If we pack the $U_{(l,m),(l',m')}$ into the matrix $\mathbf{U}$ as defined below then the following set of matrix operations produces the same sum
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} &  U_{(2,-2),(2,+0)} &  U_{(2,-2),(2,+2)} \\
-U_{(2,+0),(2,-2)} &  U_{(2,+0),(2,+0)} &  U_{(2,+0),(2,+2)} \\
-U_{(2,+2),(2,-2)} &  U_{(2,+2),(2,+0)} &  U_{(2,+2),(2,+2)}
-\end{bmatrix}
-\begin{bmatrix}
-Y_{2,-2} \\
-Y_{2,+0} \\
-Y_{2,+2}
-\end{bmatrix}
-\end{align*}
-
-because when you multiply $\mathbf{U}$ into $\vec{Y}$ this simplifies to
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} 
-\end{bmatrix}
-\end{align*}
-
-Which becomes the desired scalar. This allows us to write the second term as
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] =  - \left[\frac{D_{ref}}{2D}\right]^{2} |F^2|\vec{Y}^{*}\mathbf{U}\vec{Y}
-\end{align}
-
-We must always set up the spherical harmonic vectors based on the value of $m$ and the cross terms in row-major form based first on $m_2$ and then on $m_1$.
-If we organize the matrix $\mathbf{V}$ in the same way then the same set of steps will lead us to conclude that
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2}Re \left[F^2 \vec{Y}\mathbf{V}\vec{Y} \right]
-\end{align}
-
-Combining the results the single detector log likelihood is
-
-\begin{align}
-\ln{\mathcal{L}} = \frac{D_{ref}}{D}\Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right] - \left[\frac{D_{ref}}{2D}\right]^{2}\left[|F|^2 \vec{Y}^{*}\mathbf{U}\vec{Y} - \Re\left(F^2 \vec{Y}\mathbf{V}\vec{Y}\right) \right]
-\end{align}
-
-\subsection{GPU Implementation}
-Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures.
-This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency.
-Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level.
-A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.
-Reaching the limit of the sample throughput with conventional serial python libraries, our investigation turned to GPU based acceleration as the only means forward.
-The following is the result of this investigation.
-
-\subsection{Hardware Limitations and Logical Programming}
-GPUs work significantly differently from CPUs and must be paired with one to operate, they serve only as accelerators for complex operations and not as standalone units.
-Data must be passed back and forth from host memory on the node to RAM placed directly on the card itself through expensive transfers via the PCI bus.
-This is known as device global memory.
-While the onboard RAM on bleeding edge cards exceeds 12GB, memory transfers are the dominant cost of many GPU programs and must be minimized.
-Once the data exists on the device, it is accessed and manipulated by anywhere between nine and fifteen logical cores called Streaming Multiprocessors.
-These cores operate concurrently and contain a smaller granularity of parallel processing units that \textit{may} operate concurrently depending on the individual resources required by the requested operations.
-Programming in CUDA involves launching groups of threads called \textit{grids} that are further subdivided into \textit{blocks}.
-The blocks themselves are indexed by local variables that are rapidly accessible by all the contained threads, and the threads themselves all have access to indices placed within registers that exist for the lifetime of the thread in question.
-Both the blocks and and grids can be up to three dimensional and thus each block and thread may have access to up to three registers containing block indices and three containing thread indices.
-In addition thread blocks have access to an extremely high speed shared memory space where ideally most of the computations are performed, however it is limited in size to \textbf{SIZE} and is not designed to hold the entire target dataset.
-One major design goal for CUDA programs is implementing a process which copies portions of memory from the global space to the shared, processes it, and copies it back, while using the thread and block indices to blanket the dataset with threads and control which threads access what data.
-The current CUDA API allows for the logical launch of \textbf{HOW MANY} blocks with a maximum of 1024 threads per block, for a total of \textit{HOW MANY} possible "concurrent" threads.
-It is important to note that these threads are not \textit{physically} concurrent.
-Indeed, if each of the logical threads containing six integers in registers were to exist simultaneously, the GPU would be storing \textbf{HOW MANY} registers at once!
-Similarly perposterous amounts of shared memory would have to exist.
-In actuality, the shared memory and registers are located on the logical cores themselves, and blocks thread blocks line up to be processed.
-How many are processed at once is heavily dependent on the resources requested by the blocks themselves.
-Defining too many local variables in a kernel function can lead to \textit{register pressure} and decreased \textit{occupancy} of the multiprocessors on the GPU, a similar effect can come from requesting too much shared memory.
-Furthermore, the blocks are not guaranteed to execute in any particular order.
-While the compiler is responsible for sorting out how many blocks can be processed at a time as well as the order, the programmer is responsible for performance tuning with regard to optimizing block occupancy.
-This is often a late stage design consideration, but also leads to some of the greatest returns after algorithmic optimization has been completed.
-
-\subsection{PyCuda} 
-PyCuda is a flexible library for GPU scripting that includes a broad spectrum of GPU based array operations built to mimic the usage of NumPy arrays within a program.
-It allows for relatively easy access to the Nvidia CUDA API directly from within Python, and many of its operations are directly compatible with NumPy arrays with careful casting to the correct data types.
-Certainly one of the most challenging aspects of incorporating GPU code into a higher level language such as Python is the interface layer.
-As it stands much of the widely used LAL software consists of C code bound to Python through complex Swig interfaces, and while this is a perfectly valid way to access faster libraries, it can often be deceptively difficult to work with as many of the resulting Python objects contain read-only data structures masquerading as dynamic Python objects.
-This sometimes yields unpredictable behavior and it was decided this should be avoided as a way to interface with CUDA C.
-PyCuda is a very elegant solution to this problem as it seamlessly bridges this gap using the native GPUArray class, and a set of functions that create GPUArray copies of NumPy arrays that are built initially on the node CPU hosts.
-With a smaller group of developers, the built-in GPUArray class contained within PyCuda does not yet support \textit{all} of the rich spectrum of high level functions available within the NDArray NumPy class.
-Perhaps more importantly however, PyCuda allows for custom CUDA kernel functions to be written as pure CUDA C directly within special containers called source modules, that are contained within the main body of the Python code.
-PyCuda actually calls the real Nvidia compiler at runtime and links the resulting temporary executable to the main program.
-One may call these functions with NumPy arrays as arguments, as if they were normal Python functions.
-This is perhaps the most important feature of PyCuda as it allows the developer to tune the customization of the code to whatever level is desired.
-Those intermediate operations that may be handled with the built in methods of the GPUArray are performed as such.
-More complex operations that are not supported can be custom built with source modules.
-If desired, the entire program may consist of CUDA C with no more then basic startup routines written in actual Python meant to instantiate the original data and pass it down to the GPU.
-This particular work used mostly custom kernels, for the simple reason that applying the same matrix operation to a contiguous block of memory containing multiple slightly different copies constitutes tensor operations similar in nature to the results of section \textbf{SECTION}, for which there is no general support for in PyCuda.
-
-\subsection{General Design Considerations}
-Computing the time marginalized likelihood for a single sample involves processing the likelihood for each point in a signal time series, multiplied again by the number of associated harmonic modes.
-As the number of samples selected by the integrator as an adaptation chunk increases, the global memory required for the program to execute becomes dominated by these time series.
-Consequently the code is built upon managing and manipulating a large contiguous block of memory whose rows contain the harmonic mode time series for each sample, this lends itself naturally to using a two dimensional grid of one dimensional blocks as the launch configuration for most kernel functions.
-The one dimensional blocks lie conceptually along the rows and may line up if the number of time series samples exceeds the maximum threds per block.
-The grid extends downwards in the $y$-direction and will generally assign a row of blocks to each sample, or in some cases blanket the memory entirely in threads.
-It is the nature of CUDA programs that thread blocks operate most efficiently when the total number of threads per block is a power of two.
-Aside from the microprocessor architecture optimizing for these cases, memory reads by groups of thirty two threads (known as a \textit{Warp}) are coalesced by the compiler and recieve the value to be read as a broadcast.
-This massively boosts throughput if used correctly and it is thus advantageous in most cases to cast the problem dimensions into some workable multiple of the maximum threads per block on the hardware.
-For most Nvidia cards this is 1024.
-It is substantially easier to pad the time series with zeros out to the next greatest multiple of this number then to design specialized cleanup blocks whose sole purpose is to process the remainder.
-In the worse case this corresponds to holding \textbf{NUMBER} extra zeros in memory, which can be a painful amount of unecessary overhead in some scenarios.
-These factors come together to make memory management the main bottleneck for the entire algorithm, and a main area of focus for further development of this method.
-
-\subsection{Spherical Harmonics}
-
-The spin-weighted spherical harmonics are precomputed at startup.
-There is no analytic formula that returns the spin-weighted spherical harmonic for arbitrary quantum numbers $l, m$ and $s$.
-The most efficient way of retrieving them closely follows the method used by the current, serial LAL functions, which implement a lookup table that contains the explicit formulas.
-This is often sufficient as most waveform templates require only the $l=2$ modes, and thus only the corresponding spherical harmonics.
-Indeed, the LSC has yet to require support for $s \neq 2$, so the corresponding CUDA function does not attempt to extend this capability.
-
-\begin{wrapfig}
-\hspace{-1cm}
-\includegraphics[trim={0cm, 2cm, 2cm, 2cm}, clip]{spherical_harmonics.eps}
-\caption{Computation of Spherical Harmonics, $l=2$.}
-\label{fig:spharms}
-\end{wrapfig}
-
-Instead, it takes as argument an ordered list of $m$ values and executes the following set of steps to return a row-major ordered block of results whose rows correspond to values of $m$: Launch a one dimensional grid of one dimensional blocks, with one thread per sample.
-Compute the answer for this value of $m$ and place the result in global memory.
-Increment the local thread index by the number of samples and repeat.
-As long as it loops through the $m$ values in order, the results will be placed correctly, as per figure \ref{fig:spharms}.
-As the extrinsic parameter values used in computation are accessed relatively few times, this function does not use shared memory.
-Regardless as a standalone function it is able to compute several billion results per second, so little further optimization is necessary.
-
-\subsection{Complex Antenna Factor}
-The antenna factor requires somewhat less fancy footwork then the spherical harmonics, as there is only one per sample and not an arbirary multiple depending on the number of selected modes (which are not necessarily all the values between $-l$ and $l$.
-The position vectors corresponding to the extrinsic parameters are contracted with the detector response tensor in a thread-wise fashion: one thread for each sample.
-We avoid using BLAS based routines to perform the contraction due to the small size of the matrices involved, startup routines for BLAS might take longer then a simple loop over nine elements of the detector tensor.
-We again use a one dimensional grid of one dimensional blocks, however since the position vectors contain commonly accessed elements, we choose to store these independently in device shared memory as well as the detector tensor itself in even higher speed device constant memory.
-This makes the antenna factor computation one of the most efficient steps of the process.
-
-\subsection{Main Routine}
-
-Due to the variation in extrinsic parameters over the samples, it is necessary for each to manipulate a separate group of copies of the harmonic mode time series.
-The number of samples may be large, so the main block of memory intended to hold these time series is handed down to the GPU initialized to zeros and a GPU function is used to expanded the time series into a vertial stack of copies, per figure \ref{fig:expandrho}.
-To streamline the marginalization process this block of memory is padded with zeros out to the nearest multiple of $T_{max}$, which is generally either 512 or 1024 depending on the GPU architecture.
-
-\begin{wrapfig}
-\vspace{10mm}
-\includegraphics[trim={0cm, 13cm, 0cm, 3.0cm}, clip, scale=0.65]{expand_rhots.eps}
-\label{fig:expandrho}
-\caption{Expansion of time series. One thread per relevant item in resulting memory block. Zeros are padded out to the nearest multiple of $T_{max}$.
-\end{wrapfig}
-
-The result of this expansion of the time series into GPU memory is the desired vertical stack of time series copies.
-
-The next step is to transform these time series into term one of equation \textbf{EQUATION}.
-This involves combining the antenna factor (for which there is one complex number per sample) with the spin-weighted spherical harmonics (for which there are $N_m$ complex numbers per sample) into a single quantity and inserting them into the correct positions within the time series block.
-It is more efficient to complete the combination of the antenna factor and spherical harmonics prior to multiplying them into the time series, and this action is easily completed in an elementwise fasion using the built in machinery of PyCuda.
-It is steps like these for which PyCuda is an essential tool for performing what would otherwise be complex GPU operations with a single line of code.
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\hspace{-5cm}
-\vspace{-1cm}
-  \begin{center}
-    \includegraphics[trim={4cm, 13cm, 2cm, 3cm}, clip, width=0.75\textwidth]{FYrho.eps}
-  \end{center}
-  \caption{Insertion of the antenna factor-spherical harmonic pairs. }
-\end{wrapfigure}
-
-This step coupled with the previous is all that is necessary to build the first term of the likelihood for all of the samples on the GPU, the rest of the work is simply a series of summing operations.
-These must be handled somewhat carefully due to the possibility of race conditions and queued memory accesses, both of which can hinder or dangerously skew the results of a computation.
-The main issue stems from the fact that there is no guarantee of the order of execution of threadblocks within a program.
-This means that any program that depends upon memory reads and writes occuring in a specific order is prone to producting incorrect results.
-For this reason it is important to design the summation process in a way that does not exhibit this behavior: each thread should only read and write to memory locations unique to it.
-Note that each of the functions described thus far follows this paradigm.
-Queued memory are serialized by the CUDA compiler.
-This means that memory reads to the same location by multiple threads take place one by one.
-While not as dangerous as a race condition, it is a nontrivial performance consideration.
-Note that again the process thus far has been designed to avoid these situations.
-This is particularly important for justifying the nature of the next series of steps.
-
-\clearpage
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\vspace{-0.78cm}
-\begin{center}
-\includegraphics[trim={1.67cm, 10cm, 0, 3cm}, clip, scale=0.5]{sumrhots.eps}
-\caption{Downwards summation of the harmonic modes to form term one of the factored likelihood for a group of samples.}
-\end{center}
-\end{wrapfigure}
-
-The time series are summed downwards within memory by single threads.
-One row of threads is launched per sample, this leaves $N_m$ values for each thread to collect in a summation that resides in the bottom row.
-The memory locations accessed and written to by the individual threads in the grid are unique to that thread, avoiding races and queued memory accesses.
-The result are rows that are the same as those that would have been formed taking the product $\mathbf{Q}(F\vec{Y})$, this is the first term of the factored likelihood.
-It remains only to build the second term and subtract it from the rows of this matrix to complete the computation.
-
-\subsection{Marginalization Over Time}
-Having integrated over the extrinsic parameters numerically, it remains to marginalize over the entire time series to produce a single scalar representing the likelihood associated with a sample.
-While a simplistic operation in theory special attention was taken to optimize the summation process.
-This is due to the "width" of the data to be processed.
-The time series themselves can be thousands of entries long depending on the signal sample rate.
-Accelerating the preparation of the time series, as demonstrated in the previous steps, can be performed with brute force by flooding the calculation with threads, parallel reduction on a GPU requires finesse.
-In fact, it was found that this particular step was the main bottleneck in the serial code, thus it was expected that here the greatest returns would be achieved.
-The objective is to keep as many threads busy as possible to maintain throughput, however as we will demonstrate, the best that can be achieved without excessively complicated schemes is to start with $\frac{N_t}{2}$ threads and recursively cut the working group in half, a total of $\log_2(N_t)$ times, as per figure \ref{fig:recursive}.
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\label{fig:recursive}
-\caption{Sequential addressing for parallel reduction.}
-\begin{center}
-\includegraphics[trim={0, 13cm, 0, 0}, clip, scale=0.5]{onedmarg.eps}
-\end{center}
-\end{wrapfigure}
-
-\begin{wrapfigure}{h}{0.6\textwidth}
-\label{fig:folding}
-\caption{"Folding" of the time series block for time marginalization}
-\begin{center}
-\includegraphics[trim={0, 13cm, 0, 0}, clip, scale=0.5]{marg_fold.eps}
-\end{center}
-\end{wrapfig}
+%% This is an example first chapter.  You should put chapter/appendix that you
+%% write into a separate file, and add a line \include{yourfilename} to
+%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
+%% You can process specific files by typing their names in at the 
+%% \files=
+%% prompt when you run the file main.tex through LaTeX.
+\chapter{Results}
 
@@
 Pure gravitational wave astrophysics has exciting prospects for the future, but the possibility of advancing other classical forms of astrophysics through their interplay with these newer methods is equally compelling.
-Significant research \textbf{CITATION} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers.
+Significant research \cite{first2years} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers.
 A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration klonovae.
@@
 Gamma ray bursts (GRBs) are short duration transient events thought to occur during or shortly after the merger of compact binary systems, specifically those which include a single black hole and one neutron star.
-The presence of strong radiative emission from the poles of black holes has been confirmed by \textbf{CITATION}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs.
+The presence of strong radiative emission from the poles of black holes has been confirmed by \cite{polarjet}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs.
 The process is depicted in figure \textbf{FIGURE} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole.
@@
 It is generally accepted that to confirm this hypothesis we would need to observe direct coincidence between a GW trigger localized to some region of the sky and a GRB from the same direction.
-These phenomena have been demonstrated in numerical simulations by \textbf{CITATION} and are highly dependent on initial system parameters.
-In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors \textbf{CITATION}.
-We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
-We use rate predictions from \textbf{CITATION} to restrict the mass range to be studied to fall between \textbf{SOME RANGE}.
-These bounds corespond to a chirp mass range of \textbf{WHAT RANGE}.
-From this we can determine the expected time that the signal will spend in the Advanced LIGO band.
-In figure \textbf{FIGURE}, we have visualized this range in \textbf{SOME PLANE}.
-Using the results of our code we find that anadditional \textbf{SOME PERCENTAGE} of expected NSBH sytems expected to be picked up by Advanced LIGO are now candidates for joint GW-EM astronomy.
+These phenomena have been demonstrated in numerical simulations by \cite{kilonova} and are highly dependent on initial system parameters.
+In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors IIIIIII We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
+The frequency evolution of a compact binary system is given by the well known relationship
+
+\begin{align}
+\dot{f_{gw}}(\tau) = \frac{96}{5}\pi^{\frac{8}{3}} \left(\frac{G \mathcal{M}}{c^3}\right)^{\frac{5}{3}}f_{gw}^{\frac{11}{3}}
+\end{align}
+
+Integrating this equation we can find the total time during which a signal from such a binary remains in the Advanced LIGO frequency band:
+
+\begin{align}\label{eq:timeinband}
+\tau \approx 2.18s\left(\frac{1.21M_\odot}{\mathcal{M}}\right)^{\frac{5}{3}}\right)\left(\frac{100Hz}{f_{gw}}\right)^{\frac{8}{3}}
+\end{align}
+
+For calculational purposes the value $10Hz$ is used as $f_{gw}$ in the above equation, as this is the lower bound of reasonable sensitivity for ground based interferometers such as Advanced LIGO \textbf{CITATION}.
+
+Numerical simulations have shown that NSBH fates are strongly sensitive to the initial binary mass ratio and equation of state.
+It has been found that systems for which approximately $q \gtrsim 4$ do not exhibit tidal disruption, the NS is swallowed whole by the BH before the tidal forces can overcome the NS self gravity.
+Since the largest NS masses are predicted \textbf{CITATION} to be approximately $1.4M_\odot$, the BH can be no larger then $\approx 5.6M_\odot$.
+This is equivalent to $\mathcal{M} \approx 2.33M_\odot$.
+Plugging this value into equation \ref{eq:timeinband} we find that such a system will spend approximately $5m 40s$ in the Advanced LIGO frequency band.
 
@@
 
-%---------------------------------------------------------------------------------------
-
-\end{document}
+\begin{figure}[H!]
+\hspace{-6cm}
+\includegraphics[scale=0.5]{EMFollow.eps}
+\end{figure}

### 2016-08-28 Gradschool/thesis/final/bak/abstract.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

% $Log: abstract.tex,v $
% Revision 1.1  93/05/14  14:56:25  starflt
% Initial revision
% 
% Revision 1.1  90/05/04  10:41:01  lwvanels
% Initial revision
% 
%
%% The text of your abstract and nothing else (other than comments) goes here.
%% It will be single-spaced and the rest of the text that is supposed to go on
%% the abstract page will be generated by the abstractpage environment.  This
%% file should be \input (not \include 'd) from cover.tex.
The September 14th, 2015 detection of gravitational waves (GW) by the Advanced LIGO detectors in Hanford, Washington and Livingston, Louisiana has galvanized the GW astronomy community and ushered in a new and unexplored method of observing astrophysical phenomena. With it comes the opportunity to couple GW signals with electromagnetic (EM) observations in a manner that they may inform each other, maximizing the science returns from both types of detections. To that end it is imperative that GW signal processing techniques be developed to such a point that they may rapidly and accurately report system parameters from possible progenitors of EM signals of interest. In particular the opportunity to rapidly localize and infer crucial system parameters of Gamma Ray Burst (GRB) and Kilonova progenitors hinges on prompt reporting of the sky location and orbital angular momentum direction of compact binary coalescences (CBC). These systems are thought to be the most likely sources of detectable GW signals in the Advanced LIGO era. This thesis presents a vectorization and subsequent GPU implementation of a high performance parameter estimation algorithm which reports a Bayesian posterior probability over the CBC parameter space, given an data time series thought to contain a GW signal. We report on the speedup of the algorithm and its role in extending the capability of RapidPE to perform low latency electromagnetic follow-up as it pertains to GRBs and Kilonova afterglows.

### 2016-08-28 Gradschool/thesis/final/bak/appc.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-26 Gradschool/thesis/final/chap1.tex
@@
-%% This is an example first chapter.  You should put chapter/appendix that you
-%% write into a separate file, and add a line \include{yourfilename} to
-%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
-%% You can process specific files by typing their names in at the 
-%% \files=
-%% prompt when you run the file main.tex through LaTeX.
-\chapter{Linearized Gravity}
-Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
-The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.
+\chapter{Formulas and Derviations}\label{Appendix C}
 
-\subsection{Pieces of the Einstein Equation}
-\subsubsection{Reimann Tensor}
-The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
-The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.
+\subsection{Christoffel Symbols}
 
-\begin{figure}[h!]
-\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
-\caption{A visual depiction of the Reimann Curvature tensor.
-Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors.
-The orange vector $\vec{d}$ points between the tips.
-The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
-\label{fig:sphere}
-\end{figure}
-
-The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space.
-Here we motivate the concept with a visual explanation of parallel transport.
-Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}.
-If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$.
-This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$.
-Let this vector be denoted $\vec{d}$.
-Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector.
-From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
-Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above.
-Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop.
-It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question.
-More remarkable however is the simplicity of the formal definition within this picture:
+The Christoffel Symbols are defined in terms of the metric tensor $g_{\mu \nu}$ as
 
 \begin{align}
-R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
+\Gamma^{\rho}_{\mu \nu} = \frac{1}{2}g^{\sigma \rho}(\partial_{\nu}g_{\mu \rho} + \partial_{\mu}g_{\nu \rho} - \partial_{\sigma}g_{\mu \nu})
 \end{align}
 
-It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel symbols defined in \nameref{Appendix C}.
+\clearpage
+\newpage
 
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
-\end{align}
-
-The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
-\end{align}
-
-If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{\rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
-It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
-
-\begin{align}
-\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
-\end{align}
-
-\section{Linearized Gravity}
-
-In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation
-
-\begin{align}\label{eq:metpluspert}
-g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
-\end{align}
-
-Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources.
-The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \ref{eq:metpluspert}, we would have
-
-\begin{align}
-g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
-&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
-&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
-\end{align}
-
-The second two terms being second order in $h$, the peturbation, are neglected in linearized gravity.
-We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
-
-\begin{align}
-h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
-\end{align}
-
-The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
-&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
-\end{align}
-
+\subsection{Details on Linearized Gravity}
 Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
@@
 The same assumption allows for simplification of the Reimann and Ricci tensors:
+
+The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
+
+\begin{align}
+R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
+\end{align}
 

### 2016-08-28 Gradschool/thesis/final/bak/chap1.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-22 Gradschool/thesis/msthesis.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Short Sectioned Assignment
-% LaTeX Template
-% Version 1.0 (5/5/12)
-%
-% This template has been downloaded from:
-% http://www.LaTeXTemplates.com
-%
-% Original author:
-% Frits Wenneker (http://www.howtotex.com)
-%
-% License:
-% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% This is an example first chapter.  You should put chapter/appendix that you
+%% write into a separate file, and add a line \include{yourfilename} to
+%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
+%% You can process specific files by typing their names in at the 
+%% \files=
+%% prompt when you run the file main.tex through LaTeX.
+\chapter{Background}
+\section{Preface}
+Gravitational wave astronomy is often called a new way of observing the universe, partly due to the recent nature of its confirmation through direct observation.
+The reality of the situation is that the 2015 detection of GWs was made possible only through the careful and determined efforts of the LIGO and VIRGO scientific collaborations, which consist of thousands of scientists whose work spans decades.
+The motivation for this comes from theoretical work that is traceable back to Einstein himself and the original theoretical prediction of gravitational waves as a mathematical consequence of general relativity.
+As a result the theory as well as the methods and techniques in practical use in Advanced LIGO era are highly mature and developed and perhaps somewhat subtle to those not directly familiar with GW signal processing and data analysis.
+The purpose of this chapter is to provide a brief overview of these techniques to both remind the experienced researcher of why we use the processes we use, as well as introduce the more general scientific audience to the basics of GW physics and astronomy as it pertains to this work.
+We present an overview of the theory of linearized gravity starting (and ending) with the Einstein field equations.
+We introduce Bayesian statistics as our main tool for analysis and connect it to our work, hopefully providing enough foundation to elucidate our choices and motivations for completing this work.
 
-%----------------------------------------------------------------------------------------
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%----------------------------------------------------------------------------------------
+\section{The Einstein Field Equation}
+The Einstein Field Equations are the central objects of study in General Relativity. They are ten, coupled, nonlinear, partial differential equations in four dimensions. They are the following:
 
-\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
+\begin{align}\label{eq:fulleinst}
+R_{\mu \nu} - \frac{1}{2}Rg_{\mu \nu} = T_{\mu \nu}
+\end{align}
 
-\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
-\usepackage[adobe-utopia]{mathdesign}
-%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
-\usepackage[english]{babel} % English language/hyphenation
-\usepackage{amsmath,amsfonts,amsthm} % Math packages
+Just as a differential equation relates the forces and constraints in a physical system to, perhaps, a function (and its derivatives) describing the motion of a particle through space and time, the Einstein field equations relate the mass and energy in a region of spacetime to the \textit{metric tensor} $g_{\mu \nu}$ and its derivatives, from which the geodesic paths objects follow through space and time can be derived.
+In this way $g_{\mu \nu}$ itself can be viewed as a \textit{solution} to the field equations given some sources or sinks in the region in question.
+In the following sections we will briefly motivate the pieces of the equation itself and hopefully provide some basic intuition as to how Einstein wrote this relationship down, and how from it we may mathematically derive the existence of gravitational waves in their entirety.
 
-\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
-
-\usepackage{sectsty} % Allows customizing section commands
-
-\usepackage{braket}
-\usepackage{cancel}
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage[mathscr]{euscript}
-
-\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps
-
-\usepackage{fancyhdr} % Custom headers and footers
-\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
-\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
-\fancyfoot[L]{} % Empty left footer
-\fancyfoot[C]{} % Empty center footer
-\fancyfoot[R]{\thepage} % Page numbering for right footer
-\renewcommand{\headrulewidth}{0pt} % Remove header underlines
-\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
-\setlength{\headheight}{13.6pt} % Customize the height of the header
-
-\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-
-\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text
-
-%----------------------------------------------------------------------------------------
-%	TITLE SECTION
-%----------------------------------------------------------------------------------------
-
-\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height
-
-\title{	
-\normalfont \normalsize 
-\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
-\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
-\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
-}
-
-\author{Brandon B. Miller, Masters Thesis} % Your name
-
-\date{\normalsize\today} % Today's date or a custom date
-
-\begin{document}
-
-\maketitle % Print the title
-\newpage
-\section{Background and Motivation}
-
-Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
-The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.
-
-\subsection{Pieces of the Einstein Equation}
-\subsubsection{Reimann Tensor}
+\subsection{Reimann Tensor}
 The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
@@
 
-\begin{figure}[h!]
-\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
+\begin{figure}[h!]\label{fig:reimann}
+\includegraphics[scale=0.5]{reimann.png}
 \caption{A visual depiction of the Reimann Curvature tensor.
-Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors.
+Parallel transporting a vector originally pointing in the $\sigma$ direction both ways around the loop shown yields two different vectors.
 The orange vector $\vec{d}$ points between the tips.
-The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
+The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.} The first symmetry of the Reimann tensor is apparent - interchanging the last two indices would simply yield the same vector $\vec{d}$ the the opposite sign on all the components.
 \label{fig:sphere}
@@
 Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector.
-From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
+From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \ref{fig:reimann}.
 Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above.
-Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop.
-It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question.
-More remarkable however is the simplicity of the formal definition within this picture:
+It is sometimes written as
 
@@
 It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}
+This also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
+Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel symbols defined in \nameref{Appendix C}.
 
-\begin{align}
+\begin{align}\label{eq:reimanntnsfull}
 R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
@@
 
-The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
+\subsection{Ricci Tensor}
+
+Some experimentation with various contractions of the Reimann tensor as defined by equation \ref{eq:reimanntnsfull} over its own indices reveals that there is only one nonzero result:
 
 \begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
+R_{\mu \nu} &\equiv R^\alpha_{\sigma \alpha \nu}
 \end{align}
 
-If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
-It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
+This is called the Ricci Tensor and represents the full trace of the Reimann tensor.
+Einstein's equaiton provides the only means to a physical interpretation of this result.
+Under the assumption that equation \textbf{EINSTEIN} is the correct relation between spacetime curvature and local sources, then the traceless remnant of the Reimann tensor must describe the spacetime curvature from \textit{external} sources.
+This is named the Weyl tensor and it plays an important role in describing the propagation of gravitational radiation.
+Indeed, a spacetime carrying gravitational radiation may satisfy the vacuum Einstein equation
 
 \begin{align}
-\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
+R_{\mu \nu} = 0
 \end{align}
 
-\section{Linearized Gravity}
+But may \textit{not} satisfy
+\begin{align}
+R^\rho_{\sigma \mu \nu} = 0
+\end{align}
+
+\subsection{Ricci Scalar}
+
+The most natural line of logic to pursue next is what happens when the Ricci tensor itself is contracted over the two remaining indices.
+
+\begin{align}
+R \equiv R^{\alpha}_{\alpha} = g^{\mu \nu}R_{\mu \nu}
+\end{align}
+
+The result is the Ricci scalar of scalar curvature which is a unique scalar measuring the local curvature at every point on the manifold.
+It is only one of many ways of assigning a scalar to each point, and geometrically it is the average of the Gaussian curvatures for all two dimensional "cross sections" that exist within the tangent space at the point in question.
+
+\subsection{Stress-Energy-Momentum Tensor}
+It is the opinion of the author that the motivations behind why the stress-energy-momentum tensor falls on the right hand side of equation \ref{eq:fulleinst} are best understood by following the same line of logic taken historically by Einstein, Minkowski and Von Laue.
+This discussion has been relegated to Appendix A.
+For now it will suffice to say that $T_{\mu \nu}$ reconciles what was in the early 1900s the new formalism of four-vectors with fundamental concepts such as the continuity equation.
+It bridged the gap between the mature machinery of relativistic electric and magnetic fields and "other" types of matter, casting them as one in the same.
+For now we will simply say that it may be found by considering the properties of matter and energy in a region and plays the role of the source term in equation \ref{eq:fulleinst}.
+
+\section{Linearization}
 
@@
 The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \textbf{EQUATION}, we would have
-
-\begin{align}
-g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
-&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
-&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
-\end{align}
-
-The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity.
-We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
 
@@
 
+While the details steps are expanded in Appendix C, we state here without proof that the following statements are true.
 The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
@@
 \begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
-&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
+\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
 \end{align}
 
-Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
-Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero.
-This leaves us with three:
+These can be used to define the linearized Reimann and Ricci tensors along with the Ricci scalar:
 
 \begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
-&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
+R^{\rho}_{\sigma \mu \nu} &= \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\ 
+R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right] \\ \label{eq:linricci}
+R &= \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
 \end{align}
 
-Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself.
-It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate.
-The same assumption allows for simplification of the Reimann and Ricci tensors:
+With the linearized Reimann tensor in hand, it turns out through various gauge freedoms and coordinate transformations also expounded upon in the appendix that the linearized, vacuum Einstein field equation is simply
 
-\begin{align}
-R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
-R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
-\end{align}
-
-Plugging in equation \textbf{EQUATION} we have
-
-\begin{align}
-R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
-&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
-&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
-\end{align}
-
-The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$.
-The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$.
-The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:
-
-\begin{align}
-R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
-\end{align}
-
-Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is
-
-\begin{align}
-R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
-\end{align}
-
-Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity
-
-\begin{align}
-V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
-\end{align}
-
-With which the first two terms of the linearized Ricci tensor can be reorganized as
-
-\begin{align}
-\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
-&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
-\end{align}
-
-Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields
-
-\begin{align}
-R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
-\end{align}
-
-The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with
-
-\begin{align}
-R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
-\end{align}
-
-Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
-
-\begin{align}
-\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
-\end{align}
-
-$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics.
-However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner.
-Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e.
-$dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed.
-Thus we have
-
-\begin{align}
-h_{\mu \nu} = h_{\nu \mu}
-\end{align}
-
-Which, for a sixteen component tensor, makes six of the components redundant.
-This leaves us with ten.
-Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation.
-This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively.
-It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form
-
-\begin{align}
-\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
-V &\rightarrow V -\frac{\partial \Psi}{\partial t}
-\end{align}
-
-The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
-Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
-
-\begin{align}
-\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
-\rightarrow \partial^{\mu}A_{\mu} &= 0
-\end{align}
-
-It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden!
-While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
-It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}.
-We do not have to state \textit{what it is}, as long as we know that it exists.
-A proof that there legitimately is such a gauge function exists but is out of the scope of this outline.
-There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
-In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
-
-\begin{align}
-x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
-\end{align}
-
-In particular, we can show that there exists a coordinate transformation such that
-
-\begin{align}
-\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
-\end{align}
-
-\textbf{PUT PROOF HERE}
-
-\begin{align}
-\text{PUT PROOF HERE}
-\end{align}
-
-Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply
-
-\begin{align}
+\begin{align}\label{eq:flteinst}
 \Box h_{\mu \nu} = 0
@@
 
-This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
-Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}.
-In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
-This leaves us with six.
+Which is a wave equation in the tensor purturbation's components.
+The same freedoms serve to reveal that out of the sixteen wave equations that correspond to equation \ref{eq:flteinst}, only two are physically meaningful.
+Orienting the coordinate frame such that the wave propagation direction lies along the $z$ axis, these two degrees of freedom are manifest as the $x$ and $y$ spatial distortions
+
+\begin{align}
+h_{\mu \nu} = 
+\begin{bmatrix}
+0 & 0 & 0 & 0 \\
+0 & h_+ & h_\times & 0 \\
+0 & h_\times & -h_+ & 0 \\
+0 & 0 & 0 & 0
+\end{bmatrix}
+\end{align}
+
+In which the $+$ and $\times$ "polarizations" of the spatially propagating tensor disturbance satisfy
+
+\begin{align}
+h_p(\vec{x}) = A_p e^{i \mathbf{k}\cdot\mathbf{x}}
+\end{align}
+
+It will be convenient in the coming chapters to define the gravitational wave strain in a detector $h(t)$ subject to this disturbance to this disturbance as single complex number that encodes these two independent degrees of freedom as one object:
+
+\begin{align}
+h(t) = h_+(t) - i h_{\times}(t)
+\end{align}
 
 \section{Bayesian Statistics}
-We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
+
+We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented \cite{Volume-2} as a tool of choice for gravitational wave signal processing.
 Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
@@
 For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
-This is
+Following the treatment outlined in literature \cite{rapidpe} This is
 
@@
 Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
-Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
-These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm.
+Many schemes \cite{rapidpe} \cite{lalinference} \cite{mcmc} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
+These techniques are highly developed and at the time of writing can provide converged results on the order of days, however they are limited in their parallelism due fundamental serialization of the algorithm.
 Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
-A new and more embarassingly parallel scheme is needed.
+Methods that can scale out to the computational resources provided by the LIGO data grid are needed to reduce runtimes to minutes or less.
 
@@
 
-Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
+Where $D_r$ is the luminosity reference distance to the binary.
+This is defined similarly to the manner in which apparent magnitude is defined versus absolute magnitude in astonomy.
+We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
 
@@
 The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
-Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.
+Work by \cite{rapidpe} (RapidPE) has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing over the extrinsic parameters.
+The same work computes the posterior over the \textit{intrinsic} parameter space by spreading a grid of points across the $\mathcal{M}, \eta$ plane and computing the integral at each such point, since each of these integrations is independent each point for which the value of the posterior is desired may be assigned its own computational resources.
+This enables the necessary scaling - the posterior between the points is simply interpolated after all the cores are done.
 
-\section{Implementation}
-The main weight of the computational work described above falls under an expensive Monte Carlo integration of the evidence as part of the Bayesian posterior.
-Zooming in it quickly becomes apparent that the main challenge to be overcome is efficient evaluation of the function to be integrated, in this case the factored likelihood.
-Prior to this work, the factored likelihood is was computed with a series of looping structures that individually compiled all the terms of the equation in an independent and serial manner.
-The process took substantial advantage of many high level operations available within Python to match up the correct terms across various harmonic mode time series, spherical harmonics, and antenna patterns.
-Thanks to clever optimization and gratuitous use of fast numerical Python libraries such as NumPy, the code was able to compute approximately $10^3$ likelihood evaluations on the order of seconds, the limiting factor being the serial nature of loops and the inherent lethargy of high level languages such as Python.
-Although NumPy commonly passes target data to compiled routines written in faster languages such as C, some portions of the computation (in particular the marginalization over time) benefit little from this capability and form a bottleneck for efficient evaluation of the likelihood.
-Furthermore the original implementation heavily relied upon older structures within the existing LIGO Algorithms Library (LAL), a set of C routines bound to Python through Swig, to perform many intermediate calculations, such as computing spherical harmonics.
-The goal of the first phase of this work was thorough vectorization of the process used to build the terms of the likelihood utilizing the BLAS subroutines available through NumPy as often as possible, as well as reducing dependency on LAL.
-To that end, the components of the factored likelihod were reworked.
-
-\subsection{Complex Antenna Factor}
-The complex antenna factor is a complex number that modulates the gravitational wave signal as a function of incoming direction.
-This is necessary due to the non isotropic sensitivity of the LIGO interferometers themselves.
-$F_{+}$ and $F_{\times}$.
-
-\begin{align}
-F_{+} &= [R_{0,0}X_0 + R_{0,1}X_1 + R_{0,2}X_2]X_0 - [R_{0,0}Y_0 + R_{0,1}Y_1 + R_{0,2}Y_2]Y_0 \\
-&+ [R_{1,0}X_0 + R_{1,1}X_1 + R_{1,2}X_2]X_1 - [R_{1,0}Y_0 + R_{1,1}Y_1 + R_{1,2}Y_2]Y_1 \\
-&+ [R_{2,0}X_0 + R_{2,1}X_1 + R_{2,2}X_2]X_2 - [R_{2,0}Y_0 + R_{2,1}Y_1 + R_{2,2}Y_2]Y_2
-\end{align}
-
-Collecting all the positive terms of this expression gives
-\begin{align}
-F_{+} &= R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 \\ 
-&+ R_{0,0}X_0X_0 + R_{0,1}X_0X_1 + R_{0,2}X_0X_2 
-\end{align}
-At which point we see that the same result is computable as an outer product of the vector $\vec{X}$ using
-
-\begin{align}
-F_{+} &= \vec{X}\mathbf{R}\vec{X} - \vec{Y}\mathbf{R}\vec{Y}
-\end{align}
-
-And the complex part of the gravitational wave is thus
-
-\begin{align}
-F_{\times} &= \vec{X}\mathbf{R}\vec{Y} + \vec{Y}\mathbf{R}\vec{X}
-\end{align}
-
-One a sample-to-sample basis, the numbers that vary are the components of the vectors $\vec{X}$ and $\vec{Y}$.
-Thus we need a function that takes as input a \textit{list} vectors (or \textit{vector}) of vectors and produces a vector with the right components as output.
-To that end we define the tensor $X^{i}_{j}$ where
-
-\begin{align}
-X^i &= 
-\begin{bmatrix}
-X^{i}_0 \\
-X^{i}_1 \\
-X^{i}_2
-\end{bmatrix}
-\end{align}
-
-As well as the tensor $R^{i}_{jk}$ where
-
-\begin{align}
-R^{i} &= 
-\begin{bmatrix}
-R^{i}_{00} & R^{i}_{01} & R^{i}_{02} \\ 
-R^{i}_{10} & R^{i}_{11} & R^{i}_{12} \\ 
-R^{i}_{20} & R^{i}_{21} & R^{i}_{22}  
-\end{bmatrix}
-\end{align}
-
-The tensor $X$ is like a stack of all the different possible $\vec{X}$ coming out of the page.
-The Tensor $R$ is like $n$ copies of the matrix $\mathbf{R}$ stacked on top of each other.
-In this way the desired vector is obtainable with the tensor contraction
-
-\begin{align}
-F^{i}_{+} &= X^{lm}R^{i}_{lj}X^{j}_{m} - Y^{lm}R^{i}_{lj}Y^{j}_{m} \\ 
-F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
-\end{align}
-
-\subsection{Vectorized Single Detector Log Likelihood}
-
-With the spherical harmonics and antenna factor in hand the following set of operations yield the factored likelihood for a single detector, the network likelihood being a simple sum over all of the detectors.
-
-Equation 24 from Arxiv $15502.05370v1.pdf$ for the network log likelihood reads
-
-\begin{align}
-\ln{\mathcal{L}} = &\frac{D_{ref}}{D}Re \sum_{k}\sum_{(l,m)}\left[F_k Y_{lm}\right]^{*}Q_{k,lm} \\ 
-& - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] \\
- & - \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right]
-\end{align}
-
-Consider a single detector, thus dropping the sum over $k$. The first term is of the form $\vec{A}\cdot\vec{B} = \sum_{i=0}^{d}A_iB_i$, so if $Q_{k,(l,m)}$ were a simple vector, we could write it as
-
-\begin{equation}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2} F*\vec{Y}\cdot\vec{Q}
-\end{equation}
-
-However the $Q_{k,(l,m)}$ are actually harmonic mode time series and not single values. We desire a vector whose values are the likelihoods at each point in the time series.
-
-Consider the case where we have only the $(2,-2), (2,0)$ and $(2,2)$ modes.
-If we write all the mode time series $Q^0, Q^1, Q^2...$ as the columns of a matrix , then the desired result is obtained with
-
-\begin{align}
-F*
-\begin{bmatrix}
-Q^0_{2,-2} & Q^0_{2,+0} & Q^0_{2,+2} \\
-Q^1_{2,-2} & Q^1_{2,+0} & Q^1_{2,+2} \\ 
-Q^2_{2,-2} & Q^2_{2,+0} & Q^2_{2,+2} \\
-\vdots & \vdots & \vdots
-\end{bmatrix}
-\begin{bmatrix}
-\left(Y_{2,-2}\right)\\
-\hspace{0mm} \\
-\left(Y_{2,+0}\right) \\
-\hspace{0mm} \\
-\left(Y_{2,+2}\right) \\
-\end{bmatrix}
-=
-\begin{bmatrix}
-Q^0_{2,-2} Y_{2,-2} + Q^0_{2,-2}Y_{2,+0} + Q^0_{2,+2}Y_{2,+2} \\
-Q^1_{2,-2} Y_{2,-2} + Q^1_{2,-2}Y_{2,+0} + Q^1_{2,+2}Y_{2,+2} \\
-Q^2_{2,-2} Y_{2,-2} + Q^2_{2,-2}Y_{2,+0} + Q^2_{2,+2}Y_{2,+2} \\
-\vdots
-\end{bmatrix}
-\end{align}
-
-With $\vec{Y}$ and $\mathbf{Q}$ defined as the matrix and vector above respectively, we have for the first term
-
-\begin{equation}
-\frac{D_{ref}}{D}Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right]
-\end{equation}
-
-The second term is a sum once over all the possible combinations of $(l,m), (l',m')$ pairs using the $U_{(l,m),(l',m')}$ cross terms. Its result is a scalar quantity made up of terms like
-
-\begin{align}
-&Y_{2,-2}^{*}Y_{2,-2}U_{(2,-2),(2,-2)} + Y_{2,-2}^{*}Y_{2,+0}U_{(2,-2),(2,+0)} + Y_{2,-2}^{*}Y_{2,+2}U_{(2,-2),(2,+2)} \\ 
- +  \ &Y_{2,+0}^{*}Y_{2,-2}U_{(2,+0),(2,-2)} + Y_{2,+0}^{*}Y_{2,+0}U_{(2,+0),(2,+0)} + Y_{2,+0}^{*}Y_{2,+2}U_{(2,+0),(2,+2)} \\ 
-+ \  &Y_{2,+2}^{*}Y_{2,-2}U_{(2,+2),(2,-2)} + Y_{2,+2}^{*}Y_{2,+0}U_{(2,+2),(2,+0)} + Y_{2,+2}^{*}Y_{2,+2}U_{(2,+2),(2,+2)}
-\end{align}
-
-If we pack the $U_{(l,m),(l',m')}$ into the matrix $\mathbf{U}$ as defined below then the following set of matrix operations produces the same sum
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} &  U_{(2,-2),(2,+0)} &  U_{(2,-2),(2,+2)} \\
-U_{(2,+0),(2,-2)} &  U_{(2,+0),(2,+0)} &  U_{(2,+0),(2,+2)} \\
-U_{(2,+2),(2,-2)} &  U_{(2,+2),(2,+0)} &  U_{(2,+2),(2,+2)}
-\end{bmatrix}
-\begin{bmatrix}
-Y_{2,-2} \\
-Y_{2,+0} \\
-Y_{2,+2}
-\end{bmatrix}
-\end{align*}
-
-because when you multiply $\mathbf{U}$ into $\vec{Y}$ this simplifies to
-
-\begin{align*}
-\begin{bmatrix}
-Y_{2,-2}^{*} &Y_{2,+0}^{*}  &Y_{2,+2}^{*}   
-\end{bmatrix}
-\begin{bmatrix}
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} \\ 
-U_{(2,-2),(2,-2)} Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} + U_{(2,-2),(2,-2)}Y_{2,-2} 
-\end{bmatrix}
-\end{align*}
-
-Which becomes the desired scalar. This allows us to write the second term as
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}\left[|F_k|^2 Y_{l,m}^{*}Y_{l',m'}U_{k,(l,m),(l'm')}\right] =  - \left[\frac{D_{ref}}{2D}\right]^{2} |F^2|\vec{Y}^{*}\mathbf{U}\vec{Y}
-\end{align}
-
-We must always set up the spherical harmonic vectors based on the value of $m$ and the cross terms in row-major form based first on $m_2$ and then on $m_1$.
-If we organize the matrix $\mathbf{V}$ in the same way then the same set of steps will lead us to conclude that
-
-\begin{align}
-- \left[\frac{D_{ref}}{2D}\right]^{2}\sum_{k}\sum_{(l,m),(l',m')}Re\left[  F_{k}^{2}Y_{l,m}Y_{l'm'}V_{k,(l,m),(l'm')}\right] = - \left[\frac{D_{ref}}{2D}\right]^{2}Re \left[F^2 \vec{Y}\mathbf{V}\vec{Y} \right]
-\end{align}
-
-Combining the results the single detector log likelihood is
-
-\begin{align}
-\ln{\mathcal{L}} = \frac{D_{ref}}{D}\Re\left[\mathbf{Q}\left(F\vec{Y}\right)^{*}\right] - \left[\frac{D_{ref}}{2D}\right]^{2}\left[|F|^2 \vec{Y}^{*}\mathbf{U}\vec{Y} - \Re\left(F^2 \vec{Y}\mathbf{V}\vec{Y}\right) \right]
-\end{align}
-
-\subsection{GPU Implementation}
-Vectorization of the code produces a performance improvement due to SIMD instruction sets available on many modern microprocessor architectures.
-This allows for some on-chip parallelism that the compiler may use to unroll loops within the machine code and increase efficiency.
-Beyond this however the process of recasting the computation into clear matrix and tensor operations illuminates a higher level of parallelism possible on the hardware level.
-A substantial amount of both academc and industrial research and development has centered around efficient parallel implementations of the equivalent operations, with impressive results.
-Reaching the limit of the sample throughput with conventional serial python libraries, our investigation turned to GPU based acceleration as the only means forward.
-The following is the result of this investigation.
-
-\subsection{Hardware Limitations and Logical Programming}
-GPUs work significantly differently from CPUs and must be paired with one to operate, they serve only as accelerators for complex operations and not as standalone units.
-Data must be passed back and forth from host memory on the node to RAM placed directly on the card itself through expensive transfers via the PCI bus.
-This is known as device global memory.
-While the onboard RAM on bleeding edge cards exceeds 12GB, memory transfers are the dominant cost of many GPU programs and must be minimized.
-Once the data exists on the device, it is accessed and manipulated by anywhere between nine and fifteen logical cores called Streaming Multiprocessors.
-These cores operate concurrently and contain a smaller granularity of parallel processing units that \textit{may} operate concurrently depending on the individual resources required by the requested operations.
-Programming in CUDA involves launching groups of threads called \textit{grids} that are further subdivided into \textit{blocks}.
-The blocks themselves are indexed by local variables that are rapidly accessible by all the contained threads, and the threads themselves all have access to indices placed within registers that exist for the lifetime of the thread in question.
-Both the blocks and and grids can be up to three dimensional and thus each block and thread may have access to up to three registers containing block indices and three containing thread indices.
-In addition thread blocks have access to an extremely high speed shared memory space where ideally most of the computations are performed, however it is limited in size to \textbf{SIZE} and is not designed to hold the entire target dataset.
-One major design goal for CUDA programs is implementing a process which copies portions of memory from the global space to the shared, processes it, and copies it back, while using the thread and block indices to blanket the dataset with threads and control which threads access what data.
-The current CUDA API allows for the logical launch of \textbf{HOW MANY} blocks with a maximum of 1024 threads per block, for a total of \textit{HOW MANY} possible "concurrent" threads.
-It is important to note that these threads are not \textit{physically} concurrent.
-Indeed, if each of the logical threads containing six integers in registers were to exist simultaneously, the GPU would be storing \textbf{HOW MANY} registers at once!
-Similarly perposterous amounts of shared memory would have to exist.
-In actuality, the shared memory and registers are located on the logical cores themselves, and blocks thread blocks line up to be processed.
-How many are processed at once is heavily dependent on the resources requested by the blocks themselves.
-Defining too many local variables in a kernel function can lead to \textit{register pressure} and decreased \textit{occupancy} of the multiprocessors on the GPU, a similar effect can come from requesting too much shared memory.
-Furthermore, the blocks are not guaranteed to execute in any particular order.
-While the compiler is responsible for sorting out how many blocks can be processed at a time as well as the order, the programmer is responsible for performance tuning with regard to optimizing block occupancy.
-This is often a late stage design consideration, but also leads to some of the greatest returns after algorithmic optimization has been completed.
-
-\subsection{PyCuda} 
-PyCuda is a flexible library for GPU scripting that includes a broad spectrum of GPU based array operations built to mimic the usage of NumPy arrays within a program.
-It allows for relatively easy access to the Nvidia CUDA API directly from within Python, and many of its operations are directly compatible with NumPy arrays with careful casting to the correct data types.
-Certainly one of the most challenging aspects of incorporating GPU code into a higher level language such as Python is the interface layer.
-As it stands much of the widely used LAL software consists of C code bound to Python through complex Swig interfaces, and while this is a perfectly valid way to access faster libraries, it can often be deceptively difficult to work with as many of the resulting Python objects contain read-only data structures masquerading as dynamic Python objects.
-This sometimes yields unpredictable behavior and it was decided this should be avoided as a way to interface with CUDA C.
-PyCuda is a very elegant solution to this problem as it seamlessly bridges this gap using the native GPUArray class, and a set of functions that create GPUArray copies of NumPy arrays that are built initially on the node CPU hosts.
-With a smaller group of developers, the built-in GPUArray class contained within PyCuda does not yet support \textit{all} of the rich spectrum of high level functions available within the NDArray NumPy class.
-Perhaps more importantly however, PyCuda allows for custom CUDA kernel functions to be written as pure CUDA C directly within special containers called source modules, that are contained within the main body of the Python code.
-PyCuda actually calls the real Nvidia compiler at runtime and links the resulting temporary executable to the main program.
-One may call these functions with NumPy arrays as arguments, as if they were normal Python functions.
-This is perhaps the most important feature of PyCuda as it allows the developer to tune the customization of the code to whatever level is desired.
-Those intermediate operations that may be handled with the built in methods of the GPUArray are performed as such.
-More complex operations that are not supported can be custom built with source modules.
-If desired, the entire program may consist of CUDA C with no more then basic startup routines written in actual Python meant to instantiate the original data and pass it down to the GPU.
-This particular work used mostly custom kernels, for the simple reason that applying the same matrix operation to a contiguous block of memory containing multiple slightly different copies constitutes tensor operations similar in nature to the results of section \textbf{SECTION}, for which there is no general support for in PyCuda.
-
-\subsection{General Design Considerations}
-Computing the time marginalized likelihood for a single sample involves processing the likelihood for each point in a signal time series, multiplied again by the number of associated harmonic modes.
-As the number of samples selected by the integrator as an adaptation chunk increases, the global memory required for the program to execute becomes dominated by these time series.
-Consequently the code is built upon managing and manipulating a large contiguous block of memory whose rows contain the harmonic mode time series for each sample, this lends itself naturally to using a two dimensional grid of one dimensional blocks as the launch configuration for most kernel functions.
-The one dimensional blocks lie conceptually along the rows and may line up if the number of time series samples exceeds the maximum threds per block.
-The grid extends downwards in the $y$-direction and will generally assign a row of blocks to each sample, or in some cases blanket the memory entirely in threads.
-It is the nature of CUDA programs that thread blocks operate most efficiently when the total number of threads per block is a power of two.
-Aside from the microprocessor architecture optimizing for these cases, memory reads by groups of thirty two threads (known as a \textit{Warp}) are coalesced by the compiler and recieve the value to be read as a broadcast.
-This massively boosts throughput if used correctly and it is thus advantageous in most cases to cast the problem dimensions into some workable multiple of the maximum threads per block on the hardware.
-For most Nvidia cards this is 1024.
-It is substantially easier to pad the time series with zeros out to the next greatest multiple of this number then to design specialized cleanup blocks whose sole purpose is to process the remainder.
-In the worse case this corresponds to holding \textbf{NUMBER} extra zeros in memory, which can be a painful amount of unecessary overhead in some scenarios.
-These factors come together to make memory management the main bottleneck for the entire algorithm, and a main area of focus for further development of this method.
-
-\subsection{Spherical Harmonics}
-
-The spin-weighted spherical harmonics are precomputed at startup.
-There is no analytic formula that returns the spin-weighted spherical harmonic for arbitrary quantum numbers $l, m$ and $s$.
-The most efficient way of retrieving them closely follows the method used by the current, serial LAL functions, which implement a lookup table that contains the explicit formulas.
-This is often sufficient as most waveform templates require only the $l=2$ modes, and thus only the corresponding spherical harmonics.
-Indeed, the LSC has yet to require support for $s \neq 2$, so the corresponding CUDA function does not attempt to extend this capability.
-
-\begin{wrapfig}
-\hspace{-1cm}
-\includegraphics[trim={0cm, 2cm, 2cm, 2cm}, clip]{spherical_harmonics.eps}
-\caption{Computation of Spherical Harmonics, $l=2$.}
-\label{fig:spharms}
-\end{wrapfig}
-
-Instead, it takes as argument an ordered list of $m$ values and executes the following set of steps to return a row-major ordered block of results whose rows correspond to values of $m$: Launch a one dimensional grid of one dimensional blocks, with one thread per sample.
-Compute the answer for this value of $m$ and place the result in global memory.
-Increment the local thread index by the number of samples and repeat.
-As long as it loops through the $m$ values in order, the results will be placed correctly, as per figure \ref{fig:spharms}.
-As the extrinsic parameter values used in computation are accessed relatively few times, this function does not use shared memory.
-Regardless as a standalone function it is able to compute several billion results per second, so little further optimization is necessary.
-
-\clearpage
-
-\subsection{Complex Antenna Factor}
-The antenna factor requires somewhat less fancy footwork then the spherical harmonics, as there is only one per sample and not an arbirary multiple depending on the number of selected modes (which are not necessarily all the values between $-l$ and $l$.
-The position vectors corresponding to the extrinsic parameters are contracted with the detector response tensor in a thread-wise fashion: one thread for each sample.
-We avoid using BLAS based routines to perform the contraction due to the small size of the matrices involved, startup routines for BLAS might take longer then a simple loop over nine elements of the detector tensor.
-We again use a one dimensional grid of one dimensional blocks, however since the position vectors contain commonly accessed elements, we choose to store these independently in device shared memory as well as the detector tensor itself in even higher speed device constant memory.
-This makes the antenna factor computation one of the most efficient steps of the process.
-
-\subsection{Main Routine}
-
-Due to the variation in extrinsic parameters over the samples, it is necessary for each to manipulate a separate group of copies of the harmonic mode time series.
-The number of samples may be large, so the main block of memory intended to hold these time series is handed down to the GPU initialized to zeros and a GPU function is used to expanded the time series into a vertial stack of copies, per figure \ref{fig:expandrho}.
-To streamline the marginalization process this block of memory is padded with zeros out to the nearest multiple of $T_{max}$, which is generally either 512 or 1024 depending on the GPU architecture.
-
-\begin{wrapfig}
-\vspace{10mm}
-\includegraphics[trim={0cm, 13cm, 0cm, 3.0cm}, clip, scale=0.65]{expand_rhots.eps}
-\label{fig:expandrho}
-\caption{Expansion of time series. One thread per relevant item in resulting memory block. Zeros are padded out to the nearest multiple of $T_{max}$.
-\end{wrapfig}
-
-\clearpage
-
-The result of this expansion of the time series into GPU memory is the desired vertical stack of time series copies.
-
-The next step is to transform these time series into term one of equation \textbf{EQUATION}.
-This involves combining the antenna factor (for which there is one complex number per sample) with the spin-weighted spherical harmonics (for which there are $N_m$ complex numbers per sample) into a single quantity and inserting them into the correct positions within the time series block.
-It is more efficient to complete the combination of the antenna factor and spherical harmonics prior to multiplying them into the time series, and this action is easily completed in an elementwise fasion using the built in machinery of PyCuda.
-It is steps like these for which PyCuda is an essential tool for performing what would otherwise be complex GPU operations with a single line of code.
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\hspace{-5cm}
-\vspace{-1cm}
-  \begin{center}
-    \includegraphics[trim={4cm, 13cm, 2cm, 3cm}, clip, width=0.75\textwidth]{FYrho.eps}
-  \end{center}
-  \caption{Insertion of the antenna factor-spherical harmonic pairs. }
-\end{wrapfigure}
-
-This step coupled with the previous is all that is necessary to build the first term of the likelihood for all of the samples on the GPU, the rest of the work is simply a series of summing operations.
-These must be handled somewhat carefully due to the possibility of race conditions and queued memory accesses, both of which can hinder or dangerously skew the results of a computation.
-The main issue stems from the fact that there is no guarantee of the order of execution of threadblocks within a program.
-This means that any program that depends upon memory reads and writes occuring in a specific order is prone to producting incorrect results.
-For this reason it is important to design the summation process in a way that does not exhibit this behavior: each thread should only read and write to memory locations unique to it.
-Note that each of the functions described thus far follows this paradigm.
-Queued memory are serialized by the CUDA compiler.
-This means that memory reads to the same location by multiple threads take place one by one.
-While not as dangerous as a race condition, it is a nontrivial performance consideration.
-Note that again the process thus far has been designed to avoid these situations.
-This is particularly important for justifying the nature of the next series of steps.
-
-\clearpage
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\vspace{-0.78cm}
-\begin{center}
-\includegraphics[trim={1.67cm, 10cm, 0, 3cm}, clip, scale=0.5]{sumrhots.eps}
-\caption{Downwards summation of the harmonic modes to form term one of the factored likelihood for a group of samples.}
-\end{center}
-\end{wrapfigure}
-
-The time series are summed downwards within memory by single threads.
-One row of threads is launched per sample, this leaves $N_m$ values for each thread to collect in a summation that resides in the bottom row.
-The memory locations accessed and written to by the individual threads in the grid are unique to that thread, avoiding races and queued memory accesses.
-The result are rows that are the same as those that would have been formed taking the product $\mathbf{Q}(F\vec{Y})$, this is the first term of the factored likelihood.
-It remains only to build the second term and subtract it from the rows of this matrix to complete the computation.
-
-\subsection{Marginalization Over Time}
-
-\begin{wrapfigure}{l}{0.6\textwidth}
-\begin{center}
-\includegraphics[scale=0.7]{onedmarg.eps}
-\end{center}
-\end{wrapfigure}
-
-%---------------------------------------------------------------------------------------
-
-\end{document}
+\subsection{Precessional Effects}
+This work aims to extend and accelerate the ability for RapidPE to estimate gravitational wave parameters.
+RapidPE is written in Python, and prior to this work significant attention was given to code optimization to obtain the maximum performance possible through Python.
+Since the terms of the likelihood that may be precomputed involve computationally expensive fourier transforms, calculation of the cross terms $V$ were approximated using the properties of fourier transform to relate

### 2016-08-28 Gradschool/thesis/final/bak/chap2.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-28 Gradschool/thesis/final/bak/chap1.tex
@@
 %% prompt when you run the file main.tex through LaTeX.
-\chapter{Background}
-\section{Preface}
-Gravitational wave astronomy is often called a new way of observing the universe, partly due to the recent nature of its confirmation through direct observation.
-The reality of the situation is that the 2015 detection of GWs was made possible only through the careful and determined efforts of the LIGO and VIRGO scientific collaborations, which consist of thousands of scientists whose work spans decades.
-The motivation for this comes from theoretical work that is traceable back to Einstein himself and the original theoretical prediction of gravitational waves as a mathematical consequence of general relativity.
-As a result the theory as well as the methods and techniques in practical use in Advanced LIGO era are highly mature and developed and perhaps somewhat subtle to those not directly familiar with GW signal processing and data analysis.
-The purpose of this chapter is to provide a brief overview of these techniques to both remind the experienced researcher of why we use the processes we use, as well as introduce the more general scientific audience to the basics of GW physics and astronomy as it pertains to this work.
-We present an overview of the theory of linearized gravity starting (and ending) with the Einstein field equations.
-We introduce Bayesian statistics as our main tool for analysis and connect it to our work, hopefully providing enough foundation to elucidate our choices and motivations for completing this work.
-
-\section{The Einstein Field Equation}
-The Einstein Field Equations are the central objects of study in General Relativity. They are ten, coupled, nonlinear, partial differential equations in four dimensions. They are the following:
-
-\begin{align}\label{eq:fulleinst}
-R_{\mu \nu} - \frac{1}{2}Rg_{\mu \nu} = T_{\mu \nu}
-\end{align}
-
-Just as a differential equation relates the forces and constraints in a physical system to, perhaps, a function (and its derivatives) describing the motion of a particle through space and time, the Einstein field equations relate the mass and energy in a region of spacetime to the \textit{metric tensor} $g_{\mu \nu}$ and its derivatives, from which the geodesic paths objects follow through space and time can be derived.
-In this way $g_{\mu \nu}$ itself can be viewed as a \textit{solution} to the field equations given some sources or sinks in the region in question.
-In the following sections we will briefly motivate the pieces of the equation itself and hopefully provide some basic intuition as to how Einstein wrote this relationship down, and how from it we may mathematically derive the existence of gravitational waves in their entirety.
-
-\subsection{Reimann Tensor}
-The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
-The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.
-
-\begin{figure}[h!]\label{fig:reimann}
-\includegraphics[scale=0.5]{reimann.png}
-\caption{A visual depiction of the Reimann Curvature tensor.
-Parallel transporting a vector originally pointing in the $\sigma$ direction both ways around the loop shown yields two different vectors.
-The orange vector $\vec{d}$ points between the tips.
-The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.} The first symmetry of the Reimann tensor is apparent - interchanging the last two indices would simply yield the same vector $\vec{d}$ the the opposite sign on all the components.
-\label{fig:sphere}
-\end{figure}
-
-The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space.
-Here we motivate the concept with a visual explanation of parallel transport.
-Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}.
-If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$.
-This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$.
-Let this vector be denoted $\vec{d}$.
-Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector.
-From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \ref{fig:reimann}.
-Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above.
-It is sometimes written as
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
-\end{align}
-
-It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-This also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel symbols defined in \nameref{Appendix C}.
-
-\begin{align}\label{eq:reimanntnsfull}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
-\end{align}
-
-\subsection{Ricci Tensor}
-
-Some experimentation with various contractions of the Reimann tensor as defined by equation \ref{eq:reimanntnsfull} over its own indices reveals that there is only one nonzero result:
-
-\begin{align}
-R_{\mu \nu} &\equiv R^\alpha_{\sigma \alpha \nu}
-\end{align}
-
-This is called the Ricci Tensor and represents the full trace of the Reimann tensor.
-Einstein's equaiton provides the only means to a physical interpretation of this result.
-Under the assumption that equation \textbf{EINSTEIN} is the correct relation between spacetime curvature and local sources, then the traceless remnant of the Reimann tensor must describe the spacetime curvature from \textit{external} sources.
-This is named the Weyl tensor and it plays an important role in describing the propagation of gravitational radiation.
-Indeed, a spacetime carrying gravitational radiation may satisfy the vacuum Einstein equation
-
-\begin{align}
-R_{\mu \nu} = 0
-\end{align}
-
-But may \textit{not} satisfy
-\begin{align}
-R^\rho_{\sigma \mu \nu} = 0
-\end{align}
-
-\subsection{Ricci Scalar}
-
-The most natural line of logic to pursue next is what happens when the Ricci tensor itself is contracted over the two remaining indices.
-
-\begin{align}
-R \equiv R^{\alpha}_{\alpha} = g^{\mu \nu}R_{\mu \nu}
-\end{align}
-
-The result is the Ricci scalar of scalar curvature which is a unique scalar measuring the local curvature at every point on the manifold.
-It is only one of many ways of assigning a scalar to each point, and geometrically it is the average of the Gaussian curvatures for all two dimensional "cross sections" that exist within the tangent space at the point in question.
-
-\subsection{Stress-Energy-Momentum Tensor}
-It is the opinion of the author that the motivations behind why the stress-energy-momentum tensor falls on the right hand side of equation \ref{eq:fulleinst} are best understood by following the same line of logic taken historically by Einstein, Minkowski and Von Laue.
-This discussion has been relegated to Appendix A.
-For now it will suffice to say that $T_{\mu \nu}$ reconciles what was in the early 1900s the new formalism of four-vectors with fundamental concepts such as the continuity equation.
-It bridged the gap between the mature machinery of relativistic electric and magnetic fields and "other" types of matter, casting them as one in the same.
-For now we will simply say that it may be found by considering the properties of matter and energy in a region and plays the role of the source term in equation \ref{eq:fulleinst}.
-
-\section{Linearization}
-
-In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation
-
-\begin{align}\label{eq:metpluspert}
-g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
-\end{align}
-
-Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources.
-The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-
-\begin{align}
-h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
-\end{align}
-
-While the details steps are expanded in Appendix C, we state here without proof that the following statements are true.
-The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
-\end{align}
-
-These can be used to define the linearized Reimann and Ricci tensors along with the Ricci scalar:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} &= \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\ 
-R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right] \\ \label{eq:linricci}
-R &= \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
-\end{align}
-
-With the linearized Reimann tensor in hand, it turns out through various gauge freedoms and coordinate transformations also expounded upon in the appendix that the linearized, vacuum Einstein field equation is simply
-
-\begin{align}\label{eq:flteinst}
-\Box h_{\mu \nu} = 0
-\end{align}
-
-Which is a wave equation in the tensor purturbation's components.
-The same freedoms serve to reveal that out of the sixteen wave equations that correspond to equation \ref{eq:flteinst}, only two are physically meaningful.
-Orienting the coordinate frame such that the wave propagation direction lies along the $z$ axis, these two degrees of freedom are manifest as the $x$ and $y$ spatial distortions
-
-\begin{align}
-h_{\mu \nu} = 
-\begin{bmatrix}
-0 & 0 & 0 & 0 \\
-0 & h_+ & h_\times & 0 \\
-0 & h_\times & -h_+ & 0 \\
-0 & 0 & 0 & 0
-\end{bmatrix}
-\end{align}
-
-In which the $+$ and $\times$ "polarizations" of the spatially propagating tensor disturbance satisfy
-
-\begin{align}
-h_p(\vec{x}) = A_p e^{i \mathbf{k}\cdot\mathbf{x}}
-\end{align}
-
-It will be convenient in the coming chapters to define the gravitational wave strain in a detector $h(t)$ subject to this disturbance to this disturbance as single complex number that encodes these two independent degrees of freedom as one object:
-
-\begin{align}
-h(t) = h_+(t) - i h_{\times}(t)
-\end{align}
+\chapter{Bayesian Statistics}
 
 \section{Bayesian Statistics}
-
 We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented \cite{Volume-2} as a tool of choice for gravitational wave signal processing.

### 2016-08-28 Gradschool/thesis/final/bak/chap3.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-25 Gradschool/thesis/msthesis.tex
@@
-\documentclass[paper=a4, fontsize=11pt]{scrartcl} % A4 paper and 11pt font size
-\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
-\usepackage[adobe-utopia]{mathdesign}
-%\usepackage{fourier} % Use the Adobe Utopia font for the document - comment this line to return to the LaTeX default
-\usepackage[english]{babel} % English language/hyphenation
-\usepackage{amsmath,amsfonts,amsthm} % Math packages
-\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
-\usepackage{sectsty} % Allows customizing section commands
-
-\usepackage{braket}
-\usepackage{cancel}
-\usepackage{graphicx}
-\usepackage{wrapfig}
-\usepackage[mathscr]{euscript}
-
-\allsectionsfont{\centering \normalfont\scshape} % Make all sections centered, the default font and small caps
-
-\usepackage{fancyhdr} % Custom headers and footers
-\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
-\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
-\fancyfoot[L]{} % Empty left footer
-\fancyfoot[C]{} % Empty center footer
-\fancyfoot[R]{\thepage} % Page numbering for right footer
-\renewcommand{\headrulewidth}{0pt} % Remove header underlines
-\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
-\setlength{\headheight}{13.6pt} % Customize the height of the header
-
-\numberwithin{equation}{section} % Number equations within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{figure}{section} % Number figures within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-\numberwithin{table}{section} % Number tables within sections (i.e. 1.1, 1.2, 2.1, 2.2 instead of 1, 2, 3, 4)
-
-\setlength\parindent{0pt} % Removes all indentation from paragraphs - comment this line for an assignment with lots of text
-
-%----------------------------------------------------------------------------------------
-%	TITLE SECTION
-%----------------------------------------------------------------------------------------
-
-\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height
-
-\title{	
-\normalfont \normalsize 
-\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
-\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge GPU Accelerated Parameter Estimation of Gravitational Waves from Binary Black Hole Mergers \\ % The assignment title
-\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
-}
-
-\author{Brandon B. Miller, Masters Thesis} % Your name
-
-\date{\normalsize\today} % Today's date or a custom date
-
-\begin{document}
-
-\maketitle % Print the title
-\newpage
-\section{Background and Motivation}
-
-Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
-The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.
-
-\subsection{Pieces of the Einstein Equation}
-\subsubsection{Reimann Tensor}
-The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
-The concept of a curved spacetime is fundamentally no different from the concept of a curved space, and from examining the behavior of vectors in such a space we can address both the purpose and properties of the Reimann tensor without departing from the intuitive picture of spheres and arrows.
-
-\begin{figure}[h!]
-\includegraphics[trim={2cm, 2cm, 2cm, 2cm}, clip]{spheredone.eps}
-\caption{A visual depiction of the Reimann Curvature tensor.
-Parallel transporting a vector in the $\sigma$ direction both ways around the loop shown yields two different vectors.
-The orange vector $\vec{d}$ points between the tips.
-The magnitude of the components of this vector are $R_{\rho \sigma \mu \nu}$.}
-\label{fig:sphere}
-\end{figure}
-
-The Reimann tensor is traditionally derived by examining the local behavior of geodesics subject to a metric tensor that describes curved space.
-Here we motivate the concept with a visual explanation of parallel transport.
-Consider a vector pointing in some arbitrary direction $\sigma$ somewhere on the sphere, as depicted in figure \ref{fig:sphere}.
-If we maintain the orientation of the vector relative to the surface of the sphere as we traverse the path, we note that the vector behaves differently depending on the order in which we choose to traverse the directions $\mu$ and $\nu$.
-This difference is representable by a vector that points between the two tips - one that is dependent on the initial direction $\sigma$ and both the transport directions $\mu$ and $\nu$.
-Let this vector be denoted $\vec{d}$.
-Then the Reimann tensor, $R_{\rho \sigma \mu \nu}$, is simply $d_{\rho}$, the $\rho$'th component of this difference vector.
-From this geometric picture some of the classical symmetries that are often introduced abreast the definition of the Reimann tensor are immediately intuitive, such as the skew symmetry depicted in figure \textbf{FIGURE HERE}.
-Since it is somewhat unfair (worse - \textit{incorrect}) to characterize the entire geometry of the space with looping paths over large areas, the formal definition of the Reimann tensor is in fact the infinitesimal version of the concept above.
-Indeed, the concept of a single direction one may traverse is only valid in a small region of curved space, so the real Reimann tensor must describe the difference between two vectors as parallel transported around an infinitesimally small version of the original loop.
-It follows then how there must be a (possibly different) Reimann tensor at each point on the manifold in question.
-More remarkable however is the simplicity of the formal definition within this picture:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu}\partial_{\rho} = \left(\nabla_{\mu}\nabla{\nu} - \nabla_{\nu}\nabla_{\mu}\right)\partial_{\sigma}
-\end{align}
-
-It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
-Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel Symbols \textbf{AS DEFINED IN APPENDIX}
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
-\end{align}
-
-The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
-
-\begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
-\end{align}
-
-If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
-It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
-
-\begin{align}
-\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
-\end{align}
-
-\section{Linearized Gravity}
-
-In linearized gravity we assume the spacetime is described by a flat spacetime $g_{\mu \nu}$ plus a small perturbation
-
-\begin{align}
-g_{\mu \nu} = \eta_{\mu \nu} + h_{\mu \nu}
-\end{align}
-
-Since the other pieces of the Einstein field equations are directly derived from the metric, they must be individually linearized to form a description of spacetime around weak sources.
-The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \textbf{EQUATION}, we would have
-
-\begin{align}
-g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
-&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
-&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
-\end{align}
-
-The second two terms being second order in $h$, ther peturbation, are neglected in linearized gravity.
-We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
-
-\begin{align}
-h^\mu_\nu = \eta^{\alpha \mu} h_{\alpha \nu} \ \  \text{and} \ \ h^{\mu \nu} = \eta^{\alpha \mu}\eta^{\beta \nu}h_{\alpha \beta}
-\end{align}
-
-The next natural line of logic to pursue is to write down the Christoffel symbols for such a spacetime
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
-&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
-\end{align}
-
-Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
-Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero.
-This leaves us with three:
-
-\begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
-&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
-\end{align}
-
-Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself.
-It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate.
-The same assumption allows for simplification of the Reimann and Ricci tensors:
-
-\begin{align}
-R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
-R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
-\end{align}
-
-Plugging in equation \textbf{EQUATION} we have
-
-\begin{align}
-R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
-&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
-&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]
-\end{align}
-
-The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$.
-The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$.
-The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:
-
-\begin{align}
-R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
-\end{align}
-
-Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is
-
-\begin{align}
-R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
-\end{align}
-
-Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity
-
-\begin{align}
-V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
-\end{align}
-
-With which the first two terms of the linearized Ricci tensor can be reorganized as
-
-\begin{align}
-\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
-&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
-\end{align}
-
-Which upon substitution into the \textbf{EQUATION} for $R_{\mu \nu}$ yields
-
-\begin{align}
-R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
-\end{align}
-
-The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with
-
-\begin{align}
-R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
-\end{align}
-
-Setting the right hand side of this equation gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
-
-\begin{align}
-\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
-\end{align}
-
-$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics.
-However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner.
-Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e.
-$dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed.
-Thus we have
-
-\begin{align}
-h_{\mu \nu} = h_{\nu \mu}
-\end{align}
-
-Which, for a sixteen component tensor, makes six of the components redundant.
-This leaves us with ten.
-Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation.
-This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively.
-It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form
-
-\begin{align}
-\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
-V &\rightarrow V -\frac{\partial \Psi}{\partial t}
-\end{align}
-
-The function $Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
-Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
-
-\begin{align}
-\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
-\rightarrow \partial^{\mu}A_{\mu} &= 0
-\end{align}
-
-It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden!
-While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
-It actually is there inside the $V$ and $\vec{A}$ written in equation \textbf{EQUATION}.
-We do not have to state \textit{what it is}, as long as we know that it exists.
-A proof that there legitimately is such a gauge function exists but is out of the scope of this outline.
-There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
-In the same sense as equation \textbf{EQUATION}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
-
-\begin{align}
-x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
-\end{align}
-
-In particular, we can show that there exists a coordinate transformation such that
-
-\begin{align}
-\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
-\end{align}
-
-\textbf{PUT PROOF HERE}
-
-\begin{align}
-\text{PUT PROOF HERE}
-\end{align}
-
-Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply
-
-\begin{align}\label{eq:flteinst}
-\Box h_{\mu \nu} = 0
-\end{align}
-
-This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
-Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \textbf{EQUATION}.
-In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
-This leaves us with six.
-
-In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read
-
-\begin{align}
-\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
--\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
-\end{align}
-
-This is the general form of the time dependent wave equation whose solutions take the form
-
-\begin{align}
-f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
-\end{align}
-
-Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector.
-Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have
-
-\begin{align}
-\text{PUT EQ HERE}
-\end{align}
-
-\section{Bayesian Statistics}
-We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented by \textbf{CITATION} and \textbf{PROBABLY OTHER CITATIONS}.
-Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
-Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
-We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$.
-A perpandicular cross section would simply be $P(y|x)$.
-We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$.
-Thus we have that
-
-\begin{figure}[h!]
-\center
-\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
-\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
-\label{fig:sphere}
-\end{figure}
-
-\begin{align} \label{eq:bayes}
-P(x|y)P(y) &= P(y|x)P(x) \\
-P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
-\end{align}
-
-Which is Bayes' theorem.
-The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome.
-Reading $\neg x$ as "not x", we have
-
-\begin{align}\label{eq:continuous}
-P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
-&= \sum_{k=1}^K P(y|x_k)P(x_k) 
-\end{align}
-
-\clearpage 
-This allows us to rewrite the posterior probability as
-
-\begin{align}
-P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
-\end{align}
-
-Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has
-
-\begin{align} \label{eq:lratio}
-P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
-\end{align}
-
-Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression.
-The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur.
-In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio.
-This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.
-
-\subsection{Signal Processing}
-Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:
-
-\begin{align}
-P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
-\end{align}
-
-If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as
-
-\begin{align}
-P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
-\end{align}
-
-For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
-This is
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
-\end{align}
-
-In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$.
-If there is no signal present in the data, then the data and noise are equal, meaning
-
-\begin{align}
-d(t) &= n(t) \\
-P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
-\end{align}
-
-Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:
-
-\begin{align}
-d(t) &= n(t) + h(t) \\
-p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
-\end{align}
-
-So the likelihood ratio for gravitational wave signal processing is exactly equal to
-
-\begin{align}
-\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
-\end{align}
-
-By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
-
-\begin{align}\label{eq:parampos}
-P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
-\end{align}
-
-Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator
-
-\begin{align}
-P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
-\end{align}
-
-Where $\Omega$ indicates integration over the entire probabilistic state space.
-Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us
-
-\begin{align}\label{eq:bayeslikely}
-P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
-\end{align}
-
-Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters.
-For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source.
-We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively.
-The denominator of equation \ref{eq:bayeslikely} is thus factorizable as
-
-\begin{align}
-\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta}
-\end{align}
-
-Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
-Many schemes \textbf{CITE BEN} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
-These techniques are highly developed and at the time of writing can provide converged results in \textbf{SOME TIMESCALE}, however they are limited in their parallelism due fundamental serialization of the algorithm.
-Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
-A new and more embarassingly parallel scheme is needed.
-
-\subsection{Harmonic Mode Decomposition}
-
-For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:
-
-\begin{align}
-h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
-\end{align}
-
-Where $D_r$ is the luminosity reference distance to the binary. We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
-
-\begin{align}
-\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
-\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
-&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
-\end{align}
-
-Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,
-
-\begin{align}
-\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
-&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
-&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h^*_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y^*_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
-\end{align}
-
-Now, for any complex number $z = a + ib$, we have that $z*z + \Re(z^2) = 2z*z$, so $z*z = \frac{1}{2}(z*z + \Re z^2)$. Then the double sum over $l,m$ pairs becomes
-
-The defining feature of this form of the likelihood is the separation between intrinsic and extrinsic parameters.
-The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
-Work by \textbf{CITE CHRIS AND RICHARD} has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary and marginalizing out all of the extrinsic parameters with a Monte Carlo integration.
+%% This is an example first chapter.  You should put chapter/appendix that you
+%% write into a separate file, and add a line \include{yourfilename} to
+%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
+%% You can process specific files by typing their names in at the 
+%% \files=
+%% prompt when you run the file main.tex through LaTeX.
+\chapter{Implementation}
 
@@
 Below we have tabulated $B$ for some values of $N_s$ to be used as reference when running the code.
-
-\section{Results and Prospects for Low Latency Electromagnetic Followup}
-\begin{figure}[H!]
-\hspace{-5.0cm}
-\includegraphics[scale=0.57]{speedup.eps}
-\end{figure}
-
-The dawn of gravitational wave astronomy in the 2015-2016 era has touched off a revolution in our ability to observe and classify astronomical phenomena.
-As with any entirely novel method of gleaning information from distant sources, development of methods to process and decipher such data has been explosive.
-Pure gravitational wave astrophysics has exciting prospects for the future, but the possibility of advancing other classical forms of astrophysics through their interplay with these newer methods is equally compelling.
-Significant research \textbf{CITATION} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers.
-A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration klonovae.
-Such early warning could provide valuable insights as to the central engine of such gamma ray bursts and confirm recent models of these processes.
-However, electromagnetic observational resources are limited and expensive to operate and such it is imperative that any communication to such resources in the context of EM followup of GW triggers must be rapid and accurate to avoid wasting time and money.
-To that end it is desireable to develop existing parameter estimation codes to such a point that they can produces fast and correct results to reliably indicate potential sources of interest while minimizing false positives.
-The main goal of this work is to push this envelope and reduce the end-to-end runtime of parallel PE codes (specifically RapidPE) for the express purpose of low latency electromagnetic follow-up.
-
-\subsection{Gamma Ray Bursts}
-Gamma ray bursts (GRBs) are short duration transient events thought to occur during or shortly after the merger of compact binary systems, specifically those which include a single black hole and one neutron star.
-The presence of strong radiative emission from the poles of black holes has been confirmed by \textbf{CITATION}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs.
-The process is depicted in figure \textbf{FIGURE} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole.
-The radiation eminated from the poles of the BH impacts this disk and generates gamma rays, this is known as an internal shock.
-As the resulting cloud of matter ploughs into the surrounding interstellar medium, it slows, leaving a lower density region in its wake.
-Subsequent waves of matter may propagate through this region at a higher velocity then the initial blast wave, catching up and ramming into the back of the initial wave, producing further shocks.
-It is not yet explicitly confirmed that this is the exact process by which GRBs are generated.
-It is generally accepted that to confirm this hypothesis we would need to observe direct coincidence between a GW trigger localized to some region of the sky and a GRB from the same direction.
-These phenomena have been demonstrated in numerical simulations by \textbf{CITATION} and are highly dependent on initial system parameters.
-In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors \textbf{CITATION}.
-We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
-We use rate predictions from \textbf{CITATION} to restrict the mass range to be studied to fall between \textbf{SOME RANGE}.
-These bounds corespond to a chirp mass range of \textbf{WHAT RANGE}.
-From this we can determine the expected time that the signal will spend in the Advanced LIGO band.
-In figure \textbf{FIGURE}, we have visualized this range in \textbf{SOME PLANE}.
-Using the results of our code we find that anadditional \textbf{SOME PERCENTAGE} of expected NSBH sytems expected to be picked up by Advanced LIGO are now candidates for joint GW-EM astronomy.
-
-\begin{figure}[H!]
-\hspace{-0.5cm}
-\includegraphics[scale=0.35]{gamray.png}
-\end{figure}
-
-\subsection{Kilonovae}
-Compact binary mergers of neutron stars are thought to produce transient electromagnetic counterparts known as \textit{Kilonovae}.
-These smaller class of novae are dimmer and lower energy then regular novae or supernovae, and are hypothesized to occur due to the radiative emission of remnant matter displaced by the merger dynamics.
-It has been shown that a substantial fraction of elements heavier then Iron-56 are formed through r-process nucleosynthesis in the hearts of large stars, during which nuclei are subjected to high neutron flux as well as high electron density filling quantum states up to the Fermi energy.
-When this occurs $\beta^-$ decay of neutrons into protons is blocked \textit{CITATION} resulting in accumulation of large amounts of neutrons within the nuclei themselves.
-These nuclei later decay into heavy elements.
-These conditions are present during the merger of neutron stars or the tidal disruption of a neutron star by a black hole.
-Models of kilonovae have been successful in reproducing faint EM transients using r-process emission as a primary mechanism for radiation.
-Coincident measurements of such transients with localized gravitational waves is expected to be a smoking gun for confirming this model of kilonova genesis, thus it is important to assess the capability of modern parameter estimation to process signals from candidate systems and produce meaningful information useful for kilonova followup.
-Since merger ejecta from NSBH mergers is not subject to relativistic beaming, kilonovae emission is expected to be relatively isotropic.
-Assuming a random and uniform distribution of black hole spin directions there is thus an increased probability of detecting kilonovae of \textbf{SOLID ANGLE OVER 4 PI} relative to axially emissive GRBs.
-This translates to a better chance of finding a kilonova as a merger signature.
-Moreover, direct observation of kilonovae combined with numerical simulations can provide degeneracy-breaking information about the original progenitor through careful measurements of redshift compared to simulated spectra.
-The timescales characteristic of kilonova signatures are derived from numerical simulations, and most suggest a peak in the emission intensity roughly a day after merger.
-That said, the same simulations predict a steep rise in luminosity in the period just before the peak.
-For this reason crucial information about the astrophysics of kilonovae may be rapidly lost, and it is important to collect it as rapidly as possible.
-Figure \textbf{SOME FIGURE} depicts the timescales associated with kilonovae from \textbf{CITATION} as well as current PE methods prospects of convergence prior to certain features of the kilonova light curve.
-
-\begin{figure}[H!]
-\hspace{-6cm}
-\includegraphics[scale=0.5]{EMFollow.eps}
-\end{figure}
-
-%---------------------------------------------------------------------------------------
-
-\end{document}

### 2016-08-28 Gradschool/thesis/final/bak/chap4.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-26 Gradschool/thesis/final/chap4.tex
@@
 
-\section{Results and Prospects for Low Latency Electromagnetic Followup}
-\begin{figure}[H!]
+\section{Speedup and Scaling}
+The goal of this work is to accelerate parameter estimation of compact binaries.
+Here we describe the results of vetting the finished code for several relevant test cases.
+Since RapidPE employs a numerical Monte-Carlo integration scheme to compute the posterior probability, it relies heavily on rapid evaluations of the function to be integrated.
+In this case the function in question is the likelihood itself, thus the first comparison to be made is the ability of the GPU code to return values of the likelihood relative to the Python based code.
+In the actual workflow, likelihood evaluations come in large chunks that are used as an adaptation period for the integrator.
+We tested the code for twenty different adaptation chunk sizes averaging over ten trials to obtain relative runtimes for CUDA vs Python for the fully time-marginalized likelihood.
+The results of this set of experiements are summarized in figure \ref{fig:speedup}.
+We found that regardless of chunk size, the GPU code implemented as above can return the correct answers a factor of fifteen faster across the board.
+This includes memory transfer processes between the host and GPU.
+As a measure to reduce transfer time, the large block of memory used to hold the harmonic mode time series is reused and reinitialized to zeros before each chunk is processed, removing the single greatest barrier to high performance.
+\begin{figure}[!htb]\label{fig:speedup}
 \hspace{-5.0cm}
 \includegraphics[scale=0.57]{speedup.eps}
+\caption{Computation of the time marginalized, factored log likelihood.
+Speedup for different chunk sizes.
+The values on top of the columns are the ratio between the Python and CUDA implementations, averaged over ten trials, timed using Pythons timing library.}
 \end{figure}
 
-The dawn of gravitational wave astronomy in the 2015-2016 era has touched off a revolution in our ability to observe and classify astronomical phenomena.
+\section{Results and Prospects for Low Latency Electromagnetic Followup}
+
+The dawn of gravitational wave astronomy in the 2015-2016 era \cite{gw150914} has touched off a revolution in our ability to observe and classify astronomical phenomena.
 As with any entirely novel method of gleaning information from distant sources, development of methods to process and decipher such data has been explosive.
@@
 Significant research \cite{first2years} in th epast several years has been designed specifically to connect electromagnetic astronomy to gravitational wave observations, in particular the study of low-latency electromagnetic follow-up to gravitatioanl wave triggers.
-A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration klonovae.
+A central point of focus in these studies is the opportunity to give electromagnetic observers early warning on possible short duration transient events such as gamma ray bursts or longer duration kilonovae.
 Such early warning could provide valuable insights as to the central engine of such gamma ray bursts and confirm recent models of these processes.
@@
 Gamma ray bursts (GRBs) are short duration transient events thought to occur during or shortly after the merger of compact binary systems, specifically those which include a single black hole and one neutron star.
+Catalogs of known GRBs have resolved into two distinct categories, those known as short GRBs ($<2s$) and long GRBs ($>2s$) \cite{twogrb}.
 The presence of strong radiative emission from the poles of black holes has been confirmed by \cite{polarjet}, and while not yet directly observed, the interaction between such polar jets of radiation and the resulting post-merger NS remnant is a primary candidate for the generation of GRBs.
-The process is depicted in figure \textbf{FIGURE} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole.
+The process is depicted in figure \ref{fig:gamray} in which an infalling NS is torn apart by the BH counterpart, leaving a remnant disk of energetic matter in orbit around the black hole.
 The radiation eminated from the poles of the BH impacts this disk and generates gamma rays, this is known as an internal shock.
@@
 Subsequent waves of matter may propagate through this region at a higher velocity then the initial blast wave, catching up and ramming into the back of the initial wave, producing further shocks.
+The beaming distribution for GRBs is currently unknown, and models \cite{grbopen} attempt to characterize the effect by assigning a geometric angle $\theta_j$ to the jets that describes the jets collimation.
+Coincident observations of gravitational waves which admit accurate parameter estimation, along with observation of a GRB, can constrain this angle - if PE can recover the spin directions of the binary components, then the EM observation will directly indicate if the angle subtended from the beam to the earth is contained within $\theta_j$.
 It is not yet explicitly confirmed that this is the exact process by which GRBs are generated.
@@
 These phenomena have been demonstrated in numerical simulations by \cite{kilonova} and are highly dependent on initial system parameters.
-In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors IIIIIII We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
+In particular the currently unknown commplete NS equation of state is known to play a pivotal role in braketing the mass and mass ratio ranges associated with GRB progenitors \textbf{CITATION}.
+We now attempt to demonstrate the efficacy of the GPU based version of RapidPE for estimating these parammeters quickly and accurately.
 The frequency evolution of a compact binary system is given by the well known relationship
+
+\begin{figure}[!htb]\label{fig:gamray}
+\hspace{-0.5cm}
+\caption{A gamma ray burst. Radiation escaping the black hole is thought to be genreated via infalling matter from a neutron star remnant.} 
+\includegraphics[scale=0.35]{gamray.png}
+\end{figure}
 
@@
 \begin{align}\label{eq:timeinband}
-\tau \approx 2.18s\left(\frac{1.21M_\odot}{\mathcal{M}}\right)^{\frac{5}{3}}\right)\left(\frac{100Hz}{f_{gw}}\right)^{\frac{8}{3}}
+\tau \approx 2.18s\left(\frac{1.21M_\odot}{\mathcal{M}}\right)^{\frac{5}{3}}\left(\frac{100Hz}{f_{gw}}\right)^{\frac{8}{3}}
 \end{align}
 
-For calculational purposes the value $10Hz$ is used as $f_{gw}$ in the above equation, as this is the lower bound of reasonable sensitivity for ground based interferometers such as Advanced LIGO \textbf{CITATION}.
+For calculational purposes the value $10Hz$ is used as $f_{gw}$ in the above equation, as this is the lower bound of reasonable sensitivity for ground based interferometers such as Advanced LIGO \cite{advligo}.
 
@@
 It has been found that systems for which approximately $q \gtrsim 4$ do not exhibit tidal disruption, the NS is swallowed whole by the BH before the tidal forces can overcome the NS self gravity.
-Since the largest NS masses are predicted \textbf{CITATION} to be approximately $1.4M_\odot$, the BH can be no larger then $\approx 5.6M_\odot$.
+Since the largest NS masses are predicted \cite{nsmasslim} to be approximately $1.4M_\odot$, the BH can be no larger then $\approx 5.6M_\odot$.
 This is equivalent to $\mathcal{M} \approx 2.33M_\odot$.
 Plugging this value into equation \ref{eq:timeinband} we find that such a system will spend approximately $5m 40s$ in the Advanced LIGO frequency band.
-
-\begin{figure}[H!]
-\hspace{-0.5cm}
-\includegraphics[scale=0.35]{gamray.png}
-\end{figure}
+Figure \textbf{FIGURE} shows the relationship between corrected magnitude $R_c$ and time since merger for a large set of GRBs mostly taken from SWIFT \cite{grbcatalog} data and illustrates the prospects for EM followup.
 
@@
 It has been shown that a substantial fraction of elements heavier then Iron-56 are formed through r-process nucleosynthesis in the hearts of large stars, during which nuclei are subjected to high neutron flux as well as high electron density filling quantum states up to the Fermi energy.
-When this occurs $\beta^-$ decay of neutrons into protons is blocked \textit{CITATION} resulting in accumulation of large amounts of neutrons within the nuclei themselves.
+When this occurs $\beta^-$ decay of neutrons into protons is blocked \cite{r-proc} resulting in accumulation of large amounts of neutrons within the nuclei themselves.
 These nuclei later decay into heavy elements.
@@
 For this reason crucial information about the astrophysics of kilonovae may be rapidly lost, and it is important to collect it as rapidly as possible.
-Figure \textbf{SOME FIGURE} depicts the timescales associated with kilonovae from \textbf{CITATION} as well as current PE methods prospects of convergence prior to certain features of the kilonova light curve.
+Figure \textbf{SOME FIGURE} depicts the timescales associated with kilonovae from light curves by Kasen et al \cite{kilonova} as well as current PE methods prospects of convergence prior to certain features of the kilonova light curve.
 

### 2016-08-28 Gradschool/thesis/final/chap1.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

@@
 %% prompt when you run the file main.tex through LaTeX.
-\chapter{Linearized Gravity}
-Gravitational waves are predicted as a mathematical consequence of the linearized Einstein Field Equations.
-The Einstein Field Equations themselves are a set of ten unique, coupled, nonlinear partial differential equations.
+\chapter{Background}
+\section{Preface}
+Gravitational wave astronomy is often called a new way of observing the universe, partly due to the recent nature of its confirmation through direct observation.
+The reality of the situation is that the 2015 detection of GWs was made possible only through the careful and determined efforts of the LIGO and VIRGO scientific collaborations, which consist of thousands of scientists whose work spans decades.
+The motivation for this comes from theoretical work that is traceable back to Einstein himself and the original theoretical prediction of gravitational waves as a mathematical consequence of general relativity.
+As a result the theory as well as the methods and techniques in practical use in Advanced LIGO era are highly mature and developed and perhaps somewhat subtle to those not directly familiar with GW signal processing and data analysis.
+The purpose of this chapter is to provide a brief overview of these techniques to both remind the experienced researcher of why we use the processes we use, as well as introduce the more general scientific audience to the basics of GW physics and astronomy as it pertains to this work.
+We present an overview of the theory of linearized gravity starting (and ending) with the Einstein field equations.
+We introduce Bayesian statistics as our main tool for analysis and connect it to our work, hopefully providing enough foundation to elucidate our choices and motivations for completing this work.
 
-\subsection{Pieces of the Einstein Equation}
-\subsubsection{Reimann Tensor}
+\section{The Einstein Field Equation}
+The Einstein Field Equations are the central objects of study in General Relativity. They are ten, coupled, nonlinear, partial differential equations in four dimensions. They are the following:
+
+\begin{align}\label{eq:fulleinst}
+R_{\mu \nu} - \frac{1}{2}Rg_{\mu \nu} = T_{\mu \nu}
+\end{align}
+
+Just as a differential equation relates the forces and constraints in a physical system to, perhaps, a function (and its derivatives) describing the motion of a particle through space and time, the Einstein field equations relate the mass and energy in a region of spacetime to the \textit{metric tensor} $g_{\mu \nu}$ and its derivatives, from which the geodesic paths objects follow through space and time can be derived.
+In this way $g_{\mu \nu}$ itself can be viewed as a \textit{solution} to the field equations given some sources or sinks in the region in question.
+In the following sections we will briefly motivate the pieces of the equation itself and hopefully provide some basic intuition as to how Einstein wrote this relationship down, and how from it we may mathematically derive the existence of gravitational waves in their entirety.
+
+\subsection{Reimann Tensor}
 The Reimann Tensor is a mathematical object that encapsulates the geometric \textit{curvature} of spacetime.
@@
 It should be noted that the above formula is true only modulo a term containing $\nabla_{[\mu, \nu]}$ except in the special case of coordinate vector fields, where it is zero.
-Although it does little to redeem it as a first explanation of the concept, this also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
-Somewhat more subtle is the explicit role of the Reimann tensor in describing geodesic deviation.
+This also illuminates why the Reimann tensor is sometimes described as encoding the non-commutativity of covariant derivatives on a manifold.
 Crucial to a straightforward linearization of the Reimann tensor itself is the expression in terms of the Christoffel symbols defined in \nameref{Appendix C}.
 
-\begin{align}
+\begin{align}\label{eq:reimanntnsfull}
 R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} + \Gamma^{\rho}_{\mu \lambda}\Gamma^{\lambda}_{\nu \sigma} - \Gamma^{\rho}_{\nu \lambda}\Gamma^{\lambda}_{\mu \sigma}
@@
 
-The christoffel symbols contain some information about the curvature of space as a function of the coordinates, and thus in the framework of linearized gravity products of Christoffel symbols are second order:
+\subsection{Ricci Tensor}
+
+Some experimentation with various contractions of the Reimann tensor as defined by equation \ref{eq:reimanntnsfull} over its own indices reveals that there is only one nonzero result:
 
 \begin{align}
-R^{\rho}_{\sigma \mu \nu} = \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} 
+R_{\mu \nu} &\equiv R^\alpha_{\sigma \alpha \nu}
 \end{align}
 
-If $x^{\rho}(\tau)$ are the components of a geodesic parameterized by $\tau$, then $\dot{x^{\rho}}(\tau) \equiv T^{\rho}$ are the components of a vector tangent to the geodesic.
-Suppose there is a family such nearby geodesics indexed by some other parameter, $s$, and that we can define \textit{deviation vector} $S^{\rho}(\tau) = \partial_{s}x^{\rho}(s,\tau)$ which characterizes the separation between nearby geodesics.
-It can be shown that the second derivative of this vector with respect to $\tau$ is related to the Reimann tensor through the differential equation
+This is called the Ricci Tensor and represents the full trace of the Reimann tensor.
+Einstein's equaiton provides the only means to a physical interpretation of this result.
+Under the assumption that equation \textbf{EINSTEIN} is the correct relation between spacetime curvature and local sources, then the traceless remnant of the Reimann tensor must describe the spacetime curvature from \textit{external} sources.
+This is named the Weyl tensor and it plays an important role in describing the propagation of gravitational radiation.
+Indeed, a spacetime carrying gravitational radiation may satisfy the vacuum Einstein equation
 
 \begin{align}
-\ddot{S^{\rho}} = R^{\rho}_{\sigma \mu \nu} T^{\sigma} T^{\mu} S^{\nu}
+R_{\mu \nu} = 0
 \end{align}
 
-\section{Linearized Gravity}
+But may \textit{not} satisfy
+\begin{align}
+R^\rho_{\sigma \mu \nu} = 0
+\end{align}
+
+\subsection{Ricci Scalar}
+
+The most natural line of logic to pursue next is what happens when the Ricci tensor itself is contracted over the two remaining indices.
+
+\begin{align}
+R \equiv R^{\alpha}_{\alpha} = g^{\mu \nu}R_{\mu \nu}
+\end{align}
+
+The result is the Ricci scalar of scalar curvature which is a unique scalar measuring the local curvature at every point on the manifold.
+It is only one of many ways of assigning a scalar to each point, and geometrically it is the average of the Gaussian curvatures for all two dimensional "cross sections" that exist within the tangent space at the point in question.
+
+\subsection{Stress-Energy-Momentum Tensor}
+It is the opinion of the author that the motivations behind why the stress-energy-momentum tensor falls on the right hand side of equation \ref{eq:fulleinst} are best understood by following the same line of logic taken historically by Einstein, Minkowski and Von Laue.
+This discussion has been relegated to Appendix A.
+For now it will suffice to say that $T_{\mu \nu}$ reconciles what was in the early 1900s the new formalism of four-vectors with fundamental concepts such as the continuity equation.
+It bridged the gap between the mature machinery of relativistic electric and magnetic fields and "other" types of matter, casting them as one in the same.
+For now we will simply say that it may be found by considering the properties of matter and energy in a region and plays the role of the source term in equation \ref{eq:fulleinst}.
+
+\section{Linearization}
 
@@
 The standard treatment of linearized gravity assumes that to first order we may raise and lower indices using the Minkowski metric.
-If we tried to use the full form of equation \ref{eq:metpluspert}, we would have
-
-\begin{align}
-g^{\mu}_{\nu} &= g^{\mu \alpha}g_{\alpha \nu} \\
-&= (\eta^{\mu \alpha} + h^{\mu \alpha}) (\eta_{\alpha \nu} + h_{\alpha \nu}) \\
-&= \eta^{\mu \alpha}\eta_{\alpha \nu} + \eta^{\mu \alpha}h_{\alpha \nu} + h^{\mu \alpha}\eta_{\alpha \nu} + h^{\mu \alpha}h_{\alpha \nu}
-\end{align}
-
-The second two terms being second order in $h$, the peturbation, are neglected in linearized gravity.
-We assume from here on that indices can be freely raised and lowered to first order using only the Minkowski metric $\eta_{\mu \nu}$.
 
@@
 \begin{align}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}g^{\nu \mu}\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right] \\
-&= \frac{1}{2}\left(\eta^{\mu \nu} + h^{\mu \nu}\right)\left[\partial_{\beta}g_{\alpha \nu} + \partial_{\alpha}g_{\beta \nu} - \partial_{\nu}g_{\alpha \beta}\right]
+\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
 \end{align}
 
-Inserting the expanded form of the metric into the term within the brackets gives us six terms times two at the front, for twelve total.
-Thankfully, three correspond to the standard flat-space Christoffel symbols which can be made zero through a choice of coordinates (in fact, orthonormal Cartesian coordinates), three involve derivatives of $\eta_{\mu \nu}$ which are zero, and three are second order in $h_{\mu \nu}$, which we assume are zero.
-This leaves us with three:
+These can be used to define the linearized Reimann and Ricci tensors along with the Ricci scalar:
 
-\begin{align}\label{eq:linchrist}
-\Gamma^{\mu}_{\alpha \beta} &= \frac{1}{2}\eta^{\mu \nu}\left[\partial_{\beta}h_{\alpha \nu} + \partial_{\alpha} h_{\beta \nu} - \partial_{\nu}h_{\alpha \beta}\right] \\
-&= \frac{1}{2} \left[\partial_{\beta} h^{\mu}_{\alpha} + \partial_{\alpha}h_{\beta}^{\mu} - \partial^{\mu}h_{\alpha \beta}\right] 
+\begin{align}
+R^{\rho}_{\sigma \mu \nu} &= \partial_{\mu}\Gamma^{\rho}_{\nu \sigma} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\ 
+R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right] \\ \label{eq:linricci}
+R &= \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
 \end{align}
 
-Where in the last term $\partial^{\mu} = \eta^{\mu \nu}\partial_{\nu}$ was used to raise the index on the actual partial derivative operator itself.
-It should be noted at this point that since we neglect terms that are second order in the Christoffel symbols themselves, in linearized gravity, the covariant derivative of a metric perturbation is equivalent to the normal partial derivate.
-The same assumption allows for simplification of the Reimann and Ricci tensors:
-
-\begin{align}
-R^{\rho}_{\mu \sigma \nu} &= \partial_{\sigma}\Gamma^{\rho}_{\mu \nu} - \partial_{\nu}\Gamma^{\rho}_{\mu \sigma} \\
-R_{\mu \nu} = R^{\rho}_{\mu \rho \nu} &= \partial_{\rho}\Gamma^{\rho}_{\nu \mu} - \partial_{\mu}\Gamma^{\rho}_{\mu \rho}
-\end{align}
-
-Plugging in equation \ref{eq:linchrist} we have
-
-\begin{align}
-R_{\mu \nu} & = \frac{1}{2}\partial_{\alpha}\left[\partial_{\nu}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\nu} - \partial^{\alpha}h_{\mu \nu}\right] - \frac{1}{2}\partial_{\nu}\left[\partial_{\alpha}h^{\alpha}_{\mu} + \partial_{\mu}h^{\alpha}_{\alpha} - \partial^{\alpha}h_{\mu \alpha}\right] \\
-&= \frac{1}{2}\left[ \cancel{\partial_{\alpha}\partial_{\nu}h^{\alpha}_{\mu}} + \partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \cancel{\partial_{\nu}\partial_{\alpha}h^{\alpha}_{\mu}} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} \right] \\
-&= \frac{1}{2}\left[\partial_{\alpha}\partial_{\mu}h^{\alpha}_{\nu} + \partial_{\nu}\partial^{\alpha}h_{\mu \alpha} - \partial_{\alpha}\partial^{\alpha}h_{\mu \nu} - \partial_{\nu}\partial_{\mu}h^{\alpha}_{\alpha}\right]\label{eq:linricci}
-\end{align}
-
-The quantity $h^{\alpha}_{\alpha}$ is the trace of the pertubation metric and represents, in a sense, an overall measure of the \textit{strength} or \textit{amplitude} of the perturbation, this is often called just $h$.
-The term $\partial^{\alpha}\partial_{\alpha}$ is an inner product that when formed through the Minkowski metric becomes the wave operator, $\Box = -\partial_{t}^{2} + \nabla^2$.
-The Ricci tensor is obtained by contracting the Reimann tensor over the two free indices:
-
-\begin{align}
-R \equiv \eta^{\mu \nu}R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\alpha}\partial^{\mu}h_{\mu}^{\alpha} + \partial_{\mu} \partial^{\alpha}h^{\mu}_{\alpha} - \Box h - \Box h\right]
-\end{align}
-
-Since $\alpha$ and $\mu$ are summed indices, the first two terms are the same. Dropping the $\frac{1}{2}$, the Ricci scalar curvature is
-
-\begin{align}
-R = \partial_{\alpha}\partial_{\mu}h^{\mu \alpha} - \Box h
-\end{align}
-
-Many authors choose to simplify the expression for the Ricci Tensor by defining the quantity
-
-\begin{align}\label{eq:lorentzgauge}
-V_{\nu} = \partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\nu}h
-\end{align}
-
-With which the first two terms of the linearized Ricci tensor can be reorganized as
-
-\begin{align}
-\partial_{\mu}\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\mu}\partial_{\nu}h &= \partial_{\mu}\left[\partial_{\alpha}h_{\nu}^{\alpha} - \partial_{\nu}h\right] \\
-&= \partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h
-\end{align}
-
-Which upon substitution into equation \ref{eq:linricci} for $R_{\mu \nu}$ yields
-
-\begin{align}
-R_{\mu \nu} &= \frac{1}{2}\left[\partial_{\mu}V_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h + \partial^{\alpha}\partial_{\nu}h_{\mu \alpha} - \Box h_{\mu \nu}\right]
-\end{align}
-
-The two unsimplified terms form the quantity $\partial_{\nu}\left[\partial^{\alpha}h_{\mu \alpha} - \frac{1}{2}\partial_{\mu}h\right]$ which we recognize as $\partial_{\nu}V_{\mu}$, leaving us with
-
-\begin{align}
-R_{\mu \nu} = \frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] 
-\end{align}
-
-Setting the right hand side of this equation equal to zero gives us the linearized, vacuum Einstein Equation $R_{\mu \nu} = 0$:
-
-\begin{align}
-\frac{1}{2}\left[\partial_{\mu}V_{\nu} + \partial_{\nu}V_{\mu} - \Box h_{\mu \nu}\right] = 0 
-\end{align}
-
-$h_{\mu \nu}$ is a two index object in a four dimensional spacetime, which in the general case would present with sixteen independent components each of which would need to be explicitly determined to form a complete picture of the physics.
-However in our specific case it will turn out that there are far fewer real independent components then this, which can be observed by chiseling away at the above expression in the following manner.
-Firstly, the metric of general relativity is taken to be symmetric, because if we assume that the tensor $dx^{\mu}dx^{\nu}$ is symmetric (i.e.
-$dx^{\nu}$ and $dx^{\nu}$ commute) then only the symmetric part of any $g_{\mu \nu}$ contributes when the quantity $g_{\mu \nu}dx^{\mu}dx^{\nu}$ is computed.
-Thus we have
-
-\begin{align}
-h_{\mu \nu} = h_{\nu \mu}
-\end{align}
-
-Which, for a sixteen component tensor, makes six of the components redundant.
-This leaves us with ten.
-Furthermore, there is \textit{gauge freedom} present in the linearized vacuum Einstein equation.
-This is best illustrated with reference to the equations of classical electrodynamics, which describe the motion of charged particles subject to the electric and magnetic vector potentials, $V$ and $\vec{A}$ respectively.
-It is somewhat easily demonstrable that the same equations of motion for electrically charged particles are unchanged when $V$ and $\vec{A}$ are subject to transformations of the form
-
-\begin{align}\label{eq:emtrans}
-\vec{A} &\rightarrow \vec{A} + \nabla \Psi \\
-V &\rightarrow V -\frac{\partial \Psi}{\partial t}
-\end{align}
-
-The function $\Psi$ is known as the \textit{gauge function} and it can be absolutely anything we want, as long as we adjust the potentials accordingly.
-Within such a framework, one can impose certain conditions on the potentials that simplify the mathematics, such as requiring that
-
-\begin{align}\label{eq:emgauge}
-\nabla \cdot \vec{A} + \frac{1}{c}\frac{\partial V}{\partial t} &= 0 \\
-\rightarrow \partial^{\mu}A_{\mu} &= 0
-\end{align}
-
-It is somewhat counterintuitive why imposing this particular condition on $\vec{A}$ and $V$ themselves is equivalent to transforming the potentials via the gauge function, but indeed it is, it is just hidden!
-While we have made no reference to $\Psi$, we have \textit{implicitly} assumed that there is such a $\Psi$ that will \textit{make} the restriction on $\vec{A}$ and $V$ true.
-It actually is there inside the $V$ and $\vec{A}$ written in equation \ref{eq:emgauge}.
-We do not have to state \textit{what it is}, as long as we know that it exists.
-A proof that there legitimately is such a gauge function exists but is out of the scope of this outline.
-There is a one-to-one correspondance of this formalism with the equivalent gauge in general relativity, which is known as the Lorentz gauge.
-In the same sense as equation \ref{eq:emtrans}, it turns out that the equations of motion for massive particles in a slightly perturbed spacetime are invariant under coordinate transformations of the form
-
-\begin{align}
-x^{\mu} \rightarrow x'^{\mu} - \xi^{\mu}
-\end{align}
-
-In particular, we can show that there exists a coordinate transformation such that
-
-\begin{align}
-\partial_{\mu}V_{\nu} = \partial_{\mu}\partial_{\alpha}h^{\alpha}_{\nu} - \frac{1}{2}\partial_{\mu}\partial_{\nu}h = 0
-\end{align}
-
-\textbf{PUT PROOF HERE}
-
-\begin{align}
-\text{PUT PROOF HERE}
-\end{align}
-
-Thus coordinate transformations that obey $\Box \xi = 0$ will allow us to impose the Lorentz gauge. Within this gauge, the linearized, vacuum Einstein equation becomes simply
+With the linearized Reimann tensor in hand, it turns out through various gauge freedoms and coordinate transformations also expounded upon in the appendix that the linearized, vacuum Einstein field equation is simply
 
@@
 
-This is a four dimensional wave equation for the remaining components of $h_{\mu \nu}$.
-Moreover, we have shown that under coordinate transformations that do not change the underlying physics, there exists an explicit relationship between components of the metric given by equation \ref{eq:lorentzgauge}.
-In the same way that symmetry shows that six of the metric components to be functions of the others, The index $\nu$ indicates that in four dimensions there are four additional explicit relationships between metric components that correspond to a loss of four additional degrees of freedom.
-This leaves us with six.
-
-In flat space, equation \ref{eq:flteinst} for some arbitrary function $f(x)$ in place of the $h_{\mu \nu}$ would read
+Which is a wave equation in the tensor purturbation's components.
+The same freedoms serve to reveal that out of the sixteen wave equations that correspond to equation \ref{eq:flteinst}, only two are physically meaningful.
+Orienting the coordinate frame such that the wave propagation direction lies along the $z$ axis, these two degrees of freedom are manifest as the $x$ and $y$ spatial distortions
 
 \begin{align}
-\Box f(x) = \eta_{\alpha \beta} \frac{\partial^2 f}{\partial x^{\alpha} \partial x^{\beta}} &= 0\\
--\frac{\partial^2 f}{\partial t^2} + \nabla^2f &= 0
+h_{\mu \nu} = 
+\begin{bmatrix}
+0 & 0 & 0 & 0 \\
+0 & h_+ & h_\times & 0 \\
+0 & h_\times & -h_+ & 0 \\
+0 & 0 & 0 & 0
+\end{bmatrix}
 \end{align}
 
-This is the general form of the time dependent wave equation whose solutions take the form
+In which the $+$ and $\times$ "polarizations" of the spatially propagating tensor disturbance satisfy
 
 \begin{align}
-f(x) = a e^{i \mathbf{k} \cdot \mathbf{x}}
+h_p(\vec{x}) = A_p e^{i \mathbf{k}\cdot\mathbf{x}}
 \end{align}
 
-Where $\mathbf{x}$ is the standard positon four-vector and $\mathbf{k}$ is the four-wave-vector.
-Expanding the inner product in the exponential again using the Minkowski metric as $-k^t t + \vec{k}\cdot \vec{x}$ and plugging the result into equation \ref{eq:flteinst} we have
+It will be convenient in the coming chapters to define the gravitational wave strain in a detector $h(t)$ subject to this disturbance to this disturbance as single complex number that encodes these two independent degrees of freedom as one object:
 
 \begin{align}
-\text{PUT EQ HERE}
+h(t) = h_+(t) - i h_{\times}(t)
 \end{align}
+
+\section{Bayesian Statistics}
+
+We select a Bayesian posterior probability as our main analysis technique as its ability to quantify and propagate uncertainties in parameter estimates is well documented \cite{Volume-2} as a tool of choice for gravitational wave signal processing.
+Bayes' Theorem follows from the symmetric definition of conditional probability depicted visually in figure \ref{fig:distrib}.
+Let the joint probability evaluated at any point $(x,y)$ in the plane be denoted $P(x,y)$.
+We may take a cross section through the joint probability distribution along any direction we choose, let us call such a cross section parallel to the $x$ axis the conditional probability of $x$ given $y$, and denote it $P(x|y)$.
+A perpandicular cross section would simply be $P(y|x)$.
+We see that we can compute $P(x,y)$ in two equivalent ways: scaling $P(x|y)$ by $P(y)$ or by scaling $P(y|x)$ by $P(x)$.
+Thus we have that
+
+\begin{figure}[h!]
+\center
+\includegraphics[trim={1cm, 1cm, 1cm, 1cm}, clip]{biviariate_norm_done.eps}
+\caption{A joint probability distribution $P(x,y)$. The contours parallel with the $x$ direction represent different $P(y|x)$.}
+\label{fig:sphere}
+\end{figure}
+
+\begin{align} \label{eq:bayes}
+P(x|y)P(y) &= P(y|x)P(x) \\
+P(x|y) &= \frac{P(y|x)P(x)}{P(y)}
+\end{align}
+
+Which is Bayes' theorem.
+The denominator of the expression on the right hand side is often rewritten as a sum over all possible situations in which $y$ is an outcome.
+Reading $\neg x$ as "not x", we have
+
+\begin{align}\label{eq:continuous}
+P(y) &= P(y|x)P(x) + P(y|\neg x)P(\neg x) \\
+&= \sum_{k=1}^K P(y|x_k)P(x_k) 
+\end{align}
+
+\clearpage 
+This allows us to rewrite the posterior probability as
+
+\begin{align}
+P(x|y) = \frac{P(y|x) P(x)}{\sum_{k=1}^K P(y|x_k)P(x_k)}
+\end{align}
+
+Multiplying the right hand side of this equation by the quantity $\frac{1}{P(x)P(y|\neg x)}$ one has
+
+\begin{align} \label{eq:lratio}
+P(x|y) &= \frac{\Lambda(x|y)}{\Lambda(x|y) + \frac{P(\neg x)}{P(x)}}
+\end{align}
+
+Where we have used the \textit{likelihood ratio} $\Lambda(x|y) \equiv \frac{P(y|x)}{P(y|\neg x)}$ to simplify the expression.
+The likelihood ratio has a simple interpretation, it is simply the ratio of the probability of event $y$ occuring in the case event $x$ occurs and the probability of event $y$ occuring given that event $x$ \textit{does not} occur.
+In fact, this form of Bayes theorem indicates that the posterior probability for $x$ given $y$ is directly proportional to the likelihood ratio.
+This is particularly important in the Bayesian treatment of gravitational waves where the central question for study is the probability of observing the data in question given that there may or may not be a signal present.
+
+\subsection{Signal Processing}
+Noise in the LIGO detectors is modeled as \textit{Gaussian Noise}, which describes a length $N$ set of time series samples $\{x_j\}$ with the following probability of observation:
+
+\begin{align}
+P(\{x_j\}) = \left[\frac{1}{\sqrt{2 \pi} \sigma}\right]^N e^{-\frac{1}{2 \sigma^2} \sum_{j = 0}^{N - 1}x_j^2}
+\end{align}
+
+If we allow the $\{x_j\}$ to be weighted by the noise sensitivity in frequency space (also known as the \textit{power spectral density}) $S(f) = \sum_{k=0}^K n^*(f_k)n(f_k) \equiv \braket{n(f)|n(f)}$ then we may write the probability density for any Gaussian noise process as
+
+\begin{align}
+P_x(\{x(t)\}) \propto e^{-\frac{1}{2}\braket{x(t)|x(t)}}
+\end{align}
+
+For gravitational wave signal processing we are interested in evaluating the equivalent of the likelihood ratio  in equation \ref{eq:lratio} for the null and signal present hypotheses, $\mathcal{H}_0$ and $\mathcal{H}_1$ respectively.
+Following the treatment outlined in literature \cite{rapidpe} This is
+
+\begin{align}
+\Lambda(\mathcal{H}_1|d) &= \frac{P(d|\mathcal{H}_1)}{P(d|\mathcal{H}_0)}
+\end{align}
+
+In the above equation $\mathcal{H}_0$ is in direct correspondence with $\neg x$ and $\mathcal{H}_1$ is in direct correspondence with $x$.
+If there is no signal present in the data, then the data and noise are equal, meaning
+
+\begin{align}
+d(t) &= n(t) \\
+P(d(t)|\mathcal{H}_0) &\propto e^{-\frac{1}{2}\braket{d(t)|d(t)}}
+\end{align}
+
+Whereas if there is a signal present in the data, then the noise is the \textit{residual} signal after the gravitational wave potion of the signal $h(t)$ is subtracted from the total signal $d(t)$:
+
+\begin{align}
+d(t) &= n(t) + h(t) \\
+p(d(t)|\mathcal{H}_1) &\propto e^{-\frac{1}{2}\braket{d(t) - h(t)|d(t) - h(t)}}
+\end{align}
+
+So the likelihood ratio for gravitational wave signal processing is exactly equal to
+
+\begin{align}
+\Lambda(\mathcal{H}_1|d(t)) &= \frac{e^{-\frac{1}{2}\braket{h(t)-d(t)|h(t)-d(t)}}}{e^{-\frac{1}{2}\braket{d(t)|d(t)}}}
+\end{align}
+
+By equation \ref{eq:bayes}  the posterior probability of a gravitational wave signal with parameters $\mu$ is given by
+
+\begin{align}\label{eq:parampos}
+P[\vec{\mu}, \mathcal{H}_1|d(t)] = \frac{P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})}{P[d(t), \mathcal{H}_1]}
+\end{align}
+
+Extending equation \ref{eq:continuous} to the case of a continuous state space we have for the denominator
+
+\begin{align}
+P[d(t), \mathcal{H}_1] &= \int_{\Omega}P[d(t)|\vec{\mu}, \mathcal{H}_1]P(\vec{\mu})d\vec{\mu}
+\end{align}
+
+Where $\Omega$ indicates integration over the entire probabilistic state space.
+Dividing both the top and bottom of the right hand side of equation \ref{eq:parampos} by $P(d(t)|\vec{\mu}, \mathcal{H}_1)$ gives us
+
+\begin{align}\label{eq:bayeslikely}
+P(\vec{\mu}, \mathcal{H}_1|d(t)) = \frac{\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})}{\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)P(\vec{\mu})d\vec{\mu}}
+\end{align}
+
+Note that the integral in the denominator may be factored into components that are functions of independent groups of parameters.
+For the purposes of gravitational wave data analysis it is factored into intrinsic and extrinsic parameters, the former (masses, spins) being responsible for the actual orbital dynamics of the system, and the latter (distance, inclination, right ascension, declination) depending only on the nature of the observer relative to the source.
+We label these sets of paramters $\vec{\lambda}$ and $\vec{\theta}$ respectively.
+The denominator of equation \ref{eq:bayeslikely} is thus factorizable as
+
+\begin{align}\label{eq:sepparam}
+\int_{\Omega}\Lambda(d(t)|\vec{\mu}, \mathcal{H}_1)d\vec{\mu} &= \int_{\Omega}\Lambda(d(t)|\vec{\lambda}, \mathcal{H}_1)P(\vec{\lambda})d\vec{\lambda}\int_{\Omega}\Lambda(d(t)|\vec{\theta}, \mathcal{H}_1)P(\vec{\theta})d\vec{\theta}
+\end{align}
+
+Note that while evaluating the likelihood for a single set of parameters in the numerator of the above equation may not be any serious challenge, computing the integral in the denominator is.
+Many schemes \cite{rapidpe} \cite{lalinference} \cite{mcmc} have been developed to tackle this particular problem and generally involve simulating the posterior through MCMC sampling.
+These techniques are highly developed and at the time of writing can provide converged results on the order of days, however they are limited in their parallelism due fundamental serialization of the algorithm.
+Cutting edge implementations involving parallel tempering and some forms of ensemble MCMC have shown promising results, however none can take advantage of the many thousands of cores availble across the LIGO data grid.
+Methods that can scale out to the computational resources provided by the LIGO data grid are needed to reduce runtimes to minutes or less.
+
+\subsection{Harmonic Mode Decomposition}
+
+For a single detector $k$ We can write our gravitational wave signal $h_k(t) = h_{+,k}(t) - ih_{\times, k}(t)$ as a sum over products of harmonic mode strain values $h_{l,m}(\vec{\lambda})$ and corresponding spin-weighted spherical harmonics, scaled by the reference distance and antenna factor:
+
+\begin{align}
+h_k(t) = \frac{D_r}{D}F_k(\theta)\sum_{(l,m)}h_{k, (l,m)}(t, \vec{\lambda})Y_{(l,m)}(\vec{\theta})
+\end{align}
+
+Where $D_r$ is the luminosity reference distance to the binary.
+This is defined similarly to the manner in which apparent magnitude is defined versus absolute magnitude in astonomy.
+We substitute this expression for the signal into the expression for the likelihood ratio we find, taking the logarithm, that
+
+\begin{align}
+\mathcal{L} &= \prod_k e^{-\frac{1}{2}\braket{d - h_k|d-h_k} + \braket{d|d}} \\
+\ln\mathcal{L} &= -\frac{1}{2}\sum_{k}\left[\braket{d - h_k|d-h_k} + \braket{d|d}\right] \\
+&= \frac{1}{2}\sum_{k}\left[-\cancel{\braket{d|d}}+\braket{d|h_k}+\braket{h_k|d}+\braket{h_k|h_k} + \cancel{\braket{d|d}}\right] \\
+\end{align}
+
+Now, since for complex vector spaces $\braket{a|b} = \braket{b|a}*$, we have that $\braket{d|h_k} + \braket{h_k|d} = 2\Re\braket{h_k|d}$. Then,
+
+\begin{align}
+\ln\mathcal{L} &= \frac{1}{2}\sum_{k}\left[2\braket{h_k|d} + \braket{h_k|h_k}\right] \\
+&= \frac{1}{2} \sum_{k}\Re\left[\braket{2\frac{D_r}{D}F_k(\vec{\theta})\sum_{(l,m)}h_{k,(l,m)}(\lambda)Y_{(l,m)}(\vec{\theta})|d}\right] \\
+&+ \left[\frac{D_r}{D}\right]^2F_k^2(\vec{\theta})\sum_{(l,m),(l',m')}h^*_{(l,m)}(\vec{\lambda})h_{(l',m')}(\vec{\lambda})Y^*_{(l,m)}(\vec{\theta})Y_{(l',m')}(\vec{\theta})
+\end{align}
+
+\subsection{RapidPE}
+
+The defining feature of the factored form of the likelihood is the separation between intrinsic and extrinsic parameters.
+The intrinsic parameters of the binary exist only within the $h_{k,(l,m)}$, whereas the parameters that depend on the observer are contained entirely within the antenna factor and spherical harmonics.
+Work by \cite{rapidpe} (RapidPE) has shown that this can be exploited to accelerate computation of the likelihood by precomputing the orbital dynamics of the binary, which are functions only of the intrinsic parameters, and marginalizing over the extrinsic parameters.
+The same work computes the posterior over the \textit{intrinsic} parameter space by spreading a grid of points across the $\mathcal{M}, \eta$ plane and computing the integral at each such point, since each of these integrations is independent each point for which the value of the posterior is desired may be assigned its own computational resources.
+This enables the necessary scaling - the posterior between the points is simply interpolated after all the cores are done.
+
+\subsection{Monte Carlo Integration}
+Numerical evaluation of integrals $I$ of the form
+
+\begin{align}
+I = \int_{a}^{b}f(\vec{x})d^n\vec{x} 
+\end{align}
+
+over $n$ dimensions are generally tackled through a form of Monte-Carlo integration.
+Basic Monte-Carlo integration involves drawing $N$ random samples $\{x_i\}$ from $\Omega$ and approximating the answer to the integral as
+
+\begin{align}
+\frac{\Omega}{N}\sum_{i=0}^N f(x_i) \equiv \Omega\braket{f}
+\end{align}
+
+A brute force integration of the extrinsic portion of equation \ref{eq:sepparam} is may be computationally intractable due to the dimensionality of the problem, so modern schemes employ more advanced, adaptive versions of this process that employ importance sampling to reduce the total number of samples required.
+In particular the LePage 1978 algorithm \cite{montecarlo} is employed to compute the extrinsic integral.
+LePage's solution is to "integrate" the function $g(x) \equiv \frac{f(\vec{x})}{q(\vec{x})}$ where $q(\vec{x})$ is some probability distribution.
+
+\begin{align}
+I \sim \frac{\Omega}{N}\sum_{i = 0}^N g(x_i), \ \ x_i \sim q(\vec{x})
+\end{align}
+
+It can be shown that the result is the same if the $\{x_i\}$ are \textit{drawn} from $q(\vec{x})$.
+The effect of this is to offset the answer in an opposite sense as integrating $g(\vec{x})$ instead, which means if $q(\vec{x})$ is selected intelligently, the the algorithm has the potential to converge more quickly.
+For practical purposes, intelligent selection of the integrand corresponds to picking $g(\vec{x})$ that has higher value where $f(x)$ is expected to itself be higher.
+In RapidPE, this concept is applied iteratively.
+A group, or chunk, of samples are drawn, and their functional value is computed.
+The function values are then used to form $q(\vec{x})$ for the next chunk.
+In this way the algorithm adapts to sample the "important" regions more effectively each time.
+The point of this review of Monte-Carlo integration is to illustrate the reliance of numerical integration on large groups of function evaluations.
+If the function evaluations themselves are expensive, then intelligent integration schemes can serve only to push the bottleneck elsewhere.
+The purpose of this work is to alleviate this particular issue in the RapidPE process, and produce as many actual samples of the reduced integrand as possible in a short amount of time.
+
+\subsection{Precessional Effects}
+This work aims to extend and accelerate the ability for RapidPE to estimate gravitational wave parameters.
+RapidPE is written in Python, and prior to this work significant attention was given to code optimization to obtain the maximum performance possible through Python.
+Since the terms of the likelihood that may be precomputed involve computationally expensive fourier transforms, calculation of the cross terms $V$ were approximated using the properties of fourier transform to relate

### 2016-08-28 Gradschool/thesis/final/chap2.tex
expanded implementation chapter, merged bayesian statistics and GR chapter, relegated GR derivation to appendix

--- 2016-08-28 Gradschool/thesis/final/bak/chap3.tex
@@
 
-Where $\vec{X}$ and $\vec{Y}$ are the wave-frame axes. These can be redefined in terms of the working parameters as
+Where $\vec{X}$ and $\vec{Y}$ are the wave-frame axes. These can be redefined \cite{antennafacdefn} in terms of the working parameters as
 
 \begin{align}
-\text{PUT DEFINITIONS HERE}
+\vec{X} &=
+\begin{bmatrix}
+-\cos(\psi)\sin(h) - \sin(\psi)\cos(h)\sin(\delta) \\
+-\cos(\psi)\cos(h) + \sin(\psi)\sin(h)\sin(\delta) \\
+\sin(\psi)\cos(\delta)
+\end{bmatrix}
+, \ \ \vec{Y} = 
+\begin{bmatrix}
+\sin(\psi)\sin(h) - \cos(\psi)\cos(h)\sin(\delta) \\
+\sin(\psi)\cos(h) + \cos(\psi)\sin(h)\sin(\delta) \\
+\cos(\psi)\cos(\delta)
+\end{bmatrix}
 \end{align}
 
+In the working Python implementation of RapidPE the computation of this quantity is serialized with respect to the samples.
 Our objective is to produce this quantity for a group of samples, all at once, as a coherent vectorized operation.
@@
 F^{i}_{\times} &= X^{lm}R^{i}_{lj}Y^{j}_{m} + Y^{lm}R^{i}_{lj}X^{j}_{m} \\ 
+\end{align}
+
+\subsection{Vectorized Detector Arrival Time}
+
+The use of the geocentric signal arrival time in the above computation of the complex antenna factors $F_+$ and $F_\times$ is equivalent to making the approximation that the detectors are all physically located at the geocenter (albeit maintaining their varying orientations).
+Here we present a vectorized method for computing the more realistic temporal offsets resulting from the physical placement of the detectors on the surface of the earth.
+Let $\mathbf{A} \circ \mathbf{B}$ denote the Hadamard product of tensors, defined such that
+
+\begin{align}
+C_{ij} = A_{ij}B_{ij} \equiv (\mathbf{A} \circ \mathbf{B})_{ij}
+\end{align}
+
+Given a vector $\vec{t_{geo}}$ the components of which are all exactly the Greenwich mean sidereal arrival time at geocenter, let $\vec{h} = \vec{t_{geo}} - \vec{\alpha}$ be a vector containing the conventionally defined hour angles for the samples in question, as well as vectors $\vec{\alpha}$ and $\vec{\delta}$ denoting the sample right ascensions and declinations respectively.
+The normal vector oriented in the sample source direction is then given by
+
+\begin{align}
+\vec{V_{s}} = 
+\begin{bmatrix}
+\cos(\vec{\delta}) \circ \cos(\vec{h}) \\
+-\cos(\vec{\delta}) \circ \sin(\vec{h}) \\
+\sin(\vec{\delta})
+\end{bmatrix}
+\end{align}
+
+The above vector is relative to the standard celestial coordinate system for which the north pole is aligned along the $z$-axis and the $x$ axis points towards the Vernal equinox.
+In the same coordinate system, the vector pointing towards the detector in question $\vec{V_d}$ is available in a standard lookup table within the LAL software library.
+The sample arrival times $\vec{t_a}$ are then
+
+\begin{align}
+\vec{t_a} = \vec{t_{geo}} - \frac{1}{c}\vec{V_s} \cdot \vec{V_d}
 \end{align}

### 2017-01-09 tex/hwtemplate.tex
generic template

--- 2016-06-10 Gradschool/electrodynamics/em_hw3.tex
@@
 \usepackage{braket}
-\usepackage{cancel}
+
 \usepackage{graphicx}
-\usepackage{float}
 
@@
 \normalfont \normalsize 
-\textsc{Northwestern University} \\ [25pt] % Your us get downniversity, school and/or department name(s)
+\textsc{Retention and Customer Experience} \\ [25pt] % Your us get downniversity, school and/or department name(s)
 \horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
-\huge Electrodynamics - Homework 3 \\ % The assignment title
+\huge Know Your Customers \\ % The assignment title
 \horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
@@
 
-\author{Brandon B. Miller} % Your name
+%\author{Brandon B. Miller} % Your name
 
@@
 
-\section{Problem 1}
+\newpage
 
-\subsection{Potential and Image Potential}
+\section{Geographical Location}
+The following is a map of the geographical distribution of some random variable over the contiguous United States.
+Zip codes are not defined as restricted to a specific spatial boundary, so data are binned by zipcode tabulation area (ZCTA) - shapes developed by the US Census Bureau to represent zip codes as closely as possible.
+The mapping used to defined the areas in this figure is sourced from the USCB webpage.
 
-We are asked to investigate the inverse image charge problem, this time on the interior region of the sphere.
-We approach the problem as any standard image charge problem, and set the known charge $q$ a distance $d$ along the $\theta = 0$ axis with the image charge $q'$ at some unknown distance $x$.
-Requiring that the potential be zero at the marked points leads us to two equations in two unknowns:
-
-\begin{align}
-\frac{1}{4 \pi \epsilon_0}\frac{q}{R - d} + \frac{1}{4 \pi \epsilon_0}\frac{q'}{x - R} &= 0 \\
-\frac{1}{4 \pi \epsilon_0}\frac{q}{R + d} + \frac{1}{4 \pi \epsilon_0}\frac{q'}{x + R} &= 0
-\end{align}
-
-Solving these for $q'$ and $x$ leads to the familiar result as marked in the figure:
-
-\begin{align}
-x &= \frac{R^2}{d} \\
-q' &= -\frac{q R}{d}
-\end{align}
-
-This means that we can write the potential as the superposition of two terms, one from the real charge and the other from the image charge.
-Using the standard formula $\frac{1}{\vec{r} - \vec{r'}} = (r^2 + r'^2 - 2rr'\cos(\theta))^{-\frac{1}{2}}$, we have
-
-\begin{align}
-\Phi = \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{r^2 + d^2 - 2 r d \cos(\theta)}} - \frac{\frac{qR}{d}}{\sqrt{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}}\right] \\
-\Phi = \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{r^2 + d^2 - 2 r d \cos(\theta)}} - \frac{q}{\sqrt{\frac{r^2d^2}{R^2} + R^2 - 2 r d\cos(\theta)}}\right]\\ 
-\end{align}
-
-The desired result.
-
-\subsection{Surface Charge Density}
-
-The boundary conditions on the electric field require that the discontinuity in $\vec{E}$ be proportional to $\sigma$, or equivalently, that $\sigma = -\epsilon_0 \frac{\partial \Phi}{\partial r}$.
-We need a derivative of the potential above.
-
-\begin{align}
--4 \pi \epsilon_0 \frac{\partial \Phi}{\partial r} &= \frac{q(2r - 2d\cos(\theta))}{2(r^2 + d^2 - 2 r d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2rd^2}{R^2} - 2 d \cos(\theta)\right)}{2\left(\frac{r^2d^2}{R^2} + R^2 - 2 r d \cos(\theta)\right)^{\frac{3}{2}}} \\
-\rightarrow & \text{At $r=R$, this becomes} \rightarrow \\
-&= \frac{q(2R - 2d\cos(\theta))}{2(R^2 + d^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2Rd^2}{R^2} - 2 d \cos(\theta)\right)}{2\left(\frac{R^2d^2}{R^2} + R^2 - 2 R d \cos(\theta)\right)^{\frac{3}{2}}} \\
-&= \frac{q(2R - 2d\cos(\theta))}{2(R^2 + d^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} - \frac{q\left(\frac{2}{R}d^2 - 2 d \cos(\theta)\right)}{2\left(d^2 + R^2 - 2 R d \cos(\theta)\right)^{\frac{3}{2}}} \\
-&= \frac{q(R - \frac{d^2}{R})}{(d^2 + R^2 - 2 R d \cos(\theta))^{\frac{3}{2}}} \\
-\end{align}
-
-Multiplying both top and bottom by $R$ and rearranging we have
-
-\begin{align}
-\sigma &= -\frac{1}{4 \pi} \left[\frac{q(R^2 - d^2)}{R(d^2 + R^2 - 2 R d \cos(\theta))^{\frac{3}{2}}}\right] \\
-\end{align}
-
-\subsection{Forces}
-
-The force can be found directly from Coulomb's law taking the distance between the charges as $|\vec{r} - \vec{r'}|$:
-
-\begin{align}
-|F| &= \frac{1}{4 \pi \epsilon_0}\frac{q_1 q_2}{|\vec{r}-\vec{r'}|^2} \\
-&= \frac{1}{4 \pi \epsilon_0}\frac{-\frac{q^2 R}{d}}{(\frac{R^2}{d} - d)^2} \\
-&= -\frac{q^2}{4 \pi \epsilon_0} \frac{1}{(\frac{d}{R})(\frac{R^4}{d^2} - 2R^2 + d^2)} \\
-&= -\frac{q^2}{4 \pi \epsilon_0} \frac{1}{\left(\frac{R^3}{d} - 2 Rd + \frac{d^3}{R}\right)}
-\end{align}
-
-\subsection{Modifications}
-Like the previous homework, adding extra charge to a system that has already equilibrated through the image charge does not change the situation.
-The new added charge would have no reason to be biased towards any region of the sphere, since the surface is maintained already as an equipotential by the charge and image charge pair.
-Similarly adding a potential does not change any properties of the system as it does not affect the gradient of the potential in any location, and thus does not effect the overall dynamics of the system.
-
-\section{Problem 2}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[scale=0.5]{doubleline_geom.eps}
-\end{center}
+\begin{figure}[h!]
+\centering
+\includegraphics[trim={3.75cm, 1cm, 0, 0.60cm}, scale=0.7]{sixteennine.eps}
 \end{figure}
 
-We place two line charges a distance $a$ from the $x-$axis in the $xy$ plane as shown.
-Per the geometry in the figure, the two distances to the line charges, $r_+$ and $r_-$ are given as the following:
-
-\begin{align}
-r_+ &= \sqrt{(y-a)^2 + z^2} \\
-r_- &= \sqrt{(y+a)^2 + z^2}  
-\end{align}
-
-The known potential from a line charge is given by $V = \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{r}{r_0}\right]$, where $r_0$ is the reference distance to the zero point of the potential, which is equal for both wires due to symmetry.
-The superposition of the potentials from the two wires then is
-
-\begin{align}
-\Phi &= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{\sqrt{(y-a)^2 + z^2}}{r_0}\right] - \frac{\lambda}{2\pi\epsilon_0}\ln\left[\frac{\sqrt{(y+a)^2 + z^2}}{r_0}\right] \\
-&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\sqrt{\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}}\right] \\
-&= \frac{\lambda}{4 \pi \epsilon_0}\ln\left[\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}\right]
-\end{align}
-
-A specific equipotential is found by setting the above formula equal to a constant, $\Phi_0$.
-We absorb all the available constants into the left hand side and exponentiate, defining $C = e^{\frac{4 \pi \epsilon_0 \Phi_0}{\lambda}}$, yielding
-
-\begin{align}
-\left[\frac{(y-a)^2 + z^2}{(y+a)^2 + z^2}\right] &= C \\
-y^2 - 2 y a + a^2 + z^2 &= C(y^2 + 2ya + a^2 + z^2) \\
-y^2(1-c) - 2ya(1+c) + a^2(1-c) + z^2(1-c) &= 0 \\
-(1-c)(y^2 + z^2 + a^2) &= 2ya(1+c)\\
-y^2 + z^2 + a^2 = 2ya\left(\frac{1+c}{1-c}\right)
-\end{align}
-
-This corresponds to the equation of a circle $(y-y_0)^2 + z^2 = R^2 \rightarrow y^2 + z^2 - 2yy_0 + a^2 = R^2$ with $y_0 = a\left(\frac{1+c}{1-c}\right)$. Consider the following:
-
-\begin{align}
-(y - y_0)^2 &= y^2 - 2yy_0 + y_0^2 \\
-\end{align}
-
-Whereas our equation reads
-\begin{align}
-y^2 + z^2 + a^2 - 2ya\left(\frac{1+c}{1-c}\right) = 0
-\end{align}
-
-It would seem that if we add the quantity $a^2 \left(\frac{c+1}{c-1}\right)^2$ to both sides we can identify $y_0 = a\left(\frac{c+1}{c-1}\right)$ and do the following:
-
-\begin{align}
-y^2 - 2y\left(\frac{c + 1}{c-1}\right)a + a^2\left(\frac{c+1}{c-1}\right)^2 + a^2 + z^2 &= a^2 \left(\frac{c+1}{c-1}\right)^2 \\
-\left(y - a\left(\frac{c+1}{c-1}\right)\right)^2 + z^2 &= a^2\left(\frac{c+1}{c-1}\right)^2 - a^2 
-\end{align}
-
-These are circles with radii that are the square root of the right hand side, recalling that $c$ contains a dependence on the equipotential in question.
-
-\begin{align}
-R &= \sqrt{a^2\left(\frac{c+1}{c-1}\right)^2 - a^2} \\
-y_0 &= a\left(\frac{c+1}{c-1}\right) 
-\end{align}
-
-\section{Problem 3}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[scale=0.4]{cyl_geom.eps}
-\end{center}
+\begin{figure}[ht!]
+\centering
+\centering
+\includegraphics[trim={0, 1cm, 0, 1cm}, width=.3\textwidth]{square.eps}\hfill
+\includegraphics[trim={0, 1cm, 0, 1cm}, width=.3\textwidth]{square.eps}\hfill
+\includegraphics[trim={0, 1cm, 0, 1cm}, width=.3\textwidth]{square.eps}
+\caption{Of particular interest is the \textit{donut effect} in some of Allstate's most popular cities, possibly reflecting thinner car ownership in strictly urban areas.}
 \end{figure}
 
-\subsection{Part A - Greens Function}
+\section{Customer Age}
+The distribution of some andom variable years tied to some random company. The modes are roughly at the modes. The lower \textbf{FILL THIS IN} of the data is not shown. Note that these
 
-In the context of an infinite cylinder, the fundamental unit of charge is the infinite line charge.
-This is analogous to the point charge in the case of the sphere.
-As shown in the figure, we have a conceptually similar situation, in which we place such a line charge at $r<R$ and $\theta = 0$.
-This induces the equivalent of an \textit{image line charge} outside the boundary.
+\begin{figure}[h!]
+\centering
+\includegraphics[trim={2cm, 1cm, 0, 0.60cm}, scale=0.6]{sixteennine.eps}
+\includegraphics[trim={2cm, 1cm, 0, 0.60cm}, scale=0.6]{sixteennine.eps}
+\end{figure}
 
-\hspace{2mm}
+\section{Customer Tenure}
 
-If one looks down upon the top of the cylinder, one sees a circular cross section, in which the problem of finding the distance to the image line charge is geometrically equivalent to that of finding the distance to the image \textit{point charge} in the spherical case.
-Thus if the known distance to the real line charge is denoted $d$, the distance to the image line charge must be $d' = \frac{R^2}{d}$.
-
-\hspace{2mm}
-
-The following is the potential, found from Gauss's law, of an infinite line charge for a test particle a perpandicular distance $x$ from the line charge itself:
-
-\begin{align}
-\Phi = \frac{\lambda}{2 \pi \epsilon_0}\ln\left(\frac{r}{r_0}\right) \\
-\end{align}
-
-Where $r_0$ is the reference distance to the zero point of the potential.
-Our total potential is composed of two components, one from the real wire, and one from the image wire.
-We note that the reference distance to the zero point of the potential
-
-\begin{align}
-\Phi &= \Phi_+ + \Phi_- \\
-&= \frac{-\lambda}{2 \pi \epsilon_0}\ln\left(\frac{r_+}{d}\right) + \frac{(\lambda)}{2 \pi \epsilon_0}\ln\left(\frac{r_-}{d'}\right)  \\
-\rightarrow &  \ \ \text{(Inverted Negatives from definition of potential)} \\
-&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left(\frac{dr_-}{d'r_+}\right) \\
-&= \frac{\lambda}{2 \pi \epsilon_0}\ln\left[\frac{d^2}{R^2}\sqrt{\frac{r^2 + d'^2 - 2 r d' \cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}}\right] \\
-&= \frac{\lambda}{4 \pi \epsilon_0}\ln\left[\left(\frac{d^4}{R^4}\right)\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d} \cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right]
-\end{align}
-
-Dropping the coupling constants and the charge density we find the form of the Green's function for the cylinder:
-
-\begin{align}
-G &= \frac{1}{4 \pi \epsilon_0} \ln\left[\left(\frac{d^4}{R^4}\right)\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right] \\
-\end{align}
-
-The general form of the solution to the potential in some given region of interest, from Jackson, is
-
-\begin{align}
-\cancel{\frac{1}{4 \pi \epsilon_0}\int_v \rho(\vec{r},\vec{r'})G(\vec{r},\vec{r'})dv} + \frac{1}{4 \pi}\int_s \phi(r',\theta',z') \frac{\partial G}{\partial n'}ds'
-\end{align}
-
-With the Green's function in hand we consider the problem of an arbitrary potential distribution $\phi$ on the surface of the cylinder.
-In the above solution, the first term cancels since there is no $\rho$ in the region of interest.
-We are left with an integral term and the need to differentiate the Green's function with respect to $\hat{r}$, since the normal vector points outwards from the inside of the cylinder:
-
-\begin{align}
-4\pi\frac{\partial G}{\partial r} &= \left(\frac{d^2}{R^2}\right)\left[\frac{r^2 + d^2 - 2 r d \cos(\theta)}{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}\right]\frac{\partial}{\partial r} \left[\frac{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}{r^2 + d^2 - 2 r d \cos(\theta)}\right] \\
-&= \left(\frac{d^2}{R^2}\right)  \left[\frac{\cancel{r^2 + d^2 - 2 r d \cos(\theta)}}{r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta)}\right]  \\ &  \times  \left[\frac{(r^2 + d^2 - 2 r d \cos(\theta))(2r - 2 \frac{R^2}{d}\cos(\theta)) - (r^2 + \frac{R^4}{d^2} - 2 r \frac{R^2}{d}\cos(\theta))(2r - 2d \cos(\theta))}{(r^2 + d^2 - 2 r d \cos(\theta))^{\cancel{2}}}\right]
-\end{align}
-
-Evaluating the resulting expression at $r = R$ gives us
-
-\begin{align}
-&= \left(\frac{d^2}{R^2}\right)   \left[\frac{(R^2 + d^2 - 2 R d \cos(\theta))(2R - 2 \frac{R^2}{d}\cos(\theta)) - (R^2 + \frac{R^4}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))(2R - 2d \cos(\theta))}{(R^2 + d^2 - 2 R d \cos(\theta))(R^2 + \frac{R^2}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))}\right] \\
-%&= \frac{(2R^3 - 2 \frac{R^4}{d}\cos(\theta) + 2Rd^2 - 2R^2d\cos(\theta) - 4 R^2 d\cos(\theta) + 4 R^3 \cos^2(\theta))}{(R^2 + d^2 - 2 R d \cos(\theta))(R^2 + \frac{R^2}{d^2} - 2 R \frac{R^2}{d}\cos(\theta))} \\
-%- \frac{(2R^3 - 2 R^2 d \cos(\theta) + 2\frac{R^5}{d^2} - 2\frac{R^4}
-&= \frac{1}{R}\left[\frac{d^2 - R^2}{d^2 + R^2 - 2Rd\cos(\theta)}\right]
-\end{align}
-
-Plugging this result into the integral solution for $\Phi$ and noting that the integral over the surface \textit{along a cross section} of a cylinder is $R d\theta'$, we have
-
-\begin{align}
-\Phi = \frac{1}{2\pi}\int_s \phi(R, \theta')\frac{R^2 - r^2}{R^2 + r^2 - 2 R r \cos(\theta)}d\theta' 
-\end{align}
-
-Where we plugged in $d \rightarrow r$ due to the symmetry in the Green's function and the choice of differentiation variable.
-This is the exact same as the result we hoped to derive with $\theta_0 = 0$, or aligning the $\theta=-$ line along the axis through both line charges.
-Note that switching the terms in the numerator is allowed us to absorb the negative sign from the definition of the solution into the expression.
-
-\subsection{Part B - Split Cylinder}
-
-We wish to find the potential inside a split cylinder of the type pictured above.
-By gauge freedom, we may write the potential as the sum of a constant term plus an infinite series that we demand obeys Laplace's equation.
-For convenience, we set this constant term to be the one we need, namely
-
-\begin{align}
-\Phi = \frac{V_1 + V_2}{2} + \Phi' \\
-\rightarrow \Phi' \ \text{Solves Laplace's Equation} \\
-\end{align}
-
-But we require that $\Phi = V_1$ for $0 < \theta < \pi$ and $\Phi = V_2$ for $\pi < \theta < 2 \pi$.
-This is equivalent to requiring that $\Phi' = \frac{V_2 - V_1}{2}\text{Sign}(\theta)$ Where the \textit{Sign} function is defined as returning $1$ if $ 0 < \theta < \pi$ and $-1$ if $\pi < \theta < 2 \pi$.
-To verify that this must be correct, we examine the two cases of interest:
-
-\begin{align}
-0 < \theta < \pi: & V = \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{2} = V_1 \\
-\pi < \theta < 2\pi: & V = \frac{V_1 + V_2}{2} - \frac{V_1 - V_2}{2} = V_2 
-\end{align}
-
-In general we could expand the Sign function as an infinite mixed sine cosine series $\sum_{n=0}^{\infty} a_n r^n \sin(n \theta) + b_n r^n \cos(n \theta)$, however since the Sign function is purely odd, we know there can't be any cosine terms.
-Furthermore, if we require that $V(\theta) = V(\pi - \theta)$, we see that we can only have odd terms in the summation itself.
-This leaves us with
-
-\begin{align}
-\Phi = \frac{V_1 + V_2}{2} + \sum_{n \ odd}^{\infty}a_n r^n \sin(n \theta)
-\end{align}
-
-To get the $a_n$'s, we take the case $r = b$, multiply both sides by $\sin(m \theta)$ and integrate from $-\pi$ to $\pi$:
-
-\begin{align}
-\frac{V_1 - V_2}{2}\text{Sign}(\theta) &= \sum_{n \ odd}^{\infty}a_n b^n \sin(n \theta) \\
-\frac{V_1 - V_2}{2}\int_{-\pi}^{\pi} \text{Sign}(\theta)\sin(m \theta)d\theta &= \sum_{n \ odd}^{\infty}a_n b^n \sin(n \theta)\sin(m \theta)d\theta \\
-\end{align}
-
-Noting here that $\text{Sign}(\theta) \sin(m \theta) = |\sin(m \theta)|$, intuitive since the product of two odd functions is even, we may double the integrand on the right and take it from $0$ to $\pi$:
-
-\begin{align}
-(V_1 - V_2) \int_0^{\pi}\sin(m \theta)d\theta &= \pi a_n b^n \delta_{mn} \\
-\frac{2(V_1 - V_2)}{n} &= a_n b^n \pi \\
-a_n &= \frac{2(V_1 - V_2)}{n \pi b^n} \\
-\end{align}
-
-So we can write our solution now as
-
-\begin{align}
-\Phi &= \frac{V_1 + V_2}{2} + 2 \frac{V_1 - V_2}{\pi} \sum_{n \ odd}^{\infty} \frac{r^n}{nb^n}\sin(n \theta) \\
-\end{align}
-
-Applying Euler's formula, we note that $\sin(n \theta) = \Im [re^{i n \theta}]$ where $\Im$ denotes the imaginary part of an expression.
-Therefore $r^n \sin(n \theta) = \Im[r^n e^{i n\theta}]$.
-With that, we have
-
-\begin{align}
-\Phi = \frac{V_1 + V_2}{2} + 2 \frac{V_1 - V_2}{\pi} \Im \left[\sum_{n \ odd}^{\infty} \frac{1}{n}\left(\frac{r e^{i \theta}}{b}\right)^n \right]
-\end{align}
-
-Using now the identity that kids learn in kindergarten $\sum_{n odd}^{\infty} \frac{z^n}{n} = \frac{1}{2}\ln\frac{1 + z}{1-z}$, we have
-
-\begin{align}
-\Phi &= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 + \frac{r}{b}e^{i \theta}}{1 - \frac{r}{b}e^{i \theta}}\right] \\
-&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \frac{r}{b}e^{i \theta}}{1 - \frac{r}{b}e^{i \theta}}\right]\left[\frac{1 - \frac{r}{b}e^{-i \theta}}{1 - \frac{r}{b}e^{-i \theta}}\right] \\
-&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2 + \frac{r}{b}(e^{i \theta} - e^{- i \theta})}{1 + \left(\frac{r}{b}\right)^2 - \frac{r}{b}(e^{i \theta} + e^{-i \theta})}\right]\\
-&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi} \Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2 + \frac{r}{b}(2i)(\sin(\theta))}{1 + \left(\frac{r}{b}\right)^2 - \frac{r}{b}(2)(\cos(\theta))}\right]\\
-&= \frac{V_1 + V_2}{2} +  \frac{V_1 - V_2}{\pi}\Im\ln\left[\frac{1 - \left(\frac{r}{b}\right)^2}{1 + \left(\frac{r}{b}\right)^2 - 2 \frac{r}{b}\cos(\theta)} + \frac{2 i \frac{r}{b}\sin(\theta)}{1 + \left(\frac{r}{b}\right)^2 - 2 \frac{r}{b}\cos(\theta)}\right]
-\end{align}
-
-In the complex plane, we must have that $ln(re^{i \theta}) = ln(r) + i \theta$, so the imaginary part of the log of a complex number is that complex number's phase.
-This makes sense when one thinks about the imaginary axis of the complex plane.
-Noting that the phase of a complex number is the inverse tangent of the imaginary part over the real part, we arrive at
-
-\begin{align}
-\Phi &= \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{\pi}\tan^{-1}\left[\frac{\left(\frac{r}{b}\right) \sin(\theta)}{1 + \left(\frac{r}{b}\right)^2}\right] \\
-&= \frac{V_1 + V_2}{2} + \frac{V_1 - V_2}{\pi}\tan^{-1}\left[\frac{2 r b \sin(\theta)}{b^2 - r^2}\right]
-\end{align}
-
-Which is the same answer as what we hoped for sans the triginometric function which is different simply due to the fact that we measured $\theta$ from the plane of separation instead of using the cumbersome convention in the problem description.
-
-\section{Problem 4}
-
-We are asked to investigate the Diriclet problem consisting of an infinite plane, and a region of interest defined by $z > 0$ with the surface placed at $z=0$.
-We will find the Green's function using the elementary image charge method.
-
-\hspace{2mm}
-
-From our experience with this problem we know that a charge $q$ placed a location $z_0$ above the $xy$ plane will be balanced by an image charge of magnitude $-q$ a distance $z_0$ \textit{below} the $xy$ plane.
-Thus if we define an arbitrary point $x_0, y_0, z_0$ as the field point, the potential due to \textit{both} the real and image charges is
-
-\begin{align}
-\Phi &= \frac{1}{4 \pi \epsilon_0}\left[\frac{q}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z - z_0)^2}} - \frac{q}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z + z_0)^2}}\right] \\
-\end{align}
-
-Dropping the $q's$ and coupling constant we have the Green's function
-
-\begin{align}
-G = \frac{1}{4 \pi \epsilon_0}\left[\frac{1}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z - z_0)^2}} - \frac{1}{\sqrt{(x-x_0)^2 + (y-y_0)^2 + (z + z_0)^2}} \right]
-\end{align}
-
-\subsection{Part B - Circular Region of Potential}
-
-If the potential is specified to be $\phi_0$ on the surface $z=0$ in a small circular area of radius $a$ centered at the origin, we can use the general solution for $\rho = 0$ found in jackson and obtain
-
-\begin{align}
-\Phi &= \frac{-1}{4 \pi}\int_s \phi(r_0, \theta_0)\frac{\partial G}{\partial n_0}ds_0
-\end{align}
-
-The distances under the radicals may be expressed as the sum of the usual law of cosines type distance plus an extra term in the $z$ direction, this allows us to cast the entire green's function into cylindrical coordinates.
-We integrate only over the region of interest, outside of which the potential on the surface is zero:
-
-\begin{align}
-\Phi &= \frac{-1}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{d}{dz_0}\frac{1}{\sqrt{r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z - z_0)^2}} - \frac{1}{\sqrt{r^2 + r'^2 - 2 r r_0 \cos(\theta_0) + (z + z_0)^2}}r_0dr_0d\theta_0 \\
-&= \frac{-\phi_0}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{2z_0 - 2z}{2\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta_0) + (z - z_0)^2\right)^{\frac{3}{2}}} - \frac{2z_0 + 2_z}{2\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z + z_0)^2\right)^{\frac{3}{2}}}r_0dr_0d\theta_0 \\
-\rightarrow & z_0 = 0 \rightarrow \\
-&= \frac{-\phi_0}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{z_0 - z - z_0 - z}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
-&= \frac{-\phi_0z}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{1}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
-\end{align}
-
-This is the simplest form of the result we seek.
-
-\subsection{Part C - Potential Along the $Z$-Axis}
-
-If $x$ and $y$ are zero, then $r$ must be zero. This simplifies the result from the previous subsection as
-
-\begin{align}
-&= \frac{-\phi_0z}{4 \pi } \int_0^a \int_0^{2 \pi} \frac{1}{\left(r_0^2 + z^2\right)^{\frac{3}{2}}} r_0dr_0d\theta_0 \\
-&= \frac{-\phi_0 z}{4 \pi}\int_0^a \frac{r_0}{(r_0^2 + z^2)^{\frac{3}{2}}}dr_0
-\end{align}
-
-Let $u = r_0^2 + z^2$ and $du = 2r_0dr_0$ so $dr_0 = \frac{du}{2r_0}$.
-
-\begin{align}
-\Phi &= \frac{-\phi_0 z}{4 \pi}\int_{z^2}^{a^2 + z^2}\frac{\cancel{r_0}}{u^{\frac{3}{2}}}\frac{du}{2\cancel{r_0}} \\
-%&= \frac{\phi_0 z}{\epsilon_0} \left(\frac{1}{2}u^{-\frac{1}{2}}\right)\bigg|_{u=z^2}^{u=z^2+a^2} \\
-&= \frac{\phi_0 z}{4 \pi} \left(\frac{1}{\sqrt{u}}\right)\bigg|_{u=z^2}^{u=z^2 + a^2} \\
-&= \frac{\phi_0 z}{4 \pi} \left[\frac{1}{\sqrt{z^2 + a^2}} - \frac{1}{\sqrt{z^2}}\right] \\
-&= \frac{\phi_0 z}{4 \pi} \left[ 1 - \frac{z}{\sqrt{z^2 + a^2}}\right]\\
-\end{align}
-
-As required.
-
-\subsection{Expansion}
-
-Starting from
-\begin{align}
-&= \frac{-\phi_0z}{4 \pi} \int_0^a \int_0^{2 \pi} \frac{r_0dr_0d\theta_0}{\left(r^2 + r_0^2 - 2 r r_0 \cos(\theta) + (z)^2\right)^{\frac{3}{2}}}  \\
-\end{align}
-
-We divide both numerator and denominator by a factor of $(r^2 + z^2)^{\frac{3}{2}}$ and rearrange:
-
-\begin{align}
-\frac{-\phi_0 z}{4 \pi} \int_0^a \int_0^{2\pi}\left(1 + \frac{r_0^2 -2 r r_0 \cos(\theta_0)}{r^2 + z^2}\right)^{\frac{3}{2}}
-\end{align}
-
-Using $(1 + x)^n \sim 1 + nx + \frac{n(n-1)}{2}x^2...$ we carry the expansion to two terms as follows:
-
-\begin{align}
-\frac{-\phi_0 z}{4 \pi}\int_0^a \int_0^{2 \pi}\left[1 - \frac{3}{2(r^2 + z^2)}(r_0^2 - 2rr_0\cos(\theta_0)) + \frac{15}{8(r^2 + z^2)^2}(r_0^2 - 2rr_0\cos(\theta_0))^2\right]r_0 dr_0 d\theta_0 \\
-\end{align}
-
-Integrating term by term we have
-
-The first term, without the constants, or prefactor, is
-
-\begin{align}
-&= \int_0^a \int_0^{2 \pi}d\theta_0r_0dr_0 = \pi a^2 \\
-\end{align}
-
-The second term is
-
-\begin{align}
-&=\frac{1}{(r^2 + z^2)} \int_0^a \int_0^{2 \pi} \left(-\frac{3}{2}\right)(r_0^2 - 2rr_0^2\cos(\theta))d\theta_0dr_0 \\
-&= \frac{1}{(r^2 + z^2)} \left[\int_0^a \int_0^{2 \pi}r_0^3 d\theta_0 dr_0 - \int_0^a \int_0^{2 \pi}2rr_0^2\cos(\theta_0)d\theta_0 dr_0\right] \\
-&= \frac{1}{(r^2 + z^2)} \left[\frac{\pi a^4}{2} - \frac{2 r a^3}{3}(\sin(\theta_0 - 2 \pi) - \sin(\theta_0))\right] \\
-&= -\frac{3}{2}\frac{1}{(r^2 + z^2)}\frac{\pi a^4}{2} \\
-&= -\frac{3\pi a^4}{4(r^2 + z^2)}
-\end{align}
-
-The last integral is
-
-\begin{align}
-&=\frac{15}{8} \frac{1}{(r^2 + z^2)^2} \int_0^a \int_0^{2 \pi}\left[r_0^5 + 4 r^2 r_0^3\cos^2(\theta_0) - 4 r r_0^4\cos(\theta_0)\right]d\theta_0dr_0 \\
-&= \frac{15}{8} \frac{1}{(r^2 + z^2)^2}\bigg[\int_0^a \int_0^{2 \pi}r_0^5 d\theta_0 dr_0 + \int_0^a \int_0^{2 \pi} 4 r^2 r_0^3 \cos^2(\theta_0)d\theta_0dr_0  \\ &  - \int_0^a \int_0^{2 \pi} 4 r r_0^4 \cos(\theta_0)d\theta_0 dr_0\bigg] \\
-&= \frac{15}{8}\frac{1}{(z^2 + r^2)^2}\left[\frac{\pi a^6}{3} + \pi r^2 a^4\right] \\
-&= \frac{5a^2 \pi (3r^2 a^2 + a^4)}{8 (r^2 + z^2)^2}
-\end{align}
-
-Reinserting the prefactors and constants yields
-
-\begin{align}
-\Phi = \frac{-\phi_0  a^2}{4} \frac{z}{(z^2 + r^2)^{\frac{3}{2}}}\left[1 - \frac{3 a^2}{4(r^2 + z^2)} + \frac{5(3 r^2 a^2 + a^4)}{8(r^2 + z^2)^2}\right]
-\end{align}
-
-%Considering the figure, we see that the total vector to the field point from the origin is made up of a law of cosines like component in the plane plus a vector that accounts for the $z$ component.
-Thus to transform to cylinderical coordinates we must write the Green's function as
-
-%\begin{align}
-%G = \frac{1}{4 \pi \epsilon_)}\left[\frac{1}{
-%\end{align}
-
-\section{Problem 5}
-
-We begin with separation of variables assuming a solution of the form $\Phi  = R(r)\Theta(\theta)Z(z)$.The Laplacian in cylindrical coordinates is given by $\frac{1}{r}\frac{\partial}{\partial r}\left[r\frac{\partial}{\partial r}\right] + \frac{1}{r^2}\frac{\partial^2}{\partial \theta^2} + \frac{\partial^2}{\partial z^2}$.
-Acting upon $\Phi$ with this operator and setting the result equals to zero gives us
-
-\begin{align}
-Z\Theta\frac{1}{r} \frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{RZ}{r^2}\frac{\partial^2 \Theta}{\partial \theta^2} + R\Theta\frac{\partial^2 Z}{\partial z^2} &= 0 \\
-\frac{1}{rR}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{r^2 \Theta}\frac{\partial^2 \Theta}{\partial \theta^2} + \frac{1}{Z}\frac{\partial^2 Z}{\partial z^2} &= 0 \\
-\end{align}
-
-The only way that two functions of separate variables may be equal for all values of those variables are if they are both equal to a constant - the \textit{same} constant.
-
-\begin{align}
-\frac{1}{rR}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{r^2 \Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= -k^2 \\ 
-\frac{1}{Z}\frac{\partial^2 Z}{\partial z^2} &= k^2
-\end{align}
-
-The equation for $Z$ implies that $Z$ is proportional to real exponentials - or linear combinations of $sinh$ and $cosh$.
-
-\begin{align}
-Z \sim A\sinh(kz) + B\cosh(kz)
-\end{align}
-
-The other equation reads
-
-\begin{align}
-\frac{r}{R} \frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + \frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= -k^2r^2 \\
-\frac{r}{R}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + k^2 r^2 = -\frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} \\
-\end{align}
-
-A similar argument concerning constant functions allows us to write the two equations:
-
-\begin{align}
-\frac{r}{R}\frac{\partial}{\partial r}\left[r\frac{\partial R}{\partial r}\right] + k^2r^2 &= \mu^2 \\
-\frac{r}{R}\left[\frac{\partial R}{\partial r} + r\frac{\partial^2 R}{\partial r^2}\right] + k^2 r^2 &= \mu^2 \\
-\frac{r}{R}\frac{\partial R}{\partial r} + \frac{r^2}{R}\frac{\partial^2 R}{\partial r^2} + k^2 r^2 &= \mu^2R \\
-r^2 \frac{\partial^2 R}{\partial r^2} + r \frac{\partial R}{\partial r} + k^2 r^2 R = \mu^2R \\
-r^2 \frac{\partial^2 R}{\partial r^2} + r \frac{\partial R}{\partial r} + [k^2 r^2 - \mu^2]R &= 0 \\
--\frac{1}{\Theta}\frac{\partial^2 \Theta}{\partial \theta^2} &= \mu^2 \\
-\end{align}
-
-The equation in $R$ can be identified as Bessel's equation of order $\mu$ with argument $kr$. The $\Theta$ equation solutions are oscillatory.
-
-\begin{align}
-R &\sim J_{\mu}(kr) + \cancel{Y_\mu(kr)} \\
-\Theta &\sim \cancel{\sin(\mu\theta)}+ \cos(\mu\theta) \\
-\end{align}
-
-We may exclude Neumann functions as they diverge within the region of interest $r=0$.
-Furthermore, since $\phi_0$ is a constant on the sides of the cylinder we must have that $\Theta'' = \mu^2\Theta = 0$.
-Unless $\Theta = 0$, this requires that $\mu = 0$, which in turns restricts us to cosines in $\Theta$ and a cosine term equal to unity.
-The boundary condition at $z=0$ implies that the cosh term is also zero.
-With this information we have narrowed the solution down to something like
-
-\begin{align}
-\Phi \sim CJ_0(kr)\sinh(kz)
-\end{align}
-
-We take the top cap at $z=L$ as the one with the constant potential.
-To use the final boundary condition we need to lower the global potential by $\phi_0$ and impose the condition that $\phi = 0$ at $z=L$, or equivalently, that $\sinh(k(z-L)) = 0$.
-Additionally, $J_0(ka) = 0$, which restricts $k$ to be $\frac{x_{0n}}{a}$ where $x_{0n}$ is the $n'th$ zero of the zeroth order bessel function.
-
-\begin{align}
-\Phi &= \sum_{n=0}^{\infty} a_n J_0(\frac{x_{0n}}{a}r)\sinh(x_{0n}\frac{(z-L)}{a}) \\ 
-\text{at $z=0$} \rightarrow \phi_0 &=  \sum_{n=0}^{\infty}a_nJ_0(\frac{x_{0n}}{a}r)\sinh(-x_{0n}\frac{L}{a})\\
-\end{align}
-
-To find the coefficients we need to take a Fourier-Bessel series by multiplying both sides by $rJ_0(\frac{x_{0n'}}{a}$ and integrate from $0$ to $a$:
-
-\begin{align}
-\int_0^a \phi_0 rJ_0\left(\frac{x_{0n'}}{a}r\right)dr &= \sum_{n=0}^{\infty}a_n \sinh(-x_{0n}\frac{L}{a}) \int_0^a r J_0\left(\frac{x_{0n}}{a}r\right) J_0\left(\frac{x_{0n'}}{a}r\right)dr \\
-\int_0^a \phi_0 rJ_0\left(\frac{x_{0n'}}{a}r\right)dr &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2\delta_{nn'} \ \rightarrow \ \sinh \ \text{an odd function.} \\ 
-\end{align}
-
-The orthogonality of bessel functions was used on the right hand side.
-What we would like to do next is get the left hand side into a form that is suitable for an integral identity.
-Let $u = \frac{x_{0n}}{a}r$.
-Then $r = \frac{a}{x_{0n}}u$ and  $du = \frac{x_{0n}}{a}dr$.
-Furthermore, This allows us to rewrite the current equality as
-
-\begin{align}
-\phi_0\int_0^a \cancel{\frac{a}{x_{0n}}}uJ_0(u)\cancel{\frac{x_{0n}}{a}}du &= \sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2 \\
-\phi_0\int_0^a uJ_0(u)du &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{a^2}{2}[J_1(x_{0n})]^2 \\
-\rightarrow \  \ \text{Using an Integral identity}& \rightarrow \\
-\phi_0\frac{\cancel{a^2}}{x_{0n}}\cancel{J_1(x_{0n})} &= a_n\sinh(x_{0n}\frac{L}{a}) \frac{\cancel{a^2}}{2}[J_1(x_{0n})]^{\cancel{2}}\\
-a_n &= 2\frac{\phi_0 \ \text{cosech}(x_{0n}\frac{L}{a})}{x_{0n}J_1(x_{0n})}
-\end{align}
-
-The solution is thus as follows
-
-\begin{align}
-\Phi(r,z) = \phi_0 + 2\sum_{n=1}^{\infty}\frac{\phi_0 \ \text{cosech}(x_{0n}\frac{L}{a})}{x_{0n}J_1(x_{0n})}\sinh(x_{0n}\frac{(z-L)}{a})J_0\left(\frac{x_{0n}}{a}r\right)
-\end{align}
+%----------------------------------------------------------------------------------------
+%	PROBLEM 1
+%----------------------------------------------------------------------------------------
 
